---
title: "logFC Assessment"
author: "Nate Olson"
date: "10/4/2017"
output: html_document
---

```{r logFC_setup, include=FALSE}
library(tidyverse)
# library(broom)
# library(GGally)
library(stringr)
# library(ggExtra)
```

* Post specific features are not present in enough titration PCR replicates to use in assessment 
      * Using Post-dominant features for assessment
* Post-dominant features 
      * Some features with linear relationship when comparing the first titration to the second through 5th 
      * logFC between titrations 1-3 are not consistent with expectation
* Want to determine if the inconsistency is due to biases in the wet lab measurement method or computational.
* Things to look into  
      * logFC estimates
            * Does the expected logFC fall within the 95% confidence interval for the estimates logFC
            * Test for difference from expectation rather than non-zero logFC
      * Clustering Artifacts 
            * Unclustered logFC - to rule out clustering artifacts 
      * Wet Lab 
            * Features not observed in all four PCR replicates

## Clustering Artifacts
Looking at logFC for unclustered post-dominant features. 
Five of 22 features are well-behaved with differences in logFC between titrations decreasing when comparing non-adjacent titrations. 
For the five good features, the logFC estimates are consistent with expectations for comparisons between titrations 1,2 and 3. 

```{r}
pa_summary_unclust <- readRDS("~/Desktop/pa_summary_anno_df.RDS") %>% 
      select(pipe, biosample_id, feature_id, 
             T00, T20, pa_mixed, post_specific) %>% 
      filter(biosample_id %in% paste0("E01JH00", c("04","11","17")),
             pipe == "unclustered")

logFC_unclust <- readRDS("~/Desktop/logFC_edgeR_df.rds") %>%
      filter(biosample_id %in% paste0("E01JH00", c("04","11","17")),
             pipe == "unclustered")

logFC_unclust <- logFC_unclust %>% 
      dplyr::rename(feature_id = OTUname) %>% 
      left_join(pa_summary_unclust)
```

```{r}
logFC_prepost <- logFC_unclust %>% 
      filter(T1 == 0, T2 == 20)
```

MA Plot comparing unmixed pre- and post-exposure samples. 
```{r}
logFC_prepost %>% ggplot() + 
      geom_point(aes(x = logCPM, y = logFC, color = pa_mixed)) + 
      facet_wrap(~biosample_id) + theme_bw()
```

Only including features with logFC < -5 between pre- and post-exposure samples and present in at least 10 of the titration PCR replicates. 

```{r}
logFC_post_dom <- logFC_prepost %>% 
      filter(logFC < -5, pa_mixed > 10) %>% 
      dplyr::rename(prepost_logFC = logFC, prepost_logCPM = logCPM) %>% 
      ungroup() %>% 
      select(-T1, -T2, -PValue, -FDR) %>% 
      left_join(logFC_unclust)
```

```{r}
logFC_post_dom %>% 
      filter(T1 %in% as.character(1:5), T2 %in% as.character(1:5)) %>% 
      mutate(exp_logFC = as.numeric(as.character(T1)) - as.numeric(as.character(T2))) %>% 
             ggplot() + geom_point(aes(x = logCPM, y = logFC, color = biosample_id)) + 
      geom_hline(aes(yintercept = 0), linetype = 1) +              
      geom_hline(aes(yintercept = exp_logFC), linetype = 2) + 
                   facet_grid(T1~T2, scales = "free_y") + 
      theme_bw() + theme(legend.position = "bottom")
```




```{r}
logFC_post_dom %>% 
      filter(T1 %in% as.character(1:5), T2 %in% as.character(1:5)) %>% 
      mutate(exp_logFC = as.numeric(as.character(T1)) - as.numeric(as.character(T2))) %>% 
             ggplot() + geom_density(aes(x = logFC, 
                                         fill = biosample_id), 
                                     alpha = 0.25) + 
      geom_vline(aes(xintercept = 0), linetype = 1) +              
      geom_vline(aes(xintercept = exp_logFC), linetype = 2) + 
                   facet_grid(T1~T2, scales = "free_y") + 
      theme_bw() + theme(legend.position = "bottom")
```

Only looking at 23 total features ....
```{r}
logFC_post_dom %>% select(biosample_id, feature_id, pa_mixed) %>% 
      unique() %>% group_by(biosample_id) %>% 
      summarise(count = n(), pa_med = median(pa_mixed))
```
Good features - based on consistent decrease (more negative) in logFC for non-sequential titrations (need better descriptor)
E01JH0004 - sp468
E01JH0011 - sp2391, sp39, sp808
E01JH0017 - sp428 
```{r}
logFC_post_dom %>% 
      filter(T1 %in% as.character(1:5), T2 %in% as.character(1:5)) %>% 
      ggplot() + 
      geom_point(aes(x = T2, y = logFC, color = T1)) +
      facet_wrap(~paste(biosample_id, feature_id)) + 
      theme_bw()
```

```{r}
good_feat <- logFC_prepost %>% 
      filter((biosample_id == "E01JH0011" &
              feature_id %in% paste0("sp", c(2391, 39, 808))) |
                   (biosample_id == "E01JH0004" &
                    feature_id == "sp468") |
                   (biosample_id == "E01JH0017" &
                    feature_id == "sp428"))
good_feat %>% ungroup() %>% select(pipe, biosample_id, feature_id, Rank6, pa_mixed) %>% unique()
```


```{r}
logFC_prepost %>% ggplot() + 
      geom_point(aes(x = logCPM, y = logFC, color = pa_mixed)) + 
      geom_point(data = good_feat, aes(x = logCPM, y = logFC), color = "darkorange") + 
      facet_wrap(~biosample_id) + theme_bw()
```


```{r}
good_feat_comp <- logFC_post_dom %>% 
            filter(T1 %in% as.character(1:5), T2 %in% as.character(1:5)) %>% 
      mutate(exp_logFC = as.numeric(as.character(T1)) - as.numeric(as.character(T2))) %>% 
      filter((biosample_id == "E01JH0011" &
              feature_id %in% paste0("sp", c(2391, 39, 808))) |
                   (biosample_id == "E01JH0004" &
                    feature_id == "sp468") |
                   (biosample_id == "E01JH0017" &
                    feature_id == "sp428"))

logFC_post_dom %>% 
      filter(T1 %in% as.character(1:5), T2 %in% as.character(1:5)) %>% 
      mutate(exp_logFC = as.numeric(as.character(T1)) - as.numeric(as.character(T2))) %>% 
             ggplot() + 
      geom_point(aes(x = logCPM, y = logFC, color = biosample_id)) + 
      geom_point(data = good_feat_comp, aes(x = logCPM, y = logFC)) + 
      geom_hline(aes(yintercept = 0), linetype = 1) +              
      geom_hline(aes(yintercept = exp_logFC), linetype = 2) + 
                   facet_grid(T1~T2, scales = "free_y") + 
      theme_bw() + theme(legend.position = "bottom")
```

logFC estimates are not consistent with expection for other features present in most PCR replicates and with logCPM for pre- and post-exposure samples.
```{r}
logFC_post_dom %>% 
      filter(T1 %in% as.character(1:5), T2 %in% as.character(1:5), pa_mixed > 24, prepost_logCPM > 10) %>% 
      mutate(exp_logFC = as.numeric(as.character(T1)) - as.numeric(as.character(T2))) %>% 
             ggplot() + 
      geom_point(aes(x = logCPM, y = logFC, color = biosample_id)) + 
      geom_hline(aes(yintercept = 0), linetype = 1) +              
      geom_hline(aes(yintercept = exp_logFC), linetype = 2) + 
                   facet_grid(T1~T2, scales = "free_y") + 
      theme_bw() + theme(legend.position = "bottom")
```


```{r}
nb_unclust_11 <- readRDS("~/Desktop/nb_counts_unclustered_titration.RDS") %>% 
      filter(biosample_id == "E01JH0011")
```

```{r}
nb_wide <- nb_unclust_11 %>% filter(t_fctr %in% c(0,1,3,20)) %>% 
      mutate(t_fctr = paste0("T", str_pad(t_fctr, 2, side = "left", pad = "0"))) %>% 
      spread(t_fctr, nb_prop)
```

```{r}
nb_logfc <- nb_wide %>% mutate(prepost_logfc = log2(T20/T00), T13_logfc = log2(T03/T01)) 
```

Check values relative to expectation
```{r}
nb_logfc
```

```{r}
raw_counts <- readRDS("~/Desktop/raw_counts.RDS")
```

```{r}
raw_counts <- raw_counts %>% filter(pipe == "unclustered") %>% 
      group_by(pipe, id) %>% mutate(total_count = sum(count))
```

```{r}
post_dom_counts <- logFC_post_dom %>%
      left_join(raw_counts)
```

```{r}
post_dom_counts  %>% filter(biosample_id == "E01JH0004", t_fctr %in% c(0:5,20)) %>% 
      ggplot() + geom_point(aes(x = t_fctr, y = log2(count + 1), color = t_fctr)) + facet_wrap(~feature_id, scales = "free")
```
```{r}
post_dom_counts  %>% filter(biosample_id == "E01JH0004", t_fctr %in% c(0:5,20)) %>% 
      ggplot() + geom_point(aes(x = t_fctr, y = log2((count + 1)/total_count), color = t_fctr)) + facet_wrap(~feature_id, scales = "free")
```

```{r}
logFC_pre_dom <- logFC_prepost %>% 
      filter(logFC > 5, pa_mixed > 10) %>% 
      group_by(biosample_id) %>% 
      top_n(10, logCPM) %>% 
      dplyr::rename(prepost_logFC = logFC, prepost_logCPM = logCPM) %>% 
      ungroup() %>% 
      select(-T1, -T2, -PValue, -FDR) %>% 
      left_join(logFC_unclust)
```


```{r}
pre_dom_counts <- logFC_pre_dom %>%
      left_join(raw_counts)
```

```{r}
pre_dom_counts  %>% filter(biosample_id == "E01JH0004", t_fctr %in% c(1:5,20)) %>% 
      ggplot() + geom_point(aes(x = t_fctr, y = log2((count + 1)/total_count), color = t_fctr)) + facet_wrap(~feature_id, scales = "free")
```

```{r}
pre_dom_counts %>% select(biosample_id, feature_id, Rank6, prepost_logCPM) %>% arrange(-prepost_logCPM) %>% unique()
```

