---
title: "No Template Control Characterization"
author: "Nate Olson"
date: '`r Sys.Date()`'
always_allow_html: yes
output:
  pdf_document: default
  html_document: default
---

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd)
```

```{r echo = FALSE}
pipeline_dir <- "../../mgtst_pipelines"
mrexp <- get_mrexp(pipeline_dir)
```

```{r}
get_count_df <- function(mrobj, agg_genus = FALSE){
      if(agg_genus){
            mrobj <- aggregateByTaxonomy(mrobj, lvl = "Rank6", 
                                         norm = FALSE, log = FALSE, sl = 1)
      }
      
      mrobj <- cumNorm(mrobj, p = 0.75)
      mrobj %>%
            # not sure whether or not to normalize counts prior to analysis
            MRcounts(norm = TRUE, log = FALSE, sl = 1000) %>%  
            as.data.frame() %>% 
            rownames_to_column(var = "feature_id") %>% 
            gather("id","count", -feature_id)
}
```


```{r}
ntc_summary <- count_df %>% filter(biosample_id == "NTC") %>% filter(count != 0) %>% 
      group_by(pipe, feature_id) %>% summarise(total_count = sum(count), n_present = n())
```

Note count values are CSS normalized and scaled. 
```{r}
ntc_summary %>% ggplot() + 
      geom_histogram(aes(x = total_count)) + scale_y_log10() + theme_bw() +
      labs(x = "Total abundance across all NTC samples.")
```

```{r}
ntc_summary %>% ggplot() + 
      geom_histogram(aes(x = n_present)) + theme_bw() +
      labs(x = "Number of NTC samples feature is present in.")
```


```{r}
ntc_summary %>% ggplot() + 
      geom_point(aes(x = n_present, y = total_count)) + theme_bw() +
      labs(x = "Number of NTC Samples", y = "Total Abundance")
```


## Saving data 
```{r}
ntc_summary %>% saveRDS("data/ntc_features.rds")
```


# Session information 

## Git repo commit information
```{r}
repo <- repository(path = ".")
last_commit <- commits(repo)[[1]]
```

The current git commit of this file is `r last_commit@sha`, which is on the `r branches(repo)[[1]]@name` branch and was made by `r last_commit@committer@name` on `r when(last_commit)`. The current commit message is `r last_commit@summary`. The repository is online at https://github.com/nate-d-olson/mgtst-pub  


## Platform Information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
```


## Package Versions
```{r}
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
      knitr::kable()
```