---
title: "Titration Validation - Spike-in qPCR"
author: "Nate Olson"
date: '`r Sys.Date()`'
output:
  bookdown::pdf_document2: 
    toc: FALSE
--- 

```{r ercc_setup, warning=FALSE, message=FALSE, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
library(ProjectTemplate)
load.project()
library(modelr)
library(ggpubr)
```


```{r echo=FALSE, message=FALSE}
ercc_std <- qpcrERCC %>% filter(sample_type == "std", !grepl("NTC",sampleID)) %>% 
      mutate(sampleID = gsub("\\(.*","",sampleID), 
             Ct = as.numeric(Ct), 
             quat = as.numeric(quant),
             log_quant = log10(quant))

fit_mod <- ercc_std %>% mutate(ercc = as.numeric(ercc)) %>% 
      group_by(ercc) %>% nest() %>% 
      mutate(fit = map(data, ~lm(Ct~log_quant, data = . )))  
fit_list <- fit_mod$fit %>% set_names(fit_mod$ercc)

fit_coefs <-fit_list %>% map_df(coefficients) %>% 
      add_column(coefs = c("intercept","slope")) %>% 
      gather("ercc","stat",-coefs) %>% spread(coefs, stat) 

std_fit <- fit_list %>% map_df(broom::glance, .id = "ercc") %>% 
      select(ercc, adj.r.squared) %>%
      left_join(fit_coefs) %>% 
      mutate(amplification_factor = 10^(-1/slope), 
             efficiency = (amplification_factor - 1) * 100) 

ercc_meta <- erccMeta %>% select(ercc_id, biosample_id, treatment) %>%
    mutate(ercc = str_replace(ercc_id,"ERCC-",""),  
           ercc = as.numeric(ercc) %>% as.character())

post_assays <- c(108,12, 157, 2, 35)
ercc_sam <- qpcrERCC %>% filter(sample_type == "sam") %>% 
      mutate(Ct = as.numeric(Ct), 
             quant = as.numeric(quant), 
             ercc = as.numeric(ercc), 
             titration = gsub("._M","",sampleID),
             titration = gsub(".*\\(Pre\\)","20", titration),
             titration = gsub(".*\\(Post\\)","0", titration),
             titration = as.numeric(titration),
             pre_prop = (1 - (2^-titration)),
             assay_type = if_else(ercc %in% post_assays, "Post","Pre"))

post_fit_mod <- ercc_sam %>% filter(assay_type == "Post") %>% 
      group_by(ercc,assay_type) %>% nest() %>% 
      mutate(fit = map(data, ~lm(Ct~titration, data = . )))

post_fit_list <- post_fit_mod$fit %>% set_names(post_fit_mod$ercc)

# Extract fit parameters and calculate efficiency
post_fit_coefs <- post_fit_list %>% map_df(coefficients) %>% 
      add_column(coefs = c("intercept","slope")) %>% 
      gather("ercc","stat",-coefs) %>% spread(coefs, stat)

post_fit <- post_fit_list %>% map_df(broom::glance, .id = "ercc") %>% 
      select(ercc, adj.r.squared) %>%
      left_join(post_fit_coefs) %>% 
      mutate(amplification_factor = 10^(-1/slope), 
             efficiency = (amplification_factor - 1) * 100) 

## Pre-treatment sample fit
pre_fit_mod <- ercc_sam %>% 
      filter(assay_type == "Pre", titration %in% c(1:4,20)) %>%
      mutate(pre_log2 = log2(pre_prop)) %>% 
      group_by(ercc,assay_type) %>% nest() %>%
      mutate(fit = map(data, ~lm(Ct~pre_log2, data = . )))

pre_fit_list <- pre_fit_mod$fit %>% set_names(pre_fit_mod$ercc)

# Extract fit parameters and calculate efficiency
pre_fit_coefs <- pre_fit_list %>% map_df(coefficients) %>%
      add_column(coefs = c("intercept","slope")) %>%
      gather("ercc","stat",-coefs) %>% spread(coefs, stat)

pre_fit <- pre_fit_list %>% map_df(broom::glance, .id = "ercc") %>%
      select(ercc, adj.r.squared) %>%
      left_join(pre_fit_coefs) %>%
      mutate(amplification_factor = 10^(-1/slope),
             efficiency = (amplification_factor - 1) * 100) 


## Titration subset models ###################################
# post_fit_mod14 <- ercc_sam %>% filter(assay_type == "Post", titration %in% 1:4) %>% 
#       group_by(ercc,assay_type) %>% nest() %>% 
#       mutate(fit = map(data, ~lm(Ct~titration, data = . )))
# 
# post_fit_list14 <- post_fit_mod14$fit %>% set_names(post_fit_mod14$ercc)

# Extract fit parameters and calculate efficiency
# post_fit_coefs14 <- post_fit_list14 %>% map_df(coefficients) %>%
#       add_column(coefs = c("intercept","slope")) %>%
#       gather("ercc","stat",-coefs) %>% spread(coefs, stat)
# 
# post_fit14 <- post_fit_list14 %>% map_df(broom::glance, .id = "ercc") %>%
#       select(ercc, adj.r.squared) %>%
#       left_join(post_fit_coefs14) %>% mutate(fit = "1:4")
# 
# post_fit_mod05 <- ercc_sam %>% filter(assay_type == "Post", titration %!in% 1:4) %>%
#       group_by(ercc,assay_type) %>% nest() %>%
#       mutate(fit = map(data, ~lm(Ct~titration, data = . )))
# 
# post_fit_list05 <- post_fit_mod05$fit %>% set_names(post_fit_mod05$ercc)
# 
# # Extract fit parameters and calculate efficiency
# post_fit_coefs05 <- post_fit_list05 %>% map_df(coefficients) %>%
#       add_column(coefs = c("intercept","slope")) %>%
#       gather("ercc","stat",-coefs) %>% spread(coefs, stat)
# 
# post_fit05 <- post_fit_list05 %>% map_df(broom::glance, .id = "ercc") %>%
#       select(ercc, adj.r.squared) %>%
#       left_join(post_fit_coefs05) %>%
#       mutate(fit = "0,5,10,15")

####################### Data for qPCR summary Table ############################
ercc_stdtbl <- ercc_meta %>% 
      right_join(std_fit) %>% 
      mutate(std_R2 = round(adj.r.squared,4), E = round(efficiency, 2)) %>% 
      select(biosample_id, treatment, std_R2, E) 

ercc_vldtbl <- bind_rows(post_fit, pre_fit) %>% 
      left_join(ercc_meta) %>% 
      mutate(R2 = round(adj.r.squared, 2), Slope = round(slope, 2)) %>% 
      select(biosample_id, treatment, R2, Slope)

ercc_tbl <- left_join(ercc_stdtbl, ercc_vldtbl)


#### Code for t-tests to check slopes equal to -1 and 1
# fit_tdy <- post_fit_mod$fit[[1]] %>% tidy()
# fit_glance <- post_fit_mod$fit[[1]] %>% glance()
# resid_df <- fit_glance$df.residuals
# slope_est <- fit_tdy[2,2]
# slope_stderror <- fit_tdy[2,3]
# t_val <- (slope_est - 1)/slope_stderror
# p_val <-pt(t_val, resid_df)
# p_val

# fit_tdy <- pre_fit_mod$fit[[1]] %>% tidy()
# fit_glance <- pre_fit_mod$fit[[1]] %>% glance()
# resid_df <- fit_glance$df.residual
# slope_est <- fit_tdy[2,2]
# slope_stderror <- fit_tdy[2,3]
# 
# t_val <- (slope_est + 1)/slope_stderror
# p_val <-pt(t_val, resid_df)
# p_val


### Code for ERCC Titration Factor plot
ercc_sam <- ercc_sam %>% 
      mutate(ercc = as.character(ercc)) %>% 
      left_join(ercc_meta)

### Code for independent Pre and Post Plots
ercc_plot_dat <- ercc_sam %>% mutate(ercc = as.character(ercc)) %>% 
      left_join(ercc_meta) %>% 
      mutate(pre_log2 = log2(pre_prop))

ercc_preplot <- ercc_plot_dat %>% 
      filter(treatment == "Pre", titration %in% c(1:4)) %>% 
      ggplot() +
      geom_smooth(aes(x = pre_prop, y = Ct), method = "lm") + 
      geom_point(aes(x = pre_prop, y = Ct)) +
      facet_wrap(~biosample_id, ncol = 1) +
      theme_bw() + labs(x = "Pre-Exposure Proportion", y = "Ct")  +
      scale_x_continuous(trans = "log2")

ercc_postplot <- ercc_plot_dat %>% 
      filter(treatment == "Post") %>%
      ggplot() +
      geom_smooth(aes(x = titration, y = Ct), method = "lm") + 
      geom_point(aes(x = titration, y = Ct)) +
      facet_wrap(~biosample_id, ncol = 1) +
      theme_bw() + labs(x = "Titration Factor", y = "Ct") +
      scale_x_continuous(trans = "log2")
```

- The qPCR assay standard curves had a high level of precision with $R^2$ values close to 1 for all standard curves (Table \@ref(tab:erccTable)).  
- The amplification efficiency was outside of the ideal range (0.9 - 1.1), but within the acceptable range (0.8-1.2). Ideal and acceptable ranges based on rule of thumb community accepted guidelines.  

- The post treatment qPCR assays had good $R^2$ and slope values (Table \@ref(tab:erccTable)). The expected slope is 1, for a doubling every cycle.  
- The 1-4 titration factor samples had Ct values consistently above the regression line (Figure \@ref(fig:erccPlot)).  
- For the pre-treatment samples, should be 1 Ct difference between the unmixed post and titration factor 1 and 0.5 Ct between titration factor 0 and 1. For the other titration factors the expected difference is to small to detect using qPCR (< 0.5 Ct).  
- Figure \@ref(fig:erccPlot) has both pre- and post-exposure qPCR results on a $log_2$ scale with expected slopes of 1 and -1 (Table \@ref(tab:erccTable)). 

```{r erccTable, echo=FALSE, message=FALSE}
ercc_tbl %>% arrange(treatment) %>% 
    knitr::kable(caption = "ERCC Spike-in qPCR summary statistics. $R^2$ and Efficiency (E) for standard curves. $R^2$ and slope for titration qPCR results for ", booktabs = TRUE)
```

```{r erccPlot, fig.cap = "qPCR ERCC spike-in titration validation results.", echo=FALSE, message=FALSE}
ercc_sam %>% ggplot() + 
      geom_smooth(aes(x = titration, y = Ct), method = "lm") + 
      geom_point(aes(x = titration, y = Ct)) +
      facet_grid(treatment~biosample_id, scales = "free_y") +
      theme_bw() + labs(x = "Titration Factor", y = "Ct")
```

```{r erccPlot2, fig.cap = "qPCR ERCC spike-in results for ERCC spiked into unmixed (A) Post-exposure samples and (B) Pre-exposure samples. X-axis is on a log2 scale with expected slope of 1 and -1 for Post-exposure and Pre-exposure spike-ins respectively.", fig.height = 7, echo=FALSE, message=FALSE}
ggarrange(ercc_postplot, 
          ercc_preplot,
          labels = c("A", "B"),
          nrow = 1, ncol = 2)
```

__Potential Next Steps__
* Use a mixed effects model to account for qPCR plate effects   
* Add test (t-test) to test whether slopes are significantly different from expectation (1,-1)  
* Compare model fix results for different titration subsets (only relevant to support conclusions regarding titration specific differences.)


