---
title: "Microbiome Scale Qualitative Assessment Results"
author: "Nate Olson"
date: '`r Sys.Date()`'
output:
  bookdown::pdf_document2: 
    toc: FALSE
bibliography: [mgtst.bib, packages.bib]
---

```{r qualSetup, include=FALSE}
library(tidyverse)
library(broom)
library(ggpubr)
raw_counts <- readRDS("~/Desktop/raw_counts.RDS")
nb_counts <- readRDS("~/Desktop/nb_counts_titrations.RDS")
pa_summary_anno_df <- readRDS("~/Desktop/pa_summary_anno_df.RDS")
theta_est <- readRDS("~/Desktop/bootstrap_theta_estimates.rds")
mix_binom_tidy <- readRDS("~/Desktop/mix_bayes_test.RDS")
unmix_binom_tidy <- readRDS("~/Desktop/unmix_binom_test.RDS")
```

```{r qualMunge, message = FALSE, echo = FALSE, warning = FALSE}
## Feature-PCR rep level tidy data frame
## Relevant raw_count factors 
count_df <- raw_counts %>% 
      select(biosample_id, t_fctr, id, pipe, feature_id, count)

count_df <- count_df %>% 
      group_by(id) %>% 
      mutate(total_count = sum(count))

## Sanity Check 
# count_df$total_count %>% unique() %>% summary()

## Annotating with feature type
count_df <- pa_summary_anno_df %>% 
      select(biosample_id, pipe, feature_id, pa_specific) %>% 
      left_join(count_df)

## Inferred Feature Proportions
### Same code used in Rel_Abu_Error
pre_post_prop <- nb_counts %>% 
      ungroup() %>% 
      filter(t_fctr %in% c(0,20)) %>% 
      mutate(end_point = if_else(t_fctr == 0 , "post", "pre")) %>% 
      select(-t_fctr) %>% 
      ## setting values to 0 when one or more of the PCR replicates are 0 for titration end-points
      spread(end_point,nb_prop, fill = 0)

prop_inferred <- theta_est %>% 
      filter(pipe == "unclustered") %>% 
      ungroup() %>%
      mutate(t_fctr = factor(t_fctr, levels = c(0:5, 10, 15, 20))) %>% 
      select(biosample_id, theta_hat_mean, t_fctr) %>% 
      right_join(nb_counts) %>% right_join(pre_post_prop) %>% 
      filter(t_fctr %in% c(1:5,10,15)) %>% 
      ## Setting negative inferrred theta values to 0
      mutate(theta_hat_mean = if_else(theta_hat_mean < 0, 0, theta_hat_mean)) %>% 
      ## Using inferred theta estimates to calculate expected values
      mutate(inferred_prop = post * theta_hat_mean + pre * (1-theta_hat_mean))

count_df <- prop_inferred %>% 
      select(biosample_id, t_fctr, pipe, feature_id, 
             pre, post, inferred_prop, theta_hat_mean) %>% 
      left_join(count_df)

unmix_count_df <- filter(count_df, 
                    t_fctr %in% c(1:5, 10, 15),
                    pa_specific == "unmixed",
                    # should not have to filter by inferred_prop unmix should
                    # have non-zero inferred prop or at least non-zero pre +
                    # post
                    pre + post != 0, count == 0) 

mix_count_df <- filter(count_df,
                  t_fctr %in% c(1:5, 10, 15),
                  count != 0, pre == 0, post == 0)

## Multipannel plots
unmix_plot_df <- unmix_count_df %>% 
      filter(pipe != "unclustered") %>% 
      mutate(exp_count = inferred_prop * total_count) 
      
unmix_count_plot <- ggplot(unmix_plot_df) +
      geom_boxplot(aes(x = biosample_id, y = exp_count + 1), 
                   alpha = 0.25) + 
      geom_hline(aes(yintercept = 1 + 1), linetype = 2) + 
      scale_y_log10() + theme_bw() + 
      labs(x = "Pipeline", y = "Expected Counts", 
           fill = "Biological Replicate", color = "Biological Replicate") +
      theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) +
      facet_wrap(~pipe, nrow = 1)

## Only including p < 0.05 
# unmix_test_plot <- unmix_binom_tidy %>% 
#       filter(pipe != "unclustered") %>% 
#       mutate(pval_cat = if_else(adj.pvalue < 0.05, "p < 0.05", "p >= 0.05")) %>% 
#       ggplot() + geom_bar(aes(x = biosample_id, fill = pval_cat), position = "fill") + 
#       facet_wrap(~pipe, nrow = 1) +
#       theme_bw() + theme(axis.text.x = element_text(angle = 90), legend.position = "bottom") +
#       labs(x = "Biological Replicate", y = "Proportion", fill = "Adjusted p-value")

unmix_binom_prop <- unmix_binom_tidy %>% 
      filter(pipe != "unclustered") %>% 
      mutate(pval_cat = if_else(adj.pvalue < 0.05, "p < 0.05", "p >= 0.05")) %>% 
      group_by(pipe, biosample_id) %>% 
      mutate(count = n()) %>% 
      group_by(pipe, biosample_id, count, pval_cat) %>% 
      summarise(cat_count = n()) %>% 
      mutate(pval_prop = cat_count/count) %>% 
      filter(pval_cat == "p < 0.05")

unmix_test_plot <- unmix_binom_prop %>% 
      ggplot() + 
      geom_bar(aes(x = biosample_id, y = pval_prop), stat = "identity") + 
      facet_wrap(~pipe, nrow = 1) + 
      theme_bw() + theme(axis.text.x = element_text(angle = 90)) +
      labs(x = "Biological Replicate", y = "Proportion")

mix_count_plot <- mix_count_df %>% filter(pipe != "unclustered") %>% ggplot() + 
      geom_boxplot(aes(x = biosample_id, y = count + 1),
                   alpha = 0.25) + 
      geom_hline(aes(yintercept = 1 + 1), linetype = 2) + 
      scale_y_log10() + theme_bw() + 
      labs(x = "Pipeline", y = "Observed Counts", 
           fill = "Biological Replicate", color = "Biological Replicate") +
      theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) +
      facet_wrap(~pipe, nrow = 1)

## Only including p < 0.05 
# mix_test_plot <- mix_binom_tidy %>% filter(pipe != "unclustered") %>% 
#       mutate(pval_cat = if_else(adj.pvalue < 0.05, "p < 0.05", "p >= 0.05")) %>% 
#       ggplot() + geom_bar(aes(x = biosample_id, fill = pval_cat), position = "fill") + 
#       facet_wrap(~pipe) +
#       theme_bw() + theme(axis.text.x = element_text(angle = 90), legend.position = "bottom") +
#       labs(x = "Biological Replicate", y = "Proportion", fill = "Adjusted p-value")


mix_binom_prop <- unmix_binom_tidy %>% 
      filter(pipe != "unclustered") %>% 
      mutate(pval_cat = if_else(adj.pvalue < 0.05, "p < 0.05", "p >= 0.05")) %>% 
      group_by(pipe, biosample_id) %>% 
      mutate(count = n()) %>% 
      group_by(pipe, biosample_id, count, pval_cat) %>% 
      summarise(cat_count = n()) %>% 
      mutate(pval_prop = cat_count/count) %>% 
      filter(pval_cat == "p < 0.05")

mix_test_plot <- mix_binom_prop %>% 
      ggplot() + 
      geom_bar(aes(x = biosample_id, y = pval_prop), stat = "identity") + 
      facet_wrap(~pipe, nrow = 1) + 
      theme_bw() + theme(axis.text.x = element_text(angle = 90)) +
      labs(x = "Biological Replicate", y = "Proportion")
```

```{r qualPlot, message = FALSE, echo = FALSE, warning = FALSE, fig.cap = "Distribution of (A) observed count values for titration-specific features and (B) expected count values for unmixed-specific features by pipeline and individual. The horizontal dashed line indicates a count value of 1. (C) Proportion of unmix-specific features and (D) titration-specific features with an adjusted p-value < 0.05 for the bayesian hypothesis test and binomial test respectively. We fail to accept the null hypothesis when the p-value < 0.05, indicating that for these features the discrepancy between the feature not being observed in the titration and present in the unmixed samples is not explained by sampling alone."}
ggarrange(mix_count_plot + rremove("x.text"), unmix_count_plot + rremove("x.text"), 
          mix_test_plot, unmix_test_plot, 
          labels = c("A", "B", "C","D"),
          ncol = 2, nrow = 2, heights = c(4, 5))
#ggsave("~/Desktop/qual_assessment.png", dpi = 450)
```

There are a number of unmixed- and titration-specific features with a range of observed (titration-specific, Fig. \@ref(fig:qualPlot)A) and expected (unmix-specific, Fig. \@ref(fig:qualPlot)B) counts. 
There were unmixed-specific features with expected counts that could not be explained by sampling alone for all biological replicates and bioinformatic pipelines (Fig. \@ref(fig:qualPlot)C). 
However, the proportion of unmixed-specific features that could not be explained by sampling alone varied by bioinformatic pipeline with over half of the DADA2 unmixed-specific features could not be explained by sampling alone whereas QIIME had the lowest rate of unmixed-specific features that could not be explained by sampling alone. 
Consistent with the distribution of observed counts for titration-specific features more of the DADA2 features could not be explained by sampling alone compared to the other pipelines (Fig. \@ref(fig:qualPlot)D).  
