---
title: "MA Plot Dev"
author: "Nate Olson"
date: '`r Sys.Date()`'
output:
  pdf_document:
    toc: yes
  html_document:
    code_folding: hide
    theme: yeti
    toc: yes
    toc_float: yes
---
## TODO
* mgtst import %>
* add notes
* function for pre/post otu analysis
* look at non-pre/post specific otus

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
library(mgtst)
library(readr)
library(tidyr)
library(stringr)
library(magrittr)
library(knitr)
library(devtools)
library(sapkotaUtils)
library(ggplot2)
library(dplyr)
library(tibble)
library(purrr)
library(phyloseq)
library(biomformat)
library(vegan)
library(plotly)
library(metagenomeSeq)
opts_chunk$set(cache=TRUE)
opts_chunk$set(fig.align="center")
```

## MA Plots

```{r prj_paths, warning=FALSE, message=FALSE, echo = FALSE}
proj_dir <- "~/Projects/16S_etec_mix_study"
pipe_dir <- paste0(proj_dir, "/analysis/pipelines")
qiime_dir <- paste0(pipe_dir, "/qiime")
dada2_dir <- paste0(pipe_dir, "/dada2")
mothur_dir <- paste0(pipe_dir, "/mothur")
```

### Importing qiime output and adding metadata.
```{r load_qiime}
## importing qiime output at phyloseq object
ps_qiime <- import_biom(
    paste0(qiime_dir,"/otus_uc_fast/otu_table_mc2_w_tax_no_pynast_failures.biom"))

## generating metadata df
data(sample_sheet)
sample_sheet_ns <- sample_sheet %>%
    mutate(pos_ns = str_replace(pos, "_",""))
sample_id <- data_frame(seq_ids = sample_names(ps_qiime)) %>%
    separate(col = seq_ids,
             into = c("pcr_16S_plate","pos_ns"), sep = "-") %>%
    mutate(pcr_16S_plate = as.numeric(pcr_16S_plate))

sample_df <- left_join(sample_id, sample_sheet_ns) %>%
    filter(seq_lab == "JHU", barcode_lab == "JHU") %>%
    mutate(pcr_16S_plate = as.character(pcr_16S_plate)) %>% as.data.frame()

## defining metadata slot
samdat <- sample_data(sample_df)
sample_names(samdat) <- sample_names(ps_qiime)
sample_data(ps_qiime) <- samdat
saveRDS(ps_qiime, paste0(qiime_dir,"/otus_uc_fast/phyloseq_obj.rds"))
```

Convert to MRexperiment
```{r}
mrexp_qiime <- phyloseq_to_metagenomeSeq(ps_qiime)
# saveRDS(mrexp_qiime, paste0(qiime_dir,"/otus_uc_fast/mrexp_obj.rds"))
```

## Calculating Fold change
```{r}
sample_ids <- pData(mrexp_qiime)$sampleID %>% unique()
sample_ids <- sample_ids[sample_ids != "NTC"]
```

### logFC for individual samples
```{r}
get_ma_df <- function(mrexp, i){
    mrexp_pre_post <- mrexp[, which(pData(mrexp)$sampleID == i & 
                        pData(mrexp)$dilution %in% c(-1,0))]
    mrexp_pre_post_filt <- filterData(mrexp_pre_post, present = 4, depth = 1)
    mrexp_pre_post_filt <- cumNorm(mrexp_pre_post_filt, p = 0.75)
    normFactor <- normFactors(mrexp_pre_post_filt) %>% {log2(./median(.) + 1)}
    
    dilution <- pData(mrexp_pre_post_filt)$dilution
    mod <- model.matrix(~dilution)
    settings <- zigControl(maxit = 10, verbose = TRUE)
    fit <- fitZig(obj = mrexp_pre_post_filt, mod = mod, 
             useCSSoffset = FALSE, control = settings,
             normFactor = normFactor)
    data_frame(sampleID = i,
               otu = fit$taxa,
               abu = rowMeans(fit$counts), 
               logFC = fit$fit$coefficients[,2])
}

ma_df <- sample_ids %>% map_df(get_ma_df, mrexp = mrexp_qiime)
```

Note x axis is log10 of log2 transformed mean ...
https://en.wikipedia.org/wiki/MA_plot
```{r}
ggplot(ma_df) + geom_point(aes(x = abu, y = logFC, color = sampleID), 
                           alpha = 0.25) + scale_x_log10()
```

### LogFC calculated using normalized observed counts
```{r}
get_ma_df <- function(mrexp_qiime, i){
    mrexp_pre_post <- mrexp_qiime[, which(pData(mrexp_qiime)$sampleID == i & 
                        pData(mrexp_qiime)$dilution %in% c(-1,0))]
    mrexp_pre_post_filt <- mrexp_pre_post
    count_tbl <- MRcounts(mrexp_pre_post_filt,norm = TRUE)
    pre_sams <- rownames(pData(mrexp_pre_post_filt))[pData(mrexp_pre_post_filt)$dilution == 0]
    post_sams <- rownames(pData(mrexp_pre_post_filt))[pData(mrexp_pre_post_filt)$dilution == -1]
    rowMeans_pre <- rowMeans(count_tbl[,colnames(count_tbl) %in% pre_sams])
    rowMeans_post <- rowMeans(count_tbl[,colnames(count_tbl) %in% post_sams])
    # excluding OTUs with average counts < 1
    count_tbl <- count_tbl[rowMeans_pre > 1 & rowMeans_post > 1,]
    
    A <- rowMeans(count_tbl)
    logFC <- log10(rowMeans(count_tbl[,colnames(count_tbl) %in% pre_sams])/
             rowMeans(count_tbl[,colnames(count_tbl) %in% post_sams]))
    data_frame(sampleID = i, A, logFC)
}

ma_df <- data_frame()
for(i in sample_ids){
    ma_df <- get_ma_df(mrexp_qiime, i) %>% bind_rows(ma_df, .)
}

ggplot(ma_df) + geom_point(aes(x = A, y = logFC, color = sampleID)) + 
    geom_vline(aes(xintercept = 500), linetype = 2) +
    geom_hline(aes(yintercept = -1), linetype = 2) + 
    geom_hline(aes(yintercept = 1), linetype = 2) + 
    theme_bw() +
    scale_x_log10()
```

Four quadrants:  

1. y > 1  and x < 500   
2. y > 1  and x > 500  
3. y < -1 and x < 500  
4. y < -1 and x > 500

### Comparing All pre and post samples
Did not exclude counts with average less than 1
```{r}
get_ma_df <- function(mrexp_qiime){
    mrexp_pre_post <- mrexp_qiime[, pData(mrexp_qiime)$dilution %in% c(-1,0)]
    mrexp_pre_post_filt <- mrexp_pre_post #filterData(mrexp_pre_post, present = 4, depth = 2)
    count_tbl <- MRcounts(mrexp_pre_post_filt,norm = TRUE)
    pre_sams <- rownames(pData(mrexp_pre_post_filt))[pData(mrexp_pre_post_filt)$dilution == 0]
    post_sams <- rownames(pData(mrexp_pre_post_filt))[pData(mrexp_pre_post_filt)$dilution == -1]
    A <- rowMeans(count_tbl)
    rowMeans_pre <- rowMeans(count_tbl[,colnames(count_tbl) %in% pre_sams])
    rowMeans_post <- rowMeans(count_tbl[,colnames(count_tbl) %in% post_sams])
    count_tbl <- count_tbl[rowMeans_pre > 0 & rowMeans_post > 0,]
    
    A <- rowMeans(count_tbl)
    logFC <- log10(rowMeans(count_tbl[,colnames(count_tbl) %in% pre_sams])/
             rowMeans(count_tbl[,colnames(count_tbl) %in% post_sams]))
    data_frame(sampleID = i, A, logFC)
}


ma_df <- get_ma_df(mrexp_qiime)

ggplot(ma_df) + geom_point(aes(x = A, y = logFC)) + 
    geom_vline(aes(xintercept = 50), linetype = 2) + 
    geom_hline(aes(yintercept = -1), linetype = 2) + 
    geom_hline(aes(yintercept = 1), linetype = 2) + 
    theme_bw() +
    scale_x_log10()

```

Four quadrants:  

1. y > 1  and x < 50   
2. y > 1  and x > 50  
3. y < -1 and x < 50  
4. y < -1 and x > 50  

### LogFC calculated using normalized observed counts
Individual samples median for MA
```{r}
get_ma_df <- function(mrexp_qiime, i){
    mrexp_pre_post <- mrexp_qiime[, which(pData(mrexp_qiime)$sampleID == i & 
                        pData(mrexp_qiime)$dilution %in% c(-1,0))]
    mrexp_pre_post_filt <- mrexp_pre_post
    count_tbl <- MRcounts(mrexp_pre_post_filt,norm = TRUE)
    pre_sams <- rownames(pData(mrexp_pre_post_filt))[pData(mrexp_pre_post_filt)$dilution == 0]
    post_sams <- rownames(pData(mrexp_pre_post_filt))[pData(mrexp_pre_post_filt)$dilution == -1]
    rowMedians_pre <- rowMedians(count_tbl[,colnames(count_tbl) %in% pre_sams])
    rowMedians_post <- rowMedians(count_tbl[,colnames(count_tbl) %in% post_sams])
    # excluding OTUs with average counts < 1
    count_tbl <- count_tbl[rowMedians_pre > 0 & rowMedians_post > 0,]
    
    A <- rowMedians(count_tbl)
    logFC <- log10(rowMedians(count_tbl[,colnames(count_tbl) %in% pre_sams])/
             rowMedians(count_tbl[,colnames(count_tbl) %in% post_sams]))
    data_frame(sampleID = i, A, logFC)
}

ma_df <- data_frame()
for(i in sample_ids){
    ma_df <- get_ma_df(mrexp_qiime, i) %>% bind_rows(ma_df, .)
}

ggplot(ma_df) + geom_point(aes(x = A, y = logFC, color = sampleID)) + 
    geom_vline(aes(xintercept = 500), linetype = 2) +
    geom_hline(aes(yintercept = -1), linetype = 2) + 
    geom_hline(aes(yintercept = 1), linetype = 2) + 
    theme_bw() +
    scale_x_log10()
```

Four quadrants:  

1. y > 1  and x < 500   
2. y > 1  and x > 500  
3. y < -1 and x < 500  
4. y < -1 and x > 500

### Subset samples based on Pre-Post Presence Absence
List of Pre and Post Specific OTUs by sampleID

0 in pre or post columns indicates per or post specific OTU
```{r}
get_pre_post_specific_otus <- function(otu_mat, pre_sams, post_sams){
    rowMedians_pre <- rowMedians(otu_mat[,colnames(otu_mat) %in% pre_sams])
    rowMedians_post <- rowMedians(otu_mat[,colnames(otu_mat) %in% post_sams])
    
    data_frame(otus = rownames(otu_mat),
               pre = rowMedians_pre,
               post = rowMedians_post)
}

get_sample_otu_table <- function(mrexp, i){
    mrexp_pre_post <- mrexp[, which(pData(mrexp)$sampleID == i & 
                                    pData(mrexp)$dilution %in% c(-1,0))]
    count_tbl <- MRcounts(mrexp_pre_post,norm = TRUE)
    pre_sams <- rownames(pData(mrexp_pre_post))[
        pData(mrexp_pre_post)$dilution == 0]
    post_sams <- rownames(pData(mrexp_pre_post))[
        pData(mrexp_pre_post)$dilution == -1]
   get_pre_post_specific_otus(count_tbl, pre_sams, post_sams)
}
```

```{r}
pre_post_specific_otus <- sample_ids %>% set_names(sample_ids)%>% 
    map_df(get_sample_otu_table, mrexp = mrexp_qiime, .id = "sampleID")
```

## E01JH0038
### Pre Specific OTUs
```{r}
i <- "E01JH0038"; mrexp <- mrexp_qiime # variables for function
pre_otus <- pre_post_specific_otus %>% filter(sampleID == i,
                                              pre > 0 & post == 0) %>% 
    arrange(desc(post)) %>% top_n(20) %>% 
    .$otus

mrexp_pre_post <- mrexp[rownames(mrexp) %in% pre_otus,
                        which(pData(mrexp)$sampleID == i)]
sam_dat <- pData(mrexp_pre_post) %>% add_rownames(var = "samID")

count_tbl <- MRcounts(mrexp_pre_post,norm = TRUE) %>% as.data.frame() %>% 
    add_rownames(var = "otu") %>% 
    gather("samID","count", -otu) %>% 
    left_join(sam_dat)
```

!!! Need to check - potentially have pre and post dilution assignments switched.
```{r}
count_tbl %>% filter(otu %in% pre_otus[1:20]) %>% 
    mutate(dilution = ifelse(dilution == 0, 18, dilution),
           dilution = ifelse(dilution == -1, 0, dilution)) %>% 
    ggplot() + geom_point(aes(x = 2^-dilution, y = count)) + 
        geom_smooth(aes(x = 2^-dilution, y = count)) +
        facet_wrap(~otu, scale = "free_y")
```

## Post
```{r}
i <- "E01JH0038"; mrexp <- mrexp_qiime # variables for function
post_otus <- pre_post_specific_otus %>% filter(sampleID == i,
                                              pre == 0 & post > 0) %>% 
    arrange(desc(post)) %>% top_n(20) %>% 
    .$otus

mrexp_pre_post <- mrexp[rownames(mrexp) %in% post_otus,
                        which(pData(mrexp)$sampleID == i)]
sam_dat <- pData(mrexp_pre_post) %>% add_rownames(var = "samID")

count_tbl <- MRcounts(mrexp_pre_post,norm = TRUE) %>% as.data.frame() %>% 
    add_rownames(var = "otu") %>% 
    gather("samID","count", -otu) %>% 
    left_join(sam_dat)
```

Better linearity for pre than post
```{r}
count_tbl %>% filter(otu %in% post_otus[1:20]) %>% 
    mutate(dilution = ifelse(dilution == 0, 18, dilution),
           dilution = ifelse(dilution == -1, 0, dilution)) %>% 
    ggplot() + geom_point(aes(x = 2^-dilution, y = count)) + 
        geom_smooth(aes(x = 2^-dilution, y = count)) +
        facet_wrap(~otu, scale = "free_y")
```


## E01JH0017
### Pre
```{r}
i <- "E01JH0017"; mrexp <- mrexp_qiime # variables for function
pre_otus <- pre_post_specific_otus %>% filter(sampleID == i,
                                              pre > 0 & post == 0) %>% 
    arrange(desc(post)) %>% top_n(20) %>% 
    .$otus

mrexp_pre_post <- mrexp[rownames(mrexp) %in% pre_otus,
                        which(pData(mrexp)$sampleID == i)]
sam_dat <- pData(mrexp_pre_post) %>% add_rownames(var = "samID")

count_tbl <- MRcounts(mrexp_pre_post,norm = TRUE) %>% as.data.frame() %>% 
    add_rownames(var = "otu") %>% 
    gather("samID","count", -otu) %>% 
    left_join(sam_dat)
```

!!! Need to check - potentially have pre and post dilution assignments switched.
```{r}
count_tbl %>% filter(otu %in% pre_otus[1:20]) %>% 
    mutate(dilution = ifelse(dilution == 0, 18, dilution),
           dilution = ifelse(dilution == -1, 0, dilution)) %>% 
    ggplot() + geom_point(aes(x = 2^-dilution, y = count)) + 
        geom_smooth(aes(x = 2^-dilution, y = count)) +
        facet_wrap(~otu, scale = "free_y")
```

### Post
```{r}
post_otus <- pre_post_specific_otus %>% filter(sampleID == i,
                                              pre == 0 & post > 0) %>% 
    arrange(desc(post)) %>% top_n(20) %>% 
    .$otus

mrexp_pre_post <- mrexp[rownames(mrexp) %in% post_otus,
                        which(pData(mrexp)$sampleID == i)]
sam_dat <- pData(mrexp_pre_post) %>% add_rownames(var = "samID")

count_tbl <- MRcounts(mrexp_pre_post,norm = TRUE) %>% as.data.frame() %>% 
    add_rownames(var = "otu") %>% 
    gather("samID","count", -otu) %>% 
    left_join(sam_dat)
```

Post OTUs drop out really quickly....
```{r}
count_tbl %>% filter(otu %in% post_otus[1:20]) %>% 
    mutate(dilution = ifelse(dilution == 0, 18, dilution),
           dilution = ifelse(dilution == -1, 0, dilution)) %>% 
    ggplot() + geom_point(aes(x = 2^-dilution, y = count)) + 
        geom_smooth(aes(x = 2^-dilution, y = count)) +
        facet_wrap(~otu, scale = "free_y")
```


## E01JH0004
### Pre
```{r}
i <- "E01JH0004"; mrexp <- mrexp_qiime # variables for function
pre_otus <- pre_post_specific_otus %>% filter(sampleID == i,
                                              pre > 0 & post == 0) %>% 
    arrange(desc(post)) %>% top_n(20) %>% 
    .$otus

mrexp_pre_post <- mrexp[rownames(mrexp) %in% pre_otus,
                        which(pData(mrexp)$sampleID == i)]
sam_dat <- pData(mrexp_pre_post) %>% add_rownames(var = "samID")

count_tbl <- MRcounts(mrexp_pre_post,norm = TRUE) %>% as.data.frame() %>% 
    add_rownames(var = "otu") %>% 
    gather("samID","count", -otu) %>% 
    left_join(sam_dat)
```

!!! Need to check - potentially have pre and post dilution assignments switched.
```{r}
count_tbl %>% filter(otu %in% pre_otus[1:20]) %>% 
    mutate(dilution = ifelse(dilution == 0, 18, dilution),
           dilution = ifelse(dilution == -1, 0, dilution)) %>% 
    ggplot() + geom_point(aes(x = 2^-dilution, y = count)) + 
        geom_smooth(aes(x = 2^-dilution, y = count)) +
        facet_wrap(~otu, scale = "free_y")
```

### Post
```{r}
post_otus <- pre_post_specific_otus %>% filter(sampleID == i,
                                              pre == 0 & post > 0) %>% 
    arrange(desc(post)) %>% top_n(20) %>% 
    .$otus

mrexp_pre_post <- mrexp[rownames(mrexp) %in% post_otus,
                        which(pData(mrexp)$sampleID == i)]
sam_dat <- pData(mrexp_pre_post) %>% add_rownames(var = "samID")

count_tbl <- MRcounts(mrexp_pre_post,norm = TRUE) %>% as.data.frame() %>% 
    add_rownames(var = "otu") %>% 
    gather("samID","count", -otu) %>% 
    left_join(sam_dat)
```

Post OTUs drop out really quickly....
```{r}
count_tbl %>% filter(otu %in% post_otus[1:20]) %>% 
    mutate(dilution = ifelse(dilution == 0, 18, dilution),
           dilution = ifelse(dilution == -1, 0, dilution)) %>% 
    ggplot() + geom_point(aes(x = 2^-dilution, y = count)) + 
        geom_smooth(aes(x = 2^-dilution, y = count)) +
        facet_wrap(~otu, scale = "free_y")
```

## E01JH0016
### Pre
```{r}
i <- "E01JH0016"; mrexp <- mrexp_qiime # variables for function
pre_otus <- pre_post_specific_otus %>% filter(sampleID == i,
                                              pre > 0 & post == 0) %>% 
    arrange(desc(post)) %>% top_n(20) %>% 
    .$otus

mrexp_pre_post <- mrexp[rownames(mrexp) %in% pre_otus,
                        which(pData(mrexp)$sampleID == i)]
sam_dat <- pData(mrexp_pre_post) %>% add_rownames(var = "samID")

count_tbl <- MRcounts(mrexp_pre_post,norm = TRUE) %>% as.data.frame() %>% 
    add_rownames(var = "otu") %>% 
    gather("samID","count", -otu) %>% 
    left_join(sam_dat)
```

!!! Need to check - potentially have pre and post dilution assignments switched.
```{r}
count_tbl %>% filter(otu %in% pre_otus[1:20]) %>% 
    mutate(dilution = ifelse(dilution == 0, 18, dilution),
           dilution = ifelse(dilution == -1, 0, dilution)) %>% 
    ggplot() + geom_point(aes(x = 2^-dilution, y = count)) + 
        geom_smooth(aes(x = 2^-dilution, y = count)) +
        facet_wrap(~otu, scale = "free_y")
```

### Post
```{r}
post_otus <- pre_post_specific_otus %>% filter(sampleID == i,
                                              pre == 0 & post > 0) %>% 
    arrange(desc(post)) %>% top_n(20) %>% 
    .$otus

mrexp_pre_post <- mrexp[rownames(mrexp) %in% post_otus,
                        which(pData(mrexp)$sampleID == i)]
sam_dat <- pData(mrexp_pre_post) %>% add_rownames(var = "samID")

count_tbl <- MRcounts(mrexp_pre_post,norm = TRUE) %>% as.data.frame() %>% 
    add_rownames(var = "otu") %>% 
    gather("samID","count", -otu) %>% 
    left_join(sam_dat)
```

Post OTUs drop out really quickly....
```{r}
count_tbl %>% filter(otu %in% post_otus[1:20]) %>% 
    mutate(dilution = ifelse(dilution == 0, 18, dilution),
           dilution = ifelse(dilution == -1, 0, dilution)) %>% 
    ggplot() + geom_point(aes(x = 2^-dilution, y = count)) + 
        geom_smooth(aes(x = 2^-dilution, y = count)) +
        facet_wrap(~otu, scale = "free_y")
```

## E01JH0011
### Pre
```{r}
i <- "E01JH0011"; mrexp <- mrexp_qiime # variables for function
pre_otus <- pre_post_specific_otus %>% filter(sampleID == i,
                                              pre > 0 & post == 0) %>% 
    arrange(desc(post)) %>% top_n(20) %>% 
    .$otus

mrexp_pre_post <- mrexp[rownames(mrexp) %in% pre_otus,
                        which(pData(mrexp)$sampleID == i)]
sam_dat <- pData(mrexp_pre_post) %>% add_rownames(var = "samID")

count_tbl <- MRcounts(mrexp_pre_post,norm = TRUE) %>% as.data.frame() %>% 
    add_rownames(var = "otu") %>% 
    gather("samID","count", -otu) %>% 
    left_join(sam_dat)
```

!!! Need to check - potentially have pre and post dilution assignments switched.
```{r}
count_tbl %>% filter(otu %in% pre_otus[1:20]) %>% 
    mutate(dilution = ifelse(dilution == 0, 18, dilution),
           dilution = ifelse(dilution == -1, 0, dilution)) %>% 
    ggplot() + geom_point(aes(x = 2^-dilution, y = count)) + 
        geom_smooth(aes(x = 2^-dilution, y = count)) +
        facet_wrap(~otu, scale = "free_y")
```

### Post
```{r}
post_otus <- pre_post_specific_otus %>% filter(sampleID == i,
                                              pre == 0 & post > 0) %>% 
    arrange(desc(post)) %>% top_n(20) %>% 
    .$otus

mrexp_pre_post <- mrexp[rownames(mrexp) %in% post_otus,
                        which(pData(mrexp)$sampleID == i)]
sam_dat <- pData(mrexp_pre_post) %>% add_rownames(var = "samID")

count_tbl <- MRcounts(mrexp_pre_post,norm = TRUE) %>% as.data.frame() %>% 
    add_rownames(var = "otu") %>% 
    gather("samID","count", -otu) %>% 
    left_join(sam_dat)
```

Post OTUs drop out really quickly....
```{r}
count_tbl %>% filter(otu %in% post_otus[1:20]) %>% 
    mutate(dilution = ifelse(dilution == 0, 18, dilution),
           dilution = ifelse(dilution == -1, 0, dilution)) %>% 
    ggplot() + geom_point(aes(x = 2^-dilution, y = count)) + 
        geom_smooth(aes(x = 2^-dilution, y = count)) +
        facet_wrap(~otu, scale = "free_y")
```

Looking at all OTUs - revisit spliting by quadrants
```{r}
sam_dat <- pData(mrexp_qiime) %>% add_rownames(var = "samID")
count_tbl <- MRcounts(mrexp_qiime,norm = TRUE) %>% as.data.frame() %>% 
    add_rownames(var = "otu") %>% 
    gather("samID","count", -otu) %>% 
    left_join(sam_dat)
count_tbl <- count_tbl %>% group_by(sampleID, otu, dilution) %>% 
    summarize(count = median(count)) %>% 
    group_by(sampleID, otu) %>% 
    mutate(count_total = sum(count),count_mean = mean(count)) %>% ## removing OTUs with less than 100 observations by sample
    filter(count_total > 100)
```

```{r}
count_tbl %>% 
    ungroup() %>% 
    mutate(dilution = ifelse(dilution == 0, 18, dilution),
           dilution = ifelse(dilution == -1, 0, dilution),
           total_count_bins = cut_number(count_mean,10)) %>% 
    ggplot() + geom_line(aes(x = dilution, y = count, group = otu)) +
        facet_grid(total_count_bins~sampleID, scale = "free_y")
```

