---
title: "Normalization"
author: "Nate Olson"
date: "July 8, 2016"
output: html_document
---
  
```{r setup, warning=FALSE, message=FALSE, echo = FALSE}  
library(mgtst)
library(ggplot2)
library(dplyr)
library(tibble)
library(tidyr)
library(purrr)
library(phyloseq)
library(metagenomeSeq)
library(knitr)
opts_chunk$set(cache=TRUE)
opts_chunk$set(fig.align="center")
```

## Loading Data
QIIME Standard Pipeline  

```{r}
proj_dir <- "~/Projects/16S_etec_mix_study"
pipe_dir <- file.path(proj_dir, "analysis", "pipelines")
qiime_dir <- file.path(pipe_dir, "qiime")
qiime_std <- file.path(qiime_dir, "otus_uc_fast","mrexp_obj.rds") %>% 
    readRDS()
```

 
Excluding no template controls and low quality samples.  

```{r}
qiime_std <- qiime_std[,which(pData(qiime_std)$sampleID != "NTC" & 
                            !(pData(qiime_std)$id %in% c("B3_M7_P2_L1_S1",
                                                         "B4_M6_P1_L1_S1")))]
```

## Normalization

```{r}
qiime_raw_counts <- qiime_std@assayData$counts %>% as.data.frame() %>% 
    rownames_to_column(var = "OTU") %>% gather("samID","raw", -OTU)
qiime_norm75_counts <- cumNormMat(qiime_std,p = 0.75) %>% as.data.frame() %>% 
    rownames_to_column(var = "OTU") %>% gather("samID","cum_norm_75", -OTU)
qiime_norm_stat_counts <- cumNormMat(qiime_std,p = cumNormStatFast(qiime_std)) %>% as.data.frame() %>% 
    rownames_to_column(var = "OTU") %>% gather("samID","cum_norm_stat", -OTU)
```

Data frame with results from different normalization methods
```{r}
qiime_norm_df <- qiime_raw_counts %>% 
    right_join(qiime_norm75_counts) %>% 
    right_join(qiime_norm_stat_counts)
```

Annotate normalization data frame with experimental metadata.
```{r}
qiime_pheno <- qiime_std %>% phenoData() %>% .@data %>% 
    rownames_to_column(var = "samID") %>% as_data_frame()
qiime_anno_df <- qiime_pheno %>% right_join(qiime_norm_df)
```

```{r}
count_var <- qiime_anno_df %>% 
    select(sampleID, OTU, dilution, raw, cum_norm_75, cum_norm_stat) %>%
    gather("norm","count", -sampleID, -OTU, -dilution) %>% 
    group_by(sampleID, OTU, dilution, norm) %>% 
    summarize(count_mean = mean(count), count_var = var(count), count_sd = sd(count)) %>% 
    filter(count_mean > 0)
```

## Variance Distribution
```{r}
ggplot(count_var) + geom_boxplot(aes(x = norm, y = count_var)) + 
    scale_y_log10() + theme_bw()
```

### Conditional Variance Distributions
Variance by Normalized Counts
```{r}
ggplot(count_var) + geom_hex(aes(x = count_mean, y = count_var)) + facet_wrap(~norm)  + 
    scale_y_log10() + scale_x_log10() + theme_bw()
```

Variance by Biological Replicate - I would not expect the median variance value to be the same for the five biological replicates without any scaling or normalization.  

```{r}
ggplot(count_var) + geom_boxplot(aes(x = norm, y = count_var, color = sampleID)) +
    scale_y_log10() + theme_bw()
```

Variance by OTU
```{r}
ggplot(count_var)+ geom_point(aes(x = OTU, y = count_var)) + facet_wrap(~norm) + scale_y_log10()
```

Using linear regression to determine factors contributing to the observed variance.
```{r}

```

