---
title: "MA Plots"
author: "Nate Olson"
date: '`r Sys.Date()`'
output:
  html_document:
    code_folding: hide
    theme: yeti
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
library(mgtst)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(phyloseq)
# library(plotly)
library(metagenomeSeq)
library(knitr)
opts_chunk$set(cache=TRUE)
opts_chunk$set(fig.align="center")
```

## MA Plots
Summary ...

## Loading Datasets
```{r}
proj_dir <- "~/Projects/16S_etec_mix_study"
pipe_dir <- paste0(proj_dir, "/analysis/pipelines")
qiime_dir <- paste0(pipe_dir, "/qiime/")
dada2_dir <- paste0(pipe_dir, "/dada2/")
mothur_dir <- paste0(pipe_dir, "/mothur/")

qiime_std <- readRDS(paste0(qiime_dir, "otus_uc_fast/mrexp_obj.rds"))
qiime_denovo <- readRDS(paste0(qiime_dir, "otus_uc_fast_denovo/mrexp_obj.rds"))
mothur <- readRDS(paste0(mothur_dir, "data/process/mrexp_obj.rds"))
dada <- readRDS(paste0(dada2_dir, "/processed_data/mrexp_obj-2016-04-25.rds"))
```

## Functions
Functions for generating plots and input dataframes
```{r}
get_ma_df <- function(i, mrexp, present, depth){
    ## subset pre and post samples
    mrexp_pre_post <- mrexp[, which(pData(mrexp)$sampleID == i & 
                        pData(mrexp)$dilution %in% c(-1,0))]
    
    mrexp_pre_post_filt <- filterData(mrexp_pre_post, present = present, depth = depth)
    
    ## Normalized count table
    count_tbl <- cumNormMat(mrexp_pre_post_filt, p = 0.75)
    
    ## Pre and Post sample ids
    pre_sams <- mrexp_pre_post_filt %>% pData() %>% 
        {.[.$dilution == 0, ]} %>% rownames()
    post_sams <- mrexp_pre_post_filt %>% pData() %>%  
        {.[.$dilution == -1, ]} %>% rownames()
    
    ## Mean normalized counts by OTU
    rowMeans_pre <- count_tbl %>% {.[,colnames(.) %in% pre_sams]} %>%  rowMeans()
    rowMeans_post <- count_tbl %>% {.[,colnames(.) %in% post_sams]} %>%  rowMeans()
    
    
    ## dataframe with MA values
    ## A - mean counts across all samples
    ## logFC (M) - log10 fold change pre/post
    data_frame(sampleID = i, 
               otu = rownames(count_tbl), 
               A = rowMeans(count_tbl),
               pre_means = rowMeans_pre,
               post_means = rowMeans_post) %>% 
        mutate(logFC = log2(pre_means/post_means))
    
}

get_ma_df_by_sample <- function(mrexp, sample_ids, present=4, depth=1){
    sample_ids %>% map_df(get_ma_df, mrexp,present, depth)
}

plot_by_sample_ma <- function(ma_df){
    ggplot(ma_df) + geom_point(aes(x = A, y = logFC, group = otu), alpha = 0.5) + 
        geom_hline(aes(yintercept = -1), linetype = 2) + 
        geom_hline(aes(yintercept = 1), linetype = 2) + 
        theme_bw() +
        scale_x_log10() +
        facet_wrap(~sampleID)
}

get_quad_df <- function(ma_df, n = 3){
    ma_df %>% 
        filter(!is.infinite(logFC)) %>% 
        mutate(quad = "NO QUAD",
               quad = ifelse(A > 1000 & logFC >  1, "A >1K,M>1", quad),
               quad = ifelse(A > 1000 & logFC < -1, "A >1k,M<-1", quad),
               quad = ifelse(A > 100 & A < 1001  & logFC >  1, "A 100-1k,M>1", quad),
               quad = ifelse(A > 100 & A < 1001  & logFC < -1, "A 100-1k,M<-1", quad),
               quad = ifelse(A > 10 & A < 101  & logFC >  1, "A 10-100,M>1", quad),
               quad = ifelse(A > 10 & A < 101 & logFC < -1, "A 10-100,M<-1", quad),
               quad = ifelse(A < 10 & logFC >  1, "A <10,M>1", quad),
               quad = ifelse(A < 10 & logFC < -1, "A <10,M<-1", quad)) %>% 
        filter(quad != "NO QUAD") %>%
        group_by(quad, sampleID) %>%
        top_n(n, wt = A) %>% mutate(rep_otu = factor(1:n()))
}

get_quad_otu_tbl <- function(quad_df, mrexp){
    sam_dat <- pData(mrexp) %>% add_rownames(var = "samID")
    cumNormMat(mrexp, p = 0.75) %>% as.data.frame() %>% 
        add_rownames(var = "otu") %>% 
        gather("samID","count", -otu) %>% 
        left_join(sam_dat) %>% right_join(quad_df)
}

plot_quad_otus <- function(quad_tbl){
    quad_tbl <- quad_tbl %>% 
    mutate(dilution = ifelse(dilution == 0, 18, dilution),
           dilution = ifelse(dilution == -1, 0, dilution))
    quads <- quad_tbl$quad %>% unique()
    for(i in quads){
        pl <- quad_tbl %>% filter(quad == i) %>% 
              ggplot(aes(x = dilution, y = count,
                         group = otu)) + 
                geom_point() + geom_smooth() + scale_x_continuous(trans = "log2") +
                facet_grid(sampleID~rep_otu, scale = "free_y") + 
                theme_bw() + theme(legend.position = "bottom") + 
                labs(x = "Proportion Post", y = "Count") +
                ggtitle(paste("MA Plot Quad",i)) +
                theme(axis.text.x = element_text(angle = 90))
        print(pl)
    }
}
```

Sample IDs
```{r}
sample_ids <- pData(qiime_std)$sampleID %>% unique()
sample_ids <- sample_ids[sample_ids != "NTC"]
```

### QIIME Standard Pipeline 
```{r}
qiime_std <- qiime_std[,which(pData(qiime_std)$sampleID != "NTC" & 
                            !(pData(qiime_std)$id %in% c("B3_M7_P2_L1_S1",
                                                         "B4_M6_P1_L1_S1")))]
ma_df <- get_ma_df_by_sample(qiime_std, sample_ids) 
```

#### MA Plot
```{r}
ma_df %>% filter(!is.infinite(logFC)) %>% plot_by_sample_ma()
```

#### QUAD OTUs
```{r}
get_quad_df(ma_df, n = 6) %>% get_quad_otu_tbl(qiime_std) %>% plot_quad_otus()
```


### QIIME De Novo Pipeline 
```{r}
qiime_denovo <- qiime_denovo[,which(pData(qiime_denovo)$sampleID != "NTC" & 
                            !(pData(qiime_denovo)$id %in% c("B3_M7_P2_L1_S1",
                                                         "B4_M6_P1_L1_S1")))]
ma_df <- get_ma_df_by_sample(qiime_denovo, sample_ids) 
```

#### MA Plot
```{r}
ma_df %>% filter(!is.infinite(logFC)) %>%  plot_by_sample_ma()
```

#### QUAD OTUs
```{r}
get_quad_df(ma_df, n = 6) %>% get_quad_otu_tbl(qiime_denovo) %>% plot_quad_otus()
```


### Mothur Pipeline 
```{r}
mothur <- mothur[,which(pData(mothur)$sampleID != "NTC" & 
                            !(pData(mothur)$id %in% c("B3_M7_P2_L1_S1",
                                                         "B4_M6_P1_L1_S1")))]
ma_df <- get_ma_df_by_sample(mothur, sample_ids) 
```

#### MA Plot
```{r}
ma_df %>% filter(!is.infinite(logFC)) %>% plot_by_sample_ma()
```

#### QUAD OTUs
```{r fig.height=8}
# get_quad_df(ma_df, n = 6) %>% get_quad_otu_tbl(mothur) %>%
#     plot_quad_otus()
```

### DADA2 Pipeline
```{r}
dada <- dada[,which(pData(dada)$sampleID != "NTC" & 
                            !(pData(dada)$id %in% c("B3_M7_P2_L1_S1",
                                                         "B4_M6_P1_L1_S1")))]
ma_df <- get_ma_df_by_sample(dada, sample_ids) 
```

#### MA Plot
```{r}
ma_df %>% filter(!is.infinite(logFC)) %>% plot_by_sample_ma()
```

#### QUAD OTUs
Error - commenting out for now
```{r fig.height=8}
# get_quad_df(ma_df, n = 6) %>% get_quad_otu_tbl(dada) %>%
#     plot_quad_otus()
```


## Investigating OTUs
Looking at OTU for > -1 log fold change (OTUs with higher abundance for post-treatment samples). 

### OTUs present in all samples
OTU drop-off for post-treatment samples with greater abundance compared to pre-treatment samples. 
Filtering OTUs present in all samples with depths of at least 2 prior to normalization. 
Drop-off still observed but not consistently by biological replicates with OTU specific drop off, 
see biological replicate 11 (second row) for A 10-100, 
with representative OTU 1 and 3 having drop off but not OTU 2. 
Also biological replicate 17 A 10-100, representaive OTU 3 has a drop-off but not 1 and 2. 

```{r}
get_ma_df_by_sample(qiime_std, sample_ids, present = 8, depth = 2) %>% 
    get_quad_df(n = 3) %>% filter(logFC < -1) %>% 
    get_quad_otu_tbl(qiime_std) %>% plot_quad_otus()
```

# Comparing Observed vs. Expected Counts
Only looking at OTUs with at least of log10 fold change.
```{r}
quad_otu <- qiime_std %>%
    get_ma_df_by_sample(sample_ids, present = 4, depth = 1) %>%
    get_quad_df(n = 10000) %>%
    get_quad_otu_tbl(qiime_std)

otu_obs_vs_exp <- quad_otu %>%
    # dilution 18 selected for post as it is a real number but small enough to
    # have minimal impact on the unmixed pre samples
    mutate(dilution = ifelse(dilution == 0, 18, dilution), 
           dilution = ifelse(dilution == -1, 0, dilution),
           post_prop = 2^-dilution,
           pre_prop = 1-post_prop,
           post_exp = post_prop * post_means,
           pre_exp = pre_prop * pre_means,
           count_exp = post_exp + pre_exp) %>% 
    filter(!(dilution %in% c(18,0))) # removing unmixed samples- biases results
```

### Plot of expected vs observed
```{r}
ggplot(otu_obs_vs_exp) +
    geom_point(aes(x = count_exp, y = count, color = logFC), alpha = 0.25) +
    geom_smooth(aes(x = count_exp, y = count), color = "black") +
    facet_wrap(~sampleID) +
    scale_y_log10() + scale_x_log10() +
    theme_bw() +
    labs(x = "expected counts", y = "observed counts")
```

Observed counts less than 10
```{r}
otu_obs_vs_exp %>% filter(count < 10) %>% 
ggplot() +
    geom_point(aes(x = count_exp, y = count, color = logFC), alpha = 0.25) +
    geom_smooth(aes(x = count_exp, y = count), color = "black") +
    facet_wrap(~sampleID) +
    scale_y_log10() + scale_x_log10() +
    theme_bw() +
    labs(x = "expected counts", y = "observed counts")
```

Observed counts greater than 10
```{r}
otu_obs_vs_exp %>% filter(count > 10) %>% 
ggplot() +
    geom_point(aes(x = count_exp, y = count, color = logFC), alpha = 0.25) +
    geom_smooth(aes(x = count_exp, y = count), color = "black") +
    facet_wrap(~sampleID) +
    scale_y_log10() + scale_x_log10() +
    theme_bw() +
    labs(x = "expected counts", y = "observed counts")
```

### Expected vs Observed by logFC

OTUs with positive logFC are more abundant in pre-treatment samples and OTUs with negative logFC are more abundance in post-treatment samples.

```{r}
otu_obs_vs_exp %>%
    mutate(fc_bin = ifelse(logFC > 0, "Pre-OTU","Post-OTU")) %>%
    ggplot() +
        geom_point(aes(x = count_exp, y = count, color = logFC), alpha = 0.25) +
        geom_abline(aes(slope = 1, intercept = 0), color = "black") +
        facet_grid(fc_bin~sampleID) +
        scale_y_continuous(trans = "log2") + scale_x_continuous(trans = "log2") +
        theme_bw() +
        labs(x = "expected counts", y = "observed counts")
```

## Linear models observed vs expected values

None of the variance is explained by pcr plate and biological replicate effects

```{r}
fit <- lm(count ~ 0 + count_exp, data = otu_obs_vs_exp)
summary(fit)

## biosample effect
fit2 <- lm(count ~ 0 + count_exp + sampleID, data = otu_obs_vs_exp)
summary(fit2)

## Plate effect
fit3 <- lm(count ~ 0 + count_exp + pcr_16S_plate, data = otu_obs_vs_exp)
summary(fit3)

## plate and biosample interaction
fit4 <- lm(count ~ 0 + count_exp + sampleID*pcr_16S_plate, data = otu_obs_vs_exp)
summary(fit4)
```

Excluding Counts less than 10
```{r}
otu_obs_vs_exp_gt10 <- otu_obs_vs_exp %>% filter(count > 10)
fit <- lm(count ~ 0 + count_exp, data = otu_obs_vs_exp_gt10)
summary(fit)
```


## Models by biological replicate

Descent correlation between counts and expected counts overall
Lower correlations for sampleID 16 and for OTUs specific to post treatment samples (e.g. logFC < 1)

```{r}
sample_fit <- otu_obs_vs_exp %>% split(.$sampleID) %>%
    map(~lm(formula = count ~ 0 + count_exp, data = .))
par(mfrow = c(2,2))
for(i in 1:length(sample_fit)){
    print(names(sample_fit[i]))
    print(summary(sample_fit[[i]]))
    plot(sample_fit[[i]])
}
par(mfrow = c(1,1))
```


excluding count values less than 10
```{r}
sample_fit <- otu_obs_vs_exp %>% filter(count > 10)%>% split(.$sampleID) %>%
    map(~lm(formula = log2(count) ~ 0 + log2(count_exp), data = .))
par(mfrow = c(2,2))
for(i in 1:length(sample_fit)){
    print(names(sample_fit[i]))
    print(summary(sample_fit[[i]]))
    plot(sample_fit[[i]])
}
par(mfrow = c(1,1))
```

excluding count values less than 100
```{r}
sample_fit <- otu_obs_vs_exp %>% filter(count > 100) %>% split(.$sampleID) %>%
    map(~lm(formula = count ~ 0 + count_exp, data = .))
par(mfrow = c(2,2))
for(i in 1:length(sample_fit)){
    print(names(sample_fit[i]))
    print(summary(sample_fit[[i]]))
    plot(sample_fit[[i]])
}
par(mfrow = c(1,1))
```


## Models by OTU

```{r}
library(broom)
otu_fit <- otu_obs_vs_exp %>% split(.$otu) %>%
    map(~lm(formula = count ~ 0 + count_exp, data = .)) %>%
    map_df(glance, .id = "otu")

ggplot(otu_fit) + geom_histogram(aes(x = adj.r.squared))
```

## Correlation by OTU

biological replicate OTU correlations have a bimodal distribution

```{r}
otu_cor <- otu_obs_vs_exp %>% group_by(otu, sampleID) %>%
    summarise(corr = cor(count, count_exp))
ggplot(otu_cor) + geom_histogram(aes(x = corr)) +
    facet_wrap(~sampleID)
```

# Next Steps
* Summarize findings to date
* Look into OTUs with low correlation values  
* Check out samples with low seq counts especially filtered samples B3_M7_P2_L1_S1 and B4_M6_P1_L1_S1.


## Residuals vs. Observed
```{r}
otu_resid <- otu_obs_vs_exp %>% mutate(resid = count - count_exp)
```

Might want to exclude counts with median values < 10, instead of filtering by expected or observed counts.


### Residuals as a function of the expected counts excluding counts less than 10
```{r}
otu_resid %>% filter(count_exp > 10) %>% 
    mutate(sampleOTU = paste(otu, dilution, sampleID, sep = "_")) %>% 
    ggplot() + 
    geom_line(aes(x = count_exp, y = resid, group = sampleOTU), color = "grey70") +
    geom_point(aes(x = count_exp, y = resid), alpha = 0.25) + 
    scale_x_continuous(trans = "log2") + scale_y_continuous(trans = "log2") +
    geom_abline(aes(slope = 1, intercept = 0), color = "darkorange") +
    facet_wrap(~sampleID) + theme_bw()
```

### Residuals as a function of the observed counts excluding counts less than 10
```{r}
otu_resid %>% filter(count > 10) %>% mutate(sampleOTU = paste(otu, dilution, sampleID, sep = "_")) %>% 
    ggplot() + 
    geom_line(aes(x = count, y = resid, group = sampleOTU), color = "grey70") +
    geom_point(aes(x = count, y = resid), alpha = 0.25) + 
    scale_x_continuous(trans = "log2") + scale_y_continuous(trans = "log2") +
    geom_abline(aes(slope = 1, intercept = 0), color = "darkorange") +
    facet_wrap(~sampleID) + theme_bw()
```

```{r}
otu_resid %>% filter(sampleID == "E01JH0004") %>% 
ggplot() + 
    geom_boxplot(aes(x = samID, y = resid)) + 
    facet_grid(. ~ dilution, scales = "free") + theme_bw()
```

```{r}
otu_resid %>% filter(sampleID == "E01JH0011") %>% 
ggplot() + 
    geom_boxplot(aes(x = samID, y = resid)) + 
    facet_grid(. ~ dilution, scales = "free") + theme_bw()
```

```{r}
otu_resid %>% filter(sampleID == "E01JH0016") %>% 
ggplot() + 
    geom_boxplot(aes(x = samID, y = resid)) + 
    facet_grid(. ~ dilution, scales = "free") + theme_bw()
```

```{r}
otu_resid %>% filter(sampleID == "E01JH0017") %>% 
ggplot() + 
    geom_boxplot(aes(x = samID, y = resid)) + 
    facet_grid(. ~ dilution, scales = "free") + theme_bw()
```

```{r}
otu_resid %>% filter(sampleID == "E01JH0038") %>% 
ggplot() + 
    geom_boxplot(aes(x = samID, y = resid)) + 
    facet_grid(. ~ dilution, scales = "free") + theme_bw()
```

Might want to look at median instead of mean for predicting count values - so that expectation is not impacted by outliers

```{r}
ggplot(otu_resid) + geom_histogram(aes(x = resid))
ggplot(otu_resid) + geom_qq(aes(sample = resid))
```

Low counts not responsible for outliers
```{r}
otu_resid %>% mutate(count_bin = cut_interval(log2(count+ 1), n = 5)) %>% 
    ggplot() + geom_qq(aes(sample = resid, color = count_bin)) + 
    geom_abline(aes(slope = 1, intercept = 0)) +
    facet_wrap(~count_bin, scales = "free") +
    theme_bw()
```

Expected counts
```{r}
otu_resid %>% mutate(count_bin = cut_interval(log2(count_exp + 1), n = 5)) %>% 
    ggplot() + geom_qq(aes(sample = resid, color = count_bin)) + 
    geom_abline(aes(slope = 1, intercept = 0)) +
    facet_wrap(~count_bin, scales = "free") +
    theme_bw()
```

Pre mean
```{r}
otu_resid %>% mutate(count_bin = cut_interval(log2(pre_means + 1), n = 5)) %>% 
    ggplot() + geom_qq(aes(sample = resid, color = count_bin)) + 
    geom_abline(aes(slope = 1, intercept = 0)) +
    facet_wrap(~count_bin, scales = "free") +
    theme_bw()
```

Post Mean
```{r}
otu_resid %>% mutate(count_bin = cut_interval(log2(post_means + 1), n = 5)) %>% 
    ggplot() + geom_qq(aes(sample = resid, color = count_bin)) + 
    geom_abline(aes(slope = 1, intercept = 0)) +
    facet_wrap(~count_bin, scales = "free") +
    theme_bw()
```


### Residuals as a function of the expected counts all
```{r}
otu_resid %>% mutate(sampleOTU = paste(otu, sampleID, sep = "_")) %>% 
    ggplot() + 
    geom_line(aes(x = count_exp, y = resid, group = sampleOTU), color = "grey70") +
    geom_point(aes(x = count_exp, y = resid), alpha = 0.25) + 
    scale_x_continuous(trans = "log2") + scale_y_continuous(trans = "log2") +
    geom_abline(aes(slope = 1, intercept = 0), color = "darkorange") +
    facet_wrap(~sampleID) + theme_bw()
```

### Residuals as a function of the observed counts all
```{r}
otu_resid %>% mutate(sampleOTU = paste(otu, sampleID, sep = "_")) %>% 
    ggplot() + 
    geom_line(aes(x = count, y = resid, group = sampleOTU), color = "grey70") +
    geom_point(aes(x = count, y = resid), alpha = 0.25) + 
    scale_x_continuous(trans = "log2") + scale_y_continuous(trans = "log2") +
    geom_abline(aes(slope = 1, intercept = 0), color = "darkorange") +
    facet_wrap(~sampleID) + theme_bw()
```

Residuals are proportional to the expected and observed count value

### Normalizing by expected value
```{r}
otu_resid %>% 
    mutate(sampleOTU = paste(otu, sampleID,dilution, sep = "_")) %>% 
    ggplot() + 
    geom_line(aes(x = count_exp, y = resid/(count_exp + 1), group = sampleOTU), color = "grey70") +
    geom_point(aes(x = count_exp, y = resid/(count_exp + 1)), alpha = 0.25) + 
    scale_x_continuous(trans = "log2") + 
    #geom_abline(aes(slope = 1, intercept = 0), color = "darkorange") +
    facet_wrap(~sampleID, ncol = 1) + theme_bw()
```

What is the cause of the observed bias and how to correct it?
