---
title: "Feature Filter Data Frame Metrics"
author: "Nate Olson"
date: '`r Sys.Date()`'
output:
  bookdown::pdf_document2: default
  html_document: default
---

```{r echo=FALSE, message=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(DESeq2)
library(phyloseq)
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd)
pipeline_dir <- "../../mgtst_pipelines"
ps <- get_phyloseq(pipeline_dir)
```

##  Objective 
Generate a dataframe with metrics for use in filtering features
Metrics - Median EO, log2FC pre-post, pre-post abundance estimates

```{r}
calc_logfc_deseq <- function(ps_obj){
    ## convert to DESeq object
    dsq <- phyloseq_to_deseq2(ps_obj, ~ t_fctr) 
    
    ## calculate logFC estimates
    DESeq(dsq, test="Wald", fitType="parametric") 
}

tidy_deseq_res <- function(dsq_logfc, biosample_id){
    results(dsq_logfc, cooksCutoff = FALSE) %>% 
        as.data.frame() %>% 
        rownames_to_column(var = "feature_id") %>% 
        add_column(biosample_id = biosample_id)
}

# get_prepost_logfc <- function(ps_obj){
#     logfc_df <- data_frame()
#     
#     biosample_ids <- sample_data(ps_obj)$biosample_id %>% 
#           unique() %>% {.[. != "NTC"]}
#     print(biosample_ids)
#     for(i in biosample_ids){
#             # ps_unmix <- subset_samples(ps_obj, t_fctr %in% c(0,20) & biosample_id == i)
#             # deseq_logfc <- calc_logfc_deseq(ps_unmix)
#             # logfc_df <- tidy_deseq_res(deseq_logfc, i) %>%
#             #     bind_rows(logfc_df)
#     }
# 
#     logfc_df
# }

```

```{r}
get_prepost_logfc <- function(ps_obj){
      logfc_df <- data_frame()
      
      biosample_ids <- sample_data(ps_obj)$biosample_id %>% 
            unique() %>% {.[. != "NTC"]}
      for(i in biosample_ids){
            biosams <- as.character(get_variable(ps_obj, "biosample_id")) == i
            ps_unmix <- prune_samples(biosams, ps_obj) 
            ps_unmix <- subset_samples(ps_unmix, t_fctr %in% c(0,20))
            deseq_logfc <- calc_logfc_deseq(ps_unmix)
            logfc_df <- tidy_deseq_res(deseq_logfc, i) %>%
                  bind_rows(logfc_df)
      }
      
      logfc_df
}

logfc_df <- ps %>% map_df(get_prepost_logfc,.id = "pipe")
```

Sanity Checks
```{r}
logfc_df %>% group_by(biosample_id, pipe) %>% summarise(count = n()) %>% 
      spread(biosample_id, count) %>% knitr::kable()
```

```{r}
logfc_df %>% filter(abs(log2FoldChange) > 1, pvalue < 0.05) %>% 
      group_by(biosample_id, pipe) %>% summarise(count = n()) %>% 
      spread(biosample_id, count) %>% knitr::kable()
```

Will want to look into features in U shape and outliers
```{r}
logfc_df %>% ggplot() + geom_point(aes(x = baseMean + 1, y = log2FoldChange), alpha = 0.25) + 
    facet_grid(pipe~biosample_id) + scale_x_continuous(trans = "log2")
```


```{r}
logfc_df %>% saveRDS("../data/pre_post_deseq_logFC.RDS")
```

Investigating features in U shape - features in U are pre- and post-specific. 
```{r}
top_features <- logfc_df %>% filter(baseMean > 100, baseMean < 200) %>% 
      group_by(biosample_id, pipe) %>% top_n(5, wt = abs(log2FoldChange))
```

```{r}
logfc_df %>% ggplot() + 
      geom_point(aes(x = baseMean + 1, y = log2FoldChange), color = "grey60", alpha = 0.25) + 
      geom_point(data = top_features, aes(x = baseMean + 1, y = log2FoldChange), color = "darkorange") + 
    facet_grid(pipe~biosample_id) + scale_x_continuous(trans = "log2")
```


```{r}
# count_nb <- readRDS("../data/nb_expected_eo_metric_feature_df.rds")
# top_features %>% select(pipe, feature_id, biosample_id) %>% 
#       left_join(count_nb)
```

Investigating Outlier features - no counts for both unmixed pre and post
```{r}
outlier_fc <- logfc_df %>% filter(abs(log2FoldChange) > 18)
```

```{r}
logfc_df %>% ggplot() + 
      geom_point(aes(x = baseMean + 1, y = log2FoldChange), color = "grey60", alpha = 0.25) + 
      geom_point(data = outlier_fc, aes(x = baseMean + 1, y = log2FoldChange), color = "darkorange") + 
    facet_grid(pipe~biosample_id) + scale_x_continuous(trans = "log2")
```

```{r}
# count_nb <- readRDS("../data/nb_expected_eo_metric_feature_df.rds")
# outlier_fc %>% select(pipe, feature_id, biosample_id) %>%
#       left_join(count_nb)
```

Updated MA plot
```{r}
count_nb <- readRDS("../data/nb_expected_eo_metric_feature_df.rds")
count_nb %>% filter(pre != 0, post != 0) %>% 
      select(pipe, biosample_id, feature_id) %>% unique() %>% 
      {semi_join(logfc_df, .)} %>% 
      ggplot() + 
      geom_point(aes(x = baseMean + 1, y = log2FoldChange), alpha = 0.25) + 
    facet_grid(pipe~biosample_id) + scale_x_continuous(trans = "log2") + 
      theme_bw()
```



# Session information 

## Git repo commit information
```{r}
library(git2r)
repo <- repository(path = "../")
last_commit <- commits(repo)[[1]]
```

The current git commit of this file is `r last_commit@sha`, which is on the `r branches(repo)[[1]]@name` branch and was made by `r last_commit@committer@name` on `r when(last_commit)`. The current commit message is `r last_commit@summary`. The repository is online at https://github.com/nate-d-olson/mgtst-pub  


## Platform Information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
```


## Package Versions
```{r}
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
      knitr::kable()
```