---
title: "Relative Abundance Normalization Method Comparison"
author: "Nate Olson"
date: '`r Sys.Date()`'
output:
  bookdown::pdf_document2: 
    toc: FALSE
---

```{r setup, warning=FALSE, message = FALSE, echo = FALSE}
library(nlme)
library(multcomp)
library(tidyverse)
library(ggridges)
norm_count_df <- readRDS("~/Desktop/norm_count_df.RDS") 
pa_summary_anno_df <- readRDS("~/Desktop/to_file/mgtst_RDS/pa_summary_anno_df.RDS")
theta_est <- readRDS("~/Desktop/to_file/mgtst_RDS/bootstrap_theta_estimates.rds")
```

```{r errorRateCalc, message = FALSE, echo = FALSE }
## Calculating Relative Abundance Error Rate
pre_post_prop <- norm_count_df %>% 
      ungroup() %>% 
      filter(t_fctr %in% c(0,20)) %>% 
      mutate(end_point = if_else(t_fctr == 0 , "post", "pre")) %>% 
      select(-t_fctr, -var_count, -sd_count) %>% 
      ## setting values to 0 when one or more of the PCR replicates are 0 for titration end-points
      spread(end_point,mean_count, fill = 0)

prop_inferred <- theta_est %>% 
    filter(pipe == "unclustered") %>% 
    ungroup() %>%
    mutate(t_fctr = factor(t_fctr, levels = c(0:5, 10, 15, 20))) %>% 
    select(biosample_id, theta_hat_mean, t_fctr) %>% 
    right_join(norm_count_df) %>% 
    right_join(pre_post_prop) %>% 
    filter(t_fctr %in% c(1:5,10,15)) %>% 
    ## Using inferred theta estimates to calculate expected values
    mutate(inferred_prop = post * theta_hat_mean + pre * (1 - theta_hat_mean))

## Excluding mix and unmix specific features
## Only including features observed in all or none of the four pre- post- PCR replicates
pa_filter <- pa_summary_anno_df %>% 
      filter(pa_specific == "unspecific") %>% 
      select(biosample_id, pipe, feature_id, full_pre, T00, T20, pa_mixed) %>% 
      filter(T00 %in% c(0,4), T20 %in% c(04))

#### Error Rate Calculations
rel_abu_error <- prop_inferred %>% 
      mutate(t_fctr = factor(t_fctr, levels = c(1:5, 10, 15))) %>% 
      mutate(inferred_error = abs(mean_count - inferred_prop),
             inferred_error_rate = inferred_error/inferred_prop)
```



## Methods
To evalute the impact of different normalization methods the mean-variance relationship and error rates for mothur count tables were compared. 
Normalization methods compared were; 
1. TMM - weighted trim mean of M-values (EdgeR),    
1. RLE - relative log expression (EdgeR),  
1. CSS - cummulative sum scaling (metagenomeSeq),  
1. TSS - total sum scaling, proportions,   
1. UQ - upper quartile scaling.

## Results 
The mean-variance relationship is relevant when making assumptions regarding the distribution the data are drawn from. 
Here we compare the mean relative abundance values to the variance (Fig. \@(fig:meanVarNormComp)) and standard deviation (Fig. \@(fig:meanSdNormComp)) for four PCR replicates (Fig. \@(fig:meanVarNormComp)). 
__Not sure how to interpret these results__

The the error rates were compared across normalization methods (Fig. \@ref(fig:errorDistNormComp)). 
__Distribution comparison statistics??__
The full data were fit using a mixed effects linear model with individual as the random effect and normalization method as a fixed effect. 
As a result of the long tailed distributions the residuals were not normally distributed violating the model assumptions. 
We used a non-parametric rank based analysis to compare normalization methods. 
The median error rates for each individual were compared across normalization methods (Fig. \@ref(fig:errorMedNormComp)). 
TSS had the lowest average rank across individuals followed by CSS. 
TMM and RLE had higher ranks than non-normalized data. 




```{r meanVarNormComp, fig.cap = "Comparison of relative abundance mean and variance relationship for PCR replicates across normalization methods. RLE - relative log expression, TMM - weighted trim mean of M-values, RAW - unnormalized, CSS - cumulative sum scaling, TSS - total sum scaling, UQ - upperquartile. Orange line indicates a 1-to-1 relationship between the relative abundance mean and variance. Blue line is a GAM smoother. Points with means of zero and variance < 1e-10 were excluded from the plot.", message=FALSE, echo = FALSE}
filtered_norm <- norm_count_df %>%
    filter(mean_count != 0, var_count > 1e-10) 

ggplot(filtered_norm) +
    geom_hex(aes(x = mean_count, y = var_count)) + 
    geom_smooth(aes(x = mean_count, y = var_count)) + 
    geom_abline(aes(intercept = 0, slope = 1 ), color = "darkorange") + 
    facet_wrap(~norm_method) + 
    theme_bw() + scale_y_log10() + scale_x_log10() + 
    labs(x = "Mean", y = "Variance") + 
    # coord_equal() + 
    theme(legend.position = "bottom", axis.text.x = element_text(angle = 315))
```

```{r fig.cap = "Same plot as above but with standard deviation instead of variance.", message = FALSE, echo = FALSE}
filtered_norm <- norm_count_df %>%
    filter(mean_count != 0, var_count > 1e-10) 

ggplot(filtered_norm) +
    geom_hex(aes(x = mean_count, y = sd_count)) + 
    geom_smooth(aes(x = mean_count, y = sd_count)) + 
    geom_abline(aes(intercept = 0, slope = 1 ), color = "darkorange") + 
    facet_wrap(~norm_method) + 
    theme_bw() + scale_y_log10() + scale_x_log10() + 
    labs(x = "Mean", y = "Standard Deviation") + 
    theme(legend.position = "bottom", axis.text.x = element_text(angle = 315))
```



```{r errorDistNormComp, fig.cap = "Comparison of relative abundance error rate distribution across normalization methods. RLE - relative log expression, TMM - weighted trim mean of M-values, RAW - unnormalized, CSS - cumulative sum scaling, TSS - total sum scaling, UQ - upperquartile."}
rel_abu_ridge_df <- rel_abu_error %>% 
    filter(t_fctr %in% 1:5, theta_hat_mean > 0) %>% 
    mutate(inferred_error_rate = if_else(inferred_error_rate < 1e-10, 0, inferred_error_rate)) %>%
    filter(inferred_error_rate != 0 & mean_count != 0) %>%
    mutate(inferred_error_rate = if_else(inferred_prop == 0, NaN, inferred_error_rate))

rel_abu_ridge_df %>%
    ggplot() + 
    geom_density_ridges(aes(x =  inferred_error_rate, y = norm_method, color = norm_method), 
                        alpha = 0.5, stat = "binline", bins = 30, draw_baseline = FALSE)  + 
    facet_wrap(~biosample_id, nrow = 1) + theme_bw() + 
    scale_x_log10() +
    labs(x = "Error Rate", y = "Normalization", color = "Normalization") +
    theme(legend.position = "none", axis.text.x = element_text(angle = 315))
```


```{r errorMedNormComp, message = FALSE, echo = FALSE}
rel_abu_med <- rel_abu_ridge_df %>%
group_by(biosample_id, norm_method) %>%
summarise(med_error = median( inferred_error_rate, na.rm = TRUE)) # %>%

rel_abu_med %>% ungroup() %>% 
    mutate(biosample_id = factor(biosample_id)) %>% 
    ggplot(aes(x = biosample_id, y = med_error)) + 
    geom_blank() + 
    geom_path(aes(x = as.numeric(biosample_id), y = med_error, color = norm_method)) +
    geom_point(aes(x = biosample_id, y = med_error, fill = norm_method), shape = 21) + 
    theme_bw() + 
    labs(x = "Individual", y = "Median Error Rate", fill = "Normalization", color = "Normalization") + 
    theme(legend.position = "bottom")
```




```{r errorRankNormComp, echo = FALSE, message = FALSE}
rel_abu_rank <- rel_abu_med %>% 
    group_by(biosample_id) %>% arrange(med_error) %>% mutate(error_rank = 1:n()) %>% 
    group_by(norm_method) %>% summarize(mean_rank = mean(error_rank)) %>%
    add_column(biosample_id = "Rank") %>% 
    rename(med_error = mean_rank)

rel_abu_med <- rel_abu_ridge_df %>%
      group_by(biosample_id, norm_method) %>%
      summarise(med_error = median( inferred_error_rate, na.rm = TRUE))

rel_abu_med %>% bind_rows(rel_abu_rank) %>% 
    spread(biosample_id, med_error) %>% 
    arrange(Rank) %>% 
    knitr::kable(digits = 3)
```


##Error Stats

```{r errorStatsNormComp, echo = FALSE, message = FALSE}
rel_abu_filt <- rel_abu_ridge_df %>% #filter(!norm_method %in% c("RLE","UQ")) %>% 
    filter(!is.na(inferred_error_rate)) %>% 
    mutate(norm_method = factor(norm_method))
    
# Fitting mixed effects model with individual as the fixed effect
error_fit <- nlme::lme(inferred_error_rate ~ norm_method, random =  ~ 1 | biosample_id, 
                       data = rel_abu_filt) 
error_aov <- anova(error_fit)
```

```{r}
plot(error_fit)
```

```{r}
qqnorm(error_fit)
```


```{r}
summary(error_fit)
```

```{r}
error_aov
```


# Session information 

## Git repo commit information
```{r}
library(git2r)
repo <- repository(path = "../")
last_commit <- commits(repo)[[1]]
```

The current git commit of this file is `r last_commit@sha`, which is on the `r branches(repo)[[1]]@name` branch and was made by `r last_commit@committer@name` on `r when(last_commit)`. The current commit message is `r last_commit@summary`. The repository is online at https://github.com/nate-d-olson/mgtst-pub  


## Platform Information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
```


## Package Versions
```{r}
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
      knitr::kable()
```