---
title: "EO Metric Pipeline Biosample Eval"
author: "Nate Olson"
date: '`r Sys.Date()`'
always_allow_html: yes
output:
  pdf_document: default
  html_document: default
---

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
library(tidyverse)
library(git2r)
```



```{r}
feature_exp_df <- readRDS("../data/nb_expected_eo_metric_feature_df.rds") 

genus_exp_df <- readRDS("../data/nb_expected_eo_metric_genus_df.rds") 

exp_df <- bind_rows(feature = feature_exp_df, genus = genus_exp_df, .id = "fLvl")
```

## EO Metric Distribution
Distribution of EO metric for different aggregation levels, PCR replicate (raw), titration - median of PCR replicates, feature - median across all PCR replicates and titrations. 

### Raw metric Distribution
```{r}
exp_df %>% ggplot() + 
      geom_density(aes(x = eo_metric, fill = biosample_id), alpha = 0.25) + 
      facet_grid(fLvl~pipe, scales = "free_y") +
      theme_bw()
```

### titration aggregated metric
```{r}
exp_df %>% group_by(fLvl, pipe, biosample_id, t_fctr, feature_id) %>% 
      summarise(med_teo = median(eo_metric)) %>% 
      ggplot() + 
      geom_density(aes(x = med_teo, fill = biosample_id), alpha = 0.25) + 
      facet_grid(fLvl~pipe, scales = "free_y") +
      theme_bw()
```

### Feature aggregated metric 
At the feature level the median EO metric of 1 has the highest density. 
When aggregating features to the genus level EO metric of 1 is still the highest density but metric values near 0 are more common. 

```{r}
exp_df %>% group_by(fLvl, pipe, biosample_id, feature_id) %>% 
      summarise(med_feo = median(eo_metric)) %>% 
      ggplot() + 
      geom_density(aes(x = med_feo, fill = biosample_id), alpha = 0.25) + 
      facet_grid(fLvl~pipe, scales = "free_y") +
      theme_bw()
```

### Excluding features with Median EO of 1 and -1
When excluding features with a median EO of 1 and -1, the distribution of the EO metrics around 0 tends to skew slightly to the left and right for different biological replicates. 
For example E01JH0038 tends to skew to the left, for all pipelines and feature aggregation level, this indicates that the expected values are consistently less than the observed count values. 

```{r}
exp_df %>% group_by(fLvl, pipe, biosample_id, feature_id) %>% 
      summarise(med_feo = median(eo_metric)) %>% filter(med_feo != 1, med_feo != -1) %>% 
      ggplot() + 
      geom_density(aes(x = med_feo, fill = biosample_id), alpha = 0.25) + 
      facet_grid(fLvl~pipe, scales = "free_y") +
      theme_bw()
```

###  Comparison of EO metric values across pipeline and biological replicates 

Filtering features with median EO metrics of -1 and 1. 
```{r}
eo_med_df <- exp_df %>% group_by(fLvl, pipe, biosample_id, feature_id) %>% 
      summarise(med_feo = median(eo_metric)) %>% filter(med_feo != 1, med_feo != -1)
```

```{r}
eo_med_df %>% ggplot() +
      geom_boxplot(aes(x = biosample_id, y = med_feo, color = pipe)) +
      facet_wrap(~fLvl) +
      theme_bw() +
      theme(axis.text.x = element_text(angle = 90))
```

Testing for biological replicate and pipeline effect. 
Data does not fit a linear model. 
Tried a few different modeling approaches without any success. 
Maybe a non-parametric approach?

```{r}
fit <- lm(med_feo ~  biosample_id * pipe, data = filter(eo_med_df, fLvl == "genus"))
```

```{r}
library(ggfortify)
autoplot(fit) + theme_bw()
```

```{r}
summary(fit)
```




# Session information 

## Git repo commit information
```{r}
repo <- repository(path = "../")
last_commit <- commits(repo)[[1]]
```

The current git commit of this file is `r last_commit@sha`, which is on the `r branches(repo)[[1]]@name` branch and was made by `r last_commit@committer@name` on `r when(last_commit)`. The current commit message is `r last_commit@summary`. The repository is online at https://github.com/nate-d-olson/mgtst-pub  


## Platform Information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
```


## Package Versions
```{r}
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
      knitr::kable()
```