---
title: "logFC Error Rate Analysis"
author: "Nate Olson"
date: '`r Sys.Date()`'
output:
  bookdown::pdf_document2: 
    toc: FALSE
---

```{r setup, include=FALSE}
library(tidyverse)
library(forcats)
library(stringr)
# nb_counts <- readRDS("~/Desktop/nb_counts.RDS")
pa_summary_anno_df <- readRDS("~/Desktop/pa_summary_anno_df.RDS")
theta_est <- readRDS("~/Desktop/bootstrap_theta_estimates.rds")
logFC_edgeR_df <- readRDS("~/Desktop/logFC_edgeR_df.rds")
```

## Objective
Evaluate the logFC estimate accuracy.  

## Summary
The expected logFC between two titrations for pre- and post-specific features is $log_2{\theta_{T1}/\theta_{T2}}$, where $T1$ and $T2$ are the two titrations being compared. 
The inferred theta was used to obtain expected logFC estimates. 
The expectd logFC estimates were compared to logFC estimates calculated using edgeR. 


```{r data_munging, echo = FALSE, warning = FALSE, message = FALSE}
### Summarizing theta across pipelines
theta_est <- theta_est %>% group_by(biosample_id, t_fctr) %>% 
      summarise(theta_hat_mean = mean(theta_hat_mean))

logFC_comp <- logFC_edgeR_df %>% 
      select(pipe, biosample_id, T1, T2) %>% 
      unique() %>% 
      filter(T1 %in% c(0:5, 20), T2 %in% c(0:5,20))

### Obtaining theta estimates
logFC_exp <-logFC_comp %>% 
      mutate(t_fctr = T1) %>% 
      left_join(theta_est) %>% 
      rename(T1_theta = theta_hat_mean) %>% 
      mutate(T1_theta = case_when(T1 %in% c("1","2","3","4","5","10","15") ~ T1_theta,
                                  T1 == "0" ~ 1,
                                  T1 == "20" ~ 0)) %>% 
      # select(-theta, -theta_hat_lci, -theta_hat_uci, -theta_var, -theta_bias) %>%
      mutate(t_fctr = T2) %>%
      left_join(theta_est) %>%
      rename(T2_theta = theta_hat_mean) %>%
      mutate(T2_theta = case_when(T2 %in% c("1","2","3","4","5","10","15") ~ T2_theta,
                                  T2 == "0" ~ 1,
                                  T2 == "20" ~ 0)) %>%
      # select(-t_fctr, -theta, -theta_hat_lci, -theta_hat_uci, -theta_var, -theta_bias) %>% 
      mutate(post_exp_logFC = log2(T2_theta/T1_theta),
             pre_exp_logFC = log2((T2_theta - 1)/(T1_theta - 1)))

### Pre and post specific features
pre_post_features <- pa_summary_anno_df %>% 
      filter(pre_specific == 1 | post_specific == 1) %>% 
      filter(pa_specific == "unspecific") %>% 
      mutate(pre_post = if_else(post_specific == 1, "post","pre")) %>%
      ## Present in over half of the titration PCR replicates
      filter(pa_mixed > 14) %>% 
      select(biosample_id, pipe, feature_id, pre_post)

## Filtering logFC data frame
logFC_comp <- logFC_edgeR_df %>% 
      right_join(logFC_exp) %>%
      select(-Rank1, -Rank2, -Rank3, -Rank4, -Rank5, -Rank6, -Rank7) %>% 
      rename(feature_id = OTUname) %>% 
      right_join(pre_post_features) %>% 
      # Assign expected logFC based on pre- or post- specific 
      mutate(exp_logFC = if_else(pre_post == "pre", pre_exp_logFC, post_exp_logFC))

## Excluding unreasonable expected logFC
logFC_comp <- logFC_comp %>% filter(exp_logFC != -Inf, !is.nan(exp_logFC))

## Calculating Error Rate
logFC_error_df <- logFC_comp %>% 
      mutate(logFC_error = abs(exp_logFC - logFC)/exp_logFC)
```


## Results
Only pre- and post-specific features observed in at least half of the titration PCR replicates were use in the analysis (Table \@ref(tab:featTbl)). 
Very few pre-specific features meet the inclusion criteria and therefore excluded from the following analysis. 
Most of the logFC changes were not significantly different from 0 regardless of feature abundance (logCPM) (Fig. \@ref(fig:MAplot)). 
A 1-to-1 relationship between the logFC estimates and expected values were not observed (Fig. \@ref(fig:logFCestVexp)). 
This lack of a relationship was consistent across individuals, pipelines, and titrations comparisons for features with statistically significant logFC. 
For post-featues the logFC estimates are expected to all be negative as feature relative abundance should decrease with sequential titrations. 
However, both postive and negative logFC estimates were observed.
Some features with a 1-to-1 relationship between the logFC estimates and expected values (Fig. \@ref(fig:exampleFeatLogFC)).
The logFC estimate error rate was consistent across pipelines (excluding E01JH0038) but varied by individual (Fig. \@ref(fig:logFCerrorRate)). 
This indicates the the error rate is more dependent on the individual than pipeline indicating that a bias in the theta estimate is likely skewing the results.




There are a number of potential reasons for the overall lack of agreement between the expected value and estimated logFC. 

* logFC not calculated correctly  
* expected logFC not calculated correctly  
* bug in data munging  
* noisy features dominant  
* poor relationship between theta estimates  
* lack of consistency in direction of logFC estimates



Of the potential reasons, the theta estimates are likely a primary contibutor. 
This is evident with the postive expected logFC values for E01JH0017. 
The E01JH0017 theta estimates are not monotomically decreasing across titrations resulting in positive expected logfC values for post-exposure specific features.


```{r featTbl, echo = FALSE, warning = FALSE, message = FALSE}
pre_post_features %>% 
      group_by(biosample_id, pipe, pre_post) %>% 
      summarise(count = n()) %>% 
      spread(biosample_id, count,fill = 0) %>% 
      arrange(pre_post) %>% 
      knitr::kable(caption = "Summary Table with number of pre and post specific features.", booktabs = TRUE)
```
  

```{r MAplot, fig.cap = "MA plot of logFC estimates between titrations for post-specific features.", echo = FALSE, warning = FALSE, message = FALSE}
logFC_error_df %>% 
      filter(T1 != 0, T2 != 20, pre_post == "post") %>% 
      mutate(FDR_cat = if_else(FDR < 0.05, "<0.05",">0.05")) %>% 
      ggplot() + 
      geom_point(aes(x = logCPM, y = logFC, color = FDR_cat), alpha = 0.25) +
      facet_grid(pipe~biosample_id, scales = "free") + 
      theme_bw() 
```

```{r logFCestVexp, fig.cap = "Relationship between the estimated and observed logFC", , echo = FALSE, warning = FALSE, message = FALSE}
logFC_error_df %>% 
      filter(T1 != 0, T2 != 20, pre_post == "post") %>% 
      mutate(FDR_cat = if_else(FDR < 0.05, "<0.05",">0.05")) %>% 
      filter(biosample_id != "E01JH0017") %>% 
      ggplot() + 
      geom_point(aes(x = exp_logFC, y = logFC)) +
      geom_smooth(aes(x = exp_logFC, y = logFC)) + 
      theme_bw() + facet_wrap(~FDR_cat)
```


```{r exampleFeatLogFC, fig.cap = "Example features with a 1-to-1 relationship between the expected and estimated logFC values. The grey line indicates the expected 1-to-1 relationship. Point shape and color are used to represent the titrations being compared.  Features include those with a logCPM > 10, FDR < 0.05, for E01JH0016, excluding dada2.", echo = FALSE, message = FALSE}
logFC_error_df %>% 
      filter(logCPM > 10, FDR < 0.05, 
             pre_post == "post", 
             biosample_id == "E01JH0016",
             T1 != 0, logFC < 0, pipe != "dada2") %>%  
      ggplot() + 
      geom_point(aes(x = exp_logFC, y = logFC, color = T2, shape = T1)) + 
      geom_line(aes(x = exp_logFC, y = logFC)) + 
      geom_abline(aes(intercept = 0, slope = 1), color = "grey60") + 
      facet_wrap(~feature_id) + theme_bw()
```




```{r logFCerrorRate, fig.cap = "Distribution of logFC error rates for post-specific features by individual and pipeline. Only comparison between titrations (i.e. excluding unmixed samples) with expected logFC values < -1 were included. Orange points are comparisons with FDR < 0.05.", echo = FALSE, warning = FALSE, message = FALSE}
filtered_dat <- logFC_error_df %>% 
      filter(T1 != 0, T2 != 20, pre_post == "post", abs(exp_logFC) > 1) %>% 
      mutate(FDR_cat = if_else(FDR < 0.05, "<0.05",">0.05")) %>% 
      filter(biosample_id != "E01JH0017") 
filtered_dat %>% 
      ggplot() + 
      geom_boxplot(aes(x = pipe, y = logFC_error)) + 
      geom_jitter(data = filter(filtered_dat, FDR_cat == "<0.05"),
                  aes(x = pipe, y = logFC_error), color = "darkorange", width = 0.25) + 
      facet_grid(.~biosample_id, scales = "free") +
      theme_bw()
```


## Git repo commit information
```{r}
library(git2r)
repo <- repository(path = "../")
last_commit <- commits(repo)[[1]]
```

The current git commit of this file is `r last_commit@sha`, which is on the `r branches(repo)[[1]]@name` branch and was made by `r last_commit@committer@name` on `r when(last_commit)`. The current commit message is `r last_commit@summary`. The repository is online at https://github.com/nate-d-olson/mgtst-pub  


## Platform Information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
```


## Package Versions
```{r}
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
      knitr::kable()
```

