---
title: "LIMMA Count Estimates"
author: "Nate Olson"
date: '`r Sys.Date()`'
always_allow_html: yes
output:
  pdf_document: default
  html_document: default
---

## Objective
Calculate count estimates summarizing across PCR replicates.

## Approach
* Normalize counts using `edgeR` function `calcNormFactors` and `cpm` without log transformation 
* Summarized counts using `limma` function `lmFit` 

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(limma)
library(edgeR)
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd) 
pipeline_dir <- "../../mgtst_pipelines"
mrexp <- get_mrexp(pipeline_dir)
```

## Functions for calculating count values
```{r}
### Following functions are used to get summarize counts across titrations.

# get count 
get_norm_counts <- function(mrexp, sample_ids){
      
      ## remove low quality samples and NTC
      count_mat <- mrexp@assayData$counts
      
      ## extract counts
      count_mat <- count_mat[,colnames(count_mat) %in% sample_ids]
      
      ## Normalize count
      dge <- DGEList(counts = count_mat)
      dge <- edgeR::calcNormFactors(dge)
      norm_counts <- cpm(dge, log = FALSE)
      
      ## Return normalized counts
      norm_counts
}


get_titration_model <- function(pdat, sample_ids){
      ## get biosample and titration assignments
      mod_pdat <- pdat %>% 
            filter(id %in% sample_ids) %>% 
            dplyr::select(id, biosample_id, t_fctr) %>% 
            unite(bio_t, biosample_id, t_fctr)

      ## generate design matrix for biological replicates and titrations
      ## for pooling PCR replicates
      bio_t <- factor(mod_pdat$bio_t)
      design <- model.matrix(~bio_t)
      
      ## return design matrix and intercept term
      list(intercept_term = levels(bio_t)[1], mod = design)
}


fit_limma <- function(norm_counts, design){
      ## Fitting model
      lmFit(norm_counts, design)
}


# Extract count estimates
get_count_estimates <- function(limma_fit){
      ## Extract estimates 
      estimate_df <- limma_fit$coefficients %>% 
            as.data.frame() %>% 
            rownames_to_column(var = "feature_id")
      
      ## Adjust estimates for intercept offset
      count_est_df <- estimate_df %>% 
            rename(intercept = `(Intercept)`) %>% 
            gather("group_var", "count_fit", -feature_id, -intercept) %>% 
            mutate(count_est = intercept + count_fit) %>% 
            dplyr::select(-count_fit)
      
      ## Count estimates
      count_est_df
}

tidy_count_estimates <- function(count_est_df, intercept_term){

      ## rename intercept term
      # hard coding intercept term check for consistency with assigned name
      if(intercept_term != "E01JH0004_0"){
            warning("intercept is not consistent with assigned value")
      }

      tidy_count_est <- count_est_df %>% 
            rename(bio_tE01JH0004_0 = intercept) %>% 
            spread(group_var, count_est) %>% 
            gather("group_var", "count_est", -feature_id) %>% 
            ## remove bio_t
            mutate(group_var = str_replace(group_var, "bio_t","")) %>% 
            ## split biosample and t_fctr
            separate(group_var,c("biosample_id","t_fctr"), sep = "_") %>% 
            mutate(t_fctr = factor(t_fctr, levels = c(0:5, 10, 15, 20)))

      ## tidy counts
      tidy_count_est
}

get_norm_counts_limma <- function(mr){
      sample_ids <- pData(mr) %>% filter(biosample_id != "NTC") %>% .$id 
      
      norm_counts <- get_norm_counts(mr, sample_ids)
      
      mod_list <- get_titration_model(pData(mr), sample_ids)
      
      count_fit <- fit_limma(norm_counts, mod_list$mod)
      
      count_est_df <- get_count_estimates(count_fit) %>% 
            tidy_count_estimates(intercept_term = mod_list$intercept_term)
      
      ## return counts
      count_est_df
}

```


### Process Datasets
```{r}
count_est_df <- mrexp %>% map_df(get_norm_counts_limma, .id = "pipe")
```




### Sanity Check Plots
Need to check out large count values - attributing to cpm 
```{r}
count_est_df %>% 
      ggplot() + geom_line(aes(x = t_fctr, y = count_est, group = feature_id)) + 
      facet_grid(pipe~biosample_id)
```


```{r}
count_est_df %>% filter(count_est > 100000) %>% spread(t_fctr, count_est)
```


Very small negative values - attributing to variance in model fit...

```{r}
neg_counts <- filter(count_est_df, count_est < 0)  %>% arrange(count_est)
neg_counts %>% nrow()
```

```{r}
summary(neg_counts$count_est)
```


```{r}
count_est_df
```

### E. coli as sanity check
```{r}
count_est_df %>% filter(feature_id %in% c("SV1","Otu00001", "782953")) %>% 
      ggplot() + geom_point(aes(x = t_fctr, y = count_est)) + 
      facet_grid(pipe~biosample_id, scales = "free_y")
```


### Saving Count Data
```{r}
saveRDS(count_est_df, "../data/limma_count_est.rds")
```

