---
title: "Relative Abundance Error Rate"
author: "Nate Olson"
date: '`r Sys.Date()`'
output:
  bookdown::pdf_document2: 
    toc: FALSE
---

```{r setup, include=FALSE}
library(tidyverse)
library(forcats)
nb_counts <- readRDS("~/Desktop/nb_counts.RDS")
pa_summary_anno_df <- readRDS("~/Desktop/pa_summary_anno_df.RDS")
theta_est <- readRDS("~/Desktop/bootstrap_theta_estimates.rds")
```

The relative abundance error rate was calculated using the titration equation. 
Relative abundance estimates were obtained using a negative binomial model to summarize the observed relative abundance across the four PCR replicates. 
The pre- and post-exposure estimated relative abundance and inferred theta values were used to calculate titration and feature level error rates. 
Only features observed in all four or none of the pre- and post-exposure PCR replicates were included in the analysis. 
There is lower confidence in the relative abundance of a feature in the pre- or post-exposure unmxied samples when the feature is observed in some of the 4 PCR replicates, therefore these features were not included in the error analysis (Table \@ref(tab:relAbuFeatTbl)).

Overall aggreement between the inferred and observed relative abundance was high for all individuals and bioinformatic pipelines (Fig. \@ref(fig:relAbuOvE)). 
For all pipeline the expected relative abundance is greater than the observed relative abundance for relative abundance estiamtes less than 1e-4. 
The deviation from the expected value on the low end varies by biological replicate and pipeline. 
Outliers are observed for all pipelines and individuals. 
For E01JH0016 when processed using mothur and qiime the estimated relative abundance is less than the expected realtive abundance. 

The relative abundance error rate was used to evaluate the quantitative accuracy of the method (Fig. \@ref(fig:relAbuErrorBox)). 
Error rate calculated as follows, $error = \frac{|obs - exp|}{exp}$.
The error rates are summarized (median) across titrations to obtain a feature-level error estimate. 
Large error rates were observed with all pipelines for E01JH0016 (Table  \@ref(tab:relAbuErrorTbl)). 
Error rates were compared across pipelines and individuals using a mixed effects model. 
Features with large error rates, defined as $1.5\times IQR$ from median, were excluded from the analysis inorder to prevent outlier from biasing the comparison. 
When accounting for biological replicate effect DADA2 different from mothur and qiime, qiime and mothur are not different from each other .  
The feature-level median error rate is higher for Mothur and QIIME compared to DADA2


The feature-level error rate variability evaluated using a robust coefficient of variance (RCOV = IQR/median). 
Unlike feature-level error rates, large RCOV was observed for all individuals and pipelines (Table \@ref(tab:relAbuRCOVTbl)). 
Outlier values were also excluded from the RCOV analysis. 
The feature-level RCOV was higher for DADA2 compared to Mothur and QIIME though not significantly (Fig. \@ref(fig:relAbuRCOVBox)). 


```{r data_munge, echo = FALSE, message = FALSE, warning = FALSE}
pre_post_prop <- nb_counts %>% 
      ungroup() %>% 
      filter(t_fctr %in% c(0,20)) %>% 
      mutate(end_point = if_else(t_fctr == 0 , "post", "pre")) %>% 
      select(-t_fctr) %>% 
      ## setting values to 0 when one or more of the PCR replicates are 0 for titration end-points
      spread(end_point,nb_prop, fill = 0)

prop_inferred <- theta_est %>% 
      ungroup() %>%
      mutate(t_fctr = factor(t_fctr, levels = c(0:5, 10, 15, 20))) %>% 
      select(pipe, biosample_id, theta_hat_mean, t_fctr) %>% 
      right_join(nb_counts) %>% right_join(pre_post_prop) %>% 
      filter(t_fctr %in% c(1:5,10,15)) %>% 
      ## Using inferred theta estimates to calculate expected values
      mutate(inferred_prop = post * theta_hat_mean + pre * (1-theta_hat_mean))

## Excluding mix and unmix specific features
## Only including features observed in all or none of the four pre- post- PCR replicates
## Features with relative abundance estimates less than 1e-7, these are features that we would not expect to consistently observe in a PCR replicate for the given sequencing depth
pa_filter <- pa_summary_anno_df %>% 
      filter(pa_specific == "unspecific") %>% 
      select(biosample_id, pipe, feature_id, full_pre, T00, T20, pa_mixed) %>% 
      filter(T00 %in% c(0,4), T20 %in% c(04))

prop_inferred <- prop_inferred %>% 
      right_join(pa_filter) %>% 
      filter(nb_prop > 1e-7)


#### Error Rate Calculations
rel_abu_error <- prop_inferred %>% 
      mutate(t_fctr = factor(t_fctr, levels = c(1:5, 10, 15))) %>% 
      mutate(inferred_error = abs(nb_prop - inferred_prop),
             inferred_error_rate = inferred_error/inferred_prop) 

rel_abu_error_summary <-  rel_abu_error %>% 
      group_by(pipe, biosample_id, feature_id) %>% 
      summarise(median_error = median(inferred_error_rate),
                iqr_error = IQR(inferred_error_rate),
                rcov_error = iqr_error/median_error, 
                mean_error = mean(inferred_error_rate),
                var_error = var(inferred_error_rate),
                cov_error = var_error/mean_error) 

### Error rate boxplot and outlier annotation
error_boxplot <- rel_abu_error %>% group_by(pipe, biosample_id, feature_id) %>% 
      summarise(median_error = median(inferred_error_rate)) %>%
      ggplot() + 
      geom_boxplot(aes(x = pipe, y = median_error), outlier.shape = NA) + 
      facet_wrap(~biosample_id, nrow = 1) + 
      theme_bw() + theme(axis.text.x = element_text(angle = 90)) +
      labs(x = "Individual", y = "Median Error Rate")

## Annotating features as outliers based on boxplot
error_plot_dat <- ggplot_build(error_boxplot)$data[[1]] %>% 
      mutate(pipe = fct_recode(factor(group), 
                               dada2 = "1", 
                               mothur = "2", 
                               qiime = "3",
                               unclustered = "4"),
             biosample_id = fct_recode(PANEL, 
                                       E01JH0004 = "1", 
                                       E01JH0011 = "2", 
                                       E01JH0016 = "3", 
                                       E01JH0017 = "4", 
                                       E01JH0038 = "5"))
outlier_error_dat <- error_plot_dat %>% 
      dplyr::select(ymin, ymax, pipe, biosample_id)

rel_error_outlier_cat <- rel_abu_error_summary %>% 
      left_join(outlier_error_dat) %>% 
      mutate(outlier_cat = if_else(median_error < ymin | median_error > ymax, "outlier","inlier")) 

## Robust COV Analysis
rcov_boxplot <- rel_abu_error_summary %>%
      ggplot() + geom_boxplot(aes(x = pipe, y = rcov_error), outlier.shape = NA) + 
      facet_wrap(~biosample_id, nrow = 1) + 
      theme_bw() + theme(axis.text.x = element_text(angle = 90)) +
      labs(x = "Individual", y = "Robust Error COV") 


## Annotating features as outliers based on boxplot
rcov_plot_dat <- ggplot_build(rcov_boxplot)$data[[1]] %>% 
     mutate(pipe = fct_recode(factor(group), 
                               dada2 = "1", 
                               mothur = "2", 
                               qiime = "3",
                               unclustered = "4"),
             biosample_id = fct_recode(PANEL, 
                                       E01JH0004 = "1", 
                                       E01JH0011 = "2", 
                                       E01JH0016 = "3", 
                                       E01JH0017 = "4", 
                                       E01JH0038 = "5"))
outlier_rcov_dat <- rcov_plot_dat %>% 
      dplyr::select(ymin, ymax, pipe, biosample_id)

rcov_outlier_cat <- rel_abu_error_summary %>% 
      left_join(outlier_rcov_dat) %>% 
      mutate(outlier_cat = if_else(median_error < ymin | median_error > ymax, "outlier","inlier"))
```

```{r relAbuFeatTbl, echo = FALSE, warning = FALSE, message = FALSE}
rel_abu_error_summary %>% 
      group_by(pipe, biosample_id) %>% 
      summarise(count = n()) %>% 
      spread(biosample_id, count) %>% 
      knitr::kable(booktabs = TRUE, caption = "Number of features by pipeline and individual used in the relative abundance error rate analysis.")
```

```{r relAbuOvE, fig.cap = "Expected and observed count relationship. Orange line indicates expected 1-to-1 relationship. Blue line a smoothed regression line of the observed and expected value relationship.", echo = FALSE, warning=FALSE, message = FALSE}
prop_inferred %>% 
      ggplot() + 
      geom_point(aes(x = nb_prop, y = inferred_prop), alpha = 0.15) +
      geom_abline(aes(intercept = 0, slope = 1), color = "darkorange") +
      geom_smooth(aes(x = nb_prop, y = inferred_prop)) +
      facet_grid(pipe~biosample_id) +
      scale_y_log10() + scale_x_log10() +
      theme_bw() +
      labs(x = "Observed Relative Abundance", y = "Expected Relative Abundance")
```

```{r relAbuErrorTbl, echo = FALSE, warning = FALSE, message = FALSE}
rel_abu_error_summary %>% 
      group_by(pipe, biosample_id) %>% 
      summarise(med_med_error = median(median_error), 
                min_med_error = min(median_error), 
                max_med_error = max(median_error)) %>% 
      select(pipe, biosample_id, max_med_error) %>% 
      spread(pipe, max_med_error) %>% 
      knitr::kable(booktabs = TRUE, caption = "Maximum feature-level error rate for each pipeline and individual.")
      
```


```{r relAbuErrorBox, fig.cap = "Distribution of feature-level relative abundance error rates by individual and pipeline.", echo = FALSE, warning = FALSE, message = FALSE}
ymin <- ggplot_build(error_boxplot)$data[[1]]$ymin %>% min()
ymax <- ggplot_build(error_boxplot)$data[[1]]$ymax %>% max()


error_boxplot + coord_cartesian(ylim = c(ymin, ymax))
```




```{r relAbuRCOVTbl, echo = FALSE, warning = FALSE, message = FALSE}
rel_abu_error_summary %>% 
      group_by(pipe, biosample_id) %>% 
      summarise(med_rcov_error = median(rcov_error), 
                min_rcov_error = min(rcov_error), 
                max_rcov_error = max(rcov_error)) %>% 
      select(pipe, biosample_id, max_rcov_error) %>% 
      spread(pipe, max_rcov_error) %>% 
      knitr::kable(booktabs = TRUE, caption = "Maximum feature-level robust COV error for each pipeline and individual.")
```


```{r relAbuRCOVBox, fig.cap = "Distribution of feature-level relative abundance error rates by individual and pipeline.", echo = FALSE, warning = FALSE, message = FALSE}
ymin <- ggplot_build(rcov_boxplot)$data[[1]]$ymin %>% min()
ymax <- ggplot_build(rcov_boxplot)$data[[1]]$ymax %>% max()


rcov_boxplot + coord_cartesian(ylim = c(ymin, ymax))
```

\pagebreak  

### Relative abundance error Statistics
```{r}
error_fit <- nlme::lme(median_error ~ pipe, random =  ~ 1 | biosample_id, 
                 data = rel_error_outlier_cat %>% filter(outlier_cat == "inlier"))
summary(error_fit)
```

Pipeline larger contribution to variance than individual
```{r}
ape::varcomp(error_fit)
```

__Not sure about the fit__
```{r}
plot(error_fit)

```


```{r}
qqnorm(error_fit, ~ranef(., level=1))
```

```{r}
qqnorm(error_fit$residuals)
```


```{r}
plot(error_fit$fitted, error_fit$residuals)
```

\pagebreak  

### RCOV Error Statistics

```{r}
rcov_fit <- nlme::lme(rcov_error ~ pipe, random =  ~ 1 | biosample_id, 
                 data = rcov_outlier_cat %>% filter(outlier_cat == "inlier"))
summary(rcov_fit)
```

Pipeline larger contribution to variance than individual
```{r}
ape::varcomp(rcov_fit)
```

```{r}
plot(rcov_fit)
```


```{r}
qqnorm(rcov_fit, ~ranef(., level=1))
```

```{r}
qqnorm(rcov_fit$residuals)
```


```{r}
plot(rcov_fit$fitted, rcov_fit$residuals)
```
