---
title: "EO Metric Figures"
author: "Nate Olson"
date: "4/25/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r}
count_df <- readRDS("../data/nb_expected_eo_metric_feature_df.rds")
```

## Relationship between expected and observed count for all features and PCR replicates
```{r}
count_df %>% 
      filter(count >= 1 | exp_count >= 1) %>% 
      ggplot() + 
      geom_hex(aes(x = count + 1, y = exp_count + 1)) + 
      geom_smooth(aes(x = count + 1, y = exp_count + 1), color = "darkorange") + 
     facet_grid(pipe~biosample_id)+ theme_bw() + 
      labs(y = "Expected Count", x = "Observed Count", fill = "Abundance") + 
      scale_y_continuous(trans = "log2") + 
      scale_x_continuous(trans = "log2") + 
      theme(axis.text.x = element_text(angle = 270))
```

## EO Metric Distribution 
For most PCR replicates the expected or observed counts are 0. 
```{r}
count_df %>% filter(count >= 1 | exp_count >= 1) %>% 
      ggplot() + geom_density(aes(x = eo_metric)) + theme_bw()
```

The EO-metric can be used to obtain a feature level summary by taking the median EO metric value across PCR replicates and titrations. 
```{r}
feature_lvl_eo <- count_df %>% filter(count != 0 | exp_count != 0) %>% 
      group_by(biosample_id, pipe, feature_id) %>% 
      summarise(med_eo = median(eo_metric))
```


Can filter features that are not behaving as expected, present only in unmixed or mixed samples by excluding features with a median EO metric of -1 or 1. 
```{r}
feature_lvl_eo %>% filter(abs(med_eo) != 1) %>% 
      ggplot() + geom_boxplot(aes(x = pipe, y = med_eo)) + 
      facet_wrap(~biosample_id) + theme_bw() +
      labs(x = "Pipeline", y = "Median EO-Metric")
```

```{r}
feature_lvl_eo %>% filter(abs(med_eo) != 1) %>% 
      ggplot() + geom_boxplot(aes(x = biosample_id, y = med_eo)) + 
      theme_bw() +  
      labs(x = "Biological Replicate", y = "Median EO-Metric")
```


```{r}
feature_lvl_eo %>% filter(abs(med_eo) != 1) %>% 
      ggplot() + geom_boxplot(aes(x = pipe, y = med_eo)) + 
      theme_bw() + theme(axis.text.x = element_text(angle = 90)) + 
      labs(x = "Pipeline", y = "Median EO-Metric")
```

```{r}
fit <- lm(med_eo ~ biosample_id, data = feature_lvl_eo %>% filter(abs(med_eo) != 1, pipe == "dada2"))
```

```{r}
summary(fit)
```

```{r}
library(ggfortify)
autoplot(fit)
```

