---
title: "Feature Expected - Observed Relationship Analysis"
author: "Nate Olson"
date: '`r Sys.Date()`'
always_allow_html: yes
output:
  pdf_document: default
  html_document: default
---
```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd)
```

## Correlating Fit characteristics with feature categories 
```{r}
count_exp_fit_tidy <- readRDS("../data/expected_count_values_fit_features_tidy.rds") 
feature_cat <- readRDS("../data/feature_categories_df.rds")
```

```{r}
count_fit_anno <- left_join(count_exp_fit_tidy, feature_cat)
```

```{r}
count_fit_anno_wide <- count_fit_anno %>% 
      select(pipe, biosample_id, feature_id, term, estimate, cat) %>% 
      spread(term, estimate)
```


```{r}
count_fit_anno_wide %>% filter(cat != "cat_none", cat != "cat_null", cat != "cat_mix") %>% 
      ggplot() + 
      geom_boxplot(aes(x = pipe, y = Slope, color = pipe)) + 
      facet_wrap(~cat, scales = "free") + theme_bw()
```


```{r}
count_fit_anno_wide %>% filter(cat != "cat_none", cat != "cat_null", cat != "cat_mix") %>% 
      ggplot() + 
      geom_boxplot(aes(x = pipe, y = Intercept, color = pipe)) + 
      facet_wrap(~cat, scales = "free") + theme_bw()
```

```{r}
count_fit_anno_wide %>% filter(cat != "cat_none", cat != "cat_null", cat != "cat_mix") %>% 
      ggplot() + geom_density2d(aes(x = Intercept, y = Slope, color = cat))
```

Intercept as rows and slope at columns
```{r}
count_fit_anno_wide %>% filter(cat != "cat_null", cat != "cat_mix") %>% 
      mutate(good_int = if_else(abs(Intercept) < 1, 1,0), 
                               good_slope = if_else(abs(Slope) < 1, 1, 0)) %>% 
      group_by(pipe, biosample_id, cat, good_int, good_slope) %>% 
      #summarise(count = n()) %>% 
      ggplot() + geom_bar(aes(x = pipe, fill = cat)) + 
      facet_grid(good_int~good_slope)
```


```{r}
count_exp_df <-readRDS("../data/expected_count_values_feature_df.rds")
feature_cat <- readRDS("../data/feature_categories_df.rds")
```

```{r}
count_exp_df %>% filter(obs_count != 0) %>% ggplot() + geom_density(aes(x = obs_count))
```


```{r}
full_subset <- count_exp_df %>% left_join(count_fit_anno_wide) %>% 
      filter(cat == "cat_pre", pipe == "dada2", biosample_id == "E01JH0004", abs(Slope) < 0.25) 
```
```{r}
ggplot(full_subset) + geom_point(aes(x = exp_count, y = obs_count, color = factor(titration))) + 
      facet_wrap(~feature_id, scales = "free") + geom_abline(aes(slope = 1, intercept = 0))
```

```{r}
full_subset <- count_exp_df %>% left_join(count_fit_anno_wide) %>% 
      filter(cat == "cat_pre", biosample_id == "E01JH0004", abs(Slope) < 0.25, abs(Intercept) < 1) 
```
```{r}
ggplot(full_subset) + geom_point(aes(x = exp_count, y = obs_count, color = factor(titration))) + 
      facet_wrap(~feature_id, scales = "free") + geom_abline(aes(slope = 1, intercept = 0))
```

```{r}
count_exp_df %>% left_join(count_fit_anno_wide) %>% filter(feature_id == "SV1") %>% 
ggplot() + 
      geom_point(aes(x = exp_count, y = obs_count, color = factor(titration))) + 
      geom_abline(aes(slope = 1, intercept = 0)) + 
      facet_wrap(~biosample_id, scales = "free") +
      theme_bw() 
```

```{r}
count_fit_anno_wide %>% filter(feature_id == "SV1")
```


```{r}
count_fit_anno_wide %>% filter(feature_id == "SV2") 
```



```{r}
count_exp_df %>% left_join(count_fit_anno_wide) %>% filter(feature_id == "SV2") %>% 
ggplot() + 
      geom_point(aes(x = exp_count, y = obs_count, color = factor(titration))) + 
      geom_abline(aes(slope = 1, intercept = 0)) + 
      facet_wrap(~biosample_id, scales = "free") +
      theme_bw() + coord_fixed()
```

```{r}
count_fit_anno_wide %>% filter(feature_id == "Otu00001") 
```

```{r}
count_exp_df %>% left_join(count_fit_anno_wide) %>% filter(feature_id == "Otu00001") %>% 
ggplot() + 
      geom_point(aes(x = exp_count, y = obs_count, color = factor(titration))) + 
      geom_abline(aes(slope = 1, intercept = 0)) + 
      facet_wrap(~biosample_id) +
      theme_bw() + coord_equal()
```



```{r}
count_exp_df %>% left_join(count_fit_anno_wide) %>% filter(feature_id == "Otu00002") %>% 
ggplot() + 
      geom_point(aes(x = exp_count, y = obs_count, color = factor(titration))) + 
      geom_abline(aes(slope = 1, intercept = 0)) + 
      facet_wrap(~biosample_id) +
      theme_bw() + coord_equal()
```

```{r}
count_fit_anno_wide %>% filter(feature_id == "Otu00002") 
```

```{r}
count_exp_df %>% left_join(count_fit_anno_wide) %>% filter(cat == "cat_full") %>% 
      mutate(exp_bin = cut_interval(log10(exp_count + 1), 10)) %>% 
      ggplot() + geom_bar(aes( x= exp_bin)) + 
      theme(axis.text.x = element_text(angle = 90))
```

```{r}
count_exp_df %>% left_join(count_fit_anno_wide) %>% filter(cat == "cat_full") %>% 
      mutate(exp_bin = cut_interval(log10(obs_count + 1), 10)) %>% 
      ggplot() + geom_bar(aes( x= exp_bin)) + 
      theme(axis.text.x = element_text(angle = 90))
```


```{r}
dat <- count_exp_df %>% filter(feature_id == "SV1", biosample_id == "E01JH0038") 
lm(obs_count ~ exp_count, data = dat) %>% summary()
```

```{r}

lm(residual ~ t_fctr, data = dat) %>% summary()
```


```{r}
dat <- dat %>% mutate(error = residual/exp_count) 
lm(error ~ t_fctr, data = dat) %>% summary()
```


```{r}
dat2 <- dat %>% filter(t_fctr != 4)
lm(error ~ t_fctr, data = dat2) %>% summary()
```

```{r}
dat2 <- dat %>% filter(t_fctr != 4, t_fctr != 5)
lm(error ~ exp_count, data = dat2) %>% summary()
```

```{r}
ggplot(dat) + 
      geom_path(aes(x = exp_count, y = obs_count, color = t_fctr)) + 
      geom_point(aes( x = exp_count, y = obs_count, color = t_fctr, shape = pcr_rep)) + 
      geom_abline(aes(intercept = 0, slope = 1)) + coord_equal() + theme_bw()
```

```{r}
ggplot(dat) + 
      geom_path(aes(x = exp_count, y = residual, color = t_fctr)) + 
      geom_point(aes( x = exp_count, y = residual, color = t_fctr, shape = pcr_rep)) + 
      geom_hline(aes(yintercept = 0)) + theme_bw()
```

```{r}
ggplot(dat) + 
      geom_path(aes(x = exp_count, y = error, color = t_fctr)) + 
      geom_point(aes( x = exp_count, y = error, color = t_fctr, shape = pcr_rep)) + 
      geom_hline(aes(yintercept = 0)) + theme_bw()
```

R^2 for observed and expected with a slope of 1 and an intercept of 0 

For the assumed model only 35% of the total variability is explained.
```{r}
dat %>% mutate(ybar = mean(obs_count)) %>% 
      group_by(pipe, biosample_id, feature_id) %>% 
      summarise(ss_total = sum((obs_count - mean(obs_count))^2),
                ss_res = sum(residual^2)) %>% 
      mutate(r_squared = 1 - ss_res/ss_total)
```

```{r}
count_exp_df %>% filter(feature_id == "SV1") %>%
      mutate(ybar = mean(obs_count)) %>% 
      group_by(pipe, biosample_id, feature_id) %>% 
      summarise(ss_total = sum((obs_count - mean(obs_count))^2),
                ss_res = sum(residual^2)) %>% 
      mutate(r_squared = 1 - ss_res/ss_total)
```


```{r}
r2_df <- count_exp_df %>% #filter(pipe == "mothur") %>% 
      group_by(pipe, biosample_id, feature_id) %>% 
      summarise(ss_total = sum((obs_count - mean(obs_count))^2),
                ss_res = sum(residual^2)) %>% 
      mutate(r_squared = 1 - ss_res/ss_total)
```


```{r}
r2_df %>% arrange(desc(r_squared))
```

```{r}
dat <- count_exp_df %>% filter(pipe == "mothur", biosample_id == "E01JH0038", feature_id == "Otu00027")
ggplot(dat) + 
      geom_path(aes(x = exp_count, y = obs_count, color = t_fctr)) + 
      geom_point(aes( x = exp_count, y = obs_count, color = t_fctr, shape = pcr_rep)) + 
      geom_abline(aes(intercept = 0, slope = 1)) + coord_equal() + theme_bw()
```

```{r}
dat <- count_exp_df %>% filter(pipe == "mothur", biosample_id == "E01JH0038", feature_id == "Otu00103")
ggplot(dat) + 
      geom_path(aes(x = exp_count, y = obs_count, color = t_fctr)) + 
      geom_point(aes( x = exp_count, y = obs_count, color = t_fctr, shape = pcr_rep)) + 
      geom_abline(aes(intercept = 0, slope = 1)) + coord_equal() + theme_bw()
```

```{r}
r2_df %>% filter(r_squared > 0) %>% 
      ggplot() + geom_density(aes(x = r_squared))
```

```{r}
r2_df %>% left_join(feature_cat) %>% filter(r_squared > 0) %>% 
      ggplot() +
      geom_boxplot(aes(x = cat, y = r_squared))
```

```{r}
r2_df %>% left_join(feature_cat) %>% filter(!is.na(r_squared)) %>% 
      mutate(r2_bin = if_else(r_squared > 0, "pos","neg")) %>% 
      ggplot() +
      geom_bar(aes(x = cat,fill = r2_bin), position = "fill")
```

```{r}
genus_exp <- readRDS("../data/expected_count_values_genus_df.rds") 
genus_cat <- readRDS("../data/genus_categories_df.rds")
```

```{r}
genus_r2_df <- genus_exp %>% # filter(pipe == "mothur") %>% 
      group_by(pipe, biosample_id, feature_id) %>% 
      summarise(ss_total = sum((obs_count - mean(obs_count))^2),
                ss_res = sum(residual^2)) %>% 
      mutate(r_squared = 1 - ss_res/ss_total)
```
 
Aggregating to the genus level, a greater proportion of the full, pre, and post features have postive R^2 values. 

```{r}
genus_r2_df %>% left_join(genus_cat) %>% filter(!is.na(r_squared)) %>% 
      mutate(r2_bin = if_else(r_squared > 0, "pos","neg")) %>% 
      ggplot() +
      geom_bar(aes(x = cat,fill = r2_bin), position = "fill") + facet_wrap(~pipe) +
      theme(axis.text.x = element_text(angle = 90))
```

```{r}
genus_r2_df %>% left_join(genus_cat) %>% filter(r_squared > 0, cat != "cat_null") %>% 
      ggplot() +
      geom_histogram(aes(x = r_squared)) + theme_bw() + facet_grid(pipe~cat)
```

```{r}
genus_r2_df %>% left_join(genus_cat) %>% 
      filter(!is.na(r_squared), cat != "cat_null", cat != "cat_mix") %>%
      ungroup() %>% mutate(r_sq = if_else(r_squared < -4, -4, r_squared), 
                           r2_bin = cut_interval(r_sq, 10)) %>% 
      ggplot() + geom_bar(aes(x = cat, fill = r2_bin), position = "fill") + coord_flip()
```

```{r}
genus_r2_df %>% left_join(genus_cat) %>% 
      filter(r_squared > 0, cat != "cat_null", cat != "cat_mix") %>%
      ungroup() %>% mutate(r_sq = if_else(r_squared < -4, -4, r_squared), 
                           r2_bin = cut_interval(r_sq, 5)) %>% 
      ggplot() + geom_bar(aes(x = cat, fill = r2_bin), position = "fill") + coord_flip()
```

```{r}
genus_r2_df %>% left_join(genus_cat) %>% 
      filter(r_squared > 0, cat != "cat_null", cat != "cat_mix") %>%
      ungroup() %>% mutate(r_sq = if_else(r_squared < -4, -4, r_squared), 
                           r2_bin = cut_interval(r_sq, 5)) %>% 
      ggplot() + geom_bar(aes(x = cat, fill = r2_bin)) + coord_flip()
```

```{r}
rep_features <- genus_r2_df %>% left_join(genus_cat) %>% 
      filter(r_squared > 0, cat != "cat_null", cat != "cat_mix") %>%
      ungroup() %>% mutate(r2_bin = cut_interval(r_squared, 3)) %>% 
      group_by(cat, r2_bin) %>% sample_n(1)
```

```{r}
r2_bins <- rep_features$r2_bin %>% levels() 
```

```{r} 
print(r2_bins[1])
rep_features %>% left_join(genus_exp) %>% filter(r2_bin == r2_bins[1]) %>% 
       ggplot() +
      geom_path(aes(x = exp_count, y = obs_count, color = t_fctr)) + 
      geom_point(aes( x = exp_count, y = obs_count, color = t_fctr, shape = pcr_rep)) + 
      geom_abline(aes(intercept = 0, slope = 1)) + coord_equal() + 
      theme_bw() + facet_wrap(~cat, scales = "free")
```

```{r}
print(r2_bins[2])
rep_features %>% left_join(genus_exp) %>% filter(r2_bin == r2_bins[2]) %>% 
       ggplot() +
      geom_path(aes(x = exp_count, y = obs_count, color = t_fctr)) + 
      geom_point(aes( x = exp_count, y = obs_count, color = t_fctr, shape = pcr_rep)) + 
      geom_abline(aes(intercept = 0, slope = 1)) + coord_equal() + 
      theme_bw() + facet_wrap(~cat, scales = "free")
```

```{r}
print(r2_bins[3])
rep_features %>% left_join(genus_exp) %>% filter(r2_bin == r2_bins[3]) %>% 
       ggplot() +
      geom_path(aes(x = exp_count, y = obs_count, color = t_fctr)) + 
      geom_point(aes( x = exp_count, y = obs_count, color = t_fctr, shape = pcr_rep)) + 
      geom_abline(aes(intercept = 0, slope = 1)) + coord_equal() + 
      theme_bw() + facet_wrap(~cat, scales = "free")
```

```{r}
genus_info <- readRDS("../data/genus_info_df.rds")

genus_info %>%  
      left_join(genus_r2_df) %>% mutate(r_sq = if_else(r_squared < -4, -4, r_squared)) %>% 
      ggplot() + 
      geom_point(aes(x = prop_sum, y = r_sq)) + 
      geom_smooth(aes(x = prop_sum, y = r_sq)) + facet_grid(biosample_id~pipe)
```

```{r}
genus_info %>%  
      left_join(genus_r2_df) %>% mutate(r_sq = if_else(r_squared < -4, -4, r_squared)) %>% 
      ggplot() + 
      geom_boxplot(aes(x = factor(prop_max), y = r_sq)) + facet_grid(biosample_id~pipe)
```

```{r}
genus_info %>%  select(-starts_with("cat")) %>% 
      filter(prop_sum > 0, T0 != 0 & T20 != 0) %>%  
      left_join(genus_r2_df) %>% select(-starts_with("ss_"), -prop_sum) %>% 
      mutate(r_squared = if_else(r_squared < -1, -1, r_squared)) %>% 
      gather("parameter", "value", -pipe, -biosample_id, -feature_id, -r_squared) %>% 
      mutate(value = factor(value)) %>% 
      ggplot() + geom_boxplot(aes(x = value, y = r_squared)) + facet_wrap(~parameter, scales = "free")
```
T0 rows, T20 columns
```{r}
genus_info %>%  select(-starts_with("cat")) %>% 
      filter(prop_sum > 0, T0 != 0 & T20 != 0) %>%  
      left_join(genus_r2_df) %>% select(-starts_with("ss_"), -prop_sum) %>% 
      mutate(r_squared = if_else(r_squared < -1, -1, r_squared)) %>% 
      ggplot() + geom_histogram(aes(x = r_squared)) + facet_grid(T0~T20, scales = "free_y")
```

Even when a feature is observed in all PCR replicates for all samples, the R^2 can be low and largly negative.
```{r}
genus_info %>%  select(-starts_with("cat")) %>% 
      filter(prop_sum > 0, T0 != 0 & T20 != 0) %>%  
      left_join(genus_r2_df) %>% select(-starts_with("ss_"), -prop_sum) %>% 
      mutate(r_squared = if_else(r_squared < -1, -1, r_squared)) %>% 
      ggplot() + geom_histogram(aes(x = r_squared)) + facet_grid(prop_max~prop_min, scales = "free_y")
```

```{r}
genus_logFC <- genus_exp %>% filter(post_count != 0 & pre_count != 0) %>% 
      select(pipe, biosample_id, feature_id, pcr_rep, post_count, pre_count) %>% 
      unique() %>% 
      group_by(pipe, biosample_id, feature_id) %>% 
      summarise(logFC = log2(median(post_count)/median(pre_count)), 
                             total_unmix = median(post_count) + median(pre_count))
```

Need to figure out what experimental factors can be attribued to negative R^2 values and which are due to biases/ errors in the 16S sequence analysis methods. 

```{r}
genus_logFC %>% right_join(genus_r2_df) %>% filter(!is.na(logFC)) %>% 
      mutate(r_squared = if_else(r_squared < -1, -1, r_squared)) %>% 
      ggplot() + geom_point(aes(y = r_squared, x = logFC))
```

```{r}
genus_logFC %>% right_join(genus_r2_df) %>% filter(!is.na(logFC)) %>% 
      mutate(r_squared = if_else(r_squared < -1, -1, r_squared)) %>% 
      ggplot() + geom_point(aes(y = r_squared, x = total_unmix)) + scale_x_log10()
```


```{r}
library(rpart)
```
```{r}

dat <- genus_logFC %>% right_join(genus_r2_df) %>% filter(!is.na(logFC)) %>% 
      left_join(genus_info) %>%  select(-starts_with("cat")) %>% 
      mutate(r2_bin = if_else(r_squared > 0, "POS","NEG"))
dat
fit <- rpart(r2_bin ~ logFC + prop_min + prop_max,
  	method="class", data=dat)

```

```{r}
printcp(fit)
```

```{r}
plotcp(fit)
```

```{r}
summary(fit)
```

```{r}
plot(fit)
text(fit, use.n=TRUE, all=TRUE, cex=.8)
```


## General Approach 
Example features with high R^2 for expected and observed counts, for full, pre, and post 
Example features with low R^2 for expectd and observed count, for full, pre, and post 

## $R^2$ for Observed and Expected counts
$R^2$ value defines how much of the variability in the observed count data for the titrations can be attributed to the expected relationship between the unmixed sample counts and the titration factors. 
The $R^2$ value is expected to range from 0 to 1. 
Negative values were observed indicating that the expected values defiate from the observed count values to a greater extent than the variability observed between the mixed sample count values. 

## Example Features 

### Full Features
Features with observed counts for all PCR replicates of both unmixed and mixed samples. 
Representative features were choosen based on $R^2$ values, top features had the highest $R^2$ and bottom features had the lowest $R^2$ value. 
```{r}
top_full <- genus_r2_df %>% left_join(genus_info) %>% 
      filter(!is.na(r_squared)) %>% 
      filter(prop_sum == 9) %>% group_by(pipe) %>% top_n(1, r_squared)
top_full %>% select(pipe, biosample_id, feature_id, r_squared)
```

```{r}
top_full %>% left_join(genus_exp) %>% 
      ggplot() + 
      geom_path(aes(x = exp_count, y = obs_count, color = t_fctr)) + 
      geom_point(aes( x = exp_count, y = obs_count, color = t_fctr, shape = pcr_rep)) + 
      geom_abline(aes(intercept = 0, slope = 1)) + coord_equal() + 
      theme_bw() + facet_wrap(~pipe, scales = "free")
```

```{r}
bottom_full <- genus_r2_df %>% left_join(genus_info) %>% 
      filter(!is.na(r_squared)) %>% 
      filter(prop_sum == 9) %>% group_by(pipe) %>% top_n(1, -r_squared)
bottom_full %>% select(pipe, biosample_id, feature_id, r_squared)
```

For the features present in all samples and PCR replicates with negative $R^2$ are likely due to artifcats in the feature inference (clustering) proceedure.  
```{r}
bottom_full %>% left_join(genus_exp) %>% 
      ggplot() + 
      geom_path(aes(x = exp_count, y = obs_count, color = t_fctr)) + 
      geom_point(aes( x = exp_count, y = obs_count, color = t_fctr, shape = pcr_rep)) + 
      geom_abline(aes(intercept = 0, slope = 1)) + coord_equal() + 
      theme_bw() + facet_wrap(~pipe, scales = "free")
```


```{r}
genus_r2_df %>% left_join(genus_info) %>% 
      filter(!is.na(r_squared)) %>% ungroup() %>% 
      mutate(r_squared = if_else(r_squared < -1, -1, r_squared), 
             r2_bin = cut_interval(r_squared, n = 5)) %>% 
      filter(prop_sum == 9) %>% group_by(pipe) %>% 
      ggplot() + geom_bar(aes(x = r2_bin, fill = r2_bin), position = "dodge") + 
      facet_wrap(~pipe) + coord_flip() + theme_bw()
``` 

Want to figure out what are the characteristics of the features that results in the low $R^2$.

### Pre Features
```{r}
top_pre <- genus_r2_df %>% left_join(genus_info) %>% 
      filter(!is.na(r_squared)) %>% 
      filter(prop_sum == 8, T0 == 0, T20 == 1) %>% group_by(pipe) %>% top_n(1, r_squared)
top_pre %>% select(pipe, biosample_id, feature_id, r_squared)
```

```{r}
top_pre %>% left_join(genus_exp) %>% 
      ggplot() + 
      geom_path(aes(x = exp_count, y = obs_count, color = t_fctr)) + 
      geom_point(aes( x = exp_count, y = obs_count, color = t_fctr, shape = pcr_rep)) + 
      geom_abline(aes(intercept = 0, slope = 1)) + coord_equal() + 
      theme_bw() + facet_wrap(~pipe, scales = "free")
```

```{r}
bottom_pre <- genus_r2_df %>% left_join(genus_info) %>% 
      filter(!is.na(r_squared)) %>% 
      filter(prop_sum == 8, T0 == 0, T20 == 1) %>% group_by(pipe) %>% top_n(1, -r_squared)
bottom_pre %>% select(pipe, biosample_id, feature_id, r_squared)
```

For the features present in all samples and PCR replicates with negative $R^2$ are likely due to artifcats in the feature inference (clustering) proceedure.  
```{r}
bottom_pre %>% left_join(genus_exp) %>% 
      ggplot() + 
      geom_path(aes(x = exp_count, y = obs_count, color = t_fctr)) + 
      geom_point(aes( x = exp_count, y = obs_count, color = t_fctr, shape = pcr_rep)) + 
      geom_abline(aes(intercept = 0, slope = 1)) + coord_equal() + 
      theme_bw() + facet_wrap(~pipe, scales = "free")
```

### Post Features
```{r}
top_post <- genus_r2_df %>% left_join(genus_info) %>% 
      filter(!is.na(r_squared)) %>% 
      filter(prop_sum > 4, T0 == 1, T20 == 0) %>% group_by(pipe) %>% top_n(1, r_squared)
top_post %>% select(pipe, biosample_id, feature_id, r_squared)
```

```{r}
top_post %>% left_join(genus_exp) %>% 
      ggplot() + 
      geom_path(aes(x = exp_count, y = obs_count, color = t_fctr)) + 
      geom_point(aes( x = exp_count, y = obs_count, color = t_fctr, shape = pcr_rep)) + 
      geom_abline(aes(intercept = 0, slope = 1)) + coord_equal() + 
      theme_bw() + facet_wrap(~pipe, scales = "free")
```

```{r}
bottom_post <- genus_r2_df %>% left_join(genus_info) %>% 
      filter(!is.na(r_squared)) %>% 
      filter(prop_sum > 4, T0 == 1, T20 == 0) %>% group_by(pipe) %>% top_n(1, -r_squared)
bottom_post %>% select(pipe, biosample_id, feature_id, r_squared)
```

For the features present in all samples and PCR replicates with negative $R^2$ are likely due to artifcats in the feature inference (clustering) proceedure.  
```{r}
bottom_post %>% left_join(genus_exp) %>% 
      ggplot() + 
      geom_path(aes(x = exp_count, y = obs_count, color = t_fctr)) + 
      geom_point(aes( x = exp_count, y = obs_count, color = t_fctr, shape = pcr_rep)) + 
      geom_abline(aes(intercept = 0, slope = 1)) + coord_equal() + 
      theme_bw() + facet_wrap(~pipe, scales = "free")
```


Some of the bottom features have low $R^2$ due to noisy data, while other (post bottom) are due to observed counts being systematically lower than the expected counts. 

__TO DO__ Address this issue, identification of features with low $R^2$ for the expected one to one relationship, but a linear model with a slope and intercept other than 1 and 0, respectively fit the data well. Noise is larger than the signal, can also look at median based observed and expected counts to reduce some of the noise. 
This situation could arrise for two reasons, 1 - the samples were not mixed properly, or 2 - the sequences were not assigned to features representing independent biological units. 


__NEXT STEP__ Relate $R^2$ of lm fit to expected-observed assumed 1-to-1 fit

### Figures

```{r}
get_fit <- function(count_exp_df){
      count_exp_df %>% 
            group_by(pipe, biosample_id, feature_id) %>% 
            nest() %>% 
            mutate(fit = map(data,~lm(obs_count ~ exp_count, data = .)))
}

get_tidy_fit <- function(count_fit){
      count_fit %>% 
            mutate(fit_summary = map(fit, broom::tidy)) %>% 
            select(-data, -fit) %>% unnest() %>% 
            mutate(term = if_else(term == "(Intercept)", "Intercept","Slope"))
}

get_glance_fit <- function(count_fit){
      count_fit %>% 
            mutate(fit_summary = map(fit, broom::glance)) %>% 
            select(-data, -fit) %>% unnest() %>% 
            select(-statistic, -p.value) 
}
 
count_fit <- genus_exp %>% get_fit()
count_glance <- get_glance_fit(count_fit)
```

```{r}
fit_r2 <- count_glance %>% select(pipe, biosample_id, feature_id, adj.r.squared) %>% 
      dplyr::rename(lm_r_squared = adj.r.squared)
```

```{r}
fit_r2 %>% summary()
```


```{r}
left_join(genus_r2_df, fit_r2) %>% left_join(genus_cat) %>% 
      filter(cat != "cat_null", cat != "cat_mix") %>% 
      mutate(r_squared = if_else(r_squared < -1, -1, r_squared)) %>% 
      ggplot() + geom_point(aes(x = r_squared, y = lm_r_squared, color = cat)) + theme_bw() + facet_wrap(~cat)
```

```{r}
left_join(genus_r2_df, fit_r2) %>% left_join(genus_cat) %>% filter(cat == "cat_post")
```


```{r}
dat <- genus_exp %>% select(pipe, biosample_id, feature_id, pcr_rep, obs_count, post_count, pre_count, t_fctr) 
pre_post <- dat %>% select(pipe, biosample_id, feature_id, pcr_rep, post_count, pre_count) %>% 
      gather("count_type", "value", -pipe, -biosample_id, -feature_id, -pcr_rep) %>% 
      mutate(t_fctr = if_else(count_type == "post_count","0","20")) %>% 
      select(-count_type) %>% 
      dplyr::rename(obs_count = value) %>% unique()

dat <- bind_rows(dat, pre_post)
```

```{r}
dat <- dat %>% mutate(titration = as.numeric(as.character(t_fctr)), theta = 2^(-titration))


```


```{r}
example_dat <- dat %>% filter(pipe == "dada2", biosample_id == "E01JH0004", 
               feature_id == "[Eubacterium]_coprostanoligenes_group") %>% 
      mutate(t_fctr = fct_relevel(t_fctr, c(20, 15, 10, 5, 4, 3, 2, 1, 0)))
```

```{r}
example_dat %>% ggplot() + geom_point(aes(x = t_fctr, y = obs_count))
```

```{r}
fit <- lm(obs_count ~ -1 + theta:t_fctr, data = example_dat)

model.matrix( ~ -1 + theta:t_fctr, data=example_dat)
```

```{r}
summary(fit)
```


A good example ...
```{r}
example_dat <- dat %>% filter(pipe == "dada2", biosample_id == "E01JH0038", 
               feature_id == "Lachnoclostridium") %>% 
      mutate(t_fctr = fct_relevel(t_fctr, c(20, 15, 10, 5, 4, 3, 2, 1, 0)))
example_dat %>% ggplot() + geom_point(aes(x = t_fctr, y = obs_count))
```

```{r}
fit <- glm(obs_count ~ -1 + theta:t_fctr, data = example_dat)
summary(fit)
```

```{r}
fit %>% broom::tidy()
```

```{r}
fit %>% broom::tidy() %>% filter(term != "theta:t_fctr20") %>% 
      ggplot() + geom_point(aes(x = term, y = estimate)) +
      geom_errorbar(aes(x = term, ymin = estimate - std.error, ymax = estimate + std.error))
```

```{r}
pipeline_dir <- "../../mgtst_pipelines"
mrexp <- get_mrexp(pipeline_dir)

## Extracting a tidy dataframe with count values from MRexpiment objects
get_count_df <- function(mrobj, agg_genus = FALSE){
      if(agg_genus){
            mrobj <- aggregateByTaxonomy(mrobj, lvl = "Rank6", 
                                         norm = FALSE, log = FALSE, sl = 1)
      }
      
      mrobj <- cumNorm(mrobj, p = 0.75)
      mrobj %>%
            # not sure whether or not to normalize counts prior to analysis
            MRcounts(norm = FALSE, log = FALSE, sl = 1) %>%  
            as.data.frame() %>% 
            rownames_to_column(var = "feature_id") %>% 
            gather("id","count", -feature_id)
} 

count_df <- mrexp %>% map_df(get_count_df, .id = "pipe") %>% 
      left_join(pData(mrexp$dada2)) %>% 
      filter(biosample_id != "NTC") 
```

```{r}
count_df <- count_df %>% 
      mutate(titration = as.numeric(as.character(t_fctr)), theta = 2^(-titration)) %>% 
      group_by(pipe, biosample_id, t_fctr, pcr_rep) %>% mutate(total_count = sum(count))
```

```{r}
example_count_df <- count_df %>% filter(pipe == "dada2", biosample_id == "E01JH0038", feature_id == "SV1") %>% 
      ungroup() %>% 
      mutate(t_fctr = fct_relevel(t_fctr, c(20, 15, 10, 5, 4, 3, 2, 1, 0)))
```

```{r}
example_count_df %>% ggplot() + geom_point(aes(x = t_fctr, y = count))
```

```{r}
sdf <- example_count_df$total_count/exp(mean(log(example_count_df$total_count)))
example_count_df$total_count <- sdf
fit <- lm(count ~ -1 + theta:t_fctr, offset = total_count, data = example_count_df)
summary(fit)
```

```{r}
fit %>% broom::tidy()
```

```{r}
fit %>% broom::tidy() %>% filter(term != "theta:t_fctr20") %>% 
      ggplot() + geom_point(aes(x = term, y = estimate)) +
      geom_errorbar(aes(x = term, ymin = estimate - std.error, ymax = estimate + std.error)) +
      scale_y_log10()
```

Need to modify to test for differences in the Post-Pre estimates are being tested. 
Currently the Differences in the Post-Pre estimates from the Pre estimate is being tested. 
Modify the contrast matrix so that the test is performed as part of the model, and the model is not over parameterized.... 
```{r}

fit <- lm(count ~ -1 + theta:t_fctr, offset = total_count, data = example_count_df)
summary(fit)

# 
# grp <- example_count_df$t_fctr
# Ba <- theta:t_fctr
# Bb <-  contr.treatment(grp)
```

```{r}
dat <- fit %>% broom::tidy() %>% filter(term != "theta:t_fctr20") 
dat <- broom::augment(fit) %>% filter(t_fctr != 20)
lm(.fitted~t_fctr, data = dat) %>% summary()
```

```{r}
fit %>% broom::glance()
```


## Applying Pre-Post Modeling approach to feature level analysis
```{r}
pipeline_dir <- "../../mgtst_pipelines"
mrexp <- get_mrexp(pipeline_dir)

## Extracting a tidy dataframe with count values from MRexpiment objects
get_count_df <- function(mrobj, agg_genus = FALSE){
      if(agg_genus){
            mrobj <- aggregateByTaxonomy(mrobj, lvl = "Rank6", 
                                         norm = FALSE, log = FALSE, sl = 1)
      }
      
      mrobj <- cumNorm(mrobj, p = 0.75)
      mrobj %>%
            # not sure whether or not to normalize counts prior to analysis
            MRcounts(norm = FALSE, log = FALSE, sl = 1) %>%  
            as.data.frame() %>% 
            rownames_to_column(var = "feature_id") %>% 
            gather("id","count", -feature_id)
} 


```


```{r}
count_df <- mrexp %>% map_df(get_count_df, .id = "pipe") %>% 
      left_join(pData(mrexp$dada2)) %>% 
      filter(biosample_id != "NTC") %>% 
      ungroup() %>% 
      mutate(t_fctr = fct_relevel(t_fctr, c(20, 15, 10, 5, 4, 3, 2, 1, 0)),
             titration = as.numeric(as.character(t_fctr)), 
             theta = 2^(-titration),
             theta = if_else(theta == 2^-20, 0, theta))

## Calculate scaling factor 
count_df <- count_df %>% group_by(pipe, id) %>% 
      mutate(total_count = sum(count),
             scale_factor = total_count/ exp(mean(log(total_count))))
   
count_full_df <- left_join(count_df, feature_cat) %>% filter(cat == "cat_full") %>% 
      filter(!(t_fctr %in% c(10,15)))

nested_count_df <- count_full_df %>% group_by(pipe, biosample_id, feature_id) %>% nest()
```

```{r}
fit_df <- nested_count_df %>%
      mutate(fit = map(data, ~lm(count ~ -1 + theta:t_fctr, offset = scale_factor, data = .)))
```

```{r}
get_tidy_fit <- function(count_fit){
      count_fit %>% 
            mutate(fit_summary = map(fit, broom::tidy)) %>% 
            select(-data, -fit) %>% unnest() #%>% 
            #mutate(term = if_else(term == "(Intercept)", "Intercept","Slope"))
}

tidy_fit <- get_tidy_fit(fit_df)
```

Consistent decrease in estimates with titration factor indicates that our expectation regarding how the samples were mixed. 
Most likely due to differences in the proportion of DNA in the unmixed pre- and post-treatment samples. 

```{r}
dat <- tidy_fit %>% 
      mutate(term = fct_relevel(term, paste0("theta:t_fctr",c(20, 5:0)))) %>%
      filter(term != "theta:t_fctr20")
dat %>% ggplot() +
      geom_path(aes(x = term, y = estimate + 1, 
                    group = paste(pipe, biosample_id, feature_id)), alpha = 0.25) +
      facet_grid(pipe~biosample_id, scales = "free_y") + 
      theme(axis.text.x = element_text(angle = 90)) + scale_y_log10()
```


Next we want to identify rank features based on their bias and variance. 
Bias - variance of the estimate
Variance - mean standard error of the estimate

```{r}
fit_summary <- dat %>% group_by(pipe, biosample_id, feature_id) %>% 
      summarise(est_cov = sd(estimate)/mean(estimate), error_cov= sd(std.error)/mean(estimate))
```


```{r}
fit_summary %>%
      ggplot() + 
      geom_boxplot(aes(x = biosample_id, y = est_cov, color = pipe)) + 
      theme_bw()
```

```{r}
fit_summary %>% 
      ggplot() + 
      geom_boxplot(aes(x = biosample_id, y = error_cov, color = pipe)) + theme_bw()
```

```{r}
fit_summary %>%  ggplot() + geom_point(aes(x = error_cov, 
                                y = est_cov, 
                                color = biosample_id, 
                                shape = pipe)) + theme_bw()
```

```{r}
fit_summary %>% 
      ggplot() + geom_density2d(aes(x = est_cov, y = error_cov, color = biosample_id)) + facet_wrap(~pipe) 
```

```{r}
fit_summary %>% filter(est_cov < 0.75, error_cov < 0.25) %>% 
      left_join(count_df) %>% 
      ggplot() + 
      geom_point(aes(x = t_fctr, y = count, color = biosample_id, shape = pipe)) + 
      facet_wrap(~feature_id, scales = "free") + theme_bw()
```


```{r}
fit_summary %>% filter(est_cov > 1.75) %>% 
      left_join(count_df) %>% 
      ggplot() + 
      geom_point(aes(x = t_fctr, y = count, color = biosample_id, shape = pipe)) + 
      facet_wrap(~feature_id, scales = "free") + theme_bw()
```

```{r}
count_df %>% filter(biosample_id == "E01JH0011", feature_id == "Otu00063") %>% 
      group_by(t_fctr) %>% summarise(mean_count = mean(count))
```


```{r}
# fit_est <- fit_summary %>% filter(est_cov < 0) %>% 
#       left_join(count_df) 
# 
# fit_sum <- fit_summary %>% filter(est_cov < 0) %>% left_join(tidy_fit) %>% 
#       mutate(term = fct_relevel(term, paste0("theta:t_fctr",c(20, 15, 10, 5:0)))) %>%
#       filter(term != "theta:t_fctr20") %>% 
#       mutate(t_fctr = gsub(pattern = "theta:t_fctr", replacement = "", term))
# bind_rows(est = fit_sum, count = fit_est, .id = "var") %>% 
#       mutate(t_fctr = fct_relevel(t_fctr, c(20, 15, 10, 5:0))) %>%
# ggplot() + 
#       geom_point(aes(x = t_fctr, y = estimate), color = "orange") + 
#       geom_point(aes(x = t_fctr, y = count), color = "blue") + 
#       facet_grid(var~feature_id, scales = "free_y") + 
#       theme(axis.text.x = element_text(angle = 90))
```



```{r}
# fit_est <- fit_summary %>% filter(est_cov < 2.3, est_cov > 0) %>% 
#       left_join(count_df) 
# 
# fit_sum <- fit_summary %>% filter(est_cov < 2.3, est_cov > 0) %>% left_join(tidy_fit) %>% 
#       mutate(term = fct_relevel(term, paste0("theta:t_fctr",c(20, 15, 10, 5:0)))) %>%
#       filter(term != "theta:t_fctr20") %>% 
#       mutate(t_fctr = gsub(pattern = "theta:t_fctr", replacement = "", term))
# bind_rows(est = fit_sum, count = fit_est, .id = "var") %>% 
#       mutate(t_fctr = fct_relevel(t_fctr, c(20, 15, 10, 5:0))) %>%
# ggplot() + 
#       geom_point(aes(x = t_fctr, y = estimate), color = "orange") + 
#       geom_point(aes(x = t_fctr, y = count), color = "blue") + 
#       facet_grid(var~feature_id, scales = "free_y") + 
#       theme(axis.text.x = element_text(angle = 90))
```


```{r}
# fit_summary %>% filter(est_cov < 2.3, est_cov > 0) %>% 
#       left_join(count_df) %>% 
#       ggplot() + 
#       geom_point(aes(x = t_fctr, y = count, color = biosample_id, shape = pipe)) + 
#       facet_wrap(~feature_id, scales = "free") + theme_bw()
```

```{r}
# fit_summary %>% filter(est_cov < 2.3, est_cov > 0) %>% 
#       left_join(tidy_fit) %>% 
#       mutate(term = fct_relevel(term, paste0("theta:t_fctr",c(20, 15, 10, 5:0)))) %>%
#       filter(term != "theta:t_fctr20") %>% 
#       ggplot() + 
#       geom_point(aes(x = term, y = estimate, color = biosample_id, shape = pipe)) + 
#       facet_wrap(~feature_id, scales = "free_y") + theme_bw() + 
#       theme(axis.text.x = element_text(angle = 90))
```


```{r}
# fit_summary %>% filter(est_cov > 2.65, est_cov < 2.75, total_norm_error > -0.05, total_norm_error < 0.05) %>% 
#       left_join(tidy_fit) %>% 
#       mutate(term = fct_relevel(term, paste0("theta:t_fctr",c(20, 15, 10, 5:0)))) %>%
#       filter(term != "theta:t_fctr20") %>% 
#       ggplot() + 
#       geom_point(aes(x = term, y = estimate, color = biosample_id, shape = pipe)) + 
#       facet_wrap(~feature_id, scales = "free_y") + theme_bw() + 
#       theme(axis.text.x = element_text(angle = 90))
```

# Session information 

## Git repo commit information
```{r}
library(git2r)
repo <- repository(path = "../")
last_commit <- commits(repo)[[1]]
```

The current git commit of this file is `r last_commit@sha`, which is on the `r branches(repo)[[1]]@name` branch and was made by `r last_commit@committer@name` on `r when(last_commit)`. The current commit message is `r last_commit@summary`. The repository is online at https://github.com/nate-d-olson/mgtst-pub  


## Platform Information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
```


## Package Versions
```{r}
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
      knitr::kable()
```

