---
title: "Targeted logFC Analysis"
author: "Nate Olson"
date: '`r Sys.Date()`'
always_allow_html: yes
output:
  pdf_document: default
  html_document: default
---


## Summary
Initial evaluation of logFC estimates indicated the estimates are not consistent with our expectations. 
Specifically, for post-specific features the logFC between any two titrations is should be the difference in the titrations factors. 
This is only true when the samples are mixed according to the mixture design. 
As differences in the proportion of bacterial DNA between the unmixed samples used to generate the two-sample titrations was observed the samples were not mixed according to the mixture design. 
To account for this difference inferred estimates for the proportion of bacterial DNA in each of the titrations from the post-exposure samples were used to calculate new expected logFC values. 
However, their was still limited agreement between the observed logFC values (estimates using edgeR) to the expected logFC values (based on the inferred theta values). 
Here we explore the relationship between the logFC estimates and expected values a limited set of comparisons, E01JH0011 processed using DADA2 and unclustered for titrations 1 and 3. 
E01JH0011 and titrations 1 and 3 were chosen based on agreement between the inferred and mixture design theta. 
For post-specific features the expected logFC is 2. 
Similar to the previous analysis, the logFC estimates for this targeted analysis were not consistent with our expectations. 

```{r echo = FALSE, message = FALSE}
library(tidyverse)
library(metagenomeSeq)
```


### logFC Assessment Using logFC DF Estimates  

All logFC estimates were calculated and compiled into a single data frame to facilitate downstream analysis.  

```{r}
e16 <- readRDS("~/Desktop/logFC_edgeR_df.rds") %>% 
      filter(pipe %in% c("dada2","unclustered"),
             biosample_id == "E01JH0011",
             T1 == 1, T2 == 3)

## Adding Post specific annotations
pa_df <- readRDS("~/Desktop/pa_summary_anno_df.RDS") %>% 
      dplyr::select(biosample_id, pipe, feature_id, T00, T20, post_specific) %>% 
      filter(biosample_id == "E01JH0011", pipe %in% c("dada2","unclustered"))
e16_anno <- e16 %>% rename(feature_id = OTUname) %>% left_join(pa_df)
```

The logFC estimates for post-specific features are not consistent with the expected value of 2
```{r}
e16_anno %>% ggplot() + 
      geom_point(aes(x = logCPM, y = logFC)) + 
      geom_hline(aes(yintercept = 2)) + 
      facet_grid(post_specific~pipe) + theme_bw()
```


## Re-calculating logFC estimates
To further evaluate logFC estimates, logFC was calculated outside of analysis pipeline to ensure the lack of agreement between the estimates and expected values is not due to a error in the code used to calculate logFC. 

```{r}
mrexp_dada2 <- readRDS("~/Projects/mgtst_pipelines/dada2/dada_mrexp.rds")
mrexp_unclustered <- readRDS("~/Projects/mgtst_pipelines/mothur/unclustered_mrexp.rds")

## Extract samples of interest - E01JH0011 Titrations 1 and 3
sam_dat <- mrexp_dada2@phenoData@data %>% 
      filter(biosample_id == "E01JH0011", t_fctr %in% c(1,3))

## Subset MRexp
dada2_subset <- mrexp_dada2[,colnames(mrexp_dada2) %in% sam_dat$id]
unclustered_subset <- mrexp_unclustered[,colnames(mrexp_unclustered) %in% sam_dat$id]

## Calculate logFC estimates DADA2 ######################
dada2_counts <- dada2_subset@assayData$counts
group <- sam_dat$t_fctr
require(edgeR)
y <- DGEList(counts = dada2_counts, group = group, remove.zeros = TRUE)

## Calc normalization factors
z <- edgeR::calcNormFactors(y, method = "RLE")

## Estimate dispersions
fit <- z %>% estimateCommonDisp() %>% 
      estimateTagwiseDisp()
fit_exact <- exactTest(fit)
dada2_fit_tbl <- topTags(fit_exact, n = Inf, adjust.method = "BH") %>% 
      as.data.frame() %>% 
      rownames_to_column(var = "feature_id") %>%
      left_join(pa_df)

## Calculate logFC estimates Unclustered ######################
unclustered_counts <- unclustered_subset@assayData$counts
group <- sam_dat$t_fctr
require(edgeR)
y <- DGEList(counts = unclustered_counts, group = group, remove.zeros = TRUE)

## Calc normalization factors
z <- edgeR::calcNormFactors(y, method = "RLE")

## Estimate dispersions
fit <- z %>% estimateCommonDisp() %>% estimateTagwiseDisp()
fit_exact <- exactTest(fit)
unclustered_fit_tbl <- topTags(fit_exact, n = Inf, adjust.method = "BH") %>% 
      as.data.frame() %>% 
      rownames_to_column(var = "feature_id") %>%
      left_join(pa_df)

### Combining Data Frames
fit_tbl <- bind_rows(dada2_fit_tbl, unclustered_fit_tbl)
```



X-axis facet is the number of post-exposure PCR replicates with observed counts, post-specific features are in the top right plot.
```{r}
fit_tbl %>% mutate(T00 = as.character(T00), T20 = as.character(T20)) %>% 
      filter(pipe == "dada2") %>% 
      ggplot() + 
      geom_point(aes(x = logCPM, y = logFC), color = "grey60", alpha = 0.25) + 
      geom_hline(aes(yintercept = 2)) + 
      facet_grid(T00~T20) + 
      theme_bw()
```

```{r}
fit_tbl %>% mutate(T00 = as.character(T00), T20 = as.character(T20)) %>% 
      filter(pipe == "unclustered") %>% 
      ggplot() + 
      geom_point(aes(x = logCPM, y = logFC), color = "grey60", alpha = 0.25) + 
      geom_hline(aes(yintercept = 2)) + 
      facet_grid(T00~T20) + 
      theme_bw()
```

