---
title: "Differential Abundance Calculations"
author: "Nate Olson"
date: "January 29, 2017"
output: html_document
---

```{r message = FALSE}
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd)
```

## Overview
Calculating logFC differential abundance between pre and post-treatment samples.


### Loading Pipeline Data
```{r}
mrexp_files <- list(
      dada2  = "../data/mrexp_dada2.RDS",
      mothur = "../data/mrexp_mothur.RDS",
      qiime  = "../data/mrexp_qiime_refclus_nochimera.RDS"
)
mrexp <- mrexp_files %>% map(readRDS) 

#Extracting metadata

meta_dat <- mrexp$mothur %>% pData()

##labeling PCR replicates
half1 <- paste(rep(c("A","B","C","D","E","F","G","H"), each = 6), 1:6, sep = "_")
sam_dat <- meta_dat %>% 
      mutate(pcr_half = if_else(pos %in% half1, "1","2"),
             pcr_rep = paste0(pcr_16S_plate,":",pcr_half)) %>% 
      select(sampleID, dilution,sam_names, pcr_rep) %>% 
      dplyr::rename(samID = sam_names)
```


```{r}
E01JH004_pre_post_sams <- meta_dat %>% 
      filter(sampleID == "E01JH0004", dilution %in% c(0,-1)) %>% .$sam_names
mrexp_004_pre_post <- mrexp %>% 
      map(~.[,which(colnames(.) %in% E01JH004_pre_post_sams)]) %>% 
      map(~.[which(rowSums(MRcounts(.)) > 0), ])
mrexp_obj <- mrexp_004_pre_post$dada2
```

```{r}
pre_post_meta <- meta_dat %>% filter(sampleID == "E01JH0004", dilution %in% c(0,-1))
pre_sams <- pre_post_meta %>% filter(dilution == 0) %>% .$sam_names
post_sams <- pre_post_meta %>% filter(dilution == -1) %>% .$sam_names
```
```{r}
nrow(mrexp_004_pre_post$dada2)
```


## metagenomeSeq fitFeature Model
```{r}
mrexp_obj <- mrexp_004_pre_post$dada2
## excluding features not present in half the samples
mrexp_obj <- filterData(mrexp_obj, present = 8, depth = 1)

## Upper quartile normalization
mrexp_obj <- cumNorm(mrexp_obj, p = 0.75)
pd <- pData(mrexp_obj)
pd$dilution <- factor(pd$dilution, levels = c(0,-1))
mod <- model.matrix(~1 + dilution, data = pd)
mod
pre_post_reg <- fitFeatureModel(mrexp_obj, mod)
```

Looking at features present in all unmixed pre and post replicates. 
Only `r nrow(mrexp_obj)` out of `r nrow(mrexp_004_pre_post$dada2)` features were present in all replicates. 

```{r}
reg_coefs <- MRcoefs(pre_post_reg,number = nrow(mrexp_obj))
```

```{r}
reg_coefs %>% head() %>% knitr::kable()
```


Results interpretation  
* logFC - is this pre/post or the opposit?
* how to get se



## Next Steps
* Compare pre and post metagenomeSeq to permutation method, a kind of sanity check
* Use permutations for unmixed vs titration exected values
* Calculate fitFeatureModel unmixed vs. titration differences
* Look at pre and post specific features

* Work out methods for 
      - multiple pipelines
      - different combinations of dilutions
