---
title: "Expected and Observed Count Values - Biosample, Pipeline"
author: "Nate Olson"
date: '`r Sys.Date()`'
always_allow_html: yes
output:
  pdf_document: default
  html_document: default
---

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd) 
library(ape) 
library(nlme)
```

### Objective
Evaluate the overall fit for biological replicates and pipelines. 
Want to see which has greater variability. 
Is there a pipeline or biological sample effect. 
The pipeline effect would be due to differences in feature definition. 
The biological sample effect would indicate a potential issue titration generation. 


### Loading Data 
Feature level count data
```{r}
## Subsetting features to only include pre, post, and full features
count_exp_df <- readRDS("../data/expected_count_values_feature_df.rds") 
feature_cat <- readRDS("../data/feature_categories_df.rds")
count_exp_df <- left_join(count_exp_df, feature_cat) %>% 
      filter(cat %in% c("cat_full", "cat_pre", "cat_post")) 
```

Genus level count data
```{r}
## Subsetting features to only include pre, post, and full features
genus_exp_df <- readRDS("../data/expected_count_values_genus_df.rds") 
genus_cat <- readRDS("../data/genus_categories_df.rds")
genus_exp_df <- left_join(genus_exp_df, genus_cat) %>% 
      filter(cat %in% c("cat_full", "cat_pre", "cat_post")) 
```

## Observed - Expected By Pipeline and Biological Replicate
### Feature level analysis
Overall relationship between the observed and expected values by pipeline and biological replicate. 
Orange line is a fitted smoothing function (loess, local polynomial regression) to highlight the relationship between the observed and expected counts. 
```{r }
count_exp_df %>% ggplot() + 
      geom_hex(aes(x = obs_count + 1, y = exp_count + 1)) + 
      geom_smooth(aes(x = obs_count + 1, y = exp_count + 1), color = "darkorange") + 
     facet_grid(pipe~biosample_id)+ theme_bw() + 
      labs(y = "Expected Count", x = "Observed Count", fill = "Abundance") + 
      scale_y_continuous(trans = "log2") + 
      scale_x_continuous(trans = "log2") + 
      theme(axis.text.x = element_text(angle = 270))
```

### Genus level analysis
```{r}
genus_exp_df %>% ggplot() + 
      #geom_point(aes(x = obs_count + 1, y = exp_count + 1), 
      #           fill = "darkblue", color = "white", alpha = 0.5, shape = 21) + 
      geom_hex(aes(x = obs_count + 1, y = exp_count + 1)) + 
      geom_smooth(aes(x = obs_count + 1, y = exp_count + 1), color = "darkorange") + 
      #geom_abline(aes(intercept = 0, slope =1), color = "darkorange") +
     facet_grid(pipe~biosample_id)+ theme_bw() + 
      labs(y = "Expected Count", x = "Observed Count", fill = "Abundance") + 
      scale_y_continuous(trans = "log2") + 
      scale_x_continuous(trans = "log2") + 
      theme(axis.text.x = element_text(angle = 270))
```

### Comparison of Raw and Genus Feature Levels
Comparison of raw feature level and features aggregated to the genus level. Black line indicates the expected 1-to-1 relationship between the expected and observed counts. 

```{r}
exp_df <-bind_rows(ftu = count_exp_df, genus = genus_exp_df,.id = "feature_level")
exp_df %>% ggplot() + 
      geom_abline(aes(intercept = 0, slope = 1)) + 
      geom_smooth(aes(x = obs_count + 1, y = exp_count + 1, color = feature_level)) + 
     facet_grid(pipe~biosample_id)+ theme_bw() + 
      labs(y = "Expected Count", x = "Observed Count", fill = "Abundance") + 
      scale_y_continuous(trans = "log2") + 
      scale_x_continuous(trans = "log2") + 
      theme(axis.text.x = element_text(angle = 270))
```

## Attributing Variability 
Use a mixed effects model to characterize the impact of pipeline and biological replicate.

```{r}
count_fit <- lme(residual ~ exp_count, random =  ~ biosample_id | pipe, data = count_exp_df)
summary(count_fit)
```



```{r}
genus_fit <- lme(residual ~ exp_count, random =  ~ biosample_id | pipe, data = genus_exp_df)
summary(genus_fit)
```

Comparison of variance components
```{r}
var_count <- varcomp(count_fit, scale = TRUE)
var_genus <- varcomp(genus_fit, scale = TRUE)
data_frame(comp = names(var_count), 
           raw_feature = round(var_count * 100,2), 
           genus = round(var_genus * 100,2))
```


## Session information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
      knitr::kable()
``` 