---
title: "Theta Estimates"
author: "Nate Olson"
date: '`r Sys.Date()`'
output:
  bookdown::pdf_document2: 
    toc: FALSE
---

```{r echo=FALSE, message=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(DESeq2)
library(phyloseq)
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd)
pipeline_dir <- "../../mgtst_pipelines"
ps <- get_phyloseq(pipeline_dir)
mrexp <- get_mrexp(pipeline_dir)
```

## Objective
Use count data to estimate $\theta$. 
Deviations from expected $\theta$ are either due to biases in the count data, titration errors, or differences in the proportion of bacterial DNA in the pre- and post-exposure samples. 

## Approaches
1. $\theta$ as the log fold-change between unmixed sample and titrations for pre- and post-exposure specific samples. 
2. Calculate feature proportions for all samples, pooling replicates. Use estimated proportions to calculate $theta$, where  

$$\theta = \frac{C_{obs}- C_{pre}}{C_{post}-C_{pre}}$$ 

## $\theta$ as log-fold changes  
__NEXT STEPS__ 
1. Clean-up code
1. apply to other biological replicates
1. summarise results  
1. Identify outlier features for further exploration  
1. Look at pre-specific features

__Conclusions__
1. Unable to use logFC for post specific features to estimate theta due limited number of post-specific features with sufficient counts for each titration.  




1. Load data - use single biological replicate and pipeline as demonstration  
1. Identify pre and post-specific features  
1. Calculate log fold-change using DESeq2 for Unmixed and each titration with non-zero counts  
1. Calculate theta from log fold-change. Post-specific log fold-change is the titration factor, pre-specific __TODO__ workout equation.  

```{r}
calc_logfc_deseq <- function(ps_obj){
    ## convert to DESeq object
    dsq <- phyloseq_to_deseq2(ps_obj, ~ t_fctr)

    ## calculate logFC estimates
    DESeq(dsq, test="Wald", fitType="parametric")
}

tidy_deseq_res <- function(dsq_logfc, biosample_id){
    results(dsq_logfc, cooksCutoff = FALSE) %>%
        as.data.frame() %>%
        rownames_to_column(var = "feature_id") %>%
        add_column(biosample_id = biosample_id)
}


post_specific_features <- function(ps_obj){
    
}

subset_ps <- function(ps_obj, biosample_id, t_fctr1, t_fctr2){
    ## Get ps object with only biosample and titrations of interest
    ps_unmix <- subset_samples(ps_obj, 
                               t_fctr %in% c(get("t_fctr1"), get("t_fctr2")) & 
                                   biosample_id == get("biosample_id"))
}

get_logfc <- function(ps_obj, t_fctr1, tfctr2){
    logfc_df <- data_frame()

    biosample_ids <- sample_data(ps_obj)$biosample_id %>% unique()
    for(i in biosample_ids){
        if(i != "NTC"){
            biosams <- as.character(get_variable(ps_obj, "biosample_id")) == i
            ps_unmix <- prune_samples(biosams, ps_obj)
            ps_unmix <- subset_samples(ps_obj, t_fctr %in% c(0,20))
            deseq_logfc <- calc_logfc_deseq(ps_unmix)
            logfc_df <- tidy_deseq_res(deseq_logfc, i) %>%
                bind_rows(logfc_df)
        }
    }

    logfc_df
}

```

## calc logfc for post specific features between post and each titration  
```{r}
ps_obj <- ps$mothur
```

1. subset phyloseq by biological replicate 

```{r}
biosample_ids <- sample_data(ps_obj)$biosample_id %>% 
    unique() %>% {.[. != "NTC"]} 
ps_obj1 <- subset_samples(ps_obj, biosample_id == biosample_ids[3])
```

1. extract post specific features, mean post counts > 10 and no observed pre counts  
```{r}
ps_obj2 <- subset_samples(ps_obj1, t_fctr %in% c(0,20)) %>% 
    merge_samples("t_fctr", fun=sum)
```

```{r}
ps_obj1_df <- ps_obj1 %>% psmelt() 
sample_abun <- ps_obj1_df %>% group_by(OTU, t_fctr) %>% 
    summarise(mean_abun = mean(Abundance))  
wide_sample_abun <- sample_abun %>% spread(t_fctr, mean_abun)
```

```{r}
post_specific_df <- wide_sample_abun %>% 
    filter(`20`  == 0, `0` > 4) %>% 
    rename(post = `0`, pre = `20`) %>% 
    gather("t_fctr", "mean_abun", -OTU, -pre, -post)
```

1. list titration pairs and filter titrations with mean counts < 10  

```{r}
post_theta_est <- post_specific_df %>% filter(mean_abun >= 1)
```

Potential Issue with limited number of features per titration to use in estimating $\theta$

```{r}
post_theta_est %>% group_by(t_fctr) %>% summarise(n_features = n())
```




1. subset phyloseq object - titrations  
```{r}
feat <- post_theta_est %>% filter(t_fctr == "1") %>% .$OTU
feat
```

```{r}
ps_obj2 <- ps_obj1 %>% #prune_taxa(feat, ps_obj1) %>% 
    subset_samples(t_fctr %in% c("0","1"))
```

```{r}
ps_obj2
```

1. Calculate logFC 
```{r}
ps_obj2@sam_data$t_fctr <- factor(ps_obj2@sam_data$t_fctr, levels = c(1,0))
deseq_logfc <- calc_logfc_deseq(ps_obj2)
logfc_df <- tidy_deseq_res(deseq_logfc, "test")
```

```{r}
logfc_df %>% filter(feature_id %in% feat)
```

```{r}
post_theta_est
```


1. Compare theta estimates with expected values


#### Bulk processing 
One pipeline and biological replicate  

```{r}
ps_obj <- ps$mothur

biosample_ids <- sample_data(ps_obj)$biosample_id %>% 
    unique() %>% {.[. != "NTC"]} 

ps_obj1 <- subset_samples(ps_obj, biosample_id == biosample_ids[3])
```




Calculating logFC for each titration comparison  

```{r}
t_fctrs <- c(1:5,10,15)

logfc_df <- data.frame()

for(i in t_fctrs){
    ps_obj2 <- subset_samples(ps_obj1, t_fctr  %in% c(i, 0))
    ps_obj2@sam_data$t_fctr <- factor(ps_obj2@sam_data$t_fctr, levels = c(0, i))
    deseq_logfc <- calc_logfc_deseq(ps_obj2)
    logfc_df <- tidy_deseq_res(deseq_logfc, i) %>%
        bind_rows(logfc_df)
}
logfc_df <- logfc_df %>% rename(t = biosample_id)
```

Filtering post specific features
```{r}
theta_est_logfc <- post_theta_est %>% select(OTU, post, mean_abun, t_fctr) %>% 
    rename(feature_id = OTU) %>% 
    mutate(t = as.numeric(as.character(t_fctr))) %>%  
    left_join(logfc_df)
```



```{r}
theta_est_logfc
```

```{r}
theta_est_logfc <- theta_est_logfc %>% mutate(theta_hat = 2^log2FoldChange, theta_exp = 2^-as.numeric(t_fctr))
```

```{r}
theta_est_logfc %>% ggplot() + geom_point(aes(x = t, y = theta_hat))+
    geom_line(aes(x = t, y = theta_exp))
```

```{r}
theta_est_logfc %>% filter(theta_hat > 2)
```

```{r}
theta_est_logfc %>% filter(feature_id != "Otu00248") %>%
ggplot() + geom_point(aes(x = t, y = theta_hat))+ 
    geom_smooth(aes(x = t, y = theta_hat)) + 
geom_line(aes(x = t, y = theta_exp))
```


## Expanding to multiple biological replicates

```{r} 
ps_obj <- ps$mothur

## Variables used to subset dataset
biosample_ids <- sample_data(ps_obj)$biosample_id %>% 
    unique() %>% {.[. != "NTC"]} 

t_fctrs <- c(1:5,10,15)

logfc_df <- data.frame() 
for(i in biosample_ids){
    
    ps_obj1 <- subset_samples(ps_obj, biosample_id == i) 
    
    for(j in t_fctrs){
        ps_obj2 <- subset_samples(ps_obj1, t_fctr  %in% c(j, 0))
        ps_obj2@sam_data$t_fctr <- factor(ps_obj2@sam_data$t_fctr, levels = c(0, j))
        deseq_logfc <- calc_logfc_deseq(ps_obj2)
        logfc_df <- tidy_deseq_res(deseq_logfc, i) %>% 
            mutate(t = j) %>% 
            bind_rows(logfc_df)        
    }
} 

```


```{r}
ps_obj_df <- ps_obj %>% psmelt() 
sample_abun <- ps_obj_df %>% group_by(biosample_id, OTU, t_fctr) %>% 
    summarise(mean_abun = mean(Abundance))  
wide_sample_abun <- sample_abun %>% spread(t_fctr, mean_abun)
```

```{r}
post_specific_df <- wide_sample_abun %>% 
    filter(`20`  == 0, `0` > 4) %>% 
    rename(post = `0`, pre = `20`) %>% 
    gather("t_fctr", "mean_abun", -biosample_id, -OTU, -pre, -post)
```


```{r}
post_theta_est <- post_specific_df %>% filter(mean_abun >= 1)
```


Filtering post specific features
```{r}
theta_est_logfc <- post_theta_est %>% 
    select(OTU, biosample_id, post, mean_abun, t_fctr) %>% 
    rename(feature_id = OTU) %>% 
    mutate(t = as.numeric(as.character(t_fctr))) %>%  
    left_join(logfc_df)
```



```{r}
theta_est_logfc
```

```{r}
theta_est_logfc <- theta_est_logfc %>% 
    mutate(theta_hat = 2^log2FoldChange, 
           theta_exp = 2^-as.numeric(t_fctr))
```

```{r}
theta_est_logfc %>% 
    filter(feature_id != "Otu00248", feature_id != "Otu00335",
           t < 10) %>% 
    ggplot() + geom_point(aes(x = t, y = theta_hat))+
    geom_line(aes(x = t, y = theta_exp)) + facet_wrap(~biosample_id, nrow = 1)
```

```{r} 
theta_est_logfc %>% filter(log2FoldChange > 2)
```

### LogFC with Pre-specific features
```{r}
ps_obj <- ps$mothur

## Variables used to subset dataset
biosample_ids <- sample_data(ps_obj)$biosample_id %>% 
    unique() %>% {.[. != "NTC"]} 

t_fctrs <- c(1:5,10,15)

pre_logfc_df <- data.frame() 
for(i in biosample_ids){
    
    ps_obj1 <- subset_samples(ps_obj, biosample_id == i) 
    
    for(j in t_fctrs){
        ps_obj2 <- subset_samples(ps_obj1, t_fctr  %in% c(j, 20))
        ps_obj2@sam_data$t_fctr <- factor(ps_obj2@sam_data$t_fctr, levels = c(20, j))
        deseq_logfc <- calc_logfc_deseq(ps_obj2)
        pre_logfc_df <- tidy_deseq_res(deseq_logfc, i) %>% 
            mutate(t = j) %>% 
            bind_rows(pre_logfc_df)        
    }
} 
```


```{r}
pre_specific_df <- wide_sample_abun %>% 
    filter(`20`  > 100, `0` == 0) %>% 
    rename(post = `0`, pre = `20`) %>% 
    gather("t_fctr", "mean_abun", -biosample_id, -OTU, -pre, -post)
```


```{r}
pre_theta_est <- pre_specific_df %>% filter(mean_abun >= 4)
```


Filtering post specific features
```{r}
pre_theta_est_logfc <- pre_theta_est %>% 
    select(OTU, biosample_id, pre, post, mean_abun, t_fctr) %>% 
    rename(feature_id = OTU) %>% 
    mutate(t = as.numeric(as.character(t_fctr))) %>%  
    left_join(pre_logfc_df)
```


```{r}
pre_theta_est_logfc <- pre_theta_est_logfc %>% 
    mutate(theta_hat = 1 - 2^log2FoldChange, 
           theta_exp = 2^-as.numeric(t_fctr))
```

```{r}
pre_theta_est_logfc %>% filter(t_fctr %in% 1:3, abs(theta_hat) < 2) %>% 
    ggplot() + geom_point(aes(x = t, y = theta_hat))+
    geom_line(aes(x = t, y = theta_exp)) + 
    facet_wrap(~biosample_id, nrow = 1)
```

```{r}
outlier_features <- c("Otu00024","Otu00046", "Otu00093","Otu00146","Otu00042")
pre_theta_est_logfc %>% 
    filter(t_fctr %in% 1:3, 
           !(feature_id %in% outlier_features)) %>% 
    ggplot() + geom_boxplot(aes(x = t_fctr, y = theta_hat))+
    facet_wrap(~biosample_id, nrow = 1)
```

```{r}
pre_theta_est_logfc %>% filter(t_fctr %in% 1:3, theta_hat < -0.25) 
```


```{r}
pre_theta_est_logfc %>% filter(theta_hat < 0) %>% group_by(t_fctr) %>% summarise(count = n())
```

```{r}
pre_theta_est_logfc %>% filter(t_fctr %in% 1:3) %>% 
    ggplot() + geom_boxplot(aes(x = factor(t_fctr), y = log2FoldChange))+
    # geom_line(aes(x = t, y = theta_exp)) + 
    facet_wrap(~biosample_id, nrow = 1)
```



## $\theta$ estimate from proportions  
Initial Results don't look promising - will focus on approach 1

1. Load count data - use a single biological replicate and pipeline as demonstration  
1. Use a negative binomial glm to estimate proportions  
1. filter features with 0 pre and post proportion estimates  
1. calculate $\theta$ by feature  
1. Look at $\theta$ estimate distributions  



```{r}
get_count_df <- function(mrobj, agg_genus = FALSE){
      if(agg_genus){
            mrobj <- aggregateByTaxonomy(mrobj, lvl = "Rank6", 
                                         norm = FALSE, log = FALSE, sl = 1)
      }
      
      mrobj <- cumNorm(mrobj, p = 0.75)
      mrobj %>%
            # not sure whether or not to normalize counts prior to analysis
            MRcounts(norm = FALSE, log = FALSE, sl = 1) %>%  
            as.data.frame() %>% 
            rownames_to_column(var = "feature_id") %>% 
            gather("id","count", -feature_id)
} 

full_count_df <- mrexp %>% map_df(get_count_df, .id = "pipe") %>% 
      left_join(pData(mrexp$dada2)) %>% 
      filter(biosample_id != "NTC") %>% 
      ungroup() %>% 
      mutate(t_fctr = fct_relevel(t_fctr, c(0:5, 10, 15, 20)),
             titration = as.numeric(as.character(t_fctr)), 
             theta = 2^(-titration),
             theta = if_else(theta == 2^-20, 0, theta))  %>% 
      filter(t_fctr %in% c(0:4,20))

## Calculate scaling factor 
full_count_df <- full_count_df %>% group_by(pipe, id) %>% 
      mutate(total_count = sum(count),
             ## I don't think this is correct
             scale_factor = total_count/ exp(mean(log(total_count))),
             norm_count = count/total_count)

count_df <- full_count_df %>% filter(pipe == "mothur", 
                                     biosample_id == "E01JH0016", 
                                     feature_id == "Otu00001")
```

```{r}
fit <- MASS::glm.nb(norm_count ~ t_fctr - 1, weights = total_count, data = count_df)
```

Developing method to estimate $\theta$ from negative binomial estimates.  

```{r}
summary(fit)
```

```{r}
library(ggfortify)
autoplot(fit)
```


```{r}
tidy_fit_df <- broom::tidy(fit)
```

```{r}
tidy_fit_df
```

```{r}
prop_hat_df <- tidy_fit_df %>% mutate(prop = exp(estimate)) %>% dplyr::select(term, prop) 
prop_hat <- prop_hat_df$prop
names(prop_hat) <- prop_hat_df$term
```


```{r}
prop_hat
```

```{r}
theta1  <- (prop_hat["t_fctr1"] - prop_hat["t_fctr20"])/(prop_hat["t_fctr0"] - prop_hat["t_fctr20"])
theta1
```

```{r}
theta2  <- (prop_hat["t_fctr2"] - prop_hat["t_fctr20"])/(prop_hat["t_fctr0"] - prop_hat["t_fctr20"])
theta2
```

```{r}
theta3  <- (prop_hat["t_fctr3"] - prop_hat["t_fctr20"])/(prop_hat["t_fctr0"] - prop_hat["t_fctr20"])
theta3
```

```{r}
theta4  <- (prop_hat["t_fctr4"] - prop_hat["t_fctr20"])/(prop_hat["t_fctr0"] - prop_hat["t_fctr20"])
theta4
```


### Cross feature $\theta$ estimate 

```{r}
count_df <- full_count_df %>% filter(pipe == "mothur", biosample_id == "E01JH0016") 
null_features <- count_df %>% 
    group_by(pipe, biosample_id, feature_id) %>% 
    summarise(feature_count = sum(count)) %>% 
    filter(feature_count == 0)
```

```{r}
nested_count_df <- count_df %>% 
    anti_join(null_features) %>% 
    group_by(pipe, biosample_id, feature_id) %>% nest()
```

```{r}
fit_df <- nested_count_df %>% 
    mutate(fit = map(data, ~MASS::glm.nb(norm_count ~ t_fctr - 1, 
                                      weights = total_count, 
                                      data = .)),
        tidy = map(fit, broom::tidy))
```

Use tidy data to obtain estimates for theta
spread
calculate theta for each titration
```{r}
prop_hat_df <- fit_df %>% select(pipe, biosample_id, feature_id, tidy) %>% unnest() %>% 
    mutate(prop = exp(estimate)) %>% 
    select(pipe, biosample_id, feature_id, term, prop)  %>% 
    spread(term, prop)

```

```{r}
fit_df %>% select(pipe, biosample_id, feature_id, tidy) %>% unnest() %>% 
    mutate(prop = exp(estimate)) %>% .$prop %>% summary()
```


```{r}
theta_hat <- prop_hat_df %>% 
    mutate(theta1 = (t_fctr1 - t_fctr20)/(t_fctr0 - t_fctr20),
           theta2 = (t_fctr2 - t_fctr20)/(t_fctr0 - t_fctr20),
           theta3 = (t_fctr3 - t_fctr20)/(t_fctr0 - t_fctr20),
           theta4 = (t_fctr4 - t_fctr20)/(t_fctr0 - t_fctr20)) %>% 
    filter(!(t_fctr0 == 0 & t_fctr20 == 0), t_fctr0 != t_fctr20)
```

```{r}
theta_hat_df <- theta_hat %>% 
    select(pipe, biosample_id, feature_id, theta1, theta2, theta3, theta4) %>% 
    gather("theta","estimate", -pipe, -biosample_id, -feature_id)
```

```{r}
theta_hat_df %>% group_by(pipe, biosample_id, theta) %>% 
    summarise(est_mean = mean(estimate), 
              est_min = min(estimate), 
              est_max = max(estimate))
```


```{r}
theta_hat_df %>% filter(estimate > 0, estimate < 1) %>% 
ggplot() + 
    geom_density(aes(x = estimate, fill = theta))
```


# Session information 

## Git repo commit information
```{r}
library(git2r)
repo <- repository(path = "../")
last_commit <- commits(repo)[[1]]
```

The current git commit of this file is `r last_commit@sha`, which is on the `r branches(repo)[[1]]@name` branch and was made by `r last_commit@committer@name` on `r when(last_commit)`. The current commit message is `r last_commit@summary`. The repository is online at https://github.com/nate-d-olson/mgtst-pub  


## Platform Information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
```

## Package Versions
```{r}
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
      knitr::kable()
```