---
title: "Rank base feature categorization"
author: "Nate Olson"
date: '`r Sys.Date()`'
always_allow_html: yes
output:
  pdf_document: default
  html_document: default
---

__NEXT STEPS__
* Clean-up and revise
* Save file - to generate feature level summary df
* calculate logFC pre-post for consistency

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(tidyverse)
library(broom)
library(forcats)
```


## Objective 
Use normalized count ranks to identify features where the abundance estimates either increase or decrease with titration. 


## Approach
* Using normalized count estimates across PCR replicates generated with LIMMA calculate the count rank for titrations 1 - 4 for each feature.  
* Titrations 1-4 are used as this is where the most change is expected to occur.  
* A linear regression is fit to the titration level (1-4) and ranks. 
* The slope and adjusted R^2 values are used to categorize the features. 


## Notes
* Can use rank based regression slope and R2 values and logFC between pre- and post-exposure samples to obtain features for quantitative analysis. 


## Preparing input data
```{r}
count_est_df <- readRDS("../data/limma_count_est.rds") %>% 
      ## Small negative count values replaced with zeros
      ## small values attributed to model fit but want to make sure correct intercept term is used. 
      mutate(count_est = if_else(count_est < 0, 0, count_est))

zero_features <- count_est_df %>% 
      group_by(pipe, biosample_id, feature_id) %>% 
      summarise(total_count = sum(count_est)) %>% 
      filter(total_count > 0)

count_ranks <- count_est_df %>% 
      ## excluding features not observed in any titration for a biological replicate
      right_join(zero_features) %>% 
      ## only using titrations 1-5 as this is where the greatest change is expected
      ## the higher titrations are primarily pre-exposure with little expected change between titrations
      filter(t_fctr %in% 1:5) %>% 
      mutate(t_fctr = factor(t_fctr)) %>% 
      select(pipe, biosample_id, feature_id, t_fctr, count_est) %>% 
      group_by(pipe, biosample_id, feature_id) %>% 
      mutate(rank_dense = dense_rank(count_est)) %>% 
      mutate(t_level = as.numeric(t_fctr))
```

## Example feature set 
### Fitting linear model
```{r}
rank_fit <- count_ranks %>% 
      group_by(pipe, biosample_id, feature_id) %>% 
      nest() %>% 
      filter(feature_id %in% paste0("SV",1:9)) %>%
      mutate(fit = map(data, ~lm(rank_dense ~ t_level, data = .)))

## Extracting slope and adjusted R2 values
rank_fit_coef <- rank_fit %>% 
      mutate(fit_tidy = map(fit, tidy)) %>% 
      select(-data, -fit) %>%
      unnest()

rank_slopes <- rank_fit_coef %>% 
      filter(term == "t_level") %>% 
      select(pipe, biosample_id, feature_id, estimate) %>% 
      rename(slope = estimate)
```


### Regressions 
ranks and titration level
```{r fig.height = 8}
count_ranks %>% 
      filter(feature_id %in% paste0("SV",1:9)) %>% 
      ggplot() + 
      geom_line(aes(x = t_level, y = rank_dense)) + 
      geom_smooth(aes(x = t_level, y = rank_dense), method = "lm") + 
      facet_grid(feature_id~biosample_id)
```

High R2 and abs(slope) indicate features that titrate.
```{r}
rank_fit_summary <- rank_fit %>% 
      mutate(fit_glance = map(fit, glance)) %>% 
      select(-fit, -data) %>% 
      unnest() %>% 
      select(pipe, biosample_id, feature_id, adj.r.squared) %>% 
      left_join(rank_slopes)
```

```{r}
rank_fit_summary %>% 
      ggplot() + 
      geom_raster(aes(x = biosample_id, y = feature_id, fill = adj.r.squared)) +
      geom_text(aes(x = biosample_id, y = feature_id, label = round(slope,2)), color = "white")
```


## Larger Feature Set
```{r}
rank_fit <- count_ranks %>% 
      group_by(pipe, biosample_id, feature_id) %>% 
      nest()  %>% 
      mutate(fit = map(data, ~lm(rank_dense ~ t_level, data = .)))

rank_fit_coef <- rank_fit %>% 
      mutate(fit_tidy = map(fit, tidy)) %>% 
      select(-data, -fit) %>% 
      unnest() 

rank_slopes <- rank_fit_coef %>% 
      filter(term == "t_level") %>% 
      select(pipe, biosample_id, feature_id, estimate) %>% 
      rename(slope = estimate)

rank_fit_metrics <- rank_fit %>% 
      mutate(fit_glance = map(fit, glance)) %>% 
      select(-fit, -data) %>% 
      unnest()

feature_fit_stats <- rank_fit_metrics %>% 
      select(pipe, biosample_id, feature_id, adj.r.squared) %>% 
      left_join(rank_slopes)
```

### Rank slope and R2 relationship 
Size of the point indicates the number of features. 
```{r}
feature_fit_stats %>% 
      group_by(biosample_id, pipe, adj.r.squared, slope) %>% 
      summarise(count = n()) %>% 
      ggplot() + geom_point(aes(x = slope, y = adj.r.squared, size = count)) + facet_grid(biosample_id~pipe)
```

No clear separation between groups categorization values choosen arbitrarily - need a better method to select features.
```{r}
feature_fit_cat <- feature_fit_stats %>%  
      mutate(feat_cat = case_when(adj.r.squared <= 0.5 ~ "low-r2",
                                  slope > 0.5 & adj.r.squared > 0.5 ~ "pos-slope",
                                  slope < -0.5 & adj.r.squared > 0.5 ~ "neg-slope",
                                  abs(slope) <= 0.5 ~ "low-slope")
      )
```


```{r}
feature_fit_cat %>%
      group_by(pipe, biosample_id, feat_cat) %>% 
      summarise(count = n()) %>% 
      spread(biosample_id, count)
```


## Points to consider


Potentially exclude or annotate low abundance features, e.g. SV1131, 118, 1244, and 1607.



```{r}
feature_fit_cat %>% filter(feature_id %in% paste0("SV", c(11131,118,1244, 1607))) 

count_ranks %>% left_join(feature_fit_cat) %>% 
      filter(feature_id %in% paste0("SV", c(11131,118,1244, 1607))) %>% 
      ungroup() %>% 
      mutate(feature_id = fct_reorder(feature_id, -adj.r.squared)) %>% 
      ggplot() + 
      geom_line(aes(x = t_level, y = count_est)) + 
      geom_smooth(aes(x = t_level, y = count_est), method = "lm") + 
      facet_wrap(~feature_id, scales = "free_y")
```










Features SV271 and SV459 look okay but have low R2 values. 
Titration 5 is the outlier. 
The ERCC qPCR data indicated that titrations 5, 10, and 15 were not consistent with titrations 1-4. 
For the quantitative analysis maybe focus on titrations 1-4. 

```{r}
feature_fit_cat %>% filter(feature_id %in% c("SV271", "SV459"),
                           biosample_id == "E01JH0004") %>% 
      left_join(count_ranks) %>% 
      ungroup() %>% 
      mutate(feature_id = fct_reorder(feature_id, -abs(slope))) %>% 
      ggplot() + 
      geom_line(aes(x = t_level, y = count_est)) + 
      geom_smooth(aes(x = t_level, y = count_est, color = feat_cat), method = "lm") +
      facet_grid(.~feature_id, scales = "free_y")
```
