---
title: "logFC Microbiome-Scale Assessment"
output: html_notebook
---

```{r logFCsetup, include=FALSE}
library(tidyverse)
library(ggridges)
```

```{r logFCmunge, include = FALSE}
### Loading Data ---------------------------------------------------------------
### RDS file generated in 2017-10-19_logFC-error-metrics.RDS 
logFC_pre <- readRDS("~/Desktop/logFC_pre.RDS")
```


## Background
* Using mixture design to obtain expected log fold-change values 
* Comparing log fold-change estimates obtained using EdgeR and negative binomial 


## Agreement between logFC estimators 
* The log fold-change estimates were strongly correlated between the two methods (\@ref(fig:logFCestCor)).  
* NB log fold-change estimates or -Inf and Inf are consistent with the tails distribution of the EdgeR estimates.  

```{r logFCestCor, fig.cap = "Log fold-change estimates using EdgeR and negative binomial relative abundance values were strongly correlated for pre-specific and pre-dominant features. The blue linear indicates a linear model of the relationship between the two estimators and the orange line is the expected 1-to-1 relationship. The titration comparisons with positive and negative infinity negative log fold-change estimates, points along the left and right y-axis, were excluded from the Pearson correlation estimate."}
pearson_cor <- logFC_pre %>% filter(T1 != 0, T2 != 20) %>% 
      filter(nb_logFC != Inf, nb_logFC != -Inf) %>%
      {cor(.$logFC, .$nb_logFC)} %>% 
      {paste0("Pearson: ",round(.,2))}

spearman_cor <- logFC_pre %>% filter(T1 != 0, T2 != 20) %>% 
      {cor(.$logFC, .$nb_logFC, use = "complete.obs", method = "spearman")} %>%
      {paste0("Spearman: ",round(.,2))}

logFC_pre %>% filter(T1 != 0, T2 != 20) %>% 
      ggplot() + 
      geom_point(aes(x = nb_logFC, y = logFC), alpha = 0.05) + 
      geom_smooth(aes(x = nb_logFC, y = logFC), method = "lm") + 
      geom_abline(aes(intercept = 0, slope = 1), color = "darkorange") + 
      geom_text(aes(x = -3.5, y = 9, label = paste0(pearson_cor,"\n",spearman_cor))) + 
      theme_bw() + 
      labs(x = "Negative Binomial", y = "EdgeR", title = "log fold-change estimate comparison")
```





## Agreement between log fold-change estimates and expected values 
* When the relationship between the log fold-change estimates and expected values is summarized using a linear model the estimates are consistent with the expected values for both EdgeR and negative binomial estimates (Fig. \@ref(fig:logFCestVexp)).  
* However, there were a number of extreme outlier log fold-change estimates.  
* The log fold-change values ranged from $\pm 10$ for EdgeR estimates and $\pm \infty$ for negative binomial estimates \@ref(fig:logFCestVexp).  
* However, for pre-specific and pre-dominant features the expeted log fold-change values range from 0 to 1. 


```{r logFCestVexp, fig.cap = "Relationship between the log fold-change estimates and expected values. Blue line represents a linear model summarising the relationship and the orange line is the expected 1-to-1 relationship.", message = FALSE, warning = FALSE}
logFC_pre %>% 
      filter(T1 != 0, T2 != 20) %>% 
      dplyr::select(biosample_id, pipe, feature_id, exp_logFC, logFC, nb_logFC) %>% 
      rename(nb = nb_logFC, edgeR = logFC) %>% 
      gather("Method","logFC", -biosample_id, -pipe, -feature_id, -exp_logFC) %>% 
      ggplot() + 
      geom_point(aes(x = exp_logFC, y = logFC), alpha = 0.15) +
      geom_smooth(aes(x = exp_logFC, y = logFC), method = "lm") + 
      geom_abline(aes(intercept = 0, slope = 1), color = "darkorange") + 
      facet_wrap(~Method) + theme_bw()
```



## Individual and Pipeline Effect

* Next we wanted to see how the relationship between the expected and estimated log-fold change varies by pipeline and individual.  
* The relationship between the log fold-change estimates and expected values is more dependent on the individual than the pipeline (Fig \@ref(fig:logFCestVexpLM)).  
* For the E01JH0011 and the Mothur pipeline the log-fold change estimates were more consistent with the expected values for the edgeR estimates than the negative binomial estiamtes. Though the Mothur results are inconsistent with the other pipelines.  


```{r logFCestVexpLM, message = FALSE, warning = FALSE}
logFC_pre %>% 
      filter(T1 != 0, T2 != 20) %>% 
      dplyr::select(biosample_id, pipe, feature_id, exp_logFC, logFC, nb_logFC) %>% 
      rename(nb = nb_logFC, edgeR = logFC) %>% 
      gather("Method","logFC", -biosample_id, -pipe, -feature_id, -exp_logFC) %>%
      ggplot() + 
      geom_smooth(aes(x = exp_logFC, y = logFC, color = pipe), method = "lm") + 
      geom_abline(aes(intercept = 0, slope = 1), linetype = 2) + 
      facet_grid(Method~biosample_id) + theme_bw() + 
      labs(x = "Expected", y = "Estimate", color = "Pipeline")
```

## Evaluate parameters impacting error  

* The log fold-change estimate error (expected - estimate) was used to evaluate what if any abundance paramters were associated with the extreme/ outlier log-fold change estimates. 
* Only the log fold-change estimate was strongly correlated with the log-fold error, therefore none of the relative abundance parameters could be attributed to the extreme log fold-change estimates. 
* For edgeR log fold-change estimates the negative binomial estimates for titration 2 had more negative error rates, potentially due to unobserved count data. 
* The error rate increased with higher titration 1 negative binomial estimates, though this is likely an artifact of the experimental design and not an indicator that log fold-change estimates for higher relative abundance features are less accurate than lower relative abundance features. 
* These results indicate that the extreme log fold-change estimates are either a feature specific effect or due to an unaccounted for parameter.  

```{r}
logFC_error_param <- logFC_pre %>% 
      filter(T1 != 0, T2 != 20) %>% 
      dplyr::select(biosample_id, pipe, feature_id, exp_logFC, logFC, 
                    nb_logFC, pre_nb, post_nb, T1_nb, T2_nb) %>% 
      rename(nb = nb_logFC, edgeR = logFC) %>% 
      gather("Method","logFC", -biosample_id, -pipe, -feature_id, 
             -exp_logFC, -pre_nb, -post_nb, -T1_nb, -T2_nb) %>% 
      filter(biosample_id == "E01JH0016") %>% 
      mutate(error = logFC - exp_logFC) %>% 
      group_by(biosample_id, pipe, Method, error) %>% 
      select(exp_logFC, logFC, post_nb, pre_nb, T1_nb, T2_nb) %>% 
      gather("key", "value", -biosample_id, -pipe, -Method, -error) 

logFC_error_param %>% ggplot() + 
      geom_point(aes(x = value, y = error)) + 
      geom_smooth(aes(x = value, y = error)) + 
      facet_grid(Method~key, scales = "free") + theme_bw()
```

### Deviations from expectation by pipeline and individual
* Consistent deviation from the expected value by individual indicates that differences in proportion of bacterial DNA does account for some of the disagrement between the log fold-change estimates and expected values.  
* The consistency in the deviation of the log fold-change estimates from the expected values across pipelines and log fold-change methods indicates the deviation is due to biases in sample preparation, specifically PCR amplification. 

```{r}
logFC_error <- logFC_pre %>% 
      filter(T1 != 0, T2 != 20) %>% 
      dplyr::select(biosample_id, pipe, feature_id, exp_logFC, logFC, nb_logFC) %>% 
      rename(nb = nb_logFC, edgeR = logFC) %>% 
      filter(abs(nb) != Inf) %>% 
      gather("Method","logFC", -biosample_id, -pipe, -feature_id, -exp_logFC) %>% 
      mutate(error = logFC - exp_logFC)
      
logFC_error %>% 
      ggplot() + 
      geom_density_ridges(aes(x = error, y = pipe), alpha = 0.5) + 
      facet_grid(Method~biosample_id) + theme_bw()
```




## Conclusions
* Agreement between negative binomial and edgeR log fold-change estimates. edgeR provides estimates (extreme values in plots below) when a feature is not observed in one of the two titrations being compared. 
* Agreement between the log fold-change estimates and expected values are individual and not pipeline dependent
* Parameters impacting error metrics - features not observed in one to the two titrations being compared, others ... extreme log fold-change estimates cannot be explained by abundance alone  

