---
title: "2017-02-09-Normalization-Dev-Df"
author: "Nate Olson"
date: '`r Sys.Date()`'
output:
  pdf_document: default
  html_document: default
---

```{r setup, message = FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd)
```

## Objective
Develop a data frame with normalized count values for the different pipelines and normalization methods


## Code for analysis
### Loading Pipeline Data
```{r}
mrexp_files <- list(
      dada2  = "../data/mrexp_dada2.RDS",
      mothur = "../data/mrexp_mothur.RDS",
      qiime  = "../data/mrexp_qiime_refclus_nochimera.RDS"
)
mrexp <- mrexp_files %>% map(readRDS) 

#Extracting metadata

meta_dat <- mrexp$mothur %>% pData()

##labeling PCR replicates
half1 <- paste(rep(c("A","B","C","D","E","F","G","H"), each = 6), 1:6, sep = "_")
sam_dat <- meta_dat %>% 
      mutate(pcr_half = if_else(pos %in% half1, "1","2"),
             pcr_rep = paste0(pcr_16S_plate,":",pcr_half)) %>% 
      select(sampleID, dilution,sam_names, pcr_rep) %>% 
      dplyr::rename(samID = sam_names)
```

Adding `sam_names` to `pData` for DADA2 and QIIME MRexperiments, and converting dilutions to factors.
```{r}
pdat <- pData(mrexp$dada2)
pdat$sam_names <- rownames(pdat)
pdat$dilution <- pdat$dilution %>% factor()
pData(mrexp$dada2) <- pdat 

pdat <- pData(mrexp$mothur)
pdat$dilution <- pdat$dilution %>% factor()
pData(mrexp$mothur) <- pdat 

pdat <- pData(mrexp$qiime) 
pdat$sam_names <- rownames(pdat)
pdat$dilution <- pdat$dilution %>% factor()
pData(mrexp$qiime) <- pdat
```

#### Extracting Raw, Normalized, and Transformed Count Data
__NOTE__ Normalization and transformation order impacts results. 
What is the appropriate ordering?

__TODO__ Move to lib
```{r}
calc_raw_counts <- function(mrexp){
      mrexp@assayData$counts %>% as_tibble() %>% 
            rownames_to_column(var = "featureIndices") %>% 
            gather("samID","count",-featureIndices) %>% 
            left_join(sam_dat) 
}

calc_css_counts <- function(mrexp, norm = TRUE,log = TRUE,sl = 1000, p = 0.75){
      mrexp %>% cumNorm(p = p) %>% 
            MRcounts(norm, log, sl) %>% as_tibble() %>% 
            rownames_to_column(var = "featureIndices") %>% 
            gather("samID","count",-featureIndices) %>% 
            left_join(sam_dat) 
}

# TSS from http://mixomics.org/mixmc/normalisation/ 
calc_tss_counts <- function(mrexp){
      mrexp@assayData$counts %>% {apply(., 2, function(x){ x/sum(x) })} %>% 
            as_tibble() %>% rownames_to_column(var = "featureIndices") %>%
            gather("samID","count",-featureIndices) %>% 
            left_join(sam_dat) 
}

calc_tsslog_counts <- function(mrexp){
      mrexp@assayData$counts %>% 
            {apply(., 2, function(x){ x/sum(x) })} %>% {log2(. + 1)} %>%
            as_tibble() %>% rownames_to_column(var = "featureIndices") %>%
            gather("samID","count",-featureIndices) %>% 
            left_join(sam_dat) 
}

## DESeq method median of ratios -
## %%TODO%% replace with ref based normalization Deseq - not tmm
calc_dsq_counts <- function(mrexp){
  mrexp@assayData$counts %>% {. + 1} %>% 
            {./estimateSizeFactorsForMatrix(.)} %>% 
    as_tibble() %>% rownames_to_column(var = "featureIndices") %>%
    gather("samID","count",-featureIndices) %>% 
    left_join(sam_dat) 
}

calc_dsqlog_counts <- function(mrexp){
  mrexp@assayData$counts %>% {. + 1} %>% 
    {./estimateSizeFactorsForMatrix(.)} %>% {log2(. + 1)} %>%
    as_tibble() %>% rownames_to_column(var = "featureIndices") %>%
    gather("samID","count",-featureIndices) %>% 
    left_join(sam_dat) 
}
```

__TODO__ move to src
```{r message = FALSE}
raw_counts <- mrexp %>% map_df(calc_raw_counts, .id = "pipe")
rawlog_counts <- mrexp %>% map_df(calc_raw_counts, .id = "pipe") %>% 
      mutate(count = log2(count + 1))
```

```{r message = FALSE}
uqs_counts <- mrexp %>% map_df(calc_css_counts, p = 0.75,sl = 1, log = FALSE, .id = "pipe")
uqslog_counts <- mrexp %>% map_df(calc_css_counts, sl = 1, .id = "pipe")
```

```{r message = FALSE}
css_counts <- mrexp %>% 
  {map_df(.x=., .f=~calc_css_counts(.,log = FALSE, p = cumNormStat(.), sl = 1), .id = "pipe")}
csslog_counts <- mrexp %>% 
  {map_df(.x=., .f=~calc_css_counts(., p = cumNormStat(.), sl = 1),  .id = "pipe")}
```


```{r message = FALSE}
tss_counts <- mrexp %>% map_df(calc_tss_counts, .id = "pipe")
tsslog_counts <- mrexp %>% map_df(calc_tsslog_counts, .id = "pipe") 
```

```{r message = FALSE}
dsq_counts <- mrexp %>% map_df(calc_dsq_counts, .id = "pipe")
dsqlog_counts <- mrexp %>% map_df(calc_dsqlog_counts, .id = "pipe") 
```


Combine into a single data frame
```{r}
count_df <- list(raw = raw_counts, rawlog = rawlog_counts,
                 uqs = uqs_counts, uqslog = uqslog_counts, 
                 css = css_counts, csslog = csslog_counts, 
                 tss = tss_counts, tsslog = tsslog_counts,
                 dsq = dsq_counts, dsqlog = dsqlog_counts) %>% 
      bind_rows(.id = "norm_method")
```

### Save Data Frame as RDS
```{r}
count_df %>% saveRDS("../data/normalize_count_df.rds") 
```

