---
title: "logFC error Normalization method comparison"
output: 
      bookdown::pdf_document2: 
            toc: FALSE
---

```{r logFCsetup, include=FALSE}
library(tidyverse)
library(ggridges)
library(ggpubr)
```



```{r logFCmunge, include = FALSE}
### Loading Data ---------------------------------------------------------------
### RDS file generated in 2017-10-19_logFC-error-metrics.RDS 
logFC_pre <- readRDS("~/Desktop/norm_logFC_pre.RDS")

logFC_feature_summary <- readRDS("~/Desktop/norm_logFC_feature_summary.RDS") %>% 
      mutate(slope_error = 1 - slope)

## Calculating error rate
logFC_error <- logFC_pre %>% 
      filter(T1 != 0, T2 != 20) %>% 
      mutate(abs_error = abs(logFC - exp_logFC),
             error = logFC - exp_logFC)
```

Log fold-change correlations between normalization method
```{r}
prepost_logFC <- readRDS("~/Desktop/norm_logFC_prepost.RDS")
library(GGally)
prepost_logFC %>% 
    select(norm_method, biosample_id, feature_id, logFC) %>% 
    spread(norm_method, logFC) %>% 
    ggpairs(aes(color = biosample_id), columns = 3:8) + theme_bw()
```

Comparison of log fold-change distributions across normalization methods.
UQ has consistently lower log fold-change estimates compared to the other normalization methods. 
In general the distribution of logFC estimates is consistent across individuals for the different normalization methods excluding RLE. 

```{r}
prepost_logFC %>% ggplot() + 
    geom_boxplot(aes(x = norm_method, y = logFC, color = biosample_id)) + 
    theme_bw()
```

Few features categorized a pre-specific and dominant for UQ normalized data due to the lower distribution of log fold-change estimates compared to the other methods. 
One option is to choose a common sent of features to compare across normalization methods. 
```{r preCountTbl, echo = FALSE}
logFC_pre %>% 
      filter(T1 == 0, T2 == 20) %>%
      mutate(Type = if_else(T00 == 0, "specific", "dominant")) %>% 
      group_by(norm_method, biosample_id, Type) %>% 
      summarise(count = n()) %>% 
      spread(norm_method, count, fill = 0) %>% 
      dplyr::rename(Individual = biosample_id) %>% 
      knitr::kable(caption = "Number of pre-specific and pre-dominant features by individual and normalization method for Mothur", booktabs = TRUE)
```
Overlap in pre specific and pre-dominant features between normalization mehtods. 
Might want to consider using the features that are classified as pre-specific or pre-dominant in 5 or the 6 method. 
```{r}
library(UpSetR)
logFC_pre_set <- logFC_pre %>% select(biosample_id, norm_method, feature_id) %>% unique() %>% 
    add_column(x = 1) %>% 
    spread(norm_method, x, fill = 0)
upset(as.data.frame(logFC_pre_set), nset = 7)
```
```{r}
logFC_pre_set %>% filter(biosample_id == "E01JH0004") %>% as.data.frame() %>% 
                             upset(nset = 6)
```


```{r}
logFC_pre_set %>% filter(biosample_id == "E01JH0011") %>% as.data.frame() %>% 
                             upset(nset = 6)
```

```{r}
logFC_pre_set %>% filter(biosample_id == "E01JH0016") %>% as.data.frame() %>% 
                             upset(nset = 6)
```

```{r}
logFC_pre_set %>% filter(biosample_id == "E01JH0017") %>% as.data.frame() %>% 
                             upset(nset = 6)
```

```{r}
logFC_pre_set %>% filter(biosample_id == "E01JH0038") %>% as.data.frame() %>% 
                             upset(nset = 6)
```

```{r logfcNormComp, fig.cap = "Impact of normalization methods on the agreement between log fold-change estimates and expected values for pre-specific and pre-dominant features. (A) Linear model relating the log fold-change estimates with the expected values by individual and normalization method. (B) Distribution of log fold-change (B) absolute error and (C) error by normalization method and individual.", echo=FALSE, message = FALSE, fig.height = 8}
logFCestVexp <- logFC_error %>% 
      filter(T1 != 0, T2 != 20) %>% 
      ggplot() + 
      geom_smooth(aes(x = exp_logFC, y = logFC, color = norm_method), method = "lm") +
      geom_abline(aes(intercept = 0, slope = 1), color = "grey20", linetype = 2) + 
      facet_wrap(~biosample_id, ncol = 1, scales = "free_y") + theme_bw()  +
      labs(x = "Expected", y = "Estimate", color = "Normalization") 

logFC_abs_error_dist <- logFC_error %>% 
      group_by(biosample_id, norm_method) %>% 
      mutate(med_error = median(abs_error)) %>% 
      ggplot() + 
      geom_density_ridges(aes(x = abs_error, y = norm_method, color = norm_method), 
                          alpha = 0.5, stat = "binline", bins = 30, draw_baseline = FALSE)  + 
      geom_text(aes( x = -1, y = norm_method, label = round(med_error,2))) + 
      facet_wrap(~biosample_id, ncol = 1) + theme_bw() + 
      labs(x = "|Error|", y = "Normalization", color = "Normalization") +
      theme(legend.position = "none")

logFC_error_dist <- logFC_error %>% 
      group_by(biosample_id, norm_method) %>% 
      mutate(med_error = median(error)) %>% 
      ggplot() + 
      geom_density_ridges(aes(x = error, y = norm_method, color = norm_method), 
                          alpha = 0.5, stat = "binline", bins = 30, draw_baseline = FALSE)  + 
      geom_text(aes( x = -10, y = norm_method, label = round(med_error,2))) + 
      facet_wrap(~biosample_id, ncol = 1) + theme_bw() + 
      labs(x = "Error", y = "Normalization", color = "Normalization") +
      theme(legend.position = "none")

ggarrange(logFCestVexp, 
          logFC_error_dist, 
          logFC_abs_error_dist, 
          labels = "AUTO",
          align = "v", 
          ncol = 3, nrow = 1,
          common.legend = TRUE,
          legend = "bottom")
```

```{r}
logFC_error %>% filter(pa_mixed == 28) %>% 
    ggplot() + geom_boxplot(aes(x = biosample_id, y = abs_error, color = norm_method))
```


