---
title: "Measurement Assessment - Variance Data Frames"
author: "Nate Olson"
date: '`r Sys.Date()`'
output:
  pdf_document: default
  html_document: default
---

```{r setup, message = FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd)
```

## Objective
Calculate feature level variance for raw and normalized count table values as well as the coefficient of variation for each pipeline and normalization method.

## Loading Data 
Normalized count data from `2017-02-09-Normalization-df-Dev.Rmd`
```{r}
count_df <- readRDS("../data/normalize_count_df.rds")
```

### Count Value Variance 
```{r}
# calculating mean and variance for technical count replicates
count_var_df <- count_df %>% 
  group_by(pipe, norm_method, sampleID, dilution, featureIndices) %>% 
  summarise(mean_count  = mean(count),
            var_count   = var(count),
            cv_count = sd(count)/mean_count)
```

```{r}
glimpse(count_var_df)
```

__TODO__ Look into cv_count values that are NaN - most likely features with mean counts of 0.

#### Save Data Frame as RDS
```{r}
count_var_df %>% saveRDS("../data/normalize_count_variance_df.rds") 
```
