---
title: "Investigation of Extreme NB Prop Estimates"
author: "Nate Olson"
date: '`r Sys.Date()`'
output:
  bookdown::pdf_document2: 
    toc: FALSE
---


```{r}
library(tidyverse)
```

## Objective
Large relative abundance estimates were observed for the different sequencing runs. 
Characterize and debug extreme relative abundance estimates.

## Conclusions
No observed counts for all four replicates of the intercept point (unmixed post-exposure samples) resulted in poor model fit. 
A potential solution is to exclude the post-exposure titration pre-exposure specific samples. 

## Analysis
### JHU1
```{r}
nb_extreme <- readRDS("~/Desktop/nb_counts.RDS") %>% 
      filter(nb_prop > 1 | nb_prop < 0)
```

No values less than 0 but one values > 1
```{r}
nb_extreme
```

```{r}
nb_extreme_fit <- readRDS("~/Desktop/nb_fit.RDS") %>% right_join(nb_extreme)
```
```{r}
nb_extreme_fit$fit[[1]]
```


```{r}
nb_extreme_fit$data[[1]]
```

```{r}
nb_extreme_fit$data[[1]] %>% ggplot() + geom_point(aes(x = t_fctr, y = count/total_abu))
```


```{r}
test_dat <- nb_extreme_fit$data[[1]] %>% filter(t_fctr != 0)
```

```{r}
fit <- MASS::glm.nb(count ~ t_fctr + offset(log(total_abu)), data = test_dat, link = "log")
```

```{r}
fit
```

Estimates excluding T0 
```{r}
fit_coefs <- fit %>% coefficients()
c(fit_coefs[1], fit_coefs[1] + fit_coefs[2:9]) %>% exp() %>% plot()
```

### NIST 1
```{r}
nist1_extreme <- readRDS("~/Desktop/nist1_nb_counts.RDS") %>% 
      filter(nb_prop < 0 | nb_prop > 1)
```

Unexpected proportions, values should be between 0 and 1. 
```{r}
nist1_extreme
```

```{r}
nist1_extreme_fit <- readRDS("~/Desktop/nist1_nb_fit.RDS") %>% 
      right_join(nist1_extreme)
```

```{r}
nist1_extreme_data <- nist1_extreme_fit %>% 
      select(pipe, biosample_id, feature_id, t_fctr, data) %>% 
      unnest() %>% 
      mutate(ext_val = if_else(t_fctr == t_fctr1, T, F)) 
```

All of the titrations with extreme value estimates have no observed counts for at least one PCR at each titration and no observed counts for the unmixed post treatment sample. 

```{r}
filter(nist1_extreme_data, ext_val == F) %>% 
      ggplot() + 
      geom_point(aes(x = t_fctr1, y = count/total_abu), color = "blue") + 
      geom_point(data = filter(nist1_extreme_data, ext_val == T), 
                 aes(x = t_fctr1, y = count/total_abu, color = ext_val), color = "orange") + 
      facet_wrap(~feature_id * biosample_id, scales = "free_y")
```


### NIST 2
```{r}
nist2_extreme <- readRDS("~/Desktop/nist2_nb_counts.RDS") %>% 
      filter(nb_prop < 0 | nb_prop > 1)
```

Unexpected proportions, values should be between 0 and 1. 
```{r}
nist2_extreme
```

```{r}
nist2_extreme_fit <- readRDS("~/Desktop/nist2_nb_fit.RDS") %>% 
      right_join(nist2_extreme)
```

```{r}
nist2_extreme_data <- nist2_extreme_fit %>% 
      select(pipe, biosample_id, feature_id, t_fctr, data) %>% 
      unnest() %>% 
      mutate(ext_val = if_else(t_fctr == t_fctr1, T, F)) 
```

All of the titrations with extreme value estimates have no observed counts for at least one PCR at each titration and no observed counts for the unmixed post treatment sample. 

```{r}
filter(nist2_extreme_data, ext_val == F) %>% 
      ggplot() + 
      geom_point(aes(x = t_fctr1, y = count/total_abu), color = "blue") + 
      geom_point(data = filter(nist2_extreme_data, ext_val == T), 
                 aes(x = t_fctr1, y = count/total_abu, color = ext_val), color = "orange") + 
      facet_wrap(~feature_id * biosample_id, scales = "free_y")
```


# Session information 

## Git repo commit information
```{r}
library(git2r)
repo <- repository(path = "../")
last_commit <- commits(repo)[[1]]
```

The current git commit of this file is `r last_commit@sha`, which is on the `r branches(repo)[[1]]@name` branch and was made by `r last_commit@committer@name` on `r when(last_commit)`. The current commit message is `r last_commit@summary`. The repository is online at https://github.com/nate-d-olson/mgtst-pub  


## Platform Information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
```


## Package Versions
```{r}
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
      knitr::kable()
```
