---
title: "Measurement Assessment - Bias Data Frame"
author: "Nate Olson"
date: '`r Sys.Date()`'
output:
  pdf_document: default
  html_document: default
---

```{r setup, message = FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd)
```

## Objective
Generate a data frame with feature level expected count values for measurement assessment bias analysis. 


## Loading Data 
Normalized count data from `2017-02-09-Normalization-df-Dev.Rmd`
```{r}
count_df <- readRDS("../data/normalize_count_df.rds")
```

## Count Value Bias 
Relationship between the observed and expected count values. Expected count ($C_{exp}$)values calculated using the unmixed sample count values (unmixed pre - $C_{pre}$ and unmixed post - $C_{post}$) and proportion of unmixed post in the titration $p$. Proportion is defined as $p = 2^{-t}$, and $t$ is the titration factor. 

$$
C_{exp} = [C_{post}\times p] + [C_{pre} \times (1-p)]
$$

The expected values are calculated based on pre and post unmixed samples by replicate (defined as half of PCR plate). 

__NOTE__ The pre and post treatment samples were mixed up in the PCR plate therefore the pre-treatment sample dilution is 0, and post-treatment sample is -1. 

```{r}
pre_count <- count_df %>% filter(dilution == 0) %>% 
      dplyr::rename(pre = count) %>% select(-dilution, -samID)
post_count <- count_df %>% filter(dilution == -1) %>% 
      dplyr::rename(post = count) %>% select(-dilution, -samID)
pre_post_count <- left_join(pre_count, post_count)
rm(pre_count, post_count)

count_exp_obs <- count_df %>% 
      filter(!(dilution %in% c(0,-1))) %>%
      left_join(pre_post_count) %>% 
      mutate(p = 2^(-dilution), exp_count = pre * (1-p) + post * p)
```

```{r}
count_exp_obs %>% glimpse()
```

#### Save Data Frame as RDS
```{r}
count_exp_obs %>% saveRDS("../data/expected_counts_df.rds") 
```

## Metrics for evaluating count values 
__TODO__ Update metric - Hector's metric (until I get the real name)
$$E = \frac{\sqrt{1/n \sum_i^n(obs-exp)^2}}{\sqrt{exp^2}}$$

```{r}
count_hscore <- count_exp_obs %>% 
      group_by(pipe, norm_method, featureIndices,sampleID, dilution) %>% 
      mutate(resid = count - exp_count,
             sq_resid = resid^2) %>% 
      mutate(hscore = sqrt((mean((count - exp_count)^2)) / (mean((exp_count)^2))))
```

```{r}
count_hscore %>% filter(!is.na(hscore)) %>% glimpse()
```

#### Save Data Frame as RDS
```{r}
count_hscore %>% saveRDS("../data/expected_count_bias_metric_df.rds") 
```