---
title: "Negative Binomial Proportions"
author: "Nate Olson"
date: '`r Sys.Date()`'
always_allow_html: yes
output:
  pdf_document: default
  html_document: default
---

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd) 
pipeline_dir <- "../../mgtst_pipelines"
mrexp <- get_mrexp(pipeline_dir)
```

## Objective
Generate a data_frame with observed and predicted count values for individual PCR replicates. 
Expected values are calculated using unmixed PCR count table values from the same set of PCR replicates, same half of one of the replicate 96 well plates. 


### Negative Binomal for Weighted Count Estimates
Calculating proportion of pre and post counts using negative binomial. 

- $p_{ijk}$ is the proportion of feature $i$ in PCR $k$ of sample $j$ where a sample is defined as an individual unmixed or mixed samples for a biological replicate.  
- $r_{jk}$ is the total feature abundance for sample $j$, sum of all feature counts not the number of sequences generated for the sample.  
- $v_{ijk}$ is the variance of feature $i$ in PCR replicate $j$ of sample $k$.  

$$v_{i,j,k} = \frac{p_{ijk}r_{jk}}{(1-p_{ijk})^2}$$

- $w_{ijk}$ is the weight function

$$w_{ijk} = \frac{v_{ijk}^{ -1}}{\sum_{k \in j}v_{ijk}^{ -1}}$$


- $q_{ij}$ - the weighted count estimate for feature $ij$

$$q_{ij} = \sum_{k\in j} w_{ijk}q_{ijk}$$ 


## Functions for Processing Data 
```{r}
## Extracting a tidy dataframe with count values from MRexpiment objects
get_count_df <- function(mrobj, agg_genus = FALSE, css = TRUE){
      if(agg_genus){
            mrobj <- aggregateByTaxonomy(mrobj, lvl = "Rank6", 
                                         norm = FALSE, log = FALSE, sl = 1)
      }
      
      if(css == TRUE){
            mrobj <- cumNorm(mrobj, p = 0.75)
            count_mat <- MRcounts(mrobj, norm = TRUE, log = FALSE, sl = 1000)
      }else{
            count_mat <- MRcounts(mrobj, norm = FALSE, log = FALSE, sl = 1)
      }
       count_mat %>%  
            as.data.frame() %>% 
            rownames_to_column(var = "feature_id") %>% 
            gather("id","count", -feature_id)
}

## Calculating feature level pre and post proportions
calc_prop <- function(count_df){
      ## Estimating $q_{i}$ for titrations
      nb_est <- count_df %>% 
            mutate(prop = count/total_abu, 
                   prop_var = (prop * total_abu)/(1 - prop)^2,
                   inv_var = 1/prop_var) %>% 
            group_by(pipe, biosample_id, t_fctr, feature_id) %>% 
            mutate(weight = inv_var / sum(inv_var)) %>% 
            summarise(prop_est = sum(weight*prop)) 
}

```


### Weighted Proportions
Using negative binomial weighted proportions to calculate feature proportion across PCR replicates
```{r}
count_df <- mrexp %>% map_df(get_count_df, css = FALSE, .id = "pipe") %>% 
      left_join(pData(mrexp$dada2)) %>% 
      filter(biosample_id != "NTC") %>% 
      select(pipe, biosample_id, id, pcr_rep, feature_id, t_fctr, count) %>% 
      group_by(biosample_id, pipe, id) %>% 
      mutate(total_abu = sum(count))


prop_df <- calc_prop(count_df)

prop_df %>% saveRDS("../data/nb_feature_prop_df.rds") 

``` 

### Sanity Check
Make sure all props are between 0 and 1
```{r}
prop_df %>% summary()
```


Check E. coli
```{r}
prop_df %>% 
      filter(feature_id %in% c("SV1","Otu00001","782953")) %>% 
      mutate(titration = as.numeric(as.character(t_fctr))) %>% 
      ggplot() + geom_point(aes(x = titration, y = prop_est)) + 
      facet_grid(pipe~biosample_id)
```



Check if features sum to 1 by titration - do not necessarily need to but should be close

```{r}
total_prop_df <- prop_df %>% 
      filter(!is.nan(prop_est)) %>% 
      group_by(pipe, biosample_id, t_fctr) %>% 
      summarise(total_prop = sum(prop_est)) 
```

```{r}
total_prop_df %>% ggplot() + geom_histogram(aes(x = total_prop))
```

```{r}
total_prop_df %>% filter(total_prop < 0.9)
```



### Checking out NaN
Issue with NaN when prop_var is 0, which occurs when counts are not observed for one or more of the four PCR replicates. 
For quantitative analysis only including features observed in all 4 PCR replicates
```{r}
count_1 <- count_df %>% 
      mutate(prop = count/total_abu, 
                   prop_var = (prop * total_abu)/(1 - prop)^2,
                   inv_var = 1/prop_var)
count_1 %>% summary()
```


