---
title: "Negative Binomial Linear Model Count Estimates NIST1"
author: "Nate Olson"
date: '`r Sys.Date()`'
always_allow_html: yes
output:
  pdf_document: default
  html_document: default
---

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(MASS)
library(modelr)
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd) 
pipeline_dir <- "../../mgtst_runs/nist1" 

get_mrexp <- function(pipelines_dir){
      mrexp_files <- list(
            dada2 = file.path(pipeline_dir, "dada2/dada_mrexp.rds"),
            mothur =  file.path(pipeline_dir, "mothur/mothur_mrexp.rds"),
            qiime =  file.path(pipeline_dir, "qiime/qiime_mrexp.rds")
      )
      
      mrexp_files %>% map(readRDS)
}
mrexp <- get_mrexp(pipeline_dir)
```

## Objective
Calculate summary count values across PCR replicates values that take into account differences in sampling depth. 

## Approach
1. glm.nb - the negative binomial glm function that is part of the MASS package


## glm.nb 
```{r}
get_count_df <- function(mrobj, agg_genus = FALSE, css = TRUE){
      if(agg_genus){
            mrobj <- aggregateByTaxonomy(mrobj, lvl = "Rank6", 
                                         norm = FALSE, log = FALSE, sl = 1)
      }
      
      if(css == TRUE){
            mrobj <- cumNorm(mrobj, p = 0.75)
            count_mat <- MRcounts(mrobj, norm = TRUE, log = FALSE, sl = 1000)
      }else{
            count_mat <- MRcounts(mrobj, norm = FALSE, log = FALSE, sl = 1)
      }
       count_mat %>%  
            as.data.frame() %>% 
            rownames_to_column(var = "feature_id") %>% 
            gather("id","count", -feature_id)
}

count_df <- mrexp %>% map_df(get_count_df, css = FALSE, .id = "pipe") %>% 
      left_join(pData(mrexp$dada2)) %>% 
      filter(biosample_id != "NTC") %>% 
      select(pipe, biosample_id, id, pcr_rep, feature_id, t_fctr, count) %>% 
      group_by(biosample_id, pipe, id) %>% 
      mutate(total_abu = sum(count))

nonzero_count_features <- count_df %>% 
      mutate(t_fctr = factor(t_fctr, levels = c(0:5,10,15,20))) %>% 
      group_by(pipe, biosample_id, feature_id) %>% 
      summarise(total_count = sum(count)) %>% 
      filter(total_count > 0) %>% 
      select(-total_count)

nested_count_df <- count_df %>% 
      mutate(t_fctr = factor(t_fctr, levels = c(0:5,10,15,20))) %>% 
      group_by(pipe, biosample_id, feature_id) %>% 
      right_join(nonzero_count_features) %>% 
      nest()
```

```{r}
test_df <- nested_count_df$data[[1]]

fit_logoff <- glm.nb(count ~ t_fctr + offset(log(total_abu)), data = test_df, link = "log")
```


```{r}
fit_logoff
```
```{r}
fit_coefs <- fit_logoff %>% coefficients()
fit_coefs
```

t_fctr0 (unmixed Post-exposure) is the intercept. 
Adding the intercept term to other titration estimates and converting to proportions from log space. 
```{r}
c(fit_coefs[1], fit_coefs[1] + fit_coefs[2:9]) %>% exp() %>% plot()
```


## Fitting negative binomial model to full dataset
```{r}
get_coefs <- function(fit){
      coefs <- coefficients(fit) 
      coefs_df <- c(coefs[1], 
                    coefs[1] + coefs[2:9]) %>% 
            exp() %>% 
            as.data.frame() %>% 
            rownames_to_column(var = "t_fctr")
      
      colnames(coefs_df) <- c("t_fctr", "nb_prop")
      
      coefs_df
}

safe_glm.nb <- safely(glm.nb)

nb_fit <- nested_count_df %>%
      mutate(fit = map(data, 
                       ~safe_glm.nb(count ~ t_fctr + offset(log(total_abu)), 
                               data = ., 
                               link = "log"))) 

saveRDS(nb_fit, "~/Desktop/nist1_nb_fit.RDS")

fit_results <- nb_fit$fit %>% 
      transpose() %>% 
      as_data_frame()


nb_fit_results <- nb_fit %>% 
      select(-fit) %>% 
      bind_cols(fit_results) %>% 
      mutate(ok_model = map_lgl(error, is_null)) 

nb_counts <- nb_fit_results %>% 
      filter(ok_model == T) %>% 
      mutate(coefs = map(result, get_coefs)) %>% 
      select(-data, -result, -error, -ok_model) %>%
      unnest()

nb_counts <- nb_counts %>%
      mutate(t_fctr = if_else(t_fctr == "(Intercept)",
                              "0",
                              str_replace(t_fctr, "t_fctr","")),
             t_fctr = factor(t_fctr, levels = c(0:5,10,15,20)))

saveRDS(nb_counts, "~/Desktop/nist1_nb_counts.RDS")
```

Failed models have high proportion of zeros
```{r}
bad_models <- nb_fit_results %>% 
      filter(ok_model == F) 

bad_models
```

```{r}
bad_models$data
```



