---
title: "Pre Post Titration Mystery"
author: "Nate Olson"
date: '`r Sys.Date()`'
output:
  pdf_document: default
  html_document: default
---

```{r setup, message = FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd)
```
## Objective
Determine which wells/ samples in the PCR plate correspond to the unmixed pre and post samples. 
Based on the experimental design, the samples with dilution factors of 0 are unmixed post-treatment and -1 are unmixed pre-treatment. 
However, preliminary analysis indicates that the samples were potentially switched. 


## ERCC Spike-in results
Based on qPCR results from `titration_validation.pdf` the ERCC spiked into the unmixed post-treatment samples had the highest concentration (lowest Ct values) for titration factor 0. 
Therefore, titration factor 0 is the unmixed post-treatment sample. 
Note that the sample used in the qPCR assays and 16S PCR are in individual tubes and not in 96 well plates and different sample ids were used for qPCR. 
It is not unreasonable for the samples to be correctly labeled for the qPCR assays but switched for the 16S PCRs. 


## Expected Count Values 

Based on qPCR and ERCC spike-ins the expectd count values should be calculated as follows, with $p = 2^-t$ and $t$ is the titration factor. 

$$
C_{exp} = [C_{post}\times p] + [C_{pre} \times (1-p)]
$$
```{r}
count_df <- readRDS("../data/normalize_count_df.rds")
```


```{r}
pre_count <- count_df %>% filter(dilution == -1) %>% 
      dplyr::rename(pre = count) %>% select(-dilution, -samID)
post_count <- count_df %>% filter(dilution == 0) %>% 
      dplyr::rename(post = count) %>% select(-dilution, -samID)
pre_post_count <- left_join(pre_count, post_count)
rm(pre_count, post_count)

count_exp_obs <- count_df %>% 
      filter(!(dilution %in% c(0,-1))) %>%
      left_join(pre_post_count) %>% 
      mutate(p = 2^(-dilution),
             exp_count_pre_p = post * (1-p) + pre * p,
             exp_count_post_p = pre * (1-p) + post * p)
```


## Metrics for evaluating count values 

The overall pipeline and normalization method performance was evaluated using root mean squared error (RMSE) and the normalized RMSE (NRMSE) or coefficient of variation of the RMSE. 

```{r} 
count_rmse <- count_exp_obs %>% filter(sampleID != "NTC") %>% 
  filter(pre != 0, post != 0, count != 0) %>% 
  mutate(residual_pre_p = (exp_count_pre_p - count)^2,
         residual_post_p = (exp_count_post_p - count)^2) %>% 
      group_by(pipe, norm_method) %>% 
      summarise(mse_pre_p = mean(residual_pre_p), 
                rmse_pre_p = sqrt(mse_pre_p), 
                nrmse_pre_p = rmse_pre_p/mean(exp_count_pre_p),
                mse_post_p = mean(residual_post_p), 
                rmse_post_p = sqrt(mse_post_p), 
                nrmse_post_p = rmse_post_p/mean(exp_count_post_p))
```

Lower normalized RMSE for all normalization methods when $2^-t$ is the proportion of pre-treatment sample in the mix. 

```{r}
count_rmse %>% filter(norm_method != "dsq", norm_method != "raw") %>% 
  ggplot() + 
    geom_point(aes(x = norm_method, y = nrmse_post_p, shape = pipe), color = "darkorange") + 
    geom_point(aes(x = norm_method, y = nrmse_pre_p, shape = pipe), color = "darkblue") + 
  theme_bw()
```


## Looking at pre and post specific features

```{r}
feature_specificity_df <- readRDS("../data/feature_specificity_df.rds")
count_df <- readRDS("../data/expected_counts_df.rds")
```

```{r}
feature_anno <- feature_specificity_df %>% 
  select(pipe, featureIndices, sampleID, specific_anno) %>% 
  mutate(otuID = featureIndices) %>% unique()
```

```{r}
feature_explore <- count_df %>% filter(sampleID != "NTC") %>% 
  mutate(featureIndices = as.numeric(featureIndices)) %>% 
  left_join(feature_anno)
```

```{r}
top_pre <- feature_explore %>% filter(specific_anno == "pre_full", norm_method == "rawlog") %>% 
  group_by(pipe, norm_method, specific_anno, sampleID, otuID) %>% 
  summarise(total_count = sum(count)) %>% 
  arrange(desc(total_count)) %>% group_by(pipe, specific_anno, norm_method) %>% top_n(5, total_count)
```

```{r}
top_pre_features <- top_pre %>% left_join(feature_explore)
```

Pre-treatment specific features should increase in abundance with titration factor not decrease!
```{r}
top_pre_features %>% 
  ggplot(aes(x = dilution, y = count, color = pcr_rep)) + geom_point() + 
  facet_wrap(~otuID*sampleID, nrow = 2) + theme_bw()
```


```{r}
top_post <- feature_explore %>% filter(specific_anno == "post_full", norm_method == "rawlog") %>% 
  group_by(pipe, norm_method, specific_anno, sampleID, otuID) %>% 
  summarise(total_count = sum(count)) %>% 
  arrange(desc(total_count)) %>% group_by(pipe, specific_anno, norm_method) %>% top_n(5, total_count)
```

```{r}
top_post_features <- top_post %>% left_join(feature_explore)
```

Post-treatment specific features tend to increase in abundance or remain high, excluding feature 595 for E01JH0004 and 3119 E01JH0016. 
```{r}
top_post_features %>% 
  ggplot(aes(x = dilution, y = count, color = pcr_rep)) + geom_point() + 
  facet_wrap(~otuID*sampleID, nrow = 2) + theme_bw()
```

## Conclusion
The results indicate that the pre and post treatment unmixed samples were most likely switched in the PCR plate. 
The ERCC spike-in results indicate that this error did not impact how the titrations were made, therefore expectations regarding the count values holds. 