---
title: "Dataset with Well Behaved Features"
author: "Nate Olson"
date: '`r Sys.Date()`'
always_allow_html: yes
output:
  pdf_document: default
  html_document: default
---

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd)
```

```{r echo = FALSE}
pipeline_dir <- "../../mgtst_pipelines"
mrexp <- get_mrexp(pipeline_dir)
```


## Objective
Generate a dataset with phenotype data and count data for only informative features, thoes assigned to full, pre, and post categories. 

Extracting count data from MRexperiments
```{r}
## Extracting a tidy dataframe with count values from MRexpiment objects
get_count_df <- function(mrobj, agg_genus = FALSE){
      if(agg_genus){
            mrobj <- aggregateByTaxonomy(mrobj, lvl = "Rank6", 
                                         norm = FALSE, log = FALSE, sl = 1)
      }
      
      mrobj <- cumNorm(mrobj, p = 0.75)
      mrobj %>%
            # not sure whether or not to normalize counts prior to analysis
            MRcounts(norm = FALSE, log = FALSE, sl = 1) %>%  
            as.data.frame() %>% 
            rownames_to_column(var = "feature_id") %>% 
            gather("id","count", -feature_id)
} 

count_df <- mrexp %>% map_df(get_count_df, .id = "pipe") %>% 
      left_join(pData(mrexp$dada2)) %>% 
      filter(biosample_id != "NTC") 
```


Feature categories from `2017-03-29-Feature-Categorization-Take2.Rmd`. 
```{r}
feature_cat <- readRDS("../data/feature_categories_df.rds") 
```


```{r}
annotated_counts <- left_join(count_df, feature_cat) %>% 
      filter(cat %in% c("cat_full","cat_post","cat_pre")) %>% 
      select(-titration) %>% 
      dplyr::rename(pcr_id = id)
annotated_counts %>% saveRDS("../data/raw_counts_good_feature_categories.rds")
```

## Column description 

pipe - bioinformatic pipeline used   
feature_id - feature id assigned by the bioinformatic pipeline   
pcr_id - unique id for a sample PCR replicate, used as column names in `annotated_count_matrix`  
count - raw counts   
biosample_id - unique id for individual biological replicates   
t_fctr - titration factor   
pcr_16S_plate - id for the replicate PCR plate   
pos - well position in the PCR plate   
pcr_half - PCR plate blocking indicator, first or second half of the PCR plate  
pcr_rep - blocking indicator for the four sets of PCR replicates   
cat - feature category assignment  

```{r}
glimpse(annotated_counts)
```

## Feature Category assignment definitions  

* Full - features present in at least one PCR replicates for all samples and absent in less than 12 of the 36 total PCR replicates of a biological replicate, and pipeline.  
* Pre - present in three or more PCR replicates for unmixed pre-treatment samples, not observed in any PCR replicates of the unmixed post treatment samples, and present in at least 24 total PCR replicates.  
* Post - present in three or more PCR replicates for the unmixed post-treatment samples, not observed in any PCR replicates of the unmixed pre-treatment samples, and present in at least 12 total PCR replicates.  


Generating count matrix
```{r}
annotated_count_matrix <- annotated_counts %>% select(feature_id, pcr_id, count) %>% 
      spread(pcr_id,count,fill = 0) 
rownames(annotated_count_matrix) <- annotated_count_matrix$feature_id
annotated_count_matrix <- annotated_count_matrix %>% 
      select(-feature_id) %>% as.matrix() 

annotated_count_matrix %>% saveRDS("../data/raw_counts_matrix_good_feature_categories.rds")
``` 

pData for `annotated_count_matrix`
```{r}
anno_counts_pdata <- annotated_counts %>% select(-count, -feature_id, -cat, -pipe) %>% unique() 
rownames(anno_counts_pdata) <- anno_counts_pdata$pcr_id
```

```{r}
anno_counts_pdata %>% saveRDS("../raw_counts_pdata_good_feature_categories.rds")
```

```{r} 
annotated_counts %>% select(feature_id, pipe) %>% unique() %>% 
       saveRDS("../feature_id_pipeline_good_feature_categories.rds")
```

