---
title: "Make Feature Classification DF"
author: "Nate Olson"
date: '`r Sys.Date()`'
always_allow_html: yes
output:
  pdf_document: default
  html_document: default
---

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r echo = FALSE}
pipeline_dir <- "../../mgtst_pipelines"
mrexp_files <- list(
      dada2 = file.path(pipeline_dir, "dada2/dada_mrexp.rds"),
      mothur =  file.path(pipeline_dir, "mothur/mothur_mrexp.rds"),
      qiime =  file.path(pipeline_dir, "qiime/qiime_mrexp.rds")
)
mrexp <- mrexp_files %>% map(readRDS)
```

Code for generating a data frame with the feature classification information for all pipelines. 

## Treatment-Specific Code for all pipelines

```{r}
get_treatment_class <- function(mrexp_pData){
      ## a character vector classifying samples as pre, post, or mixed by
      ## biological replicate (biosample_id)
      
      mrexp_pData$treatment <- mrexp_pData$titration %>% 
            {case_when(
                  is.na(.)  ~ "NTC",
                  . == 0 ~ "post",
                  . == 20 ~ "pre",
                  . %in% c(1:5, 10, 15) ~ "mixed"
            )}

     mrexp_pData %>% 
           unite("bio_unmixed", biosample_id, treatment, sep = ":") %>% 
           .$bio_unmixed
}

get_unique_feature_matrix <- function(mrexp_obj){
      ## get matrix with number of samples and reads observed for each feature
      ## by biosample_id (biological replicate) and feature
      
      ## get treatment_class
      cl <- mrexp_obj %>% pData() %>% get_treatment_class()
      
      ## uniqueFeatures matrix
      uniqueFeatures(mrexp_obj, cl)
}

tidy_unique_feature_matrix <- function(unq_features){
      ## convert the uniqueFeatures matrix into a tidy data frame for downstream
      ## analysis
      unq_features %>% as.data.frame() %>% 
            rownames_to_column("otuID") %>% 
            gather("key","value",-otuID, -featureIndices) %>% 
            mutate(key = gsub(" in","", key)) %>% 
            separate(key,c("key","bio_unmixed"),sep = " ") %>%
            separate(bio_unmixed, c("biosample_id","treatment"),sep = ":") %>% 
            spread(key,value) %>% 
            dplyr::rename(Samp =`Samp.`)
}

annotate_unique_features <- function(unq_features_df){
      ## annotate features as pre, post, or unmixed specific for each biological
      ## replicate
      treatment_features <- unq_features_df %>% select(-Reads) %>% 
            spread(treatment, Samp)
      
      ## Specificity Annotations
      ## non-specific - present in both pre- and post-treatment replicates 
      ## non-specific_full - present in all 4 pre-treatment replicates and all 4 post-treatment replicates
      ## pre_partial - not present in any post and present in 1-3 of the pre-treatment replicates 
      ## pre_full - not present in any post and present in 4 of the pre-treatment replicate 
      ## post_partial - not present in any pre and present in 1-3 of the post-treatment replicates
      ## post_full - not present in any pre and present in 4 of the post-treatment replicates 
      ## mix_specific - not present in any pre or post-treatment replicates but present in titration replicates
      ## biorep_neg - not present in unmixed or mixed samples for the biological replicate 
      
      treat_anno <- treatment_features %>% ungroup() %>% 
            select(-NTC) %>% filter(biosample_id != "NTC") %>% 
            mutate(specific_anno = "non-specific",
                   specific_anno = if_else(post == 4 & pre == 4,
                                           "non-specific_full", specific_anno),
                   specific_anno = if_else(post == 0 & pre != 0, 
                                           "pre_partial", specific_anno),
                   specific_anno = if_else(post == 0 & pre == 4, 
                                           "pre_full", specific_anno),
                   specific_anno = if_else(post != 0 & pre == 0, 
                                           "post_partial", specific_anno),
                   specific_anno = if_else(post == 4 & pre == 0, 
                                           "post_full", specific_anno),
                   specific_anno = if_else(post == 0 & pre == 0 & mixed != 0, 
                                           "mix_specific", specific_anno),
                   specific_anno = if_else(post == 0 & pre == 0 & mixed == 0, 
                                           "biorep_neg", specific_anno))
      
      ## annotiting NTC features
      ntc_features <- treatment_features %>% 
            filter(biosample_id == "NTC") %>% select(otuID,NTC)
      
      unq_features_df <- unq_features_df %>% 
            filter(biosample_id != "NTC") %>% left_join(ntc_features)
      
      treat_anno %>% select(otuID, featureIndices, biosample_id, specific_anno) %>%
            right_join(unq_features_df)
}
```


### Feature Specificity
```{r}
feature_specificity_mat <- mrexp %>% map(get_unique_feature_matrix) 
feature_specificity_tidy <- feature_specificity_mat %>%  
      map(tidy_unique_feature_matrix) 
feature_specificity_df <- feature_specificity_tidy %>% 
      map_df(annotate_unique_features, .id = "pipe")
```

```{r}
glimpse(feature_specificity_df)
```


Read counts for specificity classifications 
```{r}
mixed_specific <- feature_specificity_df %>% 
      filter(specific_anno %in% c("non-specific","non-specific_full", "mixed")) %>% 
      group_by(pipe, otuID, featureIndices, biosample_id, specific_anno) %>% 
      summarise(Reads = sum(Reads), Samp = sum(Samp))

pre_specific <- feature_specificity_df %>% 
      filter(specific_anno %in% c("pre_full","pre_partial"), treatment == "pre") %>% 
      select(pipe, otuID, featureIndices,  biosample_id, specific_anno, Reads, Samp)

post_specific <- feature_specificity_df %>% 
      filter(specific_anno %in% c("post_full","post_partial"), treatment == "post") %>% 
      select(pipe, otuID, featureIndices,  biosample_id, specific_anno, Reads, Samp)

feature_specificity_counts_df <- bind_rows(mixed_specific, pre_specific, post_specific)
```

```{r}
glimpse(feature_specificity_counts_df )
```


### Saving data frames as RDS
```{r}
feature_specificity_df %>% saveRDS("../data/feature_specificity_df.rds") 
feature_specificity_counts_df %>% saveRDS("../data/feature_specificity_counts_df.rds")
```


## Feature Specificity Summary
```{r}
feature_specificity_df %>% filter(specific_anno != "biorep_neg") %>% 
      ggplot() + geom_bar(aes(x = specific_anno)) + 
      facet_wrap(~pipe, ncol = 1, scale = "free_y") +
      theme_bw() + labs(x = "Feature Classification", y = "# of Features")
# ggsave("~/Desktop/feature_specificity.png", width = 5.48, height = 6, dpi = 450)
```

Expected high accuracy features based on titration expectations are non-specific, post_full, and pre_full features. 


```{r}
feature_specificity_counts_df %>%
      ggplot() + geom_jitter(aes(x = specific_anno, y = Reads), alpha = 0.25) + 
      facet_wrap(~pipe, ncol = 1, scale = "free_y") + theme_bw() + scale_y_log10() 
```


# Session information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
      knitr::kable()
```