---
title: "OE Metric"
author: "Nate Olson"
date: '`r Sys.Date()`'
always_allow_html: yes
output:
  pdf_document: default
  html_document: default
---

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(git2r)
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd)
```

## Objective
Develop a metric for characterizing how well a observed features agree with expectation.


__OE Metric__  
 
 $$(observed - expected)/(observed + expected)$$

Values range from -1 to 1, with values of -1 for observed counts of 0 and non-zero expected counts, 1 when expected count is 0 and observed counts are non-zero. 
0 represent agreement between observed and expected counts. 

## Calculated Expected Count Values 

## Negative Binomal for Weighted Count Estimates
Calculating proportion of pre and post counts using negative binomial. 

- $q_{i,j,k}$ is the proportion of feature $i$ in PCR $k$ of sample $j$ where a sample is defined as an individual unmixed or mixed samples for a biological replicate.  
- $p_{j,k}$ is the total feature abundance for sample $j$, sum of all feature counts not the number of sequences generated for the sample.  
- $v_{i,j,k}$ is the variance of feature $i$ in PCR replicate $j$ of sample $k$.  

$$v_{i,j,k} = \frac{q_{i,j,k}(1-q_{i,j,k})}{p_j,k}$$

- $w_{i,j,k}$ is the weight function

$$w_{i,j,k} = \frac{v_{i,j,k}^{ -1}}{\sum_{k \in j}v_{i,j,k}^{ -1}}$$


- $q_{i,j}$ - the weighted count estimate for feature $i,k$

$$q_{i,j} = \sum_{k\in j} w_{i,j,k}q_{i,j,k}$$ 

## Loading Data and Calculating Expected Values

```{r echo = FALSE}
pipeline_dir <- "../../mgtst_pipelines"
mrexp <- get_mrexp(pipeline_dir)
``` 

```{r}
## Extracting a tidy dataframe with count values from MRexpiment objects
get_count_df <- function(mrobj, agg_genus = FALSE, css = TRUE){
      if(agg_genus){
            mrobj <- aggregateByTaxonomy(mrobj, lvl = "Rank6", 
                                         norm = FALSE, log = FALSE, sl = 1)
      }
      
      if(css == TRUE){
            mrobj <- cumNorm(mrobj, p = 0.75)
            count_mat <- MRcounts(mrobj, norm = TRUE, log = FALSE, sl = 1000)
      }else{
            count_mat <- MRcounts(mrobj, norm = FALSE, log = FALSE, sl = 1)
      }
       count_mat %>%  
            as.data.frame() %>% 
            rownames_to_column(var = "feature_id") %>% 
            gather("id","count", -feature_id)
}

count_df <- mrexp %>% map_df(get_count_df, css = FALSE, .id = "pipe") %>% 
      left_join(pData(mrexp$dada2)) %>% 
      filter(biosample_id != "NTC", id != "1-F9") %>% 
      select(pipe, biosample_id, id, pcr_rep, feature_id, t_fctr, count) 

count_df <- count_df %>% group_by(id) %>% mutate(total_abu = sum(count)) 
```

Subsetting `count_df`
```{r}
count_df <- count_df %>% filter(feature_id %in% c(paste0("SV", 1:3), paste0("Otu0000",1:3)))
```

Estimating $q_{i}$ for pre and post 

```{r}
nb_est <- count_df %>% filter(t_fctr %in% c(0, 20)) %>% 
      mutate(prop = count/total_abu, 
             prop_var = (prop * (1 - prop))/total_abu,
             inv_var = 1/prop_var) %>% 
      group_by(pipe, biosample_id, t_fctr, feature_id) %>% 
      mutate(weight = inv_var / sum(inv_var)) %>% 
      summarise(prop_est = sum(weight*prop)) 

# Reformatting data
pre_post_prop <- nb_est %>% ungroup() %>% 
      mutate(treat = if_else(t_fctr == "20", "pre","post")) %>% 
      select(-t_fctr) %>% 
      mutate(prop_est = if_else(is.nan(prop_est), 0, prop_est)) %>% 
      spread(treat, prop_est)
``` 


Calculating expected counts using proportion estimates  
```{r}
calc_expected_prop <- function(pre_post_prop){
      titration_list <- data_frame(titration = c(1:5,10,15)) %>% 
            mutate(post_prop = 2^-titration) %>% 
            list() %>% rep(nrow(pre_post_prop))
      
      pre_post_prop %>% ungroup() %>% 
            add_column(titration = titration_list) %>% unnest() %>% 
            mutate(exp_prop = post * post_prop + pre * (1-post_prop)) %>% 
            mutate(t_fctr = factor(titration)) %>% 
            select(-post_prop)
}

exp_prop_df <- calc_expected_prop(pre_post_prop)
```

```{r}
exp_count_df <- count_df %>% 
      filter(t_fctr %in% c(1:5, 10, 15)) %>% 
      left_join(exp_prop_df) %>% 
      mutate(exp_count = total_abu * exp_prop)
```

## Alternative Metric

```{r}
oe_metric_df <- exp_count_df %>% 
      mutate(oe_metric = (exp_count - count)/(exp_count + count))
```

```{r}
oe_metric_df %>% ggplot() + 
      geom_histogram(aes(x = oe_metric)) + theme_bw()
```

```{r}
oe_metric_df %>% ggplot() +
      geom_hex(aes(x = pre, y = oe_metric))
```

```{r}
oe_metric_df %>% ggplot() +
      geom_hex(aes(x = post, y = oe_metric))
```

```{r}
oe_metric_df %>% ggplot() + 
      geom_histogram(aes(x = oe_metric)) + theme_bw() + 
      facet_wrap(~biosample_id)
```

```{r}
oe_metric_df %>% 
      mutate(t_fctr = factor(t_fctr, levels = c(0:5, 10, 15, 20))) %>% 
      ggplot() + 
      geom_histogram(aes(x = oe_metric)) + theme_bw() + 
      facet_wrap(~t_fctr)
```

```{r}
oe_metric_df %>% ggplot() + 
      geom_point(aes(x = count + 1, y = oe_metric, color = biosample_id)) +
      theme_bw() + facet_wrap(~feature_id)
```


## Summarizing by Titration
```{r}
oe_summary <- oe_metric_df %>% 
      group_by(pipe, biosample_id, t_fctr, feature_id) %>% 
      mutate(oet_mean = mean(oe_metric),
             oet_median = median(oe_metric)) %>% 
      group_by(pipe, biosample_id, feature_id) %>% 
      mutate(oef_mean = mean(oe_metric),
             oef_median = median(oe_metric))
```

```{r}
oe_summary %>% 
      ggplot() + geom_histogram(aes(x = oet_mean)) + 
      facet_wrap(~t_fctr) + theme_bw()
```

```{r}
oe_summary %>% 
      ggplot() + geom_histogram(aes(x = oef_median)) + 
      facet_wrap(~t_fctr) + theme_bw()
```

## Metric and Scatter Plots
```{r}
oe_summary %>% ungroup() %>% 
      select(biosample_id, feature_id, oef_median) %>% unique() %>% 
      spread(biosample_id, oef_median)
```

```{r}
count_df %>% 
      mutate(t_fctr = factor(t_fctr, levels = c(0:5, 10, 15, 20))) %>% 
      ggplot() + 
      geom_point(aes(x = t_fctr, y = count)) + 
      facet_grid(feature_id~biosample_id, scales = "free")
```

# Session information 

## Git repo commit information
```{r}
repo <- repository(path = ".")
last_commit <- commits(repo)[[1]]
```

The current git commit of this file is `r last_commit@sha`, which is on the `r branches(repo)[[1]]@name` branch and was made by `r last_commit@committer@name` on `r when(last_commit)`. The current commit message is `r last_commit@summary`. The repository is online at https://github.com/nate-d-olson/mgtst-pub  


## Platform Information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
```


## Package Versions
```{r}
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
      knitr::kable()
```