---
title: "EO Metric"
author: "Nate Olson"
date: '`r Sys.Date()`'
always_allow_html: yes
output:
  pdf_document: default
  html_document: default
---

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd)
```

## Objective
Develop a metric for characterizing how well a observed features agree with expectation.


__EO Metric__  
 
 $$\frac{expected - observed}{expected + observed}$$

Values range from -1 to 1 with values of;  

- 1 for observed counts of 0 and non-zero expected counts, 
- -1 when expected count is 0 and observed counts are non-zero, 
- 0 represents agreement between observed and expected counts. 

## Expected Count Values 

## Negative Binomal for Weighted Count Estimates
Calculating proportion of pre and post counts using negative binomial. 

- $q_{i,j,k}$ is the proportion of feature $i$ in PCR $k$ of sample $j$ where a sample is defined as an individual unmixed or mixed samples for a biological replicate.  
- $p_{j,k}$ is the total feature abundance for sample $j$, sum of all feature counts not the number of sequences generated for the sample.  
- $v_{i,j,k}$ is the variance of feature $i$ in PCR replicate $j$ of sample $k$.  

$$v_{i,j,k} = \frac{q_{i,j,k}(1-q_{i,j,k})}{p_j,k}$$

- $w_{i,j,k}$ is the weight function

$$w_{i,j,k} = \frac{v_{i,j,k}^{ -1}}{\sum_{k \in j}v_{i,j,k}^{ -1}}$$


- $q_{i,j}$ - the weighted count estimate for feature $i,k$

$$q_{i,j} = \sum_{k\in j} w_{i,j,k}q_{i,j,k}$$ 

## Loading Data and Calculating Expected Values

```{r echo = FALSE}
pipeline_dir <- "../../mgtst_pipelines"
mrexp <- get_mrexp(pipeline_dir)
``` 

```{r}
## Extracting a tidy dataframe with count values from MRexpiment objects
get_count_df <- function(mrobj, agg_genus = FALSE, css = TRUE){
      if(agg_genus){
            mrobj <- aggregateByTaxonomy(mrobj, lvl = "Rank6", 
                                         norm = FALSE, log = FALSE, sl = 1)
      }
      
      if(css == TRUE){
            mrobj <- cumNorm(mrobj, p = 0.75)
            count_mat <- MRcounts(mrobj, norm = TRUE, log = FALSE, sl = 1000)
      }else{
            count_mat <- MRcounts(mrobj, norm = FALSE, log = FALSE, sl = 1)
      }
       count_mat %>%  
            as.data.frame() %>% 
            rownames_to_column(var = "feature_id") %>% 
            gather("id","count", -feature_id)
}

count_df <- mrexp %>% map_df(get_count_df, css = FALSE, .id = "pipe") %>% 
      left_join(pData(mrexp$dada2)) %>% 
      filter(biosample_id != "NTC", id != "1-F9") %>% 
      select(pipe, biosample_id, id, pcr_rep, feature_id, t_fctr, count) 

count_df <- count_df %>% group_by(id) %>% mutate(total_abu = sum(count)) 
```

Subsetting `count_df`
```{r}
count_df <- count_df %>% filter(feature_id %in% c(paste0("SV", 1:3), paste0("Otu0000",1:3)))
```

Estimating $q_{i}$ for pre and post 

```{r}
nb_est <- count_df %>% filter(t_fctr %in% c(0, 20)) %>% 
      mutate(prop = count/total_abu, 
             prop_var = (prop * (1 - prop))/total_abu,
             inv_var = 1/prop_var) %>% 
      group_by(pipe, biosample_id, t_fctr, feature_id) %>% 
      mutate(weight = inv_var / sum(inv_var)) %>% 
      summarise(prop_est = sum(weight*prop)) 

# Reformatting data
pre_post_prop <- nb_est %>% ungroup() %>% 
      mutate(treat = if_else(t_fctr == "20", "pre","post")) %>% 
      select(-t_fctr) %>% 
      mutate(prop_est = if_else(is.nan(prop_est), 0, prop_est)) %>% 
      spread(treat, prop_est)
``` 


Calculating expected counts using proportion estimates  
```{r}
calc_expected_prop <- function(pre_post_prop){
      titration_list <- data_frame(titration = c(1:5,10,15)) %>% 
            mutate(post_prop = 2^-titration) %>% 
            list() %>% rep(nrow(pre_post_prop))
      
      pre_post_prop %>% ungroup() %>% 
            add_column(titration = titration_list) %>% unnest() %>% 
            mutate(exp_prop = post * post_prop + pre * (1-post_prop)) %>% 
            mutate(t_fctr = factor(titration)) %>% 
            select(-post_prop)
}

exp_prop_df <- calc_expected_prop(pre_post_prop)
```

```{r}
exp_count_df <- count_df %>% 
      filter(t_fctr %in% c(1:5, 10, 15)) %>% 
      left_join(exp_prop_df) %>% 
      mutate(exp_count = total_abu * exp_prop) %>% 
      filter(!(pre == 0 & post == 0 & count == 0))
```

## EO Metric

```{r}
eo_metric_df <- exp_count_df %>% 
      mutate(eo_metric = (count - exp_count)/(count + exp_count))
```

Overall the metric distribution skews to the right indicating that the expected values tend to be greater than the observed values with a medain around 0. 
There are more PCR replicates with observed counts of but 0 expected counts than expected counts with 0 observed counts. 

```{r}
eo_metric_df %>% ggplot() + 
      geom_histogram(aes(x = eo_metric)) + theme_bw() +
      labs(x = "Exp-Obs Metric")
```

The EO metric tends to center around 0 with increasing Pre-Treament feature proportion but increases with Post-treatment feature proportion. 
Low pre and post treatment feature proportion have higher EO metrics indicating expected counts are overestimates for lower proportions. 
This is potentially due to a limit of detection, can look at the EO metric for feature-pcr reps with expected count values < 1 and != 0 (see table following plot). 

```{r}
eo_metric_df %>% 
      select(pipe, biosample_id, id, feature_id, t_fctr, pre, post, eo_metric) %>% 
      gather("unmix","q",-eo_metric, -pipe, -biosample_id, -id, -feature_id,-t_fctr) %>% 
      ggplot() +
      geom_hex(aes(x = q, y = eo_metric)) + 
      geom_smooth(aes(x = q, y = eo_metric), color = "darkorange") +
      facet_wrap(~unmix, scale = "free_x") +  
      theme_bw() +
      labs(x = "Treatment Feature Proportion", y = "Exp-Obs Metric")
```

```{r}
eo_metric_df %>% 
      filter(exp_count < 1, exp_count != 0) %>% 
      select(total_abu, count, exp_count, eo_metric) %>% arrange(eo_metric) %>% 
      knitr::kable()
```


For the example subset of the metric skew and -1 and 1 peaks differ by pipeline, biosample, and titration. 
Will want to look at larger set of features before drawing any conclusions. 

```{r}
eo_metric_df %>% 
      mutate(t_fctr = factor(t_fctr, levels = c(0:5, 10, 15, 20))) %>% 
      ggplot() + 
      geom_histogram(aes(x = eo_metric)) +
      facet_grid(biosample_id~t_fctr) + theme_bw() +
      labs(x = "Exp-Obs Metric")
```

The EO-metric distribution also tend to vary by feature, again a larger set of features is needed before drawing any conclusions.  

```{r}
eo_metric_df %>% ggplot() + 
      geom_point(aes(x = count, y = eo_metric, color = biosample_id)) +
      facet_wrap(~feature_id) + 
      theme_bw() +
      labs(x = "Observed Count", y = "Exp-Obs Metric")
```


## Summarizing by Titration and Feature 
To evaluate performance for a set of PCR replicates and across a feature can use the mean or median as summary metrics.  

```{r}
eo_summary <- eo_metric_df %>% 
      group_by(pipe, biosample_id, t_fctr, feature_id) %>% 
      mutate(eot_mean = mean(eo_metric),
             eot_median = median(eo_metric)) %>% 
      group_by(pipe, biosample_id, feature_id) %>% 
      mutate(eof_mean = mean(eo_metric),
             eof_median = median(eo_metric))
```

Mean EO metric for a feature titration PCR replicates.
```{r}
eo_summary %>% 
      ggplot() + geom_histogram(aes(x = eot_mean)) + 
      facet_wrap(~t_fctr) + theme_bw() +
      labs(x = "Exp-Obs Metric Feature-Titration")
```

Mean EO metric across all PCR replicates and titrations for a feature. 
```{r}
eo_summary %>% 
      ggplot() + geom_histogram(aes(x = eof_median)) + 
      facet_wrap(~t_fctr) + theme_bw() +
      labs(x = "Exp-Obs Metric Feature")
```

## Metric and Scatter Plots 
Table of feature level OE metrics and corresponding scatter plots of observed counts by titration. 
OE-metric is NA when all counts for all samples and replicates is 0.  

```{r}
eo_summary %>% ungroup() %>% 
      select(biosample_id, feature_id, eof_median) %>% unique() %>% 
      spread(biosample_id, eof_median) %>% 
      knitr::kable()
```

```{r}
count_df %>% 
      mutate(t_fctr = factor(t_fctr, levels = c(0:5, 10, 15, 20))) %>% 
      ggplot() + 
      geom_point(aes(x = t_fctr, y = count)) + 
      facet_grid(feature_id~biosample_id, scales = "free") +
      theme_bw() +
      labs(x = "Titration", y = "Observed Raw Counts")
      
```

# Session information 

## Git repo commit information
```{r}
library(git2r)
repo <- repository(path = "../")
last_commit <- commits(repo)[[1]]
```

The current git commit of this file is `r last_commit@sha`, which is on the `r branches(repo)[[1]]@name` branch and was made by `r last_commit@committer@name` on `r when(last_commit)`. The current commit message is `r last_commit@summary`. The repository is online at https://github.com/nate-d-olson/mgtst-pub  


## Platform Information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
```


## Package Versions
```{r}
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
      knitr::kable()
```