---
title: "Expected-Observed Count Model Fit"
author: "Nate Olson"
date: '`r Sys.Date()`'
always_allow_html: yes
output:
  pdf_document: default
  html_document: default
---

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd) 
```

## Objective
Fit a linear model of the residuals to the observed count values.  
Using the `expected_count_values_features_df.rds` and `expected_count_values_genus_df.rds` data frames generated in `2017-03-28-Development-Titration-Pred-Df.Rmd`. 

### Fit approach
Use `lm` with residuals (expected - observed) to the expected count values, looking for estimate deviance from slope of 0 and intercept 0
 
## Functions for Processing Data
```{r}
get_fit <- function(count_exp_df){
      count_exp_df %>% 
            group_by(pipe, biosample_id, feature_id) %>% 
            nest() %>% 
            mutate(fit = map(data,~lm(residual ~ exp_count, data = .)))
}

get_tidy_fit <- function(count_fit){
      count_fit %>% 
            mutate(fit_summary = map(fit, broom::tidy)) %>% 
            select(-data, -fit) %>% unnest() %>% 
            mutate(term = if_else(term == "(Intercept)", "Intercept","Slope"))
}

get_glance_fit <- function(count_fit){
      count_fit %>% 
            mutate(fit_summary = map(fit, broom::glance)) %>% 
            select(-data, -fit) %>% unnest() %>% 
            select(-statistic, -p.value) 
}
```


## Fitting model to individual Features
```{r}
count_exp_df <- readRDS("../data/expected_count_values_feature_df.rds")  

count_fit <- get_fit(count_exp_df)
count_fit %>% saveRDS("../data/expected_count_values_features_fit.rds") 

# Data frames with fit summary statistics and parameter estimates
get_tidy_fit(count_fit) %>% 
      saveRDS("../data/expected_count_values_fit_features_tidy.rds") 

get_glance_fit(count_fit) %>% 
      saveRDS("../data/expected_count_values_fit_features_glance.rds") 
``` 



## Fit model to Genus aggregated features 
```{r}
count_exp_df <- readRDS("../data/expected_count_values_genus_df.rds") 

count_fit <- get_fit(count_exp_df)
count_fit %>% saveRDS("../data/expected_count_values_genus_fit.rds") 

# Data frames with fit summary statistics and parameter estimates
get_tidy_fit(count_fit) %>% 
      saveRDS("../data/expected_count_values_fit_genus_tidy.rds") 

get_glance_fit(count_fit) %>% 
      saveRDS("../data/expected_count_values_fit_genus_glance.rds") 
```

## Session information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
      knitr::kable()
``` 