---
title: "Pre-Dominant and Pre-Specific logFC"
author: "Nate Olson"
date: '`r Sys.Date()`'
output:
  bookdown::pdf_document2: 
    toc: FALSE
---

```{r logFC_setup, include=FALSE}
library(tidyverse)
library(stringr)
library(broom)
```

## Objective 
Assess logFC for pre-dominant and pre-specific features.  
Only using E01JH0004, E01JH0011, and E01JH0016, as they had theta estimates that were most consistent with the expected values. 

__NOTE__ Be warry of E01JH0016 due to the odd theta and 1-theta estimates 


## Approach

* Identify pre-dominant and pre-specific features  
* calc expected logFC  
* compared observed to expected logFC  
      * Scatter plot  
      * Feature-level Bias - mixed effects model  
      * Feature-level Variance - mixed effects model


```{r processData, include = FALSE}
### Loading Data ---------------------------------------------------------------
pa_summary_df <- readRDS("~/Desktop/pa_summary_anno_df.RDS") %>% 
      select(pipe, biosample_id, feature_id, 
             T00, T20, pa_mixed) %>% 
      filter(biosample_id %in% paste0("E01JH00", c("04","11","16")))

logFC_df <- readRDS("~/Desktop/logFC_edgeR_df.rds") %>%
      filter(biosample_id %in% paste0("E01JH00", c("04","11","16")))

logFC_df <- logFC_df %>% 
      ## Fixing taxonomy for consistency
      mutate(Rank2 = str_replace(Rank2, "p__",""),
             Rank3 = str_replace(Rank3, "c__",""),
             Rank4 = str_replace(Rank4, "o__",""),
             Rank5 = str_replace(Rank5, "f__",""),
             Rank6 = str_replace(Rank6, "g__","")) %>% 
      dplyr::rename(feature_id = OTUname) %>%
      left_join(pa_summary_df)

### Extracting pre-specific and dominant features -----------------------------
logFC_prepost <- logFC_df %>% 
      filter(T1 == 0, T2 == 20)

logFC_pre <- logFC_prepost %>% 
      filter(logFC > 5, T20 == 4) %>% 
      group_by(biosample_id) %>% 
      dplyr::rename(prepost_logFC = logFC, prepost_logCPM = logCPM) %>% 
      ungroup() %>% 
      select(-T1, -T2, -PValue, -FDR) %>% 
      left_join(logFC_df)

### Calculate Expected logFC ------------------------------------------------- 
#Function for coverting numeric factors to their numeric values
as_numeric_fctr <- function(x) as.numeric(as.character(x))

logFC_pre <- logFC_pre %>% 
      mutate(T1_preprop = 1 - 2^-as_numeric_fctr(T1), 
             T2_preprop = 1 - 2^-as_numeric_fctr(T2),
            exp_logFC = log2(T2_preprop/T1_preprop)) 

### Calculating logFC Error Rate ----------------------------------------------
logFC_pre_error_rate <- logFC_pre %>% 
      filter(T1 %in% 1:5, T2 %in% 1:5) %>% 
      mutate(error_rate = abs(exp_logFC - logFC)/exp_logFC)

logFC_pre_error <- logFC_pre_error_rate %>% 
      group_by(pipe, biosample_id, feature_id, 
               prepost_logFC, prepost_logCPM,
               Rank2, Rank3, Rank4, Rank5, Rank6) %>% 
      summarise(med_logCPM = median(logCPM),
                median_error = median(error_rate, na.rm = TRUE),
                iqr_error = IQR(error_rate, na.rm = TRUE),
                rcov_error = iqr_error/median_error)
```

## Analysis
Pre-dominant and pre-specific features were identified based on logFC between pre- and post-exposure samples and number of PCR replicates the feature was observed in for pre- and post-exposure PCR replicates (Fig. \@ref(fig:preMA), Table \@ref(tab:preCountTbl)). 
Pre-dominant and pre-specific features were defined as features observed in all four pre-exposure PCR replicates and a logFC between pre- and post-exposure samples greater than 5.   
Pre-specific features were not observed in any of the post-exposure PCR replicates and pre-dominant features were observed in one or more of the post-exposure PCR replicates. 
Most of the pre-dominant and pre-specific features are present in more than 25 of the 28 titration PCR replicates (Fig. \@ref(fig:paDist)).  

To assess differential abundance logFC estimates logFC between all titrations were compared to the expected logFC for the pre-specific and pre-dominant features. 
When assuming the feature is only present in pre-exposure samples the expected logFC is independent of the observed counts for the unmixed samples. 

Expected logFC between titrations $i$ and $j$. 
$$
logFC_{ij} = log_2\left(\frac{1-\theta_i}{1-\theta_j}\right)
$$

Estimates vs. expected logFC comparison - (Fig. \@ref(fig:estVexpLogFC))


Statistics
* Using a mixed-effects models to take into account differences in individual, there was no statistical difference in the logFC error feature-level bias (median error rate) or variance (error rate RCOV).  
* An additional mixed-effects model was used to determine feature characteristics that are correlated with logFC error rate. Increased estimated logFC and logCPM were significantly related to lower error rates. 

```{r preMA, fig.cap = "MA plot of features identified as pre-dominant and pre-specific. Color is used to indicate number of post-exposure PCR replicates the feature was observed in. X-axis faceted by individual (vaccine trial participant).", echo = FALSE}
logFC_pre %>% 
      filter(T1 == 0, T2 == 20) %>% 
      ggplot() + 
      geom_point(aes(x = prepost_logCPM, 
                     y = prepost_logFC, 
                     fill = as.factor(T00)),
                 shape = 21) + 
      facet_grid(biosample_id~pipe) +
      labs(fill = "Post-Exposure PCR Replicates", x = "logCPM", y = "logFC") + theme_bw() + 
      theme(legend.position = "bottom")
``` 



```{r paDist, fig.cap = "Overall distribution of titration PCR replicates for the pre-specific and pre-domiannt  features.", echo = FALSE}
logFC_pre %>% filter(T1 == 0, T2 == 20) %>% 
      ggplot(aes(x = pa_mixed)) + 
      geom_density() + theme_bw() + 
      labs(x = "Titration PCR Replicates")
```


```{r preCountTbl, echo = FALSE}
logFC_pre %>% filter(T1 == 0, T2 == 20) %>%
      mutate(Type = if_else(T00 == 0, "specific", "dominant")) %>% 
      group_by(pipe, biosample_id, Type) %>% 
      summarise(count = n()) %>% 
      spread(pipe, count,fill = 0) %>% 
      rename(Individual = biosample_id) %>% 
      knitr::kable(caption = "Number of pre-specific and pre-dominant features by individual and pipeline", booktabs = TRUE)
```

```{r estVexpLogFC, fig.cap = "Relationship between the observed and expected logFC for pre-specific and pre-domiant features by pipeline and individual for all titration pair comparisons. Orange line indicates expected 1-to-1 relationship between the estimated and expected logFC. Blue line is a linear model was fit to the data and grey area is the models uncertainty estimate.", echo = FALSE}
logFC_pre %>% 
      filter(T1 != 0, T2 != 20) %>% 
      ggplot() + 
      geom_point(aes(x = logFC, y = exp_logFC)) +
      geom_smooth(aes(x = logFC, y = exp_logFC), method = "lm") +
      geom_abline(aes(intercept = 0, slope = 1), color = "darkorange") + 
      facet_grid(biosample_id~pipe) + theme_bw()  +
      labs(x = "Estimated logFC", y = "Expected logFC")
```


```{r}
logFC_fit <- logFC_pre %>% filter(T1 != 0, T2 != 20) %>% 
      group_by(pipe, biosample_id) %>% 
      nest() %>% 
      mutate(fit = map(data, ~lm(exp_logFC~logFC, data = .)))

logFC_glance <- logFC_fit %>% 
      mutate(fit_glance = map(fit, glance)) %>% 
      select(-data, -fit) %>% 
      unnest()

logFC_tidy <- logFC_fit %>% 
      mutate(fit_glance = map(fit, tidy)) %>% 
      select(-data, -fit) %>% 
      unnest()

logFC_slope <- logFC_tidy %>% 
      filter(term == "logFC") %>% 
      select(-term) %>% 
      rename(slope = estimate)

logFC_summary <- logFC_slope %>% 
      select(-statistic, -p.value) %>% 
      left_join(logFC_glance)
```


```{r}
logFC_summary$slope %>% summary()
```

```{r}
logFC_summary$adj.r.squared %>% summary()
```


```{r}
logFC_pre %>% 
      filter(T1 != 0, T2 != 20) %>% 
      mutate(t_comp = if_else(T1 %in% 1:5 & T2 %in% 1:5, "low-T","high-T")) %>% 
      ggplot() + 
      geom_point(aes(x = logFC, y = exp_logFC)) +
      geom_smooth(aes(x = logFC, y = exp_logFC, color = t_comp), method = "lm") +
      geom_abline(aes(intercept = 0, slope = 1), color = "darkorange") + 
      facet_grid(biosample_id~pipe) + theme_bw()
```


```{r}
logFC_pre %>% 
      filter(T1 != 0, T2 != 20) %>% 
      ggplot() + 
      geom_point(aes(x = logFC, y = exp_logFC)) +
      geom_smooth(aes(x = logFC, y = exp_logFC, color = factor(T00)), method = "lm") +
      geom_abline(aes(intercept = 0, slope = 1), color = "darkorange") + 
      facet_grid(biosample_id~pipe) + theme_bw()
```

```{r}
logFC_pre %>% 
      filter(T1 != 0, T2 != 20) %>% 
      filter(T00 %in% 0) %>% 
      ggplot() + 
      geom_point(aes(x = logFC, y = exp_logFC)) +
      geom_smooth(aes(x = logFC, y = exp_logFC), method = "lm") +
      geom_abline(aes(intercept = 0, slope = 1), color = "darkorange") + 
      facet_grid(biosample_id~pipe) + theme_bw()
```

```{r}
logFC_pre %>% 
      filter(T1 != 0, T2 != 20) %>% 
      mutate(pa_bin = cut_interval(pa_mixed,n = 4)) %>% 
      ggplot() + 
      geom_point(aes(x = logFC, y = exp_logFC)) +
      geom_smooth(aes(x = logFC, y = exp_logFC, color = pa_bin), method = "lm") +
      geom_abline(aes(intercept = 0, slope = 1), color = "darkorange") + 
      facet_grid(biosample_id~pipe) + theme_bw()
```

```{r}
logFC_pre %>% 
      filter(T1 != 0, T2 != 20) %>% 
      mutate(logFC_bin = cut_interval(prepost_logFC,n = 4)) %>% 
      ggplot() + 
      geom_point(aes(x = logFC, y = exp_logFC)) +
      geom_smooth(aes(x = logFC, y = exp_logFC, color = logFC_bin), method = "lm") +
      geom_abline(aes(intercept = 0, slope = 1), color = "darkorange") + 
      facet_grid(biosample_id~pipe) + theme_bw()
```



On a very limited set of features 
```{r}

logFC_pre %>% filter(T1 == 0, T2 == 20, T00 == 0, pa_mixed > 24, prepost_logFC > 10) %>% 
      group_by(pipe, biosample_id) %>% 
      summarise(count = n()) %>% 
      spread(pipe, count,fill = 0) %>% 
      rename(Individual = biosample_id) %>% 
      knitr::kable(caption = "Number of pre-specific and pre-dominant features by individual and pipeline", booktabs = TRUE)
```

```{r}
logFC_pre %>% 
      filter(T1 != 0, T2 != 20, T00 == 0, prepost_logFC > 10) %>% 
      ggplot() + 
      geom_point(aes(x = logFC, y = exp_logFC)) +
      geom_smooth(aes(x = logFC, y = exp_logFC), method = "lm") +
      geom_abline(aes(intercept = 0, slope = 1), color = "darkorange") + 
      facet_grid(biosample_id~pipe) + theme_bw()
```


```{r logFCerror, fig.cap = "Distribution of feature-level median error rate by individual and pipeline.", echo = FALSE}
ggplot(logFC_pre_error) + 
      geom_boxplot(aes(x = pipe, y = median_error)) + 
      facet_wrap(~biosample_id) +
      theme_bw() +
      labs(x = "Pipeline", y = "Feature-Level Median Error Rate")
```

```{r logFCcov, fig.cap = "Distribution of feature-level error rate robust coefficient of variation by individual and pipeline.", echo = FALSE}
logFC_pre_error %>% ggplot() + 
      geom_boxplot(aes(x = pipe, y = rcov_error)) + 
      facet_wrap(~biosample_id) +
      theme_bw() +
      labs(x = "Pipeline", y = "Feature-Level Robust Error COV")
```


### logFC Error Statistics
#### Bias
```{r}
error_fit <- nlme::lme(median_error ~ pipe, random =  ~ 1 | biosample_id, 
                 data = logFC_pre_error)
summary(error_fit)
```

```{r}
qqnorm(error_fit$residuals)
```

#### RCOV Error Statistics

```{r}
rcov_fit <- nlme::lme(rcov_error ~ pipe, random =  ~ 1 | biosample_id, 
                 data = logFC_pre_error)
summary(rcov_fit)
```

```{r}
qqnorm(rcov_fit$residuals)
```

#### Factors affecting error rate

```{r}
corr_fit <- nlme::lme(error_rate ~ logFC + logCPM, random =  ~ 1 | biosample_id/pipe, 
                 data = logFC_pre_error_rate %>% filter(T00 == 0, logFC < 5))
summary(corr_fit)
```

```{r}
qqnorm(rcov_fit$residuals)
```


```{r}
corr_fit <- nlme::lme(error_rate ~ logFC + logCPM + Rank3, random =  ~ 1 | biosample_id, 
                 data = logFC_pre_error_rate %>% filter(T00 == 0, logFC < 5))
summary(corr_fit)
```

```{r}
qqnorm(rcov_fit$residuals)
```


```{r}
logFC_pre %>% 
      filter(T1 != 0, T2 != 20, Rank3 %in% c("Negativicutes", "Coriobacteriia")) %>% 
      ggplot() + 
      geom_point(aes(x = logFC, y = exp_logFC, color = Rank3)) +
      geom_smooth(aes(x = logFC, y = exp_logFC, color = Rank3), method = "lm") +
      geom_abline(aes(intercept = 0, slope = 1), color = "darkorange") + 
      theme_bw() + facet_wrap(~pipe)
``` 

```{r}
logFC_pre
```

```{r}
fit <- lm(logFC ~ exp_logFC/Rank5, 
          data = logFC_pre %>% filter(T1 != 0, T2 != 20))
summary(fit) 
```
```{r}
plot(fit)
```

```{r}
logFC_feature_fit <- logFC_pre %>% 
      filter(T1 != 0, T2 != 20) %>% 
      group_by(pipe, biosample_id, prepost_logFC, prepost_logCPM, 
               feature_id, Rank2, Rank3, Rank4, Rank5, Rank6) %>% 
      nest() %>% 
      mutate(fit = map(data, ~lm(logFC ~ exp_logFC, data = .)))
```

```{r}
logFC_feature_glance <- logFC_feature_fit %>% 
      mutate(fit_glance = map(fit, glance)) %>% 
      select(-data, -fit) %>% 
      unnest()
```

```{r}
logFC_feature_tidy <- logFC_feature_fit %>% 
      mutate(fit_glance = map(fit, tidy)) %>% 
      select(-data, -fit) %>% 
      unnest()
```

```{r}
logFC_feature_slope <- logFC_feature_tidy %>% 
      filter(term == "exp_logFC") %>% 
      select(-term) %>% 
      rename(slope = estimate)
``` 

```{r}
logFC_feature_summary <- logFC_feature_slope %>% 
      select(-statistic, -p.value) %>% 
      left_join(logFC_feature_glance)
```
Fitting linear models to individual features
```{r}
logFC_feature_summary %>% 
      ggplot() + geom_point(aes(x = adj.r.squared, 
                                y = slope, 
                                color = pipe)) +
      facet_wrap(~biosample_id)
```


```{r}
logFC_feature_summary %>% 
      filter(slope > 4)
```

```{r}
logFC_feature_summary %>% 
      filter(slope > 4) %>% 
      left_join(logFC_pre) %>% 
      filter(T1 != 0, T2 != 20) %>% 
      ggplot() + geom_point(aes(x = logFC, y = exp_logFC, color = T1)) + 
      facet_wrap(~feature_id)
```

```{r}
ggplot(logFC_feature_summary) + geom_density(aes(x = adj.r.squared)) 
```

```{r}
logFC_feature_summary %>% filter(adj.r.squared > 0.25) %>% 
ggplot() + 
      geom_density(aes(x = slope, color = Rank2)) + 
      geom_vline(aes(xintercept = 1)) + 
      facet_grid(pipe~biosample_id)
```

```{r}
logFC_feature_summary %>% filter(adj.r.squared > 0.25, 
                                 Rank2 %in% c("Actinobacteria", "Proteobacteria")) %>% 
ggplot() + 
      geom_density(aes(x = slope, color = Rank2)) + 
      geom_vline(aes(xintercept = 1)) + 
      facet_grid(pipe~biosample_id)
```

```{r}
logFC_feature_summary %>% filter(adj.r.squared > 0.25, 
                                 Rank2 %in% c("Firmicutes")) %>% 
ggplot() + 
      geom_density(aes(x = slope, color = Rank3)) + 
      geom_vline(aes(xintercept = 1)) + 
      facet_grid(pipe~biosample_id, scales = "free")
```

```{r}
logFC_feature_summary %>% filter(adj.r.squared > 0.25, 
                                 Rank3 %in% c("Negativicutes")) %>% 
ggplot() + 
      geom_histogram(aes(x = slope, fill = Rank6)) + 
      geom_vline(aes(xintercept = 1)) + 
      facet_grid(pipe~biosample_id, scales = "free")
```

