---
title: "Titration and Unmix Specific Features"
author: "Nate Olson"
date: '`r Sys.Date()`'
always_allow_html: yes
output:
  bookdown::pdf_document2: 
      toc: FALSE
  html_document: default
---

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(tidyverse)
# library(ProjectTemplate)
# cwd <- getwd()
# setwd("../")
# load.project()
# setwd(cwd) 
# pipeline_dir <- "../../mgtst_pipelines"
# mrexp <- get_mrexp(pipeline_dir)
```

```{r qualData, echo = FALSE, message = FALSE, warning = FALSE}
feature_exp_df <- readRDS("../data/nb_expected_eo_metric_feature_df.rds") 
```

For the qualitative component of our measurement assessment we evaluated features only observed in either the unmixed pre- and post-exposure samples, unmixed-specific, or the titrations, titration-specific, for each biological replicate. 
Features are unmixed- or titration-specific due to differences in sampling depth (number of sequences) between the unmixed samples and titrations or an artifact of the feature inference process. 
There are a number of unmixed- and titration-specific features with a range of count values, expected counts for unmixed specific features and observed count values for titration-specific features (Fig. \@ref(fig:mixUnmixSpecific)).
We tested if sampling alone could explain feature specificity. For unmixed-specific features we used a Binomial test and for titration-specific features we used Monte-Carlo simulation and a Bayesian hypothesis test. 
For both test p-values were adjusted for multiple comparisons using the Benjamini & Hochberg (1995) method. 

```{r mixUnmixSpecific, fig.cap = "Distribution of observed count values for titration-specific features and expected count values for unmixed-specific features. The horizontal dashed line indicates a count value of 1.", echo = FALSE, message = FALSE}
feature_exp_df %>% 
    filter(count == 0 | exp_count == 0) %>% 
    mutate(feat_spec = if_else(count == 0 & exp_count != 0, "Unmixed-Specific","Titration-Specific")) %>% 
    mutate(plot_count = if_else(count == 0 & exp_count != 0,exp_count,count)) %>% 
      ggplot() +
      geom_boxplot(aes(x = pipe, y = plot_count + 1, fill = biosample_id, color = biosample_id), alpha = 0.25) + 
      geom_hline(aes(yintercept = 1 + 1), linetype = 2) + 
      scale_y_log10() + theme_bw() + 
      labs(x = "Pipeline", y = "Count Values", 
           fill = "Biological Replicate", color = "Biological Replicate") +
      facet_wrap(~feat_spec) + 
      theme(legend.position = "bottom")
``` 

To determine if sampling alone can explain unmixed-specific features the binomial test was used to test the following hypothesis;   

$H_0$ - Given no observed counts and the total abundance for a titration the true proportion of a feature is equal to the expected proportion.   

$H_1$ - Given no observed counts and the total abundance for a titration the true proportion of a feature is less than the expected proportion.   

 


There were umixed-specific features with expected proportions that could not be explained by sampling alone for all biological replicates and bioinformatic pipelines (Fig. \@ref(fig:binomTestUnmix)). 
However, the proportion of unmixed-specific features that could not be explained by sampling alone varied by bioinformatic pipeline with over half of the DADA2 unmixed-specific features could not be explained by sampling alone whereas QIIME had the lowest rate of features with 0 observed counts that could not be explained by sampling alone. 
__TODO__ Relate the number of features, diversity of features with 0 observed counts. 

To test if titration-specific features could be explained by sampling alone we used Monte-Carlo simulation and a Bayesian hypothesis test. 
For the simulation we assumed a binomial distribution given the observed total abundance and a uniform distribution of proportions, 0 to the minimum expected proportion. 
The minimum expected proportion, $\pi_{min_{exp}}$, is calculated using the mixture equation ( __TODO__ Eq.1) and the minimum observed feature proportion for unmixed pre-exposure, $\pi_{min_{pre}}$, and post-exposure $\pi_{min_{post}}$ samples for each biological replicate and pipeline. 
For features not present in unmixed samples the assumption is that the feature proportion is less than $\pi_{{min}_{exp}}$. 

We formulated our null and alternative hypothesis for the Bayesian test as follows,  

$H_0$ - Given the total abundance for a sample and minimum expected proportion the true proportion of a feature is __less than__ the minimum expected observed proportion.   
$H_1$ - Given the total abundance for a sample and minimum expected proportion the true proportion of a feature is __greater than or equal to__ the minimum expected proportion.  

The following equations were used to calculate the p-value for the Bayesian hypothesis test assuming equal priors, i.e. $P(\pi < \pi_{min_{exp}})=P(\pi \geq \pi_{min_{exp}})$.  

$$p =P(\pi < \pi_{min_{exp}} | C \geq C_{obs}) = \frac{P(C \geq C_{obs}| \pi < \pi_{min_{exp}})P(\pi < \pi_{min_{exp}})}{P(C \geq C_{obs})}$$  

$$P(C \geq C_{obs}) = P(C \geq C_{obs}| \pi < \pi_{min_{exp}})P(\pi < \pi_{min_{exp}}) + P(C \geq C_{obs}| \pi \geq \pi_{min_{exp}})P(\pi \geq \pi_{min_{exp}})$$ 




__NOTE__ 
Not sure this is appropriate due to the difference in the range of $\pi$ used for the null and alternative hypothetis. May also want to consider a different alternative hypothesis $\pi$ upper limit, potentially using the $\pi_{min_{pre}}$, and post treatment $\pi_{min_{post}}$ to calculate a more realistic upper limit._ 


Consistent with the distribution of observed counts for titration-specific features more of the DADA2 features could not be explained by sampling alone compared to the other pipelines (Fig. \@ref(fig:mcBayes)).  

```{r message = FALSE, warning = FALSE, echo = FALSE}
eo_neg1 <- feature_exp_df  %>% filter(eo_metric == -1)

eo_binom_test <- eo_neg1 %>% 
      group_by(pipe, biosample_id, id, feature_id, t_fctr) %>% 
      nest() %>% 
      mutate(binom_test = map(data, ~with(., binom.test(x = 0, 
                                                        n = total_abu, 
                                                        exp_prop, 
                                                        alternative = "less"))),
             binom_tidy = map(binom_test, broom::tidy))

eo_binom_tidy <- eo_binom_test %>% 
      select(pipe, biosample_id, id, feature_id, t_fctr, binom_tidy) %>% 
      unnest() 

eo_binom_tidy$adj.pvalue <- p.adjust(eo_binom_tidy$p.value, method = "BH") 

plot_dat <- feature_exp_df %>% right_join(eo_binom_tidy) %>% 
    filter(count == 0 | exp_count == 0) %>% 
    mutate(feat_spec = if_else(count == 0 & exp_count != 0, "Unmix","Titration")) %>% 
    mutate(plot_count = if_else(count == 0 & exp_count != 0,exp_count,count))
```

```{r binomTestUnmix, fig.cap = "Proportion of unmix specific features with an adjusted p-value < 0.05 for the binomial test by biological replicate and pipeline. We fail to accept the null hypothesis when the p-value < 0.05, indicating that for these features the discrepancy between the feature not being observed in the titration and present in the unmixed samples cannot be explained by sampling alone.", message = FALSE, echo = FALSE}
plot_dat %>% mutate(pval_cat = if_else(adj.pvalue < 0.05, "p < 0.05", "p >= 0.05")) %>% 
      ggplot() + geom_bar(aes(x = biosample_id, fill = pval_cat), position = "fill") + 
      facet_wrap(~pipe) +
      theme_bw() + theme(axis.text.x = element_text(angle = 90)) +
      labs(x = "Biological Replicate", y = "Proprtion of Features", fill = "Adjusted p-value")
```

```{r message = FALSE, warning = FALSE, echo = FALSE}
min_prop <- feature_exp_df %>% filter(pre != 0, post != 0) %>% 
      group_by(pipe, biosample_id, t_fctr) %>% 
      summarise(min_pre = min(pre),
                min_post = min(post)) %>% 
      mutate(theta = 2^-as.numeric(t_fctr),
            min_exp_prop = min_post*theta + min_pre * (1-0))

eo_pos1 <- feature_exp_df  %>% filter(eo_metric == 1) %>% left_join(min_prop)

# mc_sim <- function(max_prop, total_abundance, obs_count){
#       sim_prob <- runif(1000, min = 0, max = max_prop)
#       sim_count <- sapply(X = sim_prob, FUN = rbinom, n = 1000, size = total_abundance)
#       sum(sim_count >= obs_count)/ length(sim_count)
# }
# titration_mc <- eo_pos1 %>% 
#       group_by(pipe, biosample_id, id, feature_id, t_fctr) %>% 
#       nest() %>% 
#       mutate(mc_prop = map(data, ~with(., mc_sim(max_prop = min_exp_prop, 
#                                                     total_abundance = total_abu, 
#                                                     obs_count = count))))
# saveRDS(titration_mc,"~/Desktop/monte-carlo-titration.RDS")
#titration_mc <- readRDS("~/Desktop/monte-carlo-titration.RDS")
#```

#```{r mcProp, fig.cap = "Distribution of titration specifc features PCR replicate titrations Monte Carlo count simulations with equal to or more extreme counts than the observed counts.", message = FALSE, echo = FALSE}
# titration_mc %>% unnest() %>% 
#       ggplot() + geom_density(aes(x = mc_prop, fill = biosample_id), alpha = 0.25) + 
#       facet_wrap(~pipe, scales = "free_y", ncol = 1) + theme_bw() + 
#       labs(x = "Proportion of Counts >= Observed Counts")
```

```{r message = FALSE, warning = FALSE, echo = FALSE}
# titration_mc %>% unnest() %>% 
#       mutate(prop_cat = if_else(mc_prop < 0.05, "<0.05", ">=0.05")) %>% 
#       ggplot() + geom_bar(aes(x = biosample_id, fill = prop_cat), position = "fill") +
#       facet_wrap(~pipe) +
#       theme_bw() + theme(axis.text.x = element_text(angle = 90)) +
#       labs(x = "Biological Replicate", y = "Proprtion of Features", fill = "Probability")
```

```{r message = FALSE, warning = FALSE, echo = FALSE}
# library(MonteCarlo)
# 
# calc_bayes_mc <- function(max_prop, total_abu, obs_count = count){
#       ## simulating binomial counts from uniform distribution
#     sim_count <- function(min_prop, max_prop, total_abundance){
#       sim_prob <- runif(1, min = min_prop, max = max_prop)
#       sim_count <- rbinom(n = 1, size = total_abundance, prob = sim_prob)
#       list(sim_count = sim_count)
#     }  
#     
#     ## sim distribution for P(Pi > max_prop)
#     param_list=list("min_prop" = max_prop, "max_prop" = 1, "total_abundance" = total_abu)
#     mc_gt_max <-MonteCarlo(func=sim_count, nrep=10000, param_list=param_list)
#     mc_gt_count <- mc_gt_max$results$sim_count
#     p_gt_max <- sum(mc_gt_count >= obs_count)/length(mc_gt_count)
# 
#     ## sim distribution for P(Pi <= max_prop)
#     param_list=list("min_prop" = 0, "max_prop" = max_prop, "total_abundance" = total_abu)
#     mc_lt_max <- MonteCarlo(func=sim_count, nrep=10000, param_list=param_list)
#     mc_lt_count <- mc_lt_max$results$sim_count
#     p_lt_max <- sum(mc_lt_count >= obs_count)/length(mc_lt_count)
#     
#     ## Bayesian Hypothesis test 
#     (p_lt_max * 0.5)/sum((p_lt_max * 0.5), (p_gt_max * 0.5))
# }
# 
# 
# eo_pos1 <- feature_exp_df  %>% filter(eo_metric == 1) %>% left_join(min_prop)
# 
# unmix_binom_test <- eo_pos1 %>% 
#       rowwise() %>% 
#       mutate(bayes_test = calc_bayes_mc(max_prop = min_exp_prop, total_abu = total_abu, obs_count = count))
# 
# saveRDS(unmix_binom_test,"~/Desktop/monte-carlo-unmix_binom_test.RDS")
unmix_binom_test <- readRDS("~/Desktop/monte-carlo-unmix_binom_test.RDS")
unmix_binom_test$adj.pvalue <- p.adjust(unmix_binom_test$bayes_test, method = "BH") 
```

```{r mcBayes, fig.cap = "Proportion of titration specific features with adjusted p-values <=0.05.", message = FALSE, echo = FALSE}
unmix_binom_test %>% mutate(pval_cat = if_else(adj.pvalue < 0.05, "p < 0.05", "p >= 0.05")) %>% 
      ggplot() + geom_bar(aes(x = biosample_id, fill = pval_cat), position = "fill") + 
      facet_wrap(~pipe) +
      theme_bw() + theme(axis.text.x = element_text(angle = 90)) +
      labs(x = "Biological Replicate", y = "Proprtion of Features", fill = "Adjusted p-value")
```

# Reference 
Benjamini, Y., and Hochberg, Y. (1995). Controlling the false discovery rate: a practical and powerful approach to multiple testing. Journal of the Royal Statistical Society Series B 57, 289–300.  

# Session information 

## Git repo commit information


```{r}
library(git2r)
repo <- repository(path = "../")
last_commit <- commits(repo)[[1]]
```

The current git commit of this file is `r last_commit@sha`, which is on the `r branches(repo)[[1]]@name` branch and was made by `r last_commit@committer@name` on `r when(last_commit)`. The current commit message is `r last_commit@summary`. The repository is online at https://github.com/nate-d-olson/mgtst-pub  


## Platform Information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
```


## Package Versions
```{r}
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
      knitr::kable()
```