---
title: "Titration Abundance Expectation Calculation"
author: "Nate Olson"
date: '`r Sys.Date()`'
always_allow_html: yes
output:
  pdf_document: default
  html_document: default
---

__TODO__ Add sanity checks for expected value calculations 

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd)
```

```{r echo = FALSE}
pipeline_dir <- "../../mgtst_pipelines"
mrexp <- get_mrexp(pipeline_dir)
```

## Objective
Generate a data_frame with observed and predicted count values for individual PCR replicates. 
Expected values are calculated using unmixed PCR count table values from the same set of PCR replicates, same half of one of the replicate 96 well plates. 


## Functions for Processing Data 
```{r}
## Extracting a tidy dataframe with count values from MRexpiment objects
get_count_df <- function(mrobj, agg_genus = FALSE){
      if(agg_genus){
            mrobj <- aggregateByTaxonomy(mrobj, lvl = "Rank6", 
                                         norm = FALSE, log = FALSE, sl = 1)
      }
      
      mrobj <- cumNorm(mrobj, p = 0.75)
      mrobj %>%
            # not sure whether or not to normalize counts prior to analysis
            MRcounts(norm = TRUE, log = FALSE, sl = 1000) %>%  
            #MRcounts(sl = 1) %>% 
            as.data.frame() %>% 
            rownames_to_column(var = "feature_id") %>% 
            gather("id","count", -feature_id)
}


```  

Extracting observed count values for unmixed pre and post samples

```{r} 
get_pre_post_counts <- function(count_df){
      pre_post_df <- count_df %>% 
            filter(t_fctr %in% c(0, 20)) %>% 
            mutate(pre_post = if_else(t_fctr == 0, "post_count", "pre_count")) 
      
      ## sanity checks 
      post_check <- pre_post_df %>% filter(pre_post == "post_count") %>% .$t_fctr %>% unique()
      if(length(post_check) != 1){
            print("post t_fctr not unique")
            paste0("Post: ", post_check) %>% print()
      } 
      
      if(post_check != 0){
            print("post t_fctr not 20")
            paste0("Post t_fctr: ", post_check) %>% print()      
      }
      
      pre_check <- pre_post_df %>% filter(pre_post == "pre_count") %>% .$t_fctr %>% unique()
      if(length(pre_check) != 1){
            print("pre t_fctr not unique")
            paste0("Pre: ", pre_check) %>% print()
      } 
      
      if(pre_check != 20){
            print("pre t_fctr not 20")
            paste0("Pre t_fctr: ", pre_check) %>% print()
      }
      pre_post_df %>% 
            select(pipe, biosample_id, feature_id, pcr_rep, count, pre_post) %>% 
            spread(pre_post, count)

}

``` 

Calculating expected counts 

```{r}
calc_expected <- function(count_df, pre_post_counts){
      
      titration_list <- data_frame(titration = c(1:5,10,15)) %>% 
            mutate(post_prop = 2^-titration) %>% list() %>% rep(nrow(pre_post_counts))
      
      pre_post_counts %>% ungroup() %>% 
            add_column(titration = titration_list) %>% unnest() %>% 
            mutate(exp_count = post_count * post_prop + pre_count * (1-post_prop))
}
```

## Generating Expected Count Data Frames 
One data frame for feature level analysis and one when aggregating at the genus level. 

### Feature Level
```{r}
count_df <- mrexp %>% map_df(get_count_df, .id = "pipe")

## Annotate with metadata and exclude NTCs
count_df <- pData(mrexp$dada2) %>% right_join(count_df) %>%
      filter(biosample_id != "NTC")
``` 

Adding expected count values to observed count value data frame and calculating residuals. 
```{r}
pre_post_counts <- get_pre_post_counts(count_df) 
exp_count <- calc_expected(count_df, pre_post_counts) 


## annotating data frame 
count_exp_df <- left_join(exp_count, count_df) %>% 
      dplyr::rename(obs_count = count) %>% 
      mutate(residual = exp_count - obs_count)
```

```{r}
count_exp_df %>% saveRDS("../data/expected_count_values_feature_df.rds") 
```

### Genus Level

```{r}
count_df <- mrexp %>% map_df(get_count_df, agg_genus = TRUE, .id = "pipe")

## Annotate with metadata and exclude NTCs
count_df <- pData(mrexp$dada2) %>% right_join(count_df) %>%
      filter(biosample_id != "NTC")
``` 

Adding expected count values to observed count value data frame and calculating residuals. 
```{r}
pre_post_counts <- get_pre_post_counts(count_df) 
exp_count <- calc_expected(count_df, pre_post_counts) 

## annotating data frame 
count_exp_df <- left_join(exp_count, count_df) %>% 
      dplyr::rename(obs_count = count) %>% 
      mutate(residual = exp_count - obs_count)
```

```{r}
count_exp_df %>% saveRDS("../data/expected_count_values_genus_df.rds") 
```

## Session information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
      knitr::kable()
``` 