---
title: "Low CSSLog value check"
author: "Nate Olson"
date: '`r Sys.Date()`'
output:
  pdf_document: default
  html_document: default
---

```{r setup, message = FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd)
library(phyloseq)
```

## Objective
Check on low csslog normalized count values.

## Approach
Generate histograms of raw and log transformed count values

## Loading data
Normalized count data from `2017-02-09-Normalization-df-Dev.Rmd`
```{r}
count_df <- readRDS("../data/normalize_count_df.rds")
```

## Histograms
```{r}
filt_counts <- count_df %>% filter(norm_method %in% c("raw","rawlog"))
```

```{r}
filt_counts %>% sample_frac(size = 0.25) %>% 
      ggplot() + 
      geom_histogram(aes(x = count)) + 
      facet_grid(pipe~norm_method, scales = "free") +
      scale_y_log10() + theme_bw()
```

### CSS and CSSLog Transformation
CSS and CSS log transformation normalized count values calculated with p = 0.75 and sl = 1 (no additional scaling).

```{r}
filt_counts <- count_df %>% filter(norm_method %in% c("css","csslog"))
```

```{r}
filt_counts %>% sample_frac(size = 0.25) %>% 
      ggplot() + 
      geom_histogram(aes(x = count)) + 
      facet_grid(pipe~norm_method, scales = "free") +
      scale_y_log10() + theme_bw()
```


## Session information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
      knitr::kable()
```