---
title: "Negative Binomial Expected Counts and EO Metric"
author: "Nate Olson"
date: '`r Sys.Date()`'
always_allow_html: yes
output:
  pdf_document: default
  html_document: default
---

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd) 
pipeline_dir <- "../../mgtst_pipelines"
mrexp <- get_mrexp(pipeline_dir)
```

## Objective
Generate a data_frame with observed and predicted count values for individual PCR replicates. 
Expected values are calculated using unmixed PCR count table values from the same set of PCR replicates, same half of one of the replicate 96 well plates. 


### Negative Binomal for Weighted Count Estimates
Calculating proportion of pre and post counts using negative binomial. 

- $q_{i,j,k}$ is the proportion of feature $i$ in PCR $k$ of sample $j$ where a sample is defined as an individual unmixed or mixed samples for a biological replicate.  
- $p_{j,k}$ is the total feature abundance for sample $j$, sum of all feature counts not the number of sequences generated for the sample.  
- $v_{i,j,k}$ is the variance of feature $i$ in PCR replicate $j$ of sample $k$.  

$$v_{i,j,k} = \frac{q_{i,j,k}(1-q_{i,j,k})}{p_j,k}$$

- $w_{i,j,k}$ is the weight function

$$w_{i,j,k} = \frac{v_{i,j,k}^{ -1}}{\sum_{k \in j}v_{i,j,k}^{ -1}}$$


- $q_{i,j}$ - the weighted count estimate for feature $i,k$

$$q_{i,j} = \sum_{k\in j} w_{i,j,k}q_{i,j,k}$$ 

## EO Metric

$$\frac{expected - observed}{expected + observed}$$


## Functions for Processing Data 
```{r}
## Extracting a tidy dataframe with count values from MRexpiment objects
get_count_df <- function(mrobj, agg_genus = FALSE, css = TRUE){
      if(agg_genus){
            mrobj <- aggregateByTaxonomy(mrobj, lvl = "Rank6", 
                                         norm = FALSE, log = FALSE, sl = 1)
      }
      
      if(css == TRUE){
            mrobj <- cumNorm(mrobj, p = 0.75)
            count_mat <- MRcounts(mrobj, norm = TRUE, log = FALSE, sl = 1000)
      }else{
            count_mat <- MRcounts(mrobj, norm = FALSE, log = FALSE, sl = 1)
      }
       count_mat %>%  
            as.data.frame() %>% 
            rownames_to_column(var = "feature_id") %>% 
            gather("id","count", -feature_id)
}

## Calculating feature level pre and post proportions
calc_pre_post_prop <- function(count_df){
      ## Estimating $q_{i}$ for pre and post
      nb_est <- count_df %>% filter(t_fctr %in% c(0, 20)) %>% 
            mutate(prop = count/total_abu, 
                   prop_var = (prop * (1 - prop))/total_abu,
                   inv_var = 1/prop_var) %>% 
            group_by(pipe, biosample_id, t_fctr, feature_id) %>% 
            mutate(weight = inv_var / sum(inv_var)) %>% 
            summarise(prop_est = sum(weight*prop)) 
      
      # Reformatting data
      nb_est %>% ungroup() %>% 
            mutate(treat = if_else(t_fctr == "20", "pre","post")) %>% 
            select(-t_fctr) %>% 
            mutate(prop_est = if_else(is.nan(prop_est), 0, prop_est)) %>% 
            spread(treat, prop_est)
}

calc_expected_prop <- function(pre_post_prop){
      titration_list <- data_frame(titration = c(1:5,10,15)) %>% 
            mutate(post_prop = 2^-titration) %>% 
            list() %>% rep(nrow(pre_post_prop))
      
      pre_post_prop %>% ungroup() %>% 
            add_column(titration = titration_list) %>% unnest() %>% 
            mutate(exp_prop = post * post_prop + pre * (1-post_prop)) %>% 
            mutate(t_fctr = factor(titration)) %>% 
            select(-post_prop)
}


calc_expected_count <- function(exp_prop_df, count_df){
      count_df %>% 
            filter(t_fctr %in% c(1:5, 10, 15)) %>% 
            left_join(exp_prop_df) %>% 
            mutate(exp_count = total_abu * exp_prop) %>% 
            filter(!(pre == 0 & post == 0 & count == 0))
}

calc_eo_metric <- function(exp_count_df){
      exp_count_df %>% 
      mutate(eo_metric = (count - exp_count)/(count + exp_count))
}
```


### Feature Level Expected Counts
```{r}
count_df <- mrexp %>% map_df(get_count_df, css = FALSE, .id = "pipe") %>% 
      left_join(pData(mrexp$dada2)) %>% 
      filter(biosample_id != "NTC") %>% 
      select(pipe, biosample_id, id, pcr_rep, feature_id, t_fctr, count) 

count_df <- count_df %>% group_by(id) %>% mutate(total_abu = sum(count)) 


eo_exp_count_df <- count_df %>% 
      calc_pre_post_prop() %>% 
      calc_expected_prop() %>% 
      calc_expected_count(count_df) %>% 
      calc_eo_metric()

eo_exp_count_df %>% saveRDS("../data/nb_expected_eo_metric_feature_df.rds") 

``` 

### Genus Level Expected Counts
```{r}
count_df <- mrexp %>% map_df(get_count_df, agg_genus = TRUE, css = FALSE, .id = "pipe") %>% 
      left_join(pData(mrexp$dada2)) %>% 
      filter(biosample_id != "NTC") %>% 
      select(pipe, biosample_id, id, pcr_rep, feature_id, t_fctr, count) 

count_df <- count_df %>% group_by(id) %>% mutate(total_abu = sum(count)) 

eo_exp_count_df <- count_df %>% 
      calc_pre_post_prop() %>% 
      calc_expected_prop() %>% 
      calc_expected_count(count_df) %>% 
      calc_eo_metric()

eo_exp_count_df %>% saveRDS("../data/nb_expected_eo_metric_genus_df.rds")
```


# Session information 

## Git repo commit information
```{r}
library(git2r)
repo <- repository(path = "../")
last_commit <- commits(repo)[[1]]
```

The current git commit of this file is `r last_commit@sha`, which is on the `r branches(repo)[[1]]@name` branch and was made by `r last_commit@committer@name` on `r when(last_commit)`. The current commit message is `r last_commit@summary`. The repository is online at https://github.com/nate-d-olson/mgtst-pub  


## Platform Information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
```


## Package Versions
```{r}
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
      knitr::kable()
```