---
title: "Investigating potential outlier samples."
author: "Nate Olson"
date: '`r Sys.Date()`'
always_allow_html: yes
output:
  pdf_document: default
  html_document: default
---

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(tidyverse)
```

## Objective
Sample 1-F9, a PCR replicate of E01JH0016, had a relatively low number of reads per sample (~10,000) compared to the other samples. 
Additionally, the band was present when the 16S PCR amplicons were checked prior to library prep using gel electrophoresis. 
Sample 1-F4 had low base quality score for the reverse read. 


```{r}
count_exp_df <- readRDS("../data/expected_count_values_feature_df.rds") 
feature_cat <- readRDS("../data/feature_categories_df.rds")
count_exp_df <- left_join(count_exp_df, feature_cat) %>% 
      left_join(ntc_features) %>% 
      filter(cat %in% c("cat_full", "cat_pre", "cat_post")) %>% 
      mutate(bad_sample = if_else(id %in% c("1-F9","1-F4"),id,"0"))
```

The low quality for the reverse read did not bias the observed counts for sample 1-F4. 
The relativly low number of reads for sample 1-F9 appears to have biased the count values, will exclude this sample from analysis.  

```{r}
count_exp_df %>% filter(bad_sample != "0") %>% 
    ggplot() + 
      geom_abline(aes(intercept = 0, slope = 1)) + 
    geom_point(aes(x = obs_count + 1, y = exp_count + 1)) + 
    geom_smooth(aes(x = obs_count + 1, y = exp_count + 1), color = "darkorange") + 
    facet_grid(pipe~biosample_id)+ theme_bw() + 
    labs(y = "Expected Count", x = "Observed Count", fill = "Abundance") + 
    scale_y_continuous(trans = "log2") + 
    scale_x_continuous(trans = "log2") + 
    theme(axis.text.x = element_text(angle = 270))
```

# Session information 

## Git repo commit information
```{r}
repo <- repository(path = ".")
last_commit <- commits(repo)[[1]]
```

The current git commit of this file is `r last_commit@sha`, which is on the `r branches(repo)[[1]]@name` branch and was made by `r last_commit@committer@name` on `r when(last_commit)`. The current commit message is `r last_commit@summary`. The repository is online at https://github.com/nate-d-olson/mgtst-pub  


## Platform Information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
```


## Package Versions
```{r}
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
      knitr::kable()
```