---
title: "Feature Clustering"
author: "Nate Olson"
date: '`r Sys.Date()`'
output:
  bookdown::pdf_document2: 
    toc: FALSE
---

## Objective
Use PCA to identify group of features based titration comparison logFC estimates. 

## Conclusion
No clear separation of features not sure this approach is worth further investigation.  

```{r cluster_setup, include=FALSE}
library(tidyverse)
library(stringr)
library(ggfortify)
logFC_edgeR_df <- readRDS("~/Desktop/logFC_edgeR_df.rds")
```


```{r}
logFC_wide <- logFC_edgeR_df %>% 
      rename(feature_id = OTUname) %>% 
      #filter(pipe == "dada2") %>% 
      filter(T1 %in% c(1:4), T2 %in% c(2:5)) %>% 
      select(pipe, biosample_id, feature_id, T1, T2, logFC, Rank2) %>% 
      unite(t_comp, T1, T2, sep = ":") %>% 
      mutate(t_comp = paste0("T", t_comp)) %>%
      spread(t_comp, logFC)


logFC_mat <- logFC_wide %>% ungroup() %>% 
      select(-pipe, -biosample_id, -feature_id, -Rank2) %>% 
      as.matrix()
non_zero_rows <- rowSums(logFC_mat) != 0
# rownames(logFC_mat) <- logFC_wide$feature_id    
logFC_mat <- logFC_mat[non_zero_rows,]
logFC_wide <- logFC_wide[non_zero_rows, ]
## Number of non-zero logFC
# logFC_01 <- logFC_mat
# logFC_01[logFC_01 != 0] <- 1
# logFC_01 %>% rowSums() %>% summary()


```

```{r}
df <- logFC_wide %>% ungroup() %>% select(-pipe, -biosample_id, -feature_id, -Rank2)
logFC_rank <- logFC_wide %>% ungroup() %>% select(-feature_id) %>% 
      mutate(Rank2 = str_replace(Rank2, "p__",""))
pcs <- prcomp(df,center = FALSE, scale. = FALSE)
```

```{r}
plot(pcs)
```




```{r}
autoplot(pcs)
```

```{r}
autoplot(pcs, data = logFC_rank, colour = "Rank2") + facet_grid(pipe~biosample_id)
```



## Git repo commit information
```{r}
library(git2r)
repo <- repository(path = "../")
last_commit <- commits(repo)[[1]]
```

The current git commit of this file is `r last_commit@sha`, which is on the `r branches(repo)[[1]]@name` branch and was made by `r last_commit@committer@name` on `r when(last_commit)`. The current commit message is `r last_commit@summary`. The repository is online at https://github.com/nate-d-olson/mgtst-pub  


## Platform Information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
```


## Package Versions
```{r}
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
      knitr::kable()
```