---
title: "RIPs Talk Figures"
author: "Nate Olson"
date: "5/10/2017"
output: html_document
---

```{r echo=FALSE, message=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd)
library(modelr) 
pipeline_dir <- "../../mgtst_pipelines"
mrexp <- get_mrexp(pipeline_dir)
```

```{r}
get_count_df <- function(mrobj, agg_genus = FALSE){
      if(agg_genus){
            mrobj <- aggregateByTaxonomy(mrobj, lvl = "Rank6", 
                                         norm = FALSE, log = FALSE, sl = 1)
      }
      
      mrobj <- cumNorm(mrobj, p = 0.75)
      mrobj %>%
            # not sure whether or not to normalize counts prior to analysis
            MRcounts(norm = FALSE, log = FALSE, sl = 1) %>%  
            as.data.frame() %>% 
            rownames_to_column(var = "feature_id") %>% 
            gather("id","count", -feature_id)
} 

count_df <- mrexp %>% map_df(get_count_df, .id = "pipe") %>% 
      left_join(pData(mrexp$dada2)) %>% 
      filter(biosample_id != "NTC") %>% 
      ungroup() %>% 
      mutate(t_fctr = fct_relevel(t_fctr, c(0:5, 10, 15, 20)),
             titration = as.numeric(as.character(t_fctr)), 
             theta = 2^(-titration),
             theta = if_else(theta == 2^-20, 0, theta))  #%>% 
      # filter(t_fctr %in% c(0:4,20))

```


## Raw Expected and Observed Counts
```{r}
raw_pre_post_df <- count_df %>% filter(t_fctr %in% c(0,20)) %>% 
    group_by(pipe, feature_id, biosample_id, t_fctr) %>% 
    summarise(count = mean(count)) %>% 
    mutate(count_name = if_else(t_fctr == 0, "post_count", "pre_count")) %>% 
    dplyr::select(-t_fctr) %>% 
    spread(count_name, count)
```

```{r}
raw_count_df <- raw_pre_post_df %>% left_join(count_df) %>% 
    mutate(exp_count = theta * post_count + (1-theta) * pre_count) %>% 
    filter(exp_count != 0 | count != 0)
```

```{r}
raw_count_df %>% ggplot() +
    geom_hex(aes(x = count + 1, y = exp_count + 1)) +
    geom_smooth(aes(x = count + 1, y = exp_count + 1), color = "darkorange") + 
    facet_grid(pipe~biosample_id)+ theme_bw() +
    labs(y = "Expected Count", x = "Observed Count", fill = "Abundance") +
    scale_y_continuous(trans = "log2") +
    scale_x_continuous(trans = "log2") +
    theme(axis.text.x = element_text(angle = 270))
```


### Excluding Counts with logFC < 1
```{r}
good_logfc <- readRDS("../data/pre_post_deseq_logFC.RDS") %>% 
    filter(abs(log2FoldChange) > 1, pvalue < 0.05) 
filt_fc_count_df <- raw_count_df %>% right_join(good_logfc)
```

```{r}
filt_fc_count_df %>% ggplot() +
    geom_hex(aes(x = count + 1, y = exp_count + 1)) +
    geom_smooth(aes(x = count + 1, y = exp_count + 1), color = "darkorange") + 
    facet_grid(pipe~biosample_id)+ theme_bw() +
    labs(y = "Expected Count", x = "Observed Count", fill = "Abundance") +
    scale_y_continuous(trans = "log2") +
    scale_x_continuous(trans = "log2") +
    theme(axis.text.x = element_text(angle = 270))
```




## Filtering Mix and Unmix specific features
```{r}
## Excluding features with median EO metric of 1 or -1 and logFC between pre and post < 1
## %%TODO%% replace with heuristic filter > 14 of the PCR replicates have observed counts, and only features with observed counts for unmixed samples.  
count_nb <- readRDS("../data/nb_expected_eo_metric_feature_df.rds") %>% 
      mutate(count_post = post * total_abu, count_pre = pre * total_abu)

good_eo <- count_nb %>% group_by(pipe, biosample_id, feature_id) %>% 
      summarise(med_eo = median(eo_metric)) %>% 
      filter(abs(med_eo) != 1)

good_features <- count_nb %>% group_by(pipe, biosample_id, feature_id) %>% 
      summarise(med_eo = median(eo_metric)) %>% 
      filter(abs(med_eo) != 1)

filt_count_df <- raw_count_df %>% right_join(good_features) %>% right_join(good_logfc)
```

```{r}
filt_count_df %>% ggplot() +
    geom_hex(aes(x = count + 1, y = exp_count + 1)) +
    geom_smooth(aes(x = count + 1, y = exp_count + 1), color = "darkorange") + 
    facet_grid(pipe~biosample_id)+ theme_bw() +
    labs(y = "Expected Count", x = "Observed Count", fill = "Abundance") +
    scale_y_continuous(trans = "log2") +
    scale_x_continuous(trans = "log2") +
    theme(axis.text.x = element_text(angle = 270))
```





## Negative Binomial Count Estimates

```{r}
filt_nb_count_df <- count_nb %>% right_join(good_features) %>% right_join(good_logfc)
```

```{r}
filt_nb_count_df %>% ggplot() +
    geom_hex(aes(x = count + 1, y = exp_count + 1)) +
    geom_smooth(aes(x = count + 1, y = exp_count + 1), color = "darkorange") + 
    facet_grid(pipe~biosample_id)+ theme_bw() +
    labs(y = "Expected Count", x = "Observed Count", fill = "Abundance") +
    scale_y_continuous(trans = "log2") +
    scale_x_continuous(trans = "log2") +
    theme(axis.text.x = element_text(angle = 270))
```


## Theta Corrected Count Estimates
```{r}
theta_fit <- readRDS("../data/theta_estimates.RDS")
theta_count_df <- filt_nb_count_df %>% left_join(theta_fit) %>% 
    mutate(exp_prop = post * estimate + min(1,1- estimate) * pre,
           exp_count = exp_prop * total_abu)
```
```{r}
theta_count_df %>% ggplot() +
    geom_hex(aes(x = count + 1, y = exp_count + 1)) +
    geom_smooth(aes(x = count + 1, y = exp_count + 1), color = "darkorange") + 
    facet_grid(pipe~biosample_id)+ theme_bw() +
    labs(y = "Expected Count", x = "Observed Count", fill = "Abundance") +
    scale_y_continuous(trans = "log2") +
    scale_x_continuous(trans = "log2") +
    theme(axis.text.x = element_text(angle = 270))
```

