---
title: "No Template Control Characterization"
author: "Nate Olson"
date: '`r Sys.Date()`'
always_allow_html: yes
output:
  pdf_document: default
  html_document: default
---

## Objective
Identify and characterize features present in No Template Controls. 

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd)
library(git2r)
```

```{r}
get_count_df <- function(mrobj, agg_genus = FALSE){
      if(agg_genus){
            mrobj <- aggregateByTaxonomy(mrobj, lvl = "Rank6", 
                                         norm = FALSE, log = FALSE, sl = 1)
      }
      
      mrobj <- cumNorm(mrobj, p = 0.75)
      mrobj %>%
            # not sure whether or not to normalize counts prior to analysis
            MRcounts(norm = TRUE, log = FALSE, sl = 1000) %>%  
            as.data.frame() %>% 
            rownames_to_column(var = "feature_id") %>% 
            gather("id","count", -feature_id)
}

pipeline_dir <- "../../mgtst_pipelines"
mrexp <- get_mrexp(pipeline_dir)


```

```{r}
count_df <- mrexp %>% map_df(get_count_df, .id = "pipe") %>% 
      left_join(pData(mrexp$dada2))
```


```{r}
ntc_summary <- count_df %>% filter(biosample_id == "NTC") %>% 
      filter(count != 0) %>% 
      group_by(pipe, feature_id) %>% 
      summarise(total_count = sum(count),
                med_count = median(count), 
                n_present = n())
```

Note count values are CSS normalized and scaled. 
```{r message = FALSE, warning = FALSE}
ntc_summary %>% ggplot() + 
      geom_histogram(aes(x = total_count)) + scale_y_log10() + theme_bw() +
      labs(x = "Total abundance across all NTC samples.", y = "Number of Features")
```

```{r message = FALSE}
ntc_summary %>% ggplot() + 
      geom_histogram(aes(x = n_present)) + theme_bw() +
      labs(x = "Number of NTC samples feature is present in.", 
           y = "Number of Features")
```


```{r}
ntc_summary %>% ggplot() + 
      geom_point(aes(x = n_present, y = total_count)) + theme_bw() +
      labs(x = "Number of NTC Samples", y = "Total Abundance")
```

```{r}
ntc_summary %>% ggplot() + 
      geom_point(aes(x = n_present, y = med_count)) + theme_bw() +
      labs(x = "Number of NTC Samples", y = "Median Abundance")
```

__High Abundance Features__
* Features with a high total abundance or count and low median count are features with a high abundance count in few samples and low abundance in other samples. 

```{r}
ntc_summary %>% filter(total_count > 4000) %>% 
      arrange(desc(total_count)) %>% knitr::kable()
```

High Frequency Features
```{r}
ntc_summary %>% filter(n_present > 6) %>% 
      arrange(desc(n_present)) %>% knitr::kable()
```

__Taxonomy of high abundance and fequency features__  

* Taxonomy of high abundance no template control features is not consistent between pipelines.  
* May not want to drop all no template controls features as _Escherichia/Shigella_ features that are relevant to the study were present in NTCs.  

```{r}
feature_taxa <- mrexp %>% map_df(fData, .id = "pipe") %>% 
      dplyr::rename(feature_id = OTUname)
ntc_summary %>% filter(n_present > 6 | total_count > 4000) %>% 
      left_join(feature_taxa) %>% select(pipe, feature_id, n_present, total_count, Rank5, Rank6) %>% 
      arrange(Rank5) %>% knitr::kable()
```

## NTC Feature Bias Analysis
```{r}
count_exp_df <- readRDS("../data/expected_count_values_feature_df.rds")
feature_cat <- readRDS("../data/feature_categories_df.rds")
count_exp_df <- left_join(count_exp_df, feature_cat) %>%
      left_join(ntc_summary) %>%
      filter(cat %in% c("cat_full", "cat_pre", "cat_post")) %>%
      mutate(ntc = if_else(is.na(n_present), "0","1"),
             bad_sample = if_else(id %in% c("1-F9","1-F4"),id,"0"))
```

Features observed in at least one no template control sample overlaid on plot of all features. 
Overall the no template control features perform similarly in terms of bias relative to features not present in any no template control sample. 
```{r message=FALSE}
count_exp_df %>%
    ggplot() + 
    geom_hex(aes(x = obs_count + 1, y = exp_count + 1)) + 
  
      geom_point(data =  filter(count_exp_df, ntc == "1"),
                 aes(x = obs_count + 1, y = exp_count + 1), color = "lightgreen", alpha = 0.25) + 
        geom_smooth(aes(x = obs_count + 1, y = exp_count + 1, color = ntc)) + 
    facet_grid(pipe~biosample_id)+ theme_bw() + 
    labs(y = "Expected Count", x = "Observed Count", fill = "Abundance") + 
    scale_y_continuous(trans = "log2") + 
    scale_x_continuous(trans = "log2") + 
    theme(axis.text.x = element_text(angle = 270))
```

## Genus level No Template Control Analysis

```{r}
genus_df <- mrexp %>% map_df(get_count_df, agg_genus = TRUE, .id = "pipe") %>% 
      left_join(pData(mrexp$dada2))
```


```{r}
ntc_genus <- genus_df %>% filter(biosample_id == "NTC") %>% 
      filter(count != 0) %>% 
      group_by(pipe, feature_id) %>% 
      summarise(total_count = sum(count),
                med_count = median(count), 
                n_present = n())
```


```{r}
ntc_genus %>% ggplot() + 
      geom_point(aes(x = n_present, y = total_count)) + theme_bw() +
      labs(x = "Number of NTC Samples", y = "Total Abundance")
```


__High Frequency and Abundance Genus Level Features__
```{r}
ntc_genus %>% filter(n_present > 6 | total_count > 5000) %>% 
      arrange(desc(n_present)) %>% knitr::kable()
```

## Saving data 
```{r}
ntc_summary %>% saveRDS("../data/ntc_features.rds")
ntc_genus %>% saveRDS("../data/ntc_genus.rds")
```


# Session information 

## Git repo commit information
```{r}
repo <- repository(path = "../")
last_commit <- commits(repo)[[1]]
```

The current git commit of this file is `r last_commit@sha`, which is on the `r branches(repo)[[1]]@name` branch and was made by `r last_commit@committer@name` on `r when(last_commit)`. The current commit message is `r last_commit@summary`. The repository is online at https://github.com/nate-d-olson/mgtst-pub  


## Platform Information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
```


## Package Versions
```{r}
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
      knitr::kable()
```