---
title: "Pre and Post Specific Features"
author: "Nate Olson"
date: "February 8, 2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd)
```

## Objective
Classify features based on presence in pre and post-treatment replicates. 

### Classification scheme 

* pre_partial - not present in any post and present in 1-3 of the pre-treatment replicates 
* pre_full - not present in any post and present in 4 of the pre-treatment replicate 
* post_partial - not present in any pre and present in 1-3 of the post-treatment replicates
* post_partial - not present in any pre and present in 4 of the post-treatment replicates 
* mixed - not present in any pre or post-treatment replicates but present in titration replicates
* biorep_neg - not present in unmixed or mixed samples for the biological replicate

## Loading pipeline data
```{r}
mrexp <- list(dada2  = "../data/mrexp_dada2.RDS",
              mothur = "../data/mrexp_mothur.RDS",
              qiime  = "../data/mrexp_qiime_refclus_nochimera.RDS") %>% 
      map(readRDS)
```

Adding `sam_names` to `pData` for DADA2 and QIIME MRexperiments, and converting dilutions to factors.
```{r}
pdat <- pData(mrexp$dada2)
pdat$sam_names <- rownames(pdat)
pdat$dilution <- pdat$dilution %>% factor()
pData(mrexp$dada2) <- pdat 

pdat <- pData(mrexp$mothur)
pdat$dilution <- pdat$dilution %>% factor()
pData(mrexp$mothur) <- pdat 

pdat <- pData(mrexp$qiime) 
pdat$sam_names <- rownames(pdat)
pdat$dilution <- pdat$dilution %>% factor()
pData(mrexp$qiime) <- pdat
```

## Development  
Working out classification for DADA2 features first. 

```{r}
# pre_post_filt <- mrexp$dada2 %>% pData() %>% .$dilution %in% c("-1","0")
pre_post_dada2 <- mrexp$dada2 #%>% {.[,pre_post_filt]}# %>% 
      # filterData(present = 1, depth = 1)

bio_unmixed <- pre_post_dada2 %>% pData()

bio_unmixed$treatment <- bio_unmixed$dilution %>% 
      {case_when(
            . == -1 ~ "post",
            . == 0 ~ "pre",
            . %in% c(1:5, 10, 15) ~ "mixed"
      )}

bio_unmixed_cl <- bio_unmixed %>% 
      unite("bio_unmixed", sampleID, treatment, sep = ":") %>% 
      .$bio_unmixed

treatment_features <- uniqueFeatures(pre_post_dada2,bio_unmixed_cl) %>% as.data.frame() %>% 
      rownames_to_column("otuID") %>% gather("key","value",-otuID, -featureIndices) %>% 
      mutate(key = gsub(" in","", key)) %>% 
      separate(key,c("key","bio_unmixed"),sep = " ") %>%
      separate(bio_unmixed, c("sampleID","treatment"),sep = ":") %>% 
      spread(key,value) %>% 
      dplyr::rename(Samp =`Samp.`)
```

```{r}
treatment_features %>% glimpse()
```


### Pre specific features by sampleID
```{r}
pre_specific_features <- treatment_features %>% select(-Reads) %>% 
      spread(treatment, Samp) %>% filter(post == 0, pre != 0)
```

There are a total of `r pre_specific_features$otuID %>% unique() %>% length()` pre-treatment specific features. 
One of the biological replicates (E01JH0038) has approximately half the number of features compared to the other biological replicates. 

```{r}
pre_specific_features %>% 
      ggplot() + geom_bar(aes(x = sampleID)) + theme_bw() + 
      labs(x = "Biological Replicate", y = "Pre-Treatment Specific Features")
```

Most of the pre-treatment specific features are unique to one of the biological replicates and only one feature is specific to pre-treatment samples for all five biological replicates. 
```{r}
pre_specific_features_nbiorep <- pre_specific_features %>% 
      group_by(otuID) %>% summarise(n_biorep = n())
pre_specific_features_nbiorep %>% group_by(n_biorep) %>% summarise(count = n()) %>% kable()
```

### Post-Treatment specific features by sampleID
```{r}
post_specific_features <- treatment_features %>% select(-Reads) %>% 
      spread(treatment, Samp) %>% filter(post != 0, pre == 0)
```

There are a total of `r post_specific_features$otuID %>% unique() %>% length()` pre-treatment specific features. 

```{r}
post_specific_features %>% 
      ggplot() + geom_bar(aes(x = sampleID)) + theme_bw() + 
      labs(x = "Biological Replicate", y = "Post-Treatment Specific Features")
```

Similar to the pre-treatment specific features, most of the post-treatment specific features are unique to one of the biological replicates. 
However, five features are specific to post-treatment samples for all five biological replicates.
```{r}
post_specific_features_nbiorep <- post_specific_features %>% 
      group_by(otuID) %>% summarise(n_biorep = n())
post_specific_features_nbiorep %>% group_by(n_biorep) %>% summarise(count = n()) %>% kable()
```

### Pre-Treatment Post-Treatment Specific Features
```{r}
pre_post_absent_features <- treatment_features %>% select(-Reads) %>% 
      spread(treatment, Samp) %>% filter(post == 0, pre == 0, mixed != 0)
```

There are a total of `r pre_post_absent_features$otuID %>% unique() %>% length()` features that are not present in pre or post-treatment samples. 
No expected values for these features. 
These features should be included in the detection (presence/ absence) analysis. 

```{r}
pre_post_absent_features %>% 
      ggplot() + geom_bar(aes(x = sampleID)) + theme_bw() + 
      labs(x = "Biological Replicate", y = "Pre and Post-Treatment Absent Features")
```

```{r}
pre_post_absent_features_nbiorep <- pre_post_absent_features %>% 
      group_by(otuID) %>% summarise(n_biorep = n())
pre_post_absent_features_nbiorep %>% group_by(n_biorep) %>% summarise(count = n()) %>% kable()
```

## Treatment-Specific Code for all pipelines

```{r}
get_treatment_class <- function(mrexp_pData){
      ## a character vector classifying samples as pre, post, or mixed by
      ## biological replicate (sampleID)
      
      mrexp_pData$treatment <- mrexp_pData$dilution %>% 
            {case_when(
                  is.na(.)  ~ "NTC",
                  . == 0 ~ "post",
                  . == -1 ~ "pre",
                  . %in% c(1:5, 10, 15) ~ "mixed"
            )}

     mrexp_pData %>% 
           unite("bio_unmixed", sampleID, treatment, sep = ":") %>% 
           .$bio_unmixed
}

get_unique_feature_matrix <- function(mrexp_obj){
      ## get matrix with number of samples and reads observed for each feature
      ## by sampleID (biological replicate) and feature
      
      ## get treatment_class
      cl <- mrexp_obj %>% pData() %>% get_treatment_class()
      
      ## uniqueFeatures matrix
      uniqueFeatures(mrexp_obj, cl)
}

tidy_unique_feature_matrix <- function(unq_features){
      ## convert the uniqueFeatures matrix into a tidy data frame for downstream
      ## analysis
      unq_features %>% as.data.frame() %>% 
            rownames_to_column("otuID") %>% 
            gather("key","value",-otuID, -featureIndices) %>% 
            mutate(key = gsub(" in","", key)) %>% 
            separate(key,c("key","bio_unmixed"),sep = " ") %>%
            separate(bio_unmixed, c("sampleID","treatment"),sep = ":") %>% 
            spread(key,value) %>% 
            dplyr::rename(Samp =`Samp.`)
}

annotate_unique_features <- function(unq_features_df){
      ## annotate features as pre, post, or unmixed specific for each biological
      ## replicate
      treatment_features <- unq_features_df %>% select(-Reads) %>% 
            spread(treatment, Samp)
      
      ## Specificity Annotations
      ## pre_partial - not present in any post and present in 1-3 of the pre-treatment replicates 
      ## pre_full - not present in any post and present in 4 of the pre-treatment replicate 
      ## post_partial - not present in any pre and present in 1-3 of the post-treatment replicates
      ## post_full - not present in any pre and present in 4 of the post-treatment replicates 
      ## mix_specific - not present in any pre or post-treatment replicates but present in titration replicates
      ## biorep_neg - not present in unmixed or mixed samples for the biological replicate
      
      treat_anno <- treatment_features %>% ungroup() %>% 
            select(-NTC) %>% filter(sampleID != "NTC") %>% 
            mutate(specific_anno = "non-specific",
                   specific_anno = if_else(post == 0 & pre != 0, 
                                           "pre_partial", specific_anno),
                   specific_anno = if_else(post == 0 & pre == 4, 
                                           "pre_full", specific_anno),
                   specific_anno = if_else(post != 0 & pre == 0, 
                                           "post_partial", specific_anno),
                   specific_anno = if_else(post == 4 & pre == 0, 
                                           "post_full", specific_anno),
                   specific_anno = if_else(post == 0 & pre == 0 & mixed != 0, 
                                           "mix_specific", specific_anno),
                   specific_anno = if_else(post == 0 & pre == 0 & mixed == 0, 
                                           "biorep_neg", specific_anno))
      
      ## annotiting NTC features
      ntc_features <- treatment_features %>% 
            filter(sampleID == "NTC") %>% select(otuID,NTC)
      
      unq_features_df <- unq_features_df %>% 
            filter(sampleID != "NTC") %>% left_join(ntc_features)
      
      treat_anno %>% select(otuID, featureIndices, sampleID, specific_anno) %>%
            right_join(unq_features_df)
}
```


### Feature Specificity
```{r}
feature_specificity_mat <- mrexp %>% map(get_unique_feature_matrix) 
feature_specificity_tidy <- feature_specificity_mat %>%  map(tidy_unique_feature_matrix) 
feature_specificity_df <- feature_specificity_tidy %>% map_df(annotate_unique_features, .id = "pipe")
```

```{r}
glimpse(feature_specificity_df)
```


Read counts for specificity classifications 
```{r}
mixed_specific <- feature_specificity_df %>% 
      filter(specific_anno %in% c("non-specific","mixed")) %>% 
      group_by(pipe, otuID, featureIndices, sampleID, specific_anno) %>% 
      summarise(Reads = sum(Reads), Samp = sum(Samp))

pre_specific <- feature_specificity_df %>% 
      filter(specific_anno %in% c("pre_full","pre_partial"), treatment == "pre") %>% 
      select(pipe, otuID, featureIndices,  sampleID, specific_anno, Reads, Samp)

post_specific <- feature_specificity_df %>% 
      filter(specific_anno %in% c("post_full","post_partial"), treatment == "post") %>% 
      select(pipe, otuID, featureIndices,  sampleID, specific_anno, Reads, Samp)

feature_specificity_counts_df <- bind_rows(mixed_specific, pre_specific, post_specific)
```

```{r}
glimpse(feature_specificity_counts_df )
```


### Saving data frames as RDS
```{r}
feature_specificity_df %>% saveRDS("../data/feature_specificity_df.rds") 
feature_specificity_counts_df %>% saveRDS("../data/feature_specificity_counts_df.rds")
```


## Feature Specificity Summary
```{r}
feature_specificity_df %>% filter(specific_anno != "biorep_neg") %>% 
      ggplot() + geom_bar(aes(x = specific_anno)) + 
      facet_wrap(~pipe, ncol = 1, scale = "free_y") +
      theme_bw()
# ggsave("~/Desktop/feature_specificity.png", width = 5.48, height = 6, dpi = 450)
```

Expected high accuracy features based on titration expectations are non-specific, post_full, and pre_full features. 




```{r}
feature_specificity_counts_df %>%
      ggplot() + geom_jitter(aes(x = specific_anno, y = Reads), alpha = 0.25) + 
      facet_wrap(~pipe, ncol = 1, scale = "free_y") + theme_bw() + scale_y_log10() 
```
 
