---
title: "Example Features"
author: "Nate Olson"
date: '`r Sys.Date()`'
always_allow_html: yes
output:
  html_document: default
---

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd) 
```

### Objective 
Generate a set of example features and provide summary metrics for these features based on the different models and metrics being evaluated. 

```{r}
feature_cat <- readRDS("../data/feature_categories_df.rds")

pipeline_dir <- "../../mgtst_pipelines"
mrexp <- get_mrexp(pipeline_dir)

## Extracting a tidy dataframe with count values from MRexpiment objects
get_count_df <- function(mrobj, agg_genus = FALSE){
      if(agg_genus){
            mrobj <- aggregateByTaxonomy(mrobj, lvl = "Rank6", 
                                         norm = FALSE, log = FALSE, sl = 1)
      }
      
      mrobj <- cumNorm(mrobj, p = 0.75)
      mrobj %>%
            # not sure whether or not to normalize counts prior to analysis
            MRcounts(norm = TRUE, sl = 1000) %>%  
            as.data.frame() %>% 
            rownames_to_column(var = "feature_id") %>% 
            gather("id","count", -feature_id)
} 

count_df <- mrexp %>% map_df(get_count_df, .id = "pipe") %>% 
      left_join(pData(mrexp$dada2)) %>% 
      filter(biosample_id != "NTC") %>% 
      mutate(t_fctr = fct_relevel(t_fctr, c(20, 15, 10, 5:0))) %>%
      left_join(feature_cat) #%>% filter(cat %in% c("cat_full", "cat_pre","cat_post"))
```

## Escherichia Features
### DADA2
```{r}
count_df %>% filter(feature_id == "SV1") %>% 
      ggplot() + 
      geom_point(aes(x = t_fctr, y = count)) + 
      facet_wrap(~biosample_id, scales = "free") + theme_bw()

```

### Mothur 
```{r}
count_df %>% filter(feature_id == "Otu00001") %>% 
      ggplot() + 
      geom_point(aes(x = t_fctr, y = count)) + 
      facet_wrap(~biosample_id, scales = "free") + theme_bw()
```




### QIIME 
No QIIME features classified to the genus level as _Escherichia_ were categorized as pre, post, or full. 
```{r}
mrexp$qiime %>% fData() %>% filter(Rank6 == "g__Escherichia")
```

```{r}
count_df %>% filter(feature_id == "New.CleanUp.ReferenceOTU7975") %>% 
      ggplot() + 
      geom_point(aes(x = t_fctr, y = count)) + 
      facet_wrap(~biosample_id, scales = "free") + theme_bw()
```

```{r}
count_df %>% filter(feature_id == "114510") %>% 
      ggplot() + 
      geom_point(aes(x = t_fctr, y = count)) + 
      facet_wrap(~biosample_id, scales = "free") + theme_bw()
```

```{r}
count_df %>% filter(feature_id == "744023") %>% 
      ggplot() + 
      geom_point(aes(x = t_fctr, y = count)) + 
      facet_wrap(~biosample_id, scales = "free") + theme_bw()
```

## Post Treatment Drop-out
```{r}
count_df %>% filter(feature_id == "1040889", biosample_id == "E01JH0004", pipe == "qiime") %>% 
      ggplot() + 
      geom_point(aes(x = t_fctr, y = count)) + theme_bw()
```

## Good Feature
```{r}
count_df %>% filter(feature_id == "SV1", biosample_id == "E01JH0038", pipe == "dada2") %>% 
      ggplot() + 
      geom_point(aes(x = t_fctr, y = count)) + theme_bw()
```

## Mix Dropout 
Feature present in unmixed post-treatment but at relatively low abundance for the titrations.
```{r}
count_df %>% filter(feature_id == "198151", biosample_id == "E01JH0004", pipe == "qiime") %>% 
      ggplot() + 
      geom_point(aes(x = t_fctr, y = count)) + theme_bw()
```

## Random Outliers 
Mostly low count values with a few PCR replicates with high count values. 
```{r}
count_df %>% filter(feature_id == "Otu00133", biosample_id == "E01JH0011", pipe == "mothur") %>% 
      ggplot() + 
      geom_point(aes(x = t_fctr, y = count)) + theme_bw()
```

## Noisy Feature

```{r}
count_df %>% filter(feature_id == "Otu00060", biosample_id == "E01JH0017", pipe == "mothur") %>% 
      ggplot() + 
      geom_point(aes(x = t_fctr, y = count)) + theme_bw()
```

## Partial Dropout 
Multiple titrations and replicates with low abundance counts. 
```{r}
count_df %>% filter(feature_id == "Otu00143", biosample_id == "E01JH0016", pipe == "mothur") %>% 
      ggplot() + 
      geom_point(aes(x = t_fctr, y = count)) + theme_bw()
```

```{r}
feature_subset <- count_df %>% filter(cat %in% c("cat_full", "cat_pre","cat_post")) %>% 
      select(pipe, feature_id, biosample_id, cat) %>% unique() %>% 
      group_by(pipe, cat) %>% sample_n(5, replace = F)
```

Breakdown of feature subset
```{r}
feature_subset %>% ggplot() + 
      geom_bar(aes(x = cat, fill = biosample_id)) + theme_bw()
```

```{r}
feature_subset_count <- count_df %>% right_join(feature_subset) 
large_feature_plot <- feature_subset_count %>% ggplot() +
      geom_point(aes(x = t_fctr, y = count, shape = cat, color = biosample_id)) + theme_bw() +
      facet_wrap(~feature_id, scales = "free_y")

# ggsave(plot = large_feature_plot, 
#        filename = "large_feature_plot.pdf", 
#        width = 24, height = 18)
```

## Feature Metrics for Feature Subsets
```{r}
count_exp_df <- readRDS("../data/expected_count_values_feature_df.rds") 
```

```{r}
count_exp_subset <- count_exp_df %>% right_join(feature_subset) 
```


### R^2 Expected - Observed Counts
```{r}
subset_r2_df <- count_exp_subset %>% 
      group_by(pipe, biosample_id, feature_id) %>% 
      summarise(ss_total = sum((obs_count - mean(obs_count))^2),
                ss_res = sum(residual^2)) %>% 
      mutate(r_squared = 1 - ss_res/ss_total)
```

```{r}
subset_r2_df %>% ungroup() %>% 
      mutate(feature_id = fct_reorder(feature_id, r_squared, .desc = TRUE)) %>% 
      ggplot() + 
      geom_hline(aes(yintercept = -10)) + 
            geom_hline(aes(yintercept = -1)) + 
      geom_hline(aes(yintercept = 0)) + 
      geom_point(aes(x = feature_id, y = r_squared)) + coord_flip() + theme_bw()
```

```{r}
neg10_subset_r2_df <- subset_r2_df %>% 
      select(pipe, biosample_id, feature_id, r_squared) %>% 
      unique() %>% filter(r_squared < -10)
```

```{r}
count_df %>% right_join(neg10_subset_r2_df) %>% ggplot() +
      geom_point(aes(x = t_fctr, y = count, shape = cat, color = biosample_id)) + theme_bw() +
      facet_wrap(~feature_id, scales = "free_y")
```


```{r}
neg1_10_subset_r2_df <- subset_r2_df %>% select(pipe, biosample_id, feature_id, r_squared) %>% 
      unique() %>% filter(r_squared > -10, r_squared < -1)
```

```{r}
neg1_10_subset_r2_df %>% arrange(r_squared)
```

```{r}
count_df %>% right_join(neg1_10_subset_r2_df) %>% 
      ungroup() %>% mutate(feature_id = fct_reorder(feature_id, r_squared)) %>% 
      ggplot() +
      geom_point(aes(x = t_fctr, y = count, shape = cat, color = biosample_id)) + theme_bw() +
      facet_wrap(~feature_id, scales = "free_y")
```

```{r}
pos_subset_r2_df <- subset_r2_df %>% select(pipe, biosample_id, feature_id, r_squared) %>% 
      unique() %>% filter(r_squared > 0)
```

```{r}
pos_subset_r2_df %>% arrange(r_squared)
```

```{r}
count_df %>% right_join(pos_subset_r2_df) %>% 
      ungroup() %>%  mutate(feature_id = fct_reorder(feature_id, r_squared)) %>%  
      ggplot() +
      geom_point(aes(x = t_fctr, y = count, shape = cat, color = biosample_id)) + theme_bw() +
      facet_wrap(~feature_id, scales = "free_y")
```

## Residual Linear Model
```{r}
get_fit <- function(count_exp_df){
      count_exp_df %>% 
            group_by(pipe, biosample_id, feature_id) %>% 
            nest() %>% 
            mutate(fit = map(data,~lm(residual ~ exp_count, data = .)))
}

get_tidy_fit <- function(count_fit){
      count_fit %>% 
            mutate(fit_summary = map(fit, broom::tidy)) %>% 
            select(-data, -fit) %>% unnest() %>% 
            mutate(term = if_else(term == "(Intercept)", "Intercept","Slope"))
}

get_glance_fit <- function(count_fit){
      count_fit %>% 
            mutate(fit_summary = map(fit, broom::glance)) %>% 
            select(-data, -fit) %>% unnest() %>% 
            select(-statistic, -p.value) 
}

resid_fit <- get_fit(count_exp_subset)
# Data frames with fit summary statistics and parameter estimates
resid_tidy <- get_tidy_fit(resid_fit) %>% group_by(pipe, biosample_id, feature_id) %>% 
      mutate(euclid_dist = estimate^2 %>% sum() %>% sqrt())
resid_glance <- get_glance_fit(resid_fit)
resid_tidy
```

```{r}
resid_tidy %>% ungroup() %>% group_by(term) %>%  
      mutate(feature_id = fct_reorder(feature_id, euclid_dist)) %>% 
      ggplot() + 
      # geom_hline(aes(yintercept = -10)) + 
      #       geom_hline(aes(yintercept = -1)) + 
      # geom_hline(aes(yintercept = 0)) + 
      geom_point(aes(x = feature_id, y = estimate)) + facet_wrap(~term, scales = "free_x") + coord_flip() + theme_bw()
```

```{r}
ed_quants <- resid_tidy$euclid_dist %>% quantile()
```

```{r}
ed_quants
```


```{r}
firstq <- resid_tidy %>% filter(euclid_dist < ed_quants[2])
firstq %>% select(pipe, biosample_id, feature_id, term, estimate, euclid_dist) %>% 
      spread(term, estimate) %>% arrange(euclid_dist)
```

```{r}
count_df %>% right_join(firstq) %>% 
      ungroup() %>%  mutate(feature_id = fct_reorder(feature_id, euclid_dist)) %>%  
      ggplot() +
      geom_point(aes(x = t_fctr, y = count, shape = cat, color = biosample_id)) + theme_bw() +
      facet_wrap(~feature_id, scales = "free_y")
```


```{r}
secondq <- resid_tidy %>% filter(euclid_dist > ed_quants[2], euclid_dist < ed_quants[3])
secondq %>% select(pipe, biosample_id, feature_id, term, estimate, euclid_dist) %>% 
      spread(term, estimate) %>% arrange(euclid_dist)
```

```{r}
count_df %>% right_join(secondq) %>% 
      ungroup() %>%  mutate(feature_id = fct_reorder(feature_id, euclid_dist)) %>%  
      ggplot() +
      geom_point(aes(x = t_fctr, y = count, shape = cat, color = biosample_id)) + theme_bw() +
      facet_wrap(~feature_id, scales = "free_y")
```
```{r}
ed_quants
```


```{r}
thirdq <- resid_tidy %>% filter(euclid_dist > ed_quants[3], euclid_dist < ed_quants[4])
thirdq %>% select(pipe, biosample_id, feature_id, term, estimate, euclid_dist) %>% 
      spread(term, estimate) %>% arrange(euclid_dist)
```


```{r}
count_df %>% right_join(thirdq) %>% 
      ungroup() %>%  mutate(feature_id = fct_reorder(feature_id, euclid_dist)) %>%  
      ggplot() +
      geom_point(aes(x = t_fctr, y = count, shape = cat, color = biosample_id)) + theme_bw() +
      facet_wrap(~feature_id, scales = "free_y")
```

```{r}
fourthq <- resid_tidy %>% filter(euclid_dist > ed_quants[4])
fourthq %>% select(pipe, biosample_id, feature_id, term, estimate, euclid_dist) %>% 
      spread(term, estimate) %>% arrange(euclid_dist)
```

```{r}
count_df %>% right_join(fourthq) %>% 
      ungroup() %>%  mutate(feature_id = fct_reorder(feature_id, euclid_dist)) %>%  
      ggplot() +
      geom_point(aes(x = t_fctr, y = count, shape = cat, color = biosample_id)) + theme_bw() +
      facet_wrap(~feature_id, scales = "free_y")
```

## Observed - Expected Linear Model
```{r}
get_fit <- function(count_exp_df){
      count_exp_df %>% 
            group_by(pipe, biosample_id, feature_id) %>% 
            nest() %>% 
            mutate(fit = map(data,~lm(obs_count ~ exp_count, data = .)))
}

get_tidy_fit <- function(count_fit){
      count_fit %>% 
            mutate(fit_summary = map(fit, broom::tidy)) %>% 
            select(-data, -fit) %>% unnest() %>% 
            mutate(term = if_else(term == "(Intercept)", "Intercept","Slope"))
}

get_glance_fit <- function(count_fit){
      count_fit %>% 
            mutate(fit_summary = map(fit, broom::glance)) %>% 
            select(-data, -fit) %>% unnest() %>% 
            select(-statistic, -p.value) 
}

exp_fit <- get_fit(count_exp_subset)
# Data frames with fit summary statistics and parameter estimates
exp_tidy <- get_tidy_fit(exp_fit) %>% group_by(pipe, biosample_id, feature_id) 
exp_glance <- get_glance_fit(exp_fit)
```

When fitting model to observed and expected counts mix dropout (when post-treatment high and rest are low) perform well. 

```{r}
exp_tidy %>% select(pipe, biosample_id, feature_id, term, estimate) %>% spread(term, estimate) %>% DT::datatable()
```

```{r}
resid_tidy %>% ungroup() %>% group_by(term) %>%  
      mutate(feature_id = fct_reorder(feature_id, estimate)) %>% 
      ggplot() + 
      # geom_hline(aes(yintercept = -10)) + 
      #       geom_hline(aes(yintercept = -1)) + 
      # geom_hline(aes(yintercept = 0)) + 
      geom_point(aes(x = feature_id, y = estimate)) + facet_wrap(~term, scales = "free_x") + coord_flip() + theme_bw()
```

## Pre - Post Modeling Approach
```{r}
pipeline_dir <- "../../mgtst_pipelines"
mrexp <- get_mrexp(pipeline_dir)

## Extracting a tidy dataframe with count values from MRexpiment objects
get_count_df <- function(mrobj, agg_genus = FALSE){
      if(agg_genus){
            mrobj <- aggregateByTaxonomy(mrobj, lvl = "Rank6", 
                                         norm = FALSE, log = FALSE, sl = 1)
      }
      
      mrobj <- cumNorm(mrobj, p = 0.75)
      mrobj %>%
            # not sure whether or not to normalize counts prior to analysis
            MRcounts(norm = FALSE, log = FALSE, sl = 1) %>%  
            as.data.frame() %>% 
            rownames_to_column(var = "feature_id") %>% 
            gather("id","count", -feature_id)
} 
```


```{r}
count_df <- mrexp %>% map_df(get_count_df, .id = "pipe") %>% 
      left_join(pData(mrexp$dada2)) %>% 
      filter(biosample_id != "NTC") %>% 
      ungroup() %>% 
      mutate(t_fctr = fct_relevel(t_fctr, c(20, 15, 10, 5, 4, 3, 2, 1, 0)),
             titration = as.numeric(as.character(t_fctr)), 
             theta = 2^(-titration),
             theta = if_else(theta == 2^-20, 0, theta)) %>% 
      right_join(feature_subset)

## Calculate scaling factor 
count_df <- count_df %>% group_by(pipe, id) %>% 
      mutate(total_count = sum(count),
             scale_factor = total_count/ exp(mean(log(total_count))))
   
nested_count_df <- count_df %>% group_by(pipe, biosample_id, feature_id) %>% nest()
```

```{r}
fit_df <- nested_count_df %>%
      mutate(fit = map(data, ~lm(count ~ -1 + theta:t_fctr, offset = scale_factor, data = .)))
```

```{r}
nested_count_df
```

```{r}
get_tidy_fit <- function(count_fit){
      count_fit %>% 
            mutate(fit_summary = map(fit, broom::tidy)) %>% 
            select(-data, -fit) %>% unnest() #%>% 
            #mutate(term = if_else(term == "(Intercept)", "Intercept","Slope"))
}

tidy_fit <- get_tidy_fit(fit_df) %>% filter(term != "theta:t_fctr20")
```

```{r}
tidy_fit
```

```{r}
fit_summary <- tidy_fit %>% group_by(pipe, biosample_id, feature_id) %>% 
      summarise(est_cov = sd(estimate)/mean(estimate), error_cov= sd(std.error)/mean(estimate))
```

```{r}
fit_summary %>% ungroup() %>%
      mutate(feature_id = fct_reorder(feature_id, est_cov)) %>% 
      gather("key", "value", est_cov, error_cov) %>% 
      ggplot() + 
      # geom_hline(aes(yintercept = -10)) + 
      #       geom_hline(aes(yintercept = -1)) + 
      # geom_hline(aes(yintercept = 0)) + 
      geom_point(aes(x = feature_id, y = value)) + facet_wrap(~key, scales = "free_x") + coord_flip() + theme_bw()
```

## Variance Metric
```{r}
var_metric <- feature_subset_count %>% 
      group_by(pipe, feature_id, biosample_id, t_fctr) %>% 
      summarise(count_cov = sd(count)/mean(count),
                med_count = median(count)) %>% 
      group_by(pipe, feature_id, biosample_id) %>% 
      mutate(mean_count_cov = mean(count_cov, na.rm = TRUE)) %>% 
      ungroup() %>% 
      mutate(feature_id = fct_reorder(feature_id, mean_count_cov))
```

```{r}
ggplot(var_metric) + 
      geom_point(aes(x = feature_id, y = count_cov)) + coord_flip() + theme_bw()
```




```{r}
var_metric %>% ggplot() +
      geom_point(aes(y = med_count, x = count_cov, color = biosample_id, shape = pipe)) + 
      geom_smooth(aes(x = count_cov, y = med_count, color = biosample_id)) +
      scale_y_log10()
```

```{r}
var_metric %>% ggplot() +
      geom_point(aes(y = med_count+ 1, x = count_cov, color = pipe)) + 
      geom_smooth(aes(x = count_cov, y = med_count + 1, color = pipe)) +
      scale_y_log10() + facet_wrap(~biosample_id, scales = "free_y")
```


### Investigating the relationship between Metrics


```{r}
metric_subset_count <- var_metric %>% right_join(subset_r2_df) %>% 
      select(pipe, feature_id, biosample_id, mean_count_cov, r_squared) %>% 
      unique() %>% left_join(feature_subset_count) %>% 
      group_by(pipe, feature_id, biosample_id) %>% 
      mutate(mean_count = mean(count), sd_count = sd(count))

large_feature_plot_metrics <- metric_subset_count %>% ggplot() +
      geom_point(aes(x = t_fctr, y = count, shape = cat, color = biosample_id)) + 
      geom_text(aes(x = "4", y = mean_count, label = round(r_squared,2))) +
      geom_text(aes(x = "4", y = mean_count - sd_count, label = round(mean_count_cov, 2))) + 
      theme_bw() +
      facet_wrap(~feature_id, scales = "free_y")

ggsave(plot = large_feature_plot_metrics,
       filename = "large_feature_plot_metrics.pdf",
       width = 24, height = 18)
```

```{r}
var_metric %>% right_join(subset_r2_df) %>% filter(r_squared > -3) %>% 
      ggplot() + geom_point(aes(x = scale(mean_count_cov), y = scale(r_squared)))
```

```{r}
var_metric %>% select(pipe, biosample_id, feature_id, mean_count_cov) %>% 
      unique() %>% right_join(subset_r2_df) %>% select(-ss_total, -ss_res) %>% arrange(desc(r_squared))
```


```{r}
resid_tidy %>% select(pipe, biosample_id, feature_id,term, estimate,euclid_dist) %>% 
      spread(term, estimate) %>% right_join(subset_r2_df) %>% DT::datatable()
```

```{r}
resid_tidy %>% select(pipe, biosample_id, feature_id,term, estimate,euclid_dist) %>% 
      spread(term, estimate) %>% right_join(subset_r2_df) %>% ungroup() %>% 
      select(-feature_id, -pipe,-biosample_id) %>% 
GGally::ggpairs()
```




### Conclusions
R^2 for observd and expected counts is a good metric for bias



## Session information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
      knitr::kable()
``` 