---
title: "Bias Variance Fit Metric"
author: "Nate Olson"
date: '`r Sys.Date()`'
output:
  pdf_document: default
  html_document: default
---

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
library(ggfortify)
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd)
pipeline_dir <- "../../mgtst_pipelines"
mrexp <- get_mrexp(pipeline_dir)
```

## Sample Results
```{r}
## Extracting a tidy dataframe with count values from MRexpiment objects
get_count_df <- function(mrobj, agg_genus = FALSE){
      if(agg_genus){
            mrobj <- aggregateByTaxonomy(mrobj, lvl = "Rank6", 
                                         norm = FALSE, log = FALSE, sl = 1)
      }
      
      mrobj <- cumNorm(mrobj, p = 0.75)
      mrobj %>%
            # not sure whether or not to normalize counts prior to analysis
            MRcounts(norm = TRUE, log = FALSE, sl = 1000) %>%  
            as.data.frame() %>% 
            rownames_to_column(var = "feature_id") %>% 
            gather("id","count", -feature_id)
} 

count_df <- mrexp %>% map_df(get_count_df, .id = "pipe") %>% 
      left_join(pData(mrexp$dada2)) %>% 
      filter(biosample_id != "NTC") %>% 
      ungroup() %>% 
      mutate(t_fctr = fct_relevel(t_fctr, c(0:5, 10, 15, 20)),
             titration = as.numeric(as.character(t_fctr)), 
             theta = 2^(-titration),
             theta = if_else(theta == 2^-20, 0, theta))  %>% 
      filter(t_fctr %in% c(0:4,20))

## Calculate scaling factor 
count_df <- count_df %>% group_by(pipe, id) %>% 
      mutate(total_count = sum(count),
             scale_factor = total_count/ exp(mean(log(total_count))),
             norm_count = count/scale_factor)

## Excluding features with median EO metric of 1 or -1 and logFC between pre and post < 1
## %%TODO%% replace with heuristic filter > 14 of the PCR replicates have observed counts, and only features with observed counts for unmixed samples.  
count_nb <- readRDS("../data/nb_expected_eo_metric_feature_df.rds") %>% 
      mutate(count_post = post * total_abu, count_pre = pre * total_abu)

good_eo <- count_nb %>% group_by(pipe, biosample_id, feature_id) %>% 
      summarise(med_eo = median(eo_metric)) %>% 
      filter(abs(med_eo) != 1)

good_logfc <- readRDS("../data/pre_post_deseq_logFC.RDS") %>% 
    filter(abs(log2FoldChange) > 1, pvalue < 0.05) 



good_features <- count_nb %>% group_by(pipe, biosample_id, feature_id) %>% 
      summarise(med_eo = median(eo_metric)) %>% 
      filter(abs(med_eo) != 1)

theta_estimates <- readRDS("../data/theta_estimates.RDS")
count_full_df <- count_df %>% right_join(good_features) %>% left_join(theta_estimates) %>% 
    mutate(theta_estimate = max(estimate, 0))


nested_count_df <- count_full_df %>% group_by(pipe, biosample_id, feature_id) %>% nest()
```

Fitting model to real data
```{r}
fit_df <- nested_count_df %>%
      mutate(fit = map(data, ~lm(count ~ theta:t_fctr, data = .)),
             fit_thetahat = map(data, ~lm(count ~ theta_estimate:t_fctr, data = .)))
```

```{r}
get_tidy_fit <- function(count_fit){
      count_fit %>% 
            mutate(fit_summary = map(fit, broom::tidy)) %>% 
            dplyr::select(-data, -fit, -fit_thetahat) %>% unnest()
}

tidy_fit <- get_tidy_fit(fit_df)
```

Consistent decrease in estimates with titration factor.  

Two potential reasons;  

1. model not estimating Post-Pre as expected,  
2. assumptions about how the samples were mixed is not valid. 

```{r}
dat <- tidy_fit %>% 
      mutate(term = factor(term, levels = c( paste0("theta:t_fctr",c(15,10,5:0)), "(Intercept)"))) %>%
      filter(term != "theta:t_fctr20", term != "(Intercept)") %>% 
      group_by(pipe, biosample_id, feature_id) %>% mutate(min_est = min(estimate)) 

dat %>% ggplot() +
      geom_path(aes(x = term, y = estimate, group = feature_id), alpha = 0.25) +
      facet_grid(pipe~biosample_id, scales = "free_y") + theme_bw() +
      theme(axis.text.x = element_text(angle = 90))
```


### Model Diagnostics for Example Features
```{r}
fit_plots <- fit_df %>% filter(feature_id == "SV1") %>% .$fit %>% map(autoplot)
```


```{r}
print(fit_plots)
```

The E. coli feature for the mothur and DADA2 pipeline look very similar but the numbers different
```{r}
sv1_tidy_df <- fit_df %>% filter(feature_id == "SV1") %>% 
      mutate(fit_tidy = map(fit, broom::tidy)) %>% 
      dplyr::select(-fit,-fit_thetahat, -data) %>% unnest()

sv1_tidy_df %>% 
      mutate(term = factor(term, 
                           levels = c( paste0("theta:t_fctr",c(15,10,5:0)),
                                       "(Intercept)"))) %>% 
      filter(term != "(Intercept)") %>% 
      ggplot() + 
      geom_point(aes(x = term, y = estimate)) +
      geom_errorbar(aes(x = term, ymin = estimate - 2*std.error, ymax = estimate + 2*std.error)) +
      #scale_y_log10() +
      facet_wrap(~biosample_id) + theme_bw() +
      theme(axis.text.x = element_text(angle = 90))

```

```{r}
fit_df %>% filter(feature_id == "SV1") %>% .$fit %>% map(summary)
```


```{r}
otu1_tidy_df <- fit_df %>% filter(feature_id == "Otu00001") %>% 
      mutate(fit_tidy = map(fit, broom::tidy)) %>% 
      dplyr::select(-fit,-fit_thetahat, -data) %>% unnest()

otu1_tidy_df %>% 
      mutate(term = factor(term, levels = c( paste0("theta:t_fctr",c(15,10,5:0)), "(Intercept)"))) %>%
            filter(term != "(Intercept)") %>% 
      ggplot() + 
      geom_point(aes(x = term, y = estimate)) +
      geom_errorbar(aes(x = term, ymin = estimate - 2*std.error, ymax = estimate + 2*std.error)) +
      facet_wrap(~biosample_id) + theme_bw() +
      theme(axis.text.x = element_text(angle = 90))
```

## Bias - Variance Relationship
```{r}
fit_summary <- dat %>% group_by(pipe, biosample_id, feature_id) %>% 
      summarise(mean_est = mean(estimate), cov_est = sd(estimate)/mean(estimate))

get_glance_fit <- function(count_fit){
      count_fit %>% 
            mutate(fit_summary = map(fit, broom::glance)) %>% 
            dplyr::select(-data, -fit, -fit_thetahat) %>% unnest()
}

glance_fit <- get_glance_fit(fit_df)

fit_summary <- glance_fit %>% 
    dplyr::select(pipe, biosample_id, feature_id, sigma) %>% 
    left_join(fit_summary) %>% 
      mutate(cov_sigma = sigma/mean_est)

``` 

```{r}
fit_summary %>% 
      ggplot() + 
      geom_point(aes(x = cov_est, 
                     y = cov_sigma)) + 
      theme_bw() +
    labs(x = "COV(Beta1)", y = "COV(sigma)", subtitle = "CSS Normalized Counts Bias-Variance Metrics") 
```

```{r}
fit_summary %>% filter(abs(mean_est) > 1) %>% 
      ggplot() + 
      geom_point(aes(x = abs(cov_est), 
                     y = abs(cov_sigma))) + 
      theme_bw() +
    labs(x = "|COV(Beta1)|", y = "|COV(sigma)|", subtitle = "CSS Normalized Counts Bias-Variance Metrics: mean Beta1 > 1") 
```

```{r}
fit_summary %>% filter(abs(mean_est) > 1) %>% 
      ggplot() + 
      geom_point(aes(x = abs(cov_est), 
                     y = abs(cov_sigma), color = pipe), alpha = 0.25) + 
      theme_bw() + facet_wrap(~biosample_id, nrow = 1) + 
    labs(x = "COV(Beta1)", y = "COV(sigma)", subtitle = "CSS Normalized Counts Bias-Variance Metrics: mean Beta1 > 1") 
```


```{r}
fit_summary %>% filter(abs(mean_est) > 1) %>% 
    ggplot() + geom_boxplot(aes(x = pipe, y = abs(cov_est))) 
```

```{r}
fit_summary %>% filter(abs(mean_est) > 1) %>% 
    ggplot() + geom_boxplot(aes(x = biosample_id, y = abs(cov_est), color = pipe)) +
    labs(x = "Biological Replicate", y = "|COV(Beta1)|", color = "Pipeline") + theme_bw()
```

```{r}
fit_summary %>% filter(abs(mean_est) > 1) %>% 
    ggplot() + geom_boxplot(aes(x = biosample_id, y = abs(cov_sigma), color = pipe)) +
    labs(x = "Biological Replicate", y = "|COV(sigma)|", color = "Pipeline") + theme_bw()
```



```{r}
fit_summary %>% filter(abs(mean_est) > 1) %>% 
      ggplot() + 
      geom_point(aes(x = abs(mean_est) + 1, 
                     y = abs(cov_sigma), 
                     color = biosample_id, shape = pipe)) + 
      scale_x_continuous(trans = "log2") + 
      theme_bw() + facet_grid(biosample_id~pipe, scales = "free_y") + 
    labs(x = "|mean(Beta1)|", y = "|COV(sigma)|")
    
```

```{r}
fit_summary %>% filter(abs(mean_est) > 1) %>% 
      ggplot() + 
      geom_point(aes(x = abs(mean_est) + 1, 
                     y = abs(cov_est))) + 
    geom_smooth(aes(x = abs(mean_est) + 1, y = abs(cov_est))) +
      theme_bw() + facet_grid(biosample_id~pipe, )  + scale_x_continuous(trans = "log2") 
```





```{r}
fit_summary %>% filter(abs(cov_sigma) < 50) %>% 
      mutate(cov_est = if_else(cov_est < 0, -log2(abs(cov_est)), log2(cov_est)),
             cov_sig_sign = if_else(cov_sigma < 0, "neg","pos")) %>% 
      ggplot() + geom_point(aes(x = cov_est, y = abs(cov_sigma), color = cov_sig_sign), alpha = 0.25) + 
      theme_bw() + scale_y_continuous(trans = "log2") + 
      labs(x = "log2[COV(Post - Pre)]", 
           y = "Absolute Value COV Sigma", 
           color = "COV Sigma Sign") +
      facet_grid(biosample_id~pipe)
```



## Theta Hat Estimates
```{r}
get_tidy_fit <- function(count_fit){
      count_fit %>% 
            mutate(fit_summary = map(fit_thetahat, broom::tidy)) %>% 
            dplyr::select(-data, -fit, -fit_thetahat) %>% unnest()
}

tidy_fit <- get_tidy_fit(fit_df)
```

Consistent decrease in estimates with titration factor.  

Two potential reasons;  

1. model not estimating Post-Pre as expected,  
2. assumptions about how the samples were mixed is not valid. 

```{r}
get_tidy_fit <- function(count_fit){
      count_fit %>% 
            mutate(fit_summary = map(fit_thetahat, broom::tidy)) %>% 
            dplyr::select(-data, -fit, -fit_thetahat) %>% unnest()
}

tidy_fit <- get_tidy_fit(fit_df)

dat <- tidy_fit %>% 
      mutate(term = factor(term, levels = c( paste0("theta_estimate:t_fctr",c(15,10,5:0)), "(Intercept)"))) %>%
      filter(term != "theta_estimate:t_fctr20", term != "(Intercept)") %>% 
      group_by(pipe, biosample_id, feature_id) %>% mutate(min_est = min(estimate)) 

dat %>% ggplot() +
      geom_path(aes(x = term, y = estimate, group = feature_id), alpha = 0.25) +
      facet_grid(pipe~biosample_id, scales = "free_y") + theme_bw() +
      theme(axis.text.x = element_text(angle = 90))
```

### Theta Hat Bias Variance Estimates
```{r} 
fit_summary <- dat %>% group_by(pipe, biosample_id, feature_id) %>% 
      summarise(mean_est = mean(estimate), cov_est = sd(estimate)/mean(estimate))

get_glance_fit <- function(count_fit){
      count_fit %>% 
            mutate(fit_summary = map(fit, broom::glance)) %>% 
            dplyr::select(-data, -fit, -fit_thetahat) %>% unnest()
}

glance_fit <- get_glance_fit(fit_df)

fit_summary <- glance_fit %>% 
    dplyr::select(pipe, biosample_id, feature_id, sigma) %>% 
    left_join(fit_summary) %>% 
      mutate(cov_sigma = sigma/mean_est)

``` 

```{r}
fit_summary %>% 
      ggplot() + 
      geom_point(aes(x = cov_est, 
                     y = cov_sigma)) + 
      theme_bw() +
    labs(x = "COV(Beta1)", y = "COV(sigma)", subtitle = "CSS Normalized Counts Bias-Variance Metrics") 
```

```{r}
fit_summary %>% filter(abs(mean_est) > 1) %>% 
      ggplot() + 
      geom_point(aes(x = abs(cov_est), 
                     y = abs(cov_sigma))) + 
      theme_bw() +
    labs(x = "|COV(Beta1)|", y = "|COV(sigma)|", subtitle = "CSS Normalized Counts Bias-Variance Metrics: mean Beta1 > 1") 
```

```{r}
fit_summary %>% filter(abs(mean_est) > 1) %>% 
      ggplot() + 
      geom_point(aes(x = abs(cov_est), 
                     y = abs(cov_sigma), color = pipe), alpha = 0.25) + 
      theme_bw() + facet_wrap(~biosample_id, nrow = 1) + 
    labs(x = "COV(Beta1)", y = "COV(sigma)", subtitle = "CSS Normalized Counts Bias-Variance Metrics: mean Beta1 > 1") 
```