---
title: "JHU Lib Size Analysis"
author: "Nate Olson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(mgtst)
```


```{r}
barcode_prop <- read_csv("~/mgtst/inst/extdata/read_index_percent - Sheet1.csv") %>% as_data_frame()
jhu_run <- sample_sheet %>% filter(seq_lab == "JHU", kit_version %in% c("A","D"))
illumina_index <- read_csv("~/mgtst/inst/extdata/illumina_index.csv", comment = "#") %>% rename(sample_sheet_seq = `sample_sheet`)
barcode_prop_idx_name <- barcode_prop %>% 
    left_join(illumina_index, by = c("INDEX 1 (I7)" = "sample_sheet_seq")) %>% 
    select(-barcode_seq, -`Index`,-`Kit Version`) %>% rename(Rev_Index_ID = `i7 index name`) %>%
    left_join(illumina_index,  by = c("INDEX 2 (I5)" = "sample_sheet_seq")) %>% 
    select(-barcode_seq, -`Index`,-`Kit Version`) %>% rename(For_Index_ID = `i7 index name`) %>% 
    unique()
jhu_run_barcode_prop <-left_join(sample_sheet, barcode_prop_idx_name) %>% 
    rename(prop_barcode = `% READS IDENTIFIED (PF)`) %>% filter(!is.na(prop_barcode), seq_lab == "JHU")
```

Need to fix sample sheet for and rev index reversed...

## barcode proportion distribution
```{r}
ggplot(jhu_run_barcode_prop) + geom_histogram(aes(x = prop_barcode)) + 
    geom_rug(aes(x = prop_barcode, color = sampleID))
```

One non-template sample with library proportion < 0.2
```{r}
jhu_run_barcode_prop %>% filter(prop_barcode < 0.2, sampleID != "NTC") %>% 
    select(id, sampleID, prop_barcode)
```

## Estimated Read Counts
15,632,954 reads passed Illumina QC filter

```{r}
jhu_read_cnt_est <- jhu_run_barcode_prop %>% mutate(lib_est = prop_barcode * 15632954)
```

```{r}
ggplot(jhu_read_cnt_est) + 
    geom_jitter(aes(x = sampleID, y = lib_est, 
                    color = as.factor(dilution)), width = 0.2) +
    labs(x = "Biological Replicate", y = "Estimated Read Count", color = "Titration") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90))
```

No indication of potential biological sample or titration/ dilution factor specific effect on estimated library size.
```{r}
ggplot(jhu_read_cnt_est) + 
    geom_jitter(aes(x = as.factor(dilution), y = lib_est, 
                    color = as.factor(pcr_16S_plate)), width = 0.2) +
    facet_wrap(~sampleID, scales = "free_x") +
    labs(x = "Titration", y = "Estimated Read Count", color = "PCR Plate") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90))
```

Estimated read count distribution - potential batch effect
```{r}
ggplot(jhu_read_cnt_est) + 
    geom_density(aes(x = lib_est,fill = as.character(pcr_16S_plate)), 
                 alpha = 0.25) +
    theme_bw() + labs(x = "Estimated Read Count", fill = "PCR Plate")
```

