---
title: "Mixture Design V3"
author: "Nate Olson"
date: "February 15, 2016"
output: html_document
---
<!-- Fill in title, author, and date above -->

<!-- Resources for creating Rmarkdown documents
    https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf
    https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf
-->

### Purpose
<!-- Reason for performing the analysis -->
Generate a PCR and sequencing sample sheet for mixture study experimental design.

#### Software Versions
```{r warning=FALSE, message=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(magrittr)
library(knitr)
library(devtools)
library(sapkotaUtils)
library(DT)
opts_chunk$set(cache=TRUE)
```



### Experimental Factors
* sampleID - five biological replicate samples  
* treatment - either "pre" or "post" treatment sample  
* labID - individual biological sample id assigned to a specific replicate and timepoint relative to treatment  
* dilution - post treatment sample dilution factor $2^{-x}$  
* sample_type - "unmixed" samples are the pure pre and post treatment samples, "titration" samples are the mixtures.    
* seq_center - the two planned places where the samples will be sequenced, "JHU" and "NIST"  
```{r}
## five biological replicates
sampleID <- c("E01JH0004","E01JH0011","E01JH0016","E01JH0017","E01JH0038")
labID = c(3,8,12,13,28,138,115,999,177,105)
treatment = c(rep("pre",5),rep("post",5))
timepoint = c(rep(-1,5), c(4,2,2,5,2))
biosample_df <- data_frame(sampleID= rep(sampleID,2), labID, treatment, timepoint)
```

Sample id information
```{r}
biosample_df %>% arrange(sampleID) %>% kable()
```


### Experimental Design 

To eliminate confounding variability at different steps in the protocol; amplicon PCR, barcoded amplicons, and pooled libraries are processed at both laboratories (e.g. before barcoding and sequencing).

```{r}
make_plan_a_96_df <- function(sampleID, dilution){
    ## experimental design for 96 sample layout
    bio_replicates <- rep(sampleID,each = 2*length(dilution))
    dilution_96 <- rep(dilution, times = 2*length(sampleID))
    
    ntc <- data_frame(sampleID = rep("NTC",6),
                      sample_type = "control", dilution = NA)
    
    data_frame(sampleID = bio_replicates, 
                sample_type = "titration", dilution = dilution_96) %>% 
        bind_rows(ntc) %>% mutate(sample_type = ifelse(dilution %in% c(0,-1), 
                                                       "unmixed",sample_type)) ->df_96
}

make_plan_a_df <- function(sampleID, dilution){
    ## PCRs
    plate_1 <- make_plan_a_96_df(sampleID, dilution) %>% 
        mutate(pcr_16S_plate = 1, pcr_16S_id = 1:n())
    plate_2 <- make_plan_a_96_df(sampleID, dilution) %>% 
        mutate(pcr_16S_plate = 2, pcr_16S_id = (n() + 1):(2*n()))
    pcr_plates <- bind_rows(plate_1, plate_2) 
    
    ## Barcode
    barcode_jhu <- pcr_plates %>% mutate(barcode_lab = "JHU", 
                                          barcode_id = 1:n())
    barcode_nist <- pcr_plates %>% mutate(barcode_lab = "NIST", 
                                          barcode_id = (n() + 1):(2*n()))
    seq_plates <- bind_rows(barcode_jhu, barcode_nist)

    seq_jhu <- seq_plates %>% mutate(seq_lab = "JHU")
    seq_nist <- seq_plates %>% mutate(seq_lab = "NIST")
    return(list(pcr_sample_sheet = pcr_plates, 
                seq_sample_sheet = bind_rows(seq_jhu, seq_nist)))
}
```

```{r}
dilution <- c(-1,0:4,5,10,15)
plan_a_sample_sheets <- make_plan_a_df(sampleID, dilution)
plan_a_pcr <- plan_a_sample_sheets$pcr_sample_sheet
plan_a <- plan_a_sample_sheets$seq_sample_sheet
```

### Adding barcode index information
```{r}
pcr_plate_layout<- plan_a_pcr %>% 
    filter(pcr_16S_plate == 1, dilution != -1 | is.na(dilution)) %>% 
    mutate(half = c(rep(c(rep(0,8),rep(6,8)),5),rep(c(0,6), each = 3)), 
           col = half + as.numeric(factor(sampleID)), 
           row = c(rep(c("A","B","C","D","E","F","G","H"),10),
                   rep(c("A","D","H"), 2)))
pcr_plate_layout <- plan_a_pcr %>% filter(pcr_16S_plate == 1, dilution == -1) %>% 
    mutate(half = rep(c(0,6), 5), col = rep(c(6,12), 5), 
           row = rep(c("F","B","C","E","G"),each = 2)) %>%
    bind_rows(pcr_plate_layout) %>% select(-pcr_16S_plate)
pcr_plates <- pcr_plate_layout %>% mutate(pcr_16S_id = pcr_16S_id + 96) %>%
    bind_rows(pcr_plate_layout, .)
```

JHU barcode kit A for PCR plate 1 and D for plate 2, NIST barcode kit B for plate 1 and C for plate 2

```{r}
illumina_index <- read_csv("illumina_index - Sheet1.csv", comment = "#") %>% 
    rename(kit_version = `Kit Version`, index_name = `i7 index name`) 
assign_plate <- c(A= 1, B= 1, C= 2, D = 2)
assign_lab <- c(A= "JHU", B= "NIST", C= "NIST", D = "JHU")

## split and assign plate position
forward_index <- illumina_index %>% filter(Index == "i5") %>% 
    group_by(kit_version) %>% mutate(row = LETTERS[1:8]) %>% 
    mutate(pcr_16S_plate = assign_plate[kit_version], 
           barcode_lab = assign_lab[kit_version]) %>% 
    rename(For_Index = Index, For_Index_ID = index_name, 
           For_sample_sheet = sample_sheet, For_barcode_seq = barcode_seq) %>% 
    select(kit_version, For_Index_ID, pcr_16S_plate, barcode_lab, row)


reverse_index <- illumina_index %>% filter(Index == "i7")%>% 
    group_by(kit_version) %>% mutate(col = 1:12) %>% 
    mutate(pcr_16S_plate = assign_plate[kit_version], 
           barcode_lab = assign_lab[kit_version]) %>% 
    rename(Rev_Index = Index, Rev_Index_ID = index_name, 
           Rev_sample_sheet = sample_sheet, Rev_barcode_seq = barcode_seq) %>% 
    select(kit_version, Rev_Index_ID, pcr_16S_plate, barcode_lab, col)

annotated_index <- full_join(forward_index, reverse_index)
```

Combine PCR layout, index, and plan a data frames
```{r}
full_sample_sheet <- left_join(plan_a, pcr_plates) %>% left_join(annotated_index)
```



### Sample ID format
`B#_M#_P#_L#_S#`
* B - biological replicate, 1-5  
* M - mix, 1-9  
* P - PCR replicate 1-4  
* L - lab performing library prep 1 or 2  
* S - sequencing lab, 1 or 2  

```{r}
sam_id <- paste0("B", 0:5)
names(sam_id) <- c("NTC","E01JH0004","E01JH0011","E01JH0016",
                   "E01JH0017","E01JH0038")

mix_id <- paste0("M", 0:9)
mix_id_df <- data_frame(mix_id, dilution = c(NA, dilution))
pcr_id_df <- data_frame(pcr_id = paste0("P", 1:4), 
                     pcr_16S_plate = rep(1:2, each = 2), half = rep(c(0,6), 2))
lib_id <- c("JHU" = "L1","NIST" = "L2")
seq_id <- c("JHU" = "S1","NIST" = "S2")

sample_sheet <- full_sample_sheet %>% 
    left_join(pcr_id_df) %>% left_join(mix_id_df) %>%  
    mutate(sam_id = sam_id[sampleID],
           lib_id = lib_id[barcode_lab], seq_id = seq_id[seq_lab],
           id = paste(sam_id, mix_id, pcr_id, lib_id, seq_id, sep ="_")) %>%
    unite(pos, row, col) %>% 
    select(id, sampleID, dilution, pcr_16S_plate, pos, barcode_lab,kit_version, 
           For_Index_ID, Rev_Index_ID, seq_lab)
```

#### Experimental Design summary
note no overlap between barcode labs and barcode kits
```{r}
ggplot(sample_sheet) + 
    geom_bar(aes(x = sampleID, fill = as.character(dilution))) +
    facet_grid(pcr_16S_plate*kit_version*barcode_lab~seq_lab) + theme_bw() +
    theme(axis.text.x = element_text(angle = 90)) + labs(fill = "Dilution\nFactor")
```

Total replicates per mixture, 
2 pcr replicates x 2 plates x 2 library prep x 2 sequencing labs

```{r}
sample_sheet %>% group_by(sampleID, dilution) %>% 
    summarise(count = n()) %>% spread(sampleID, count)
```

Note NTC sample IDs are not unique
```{r}
sample_sheet %>% group_by(id) %>% summarize(count = n()) %>% filter(count > 1)
```

Breakdown of replicates
```{r}
sample_sheet %>% 
    ggplot() + geom_bar(aes(x = as.factor(dilution), fill = as.factor(pcr_16S_plate))) +
    facet_grid(sampleID~barcode_lab*seq_lab, scale = "free_y") + 
    labs(x = "Dilution Factor 2^-x", y = "Number of Technical Replicates", fill = "Replicate PCR Plate")
ggsave("mixture_design_v3.png",width = 8, height = 3.5, dpi = 450)
```


### PCR plate layout
Diagram of 96 well plate layout
```{r}
pcr_plate_layout<- plan_a_pcr %>% 
    filter(pcr_16S_plate == 1, dilution != -1 | is.na(dilution)) %>% 
    mutate(half = c(rep(c(rep(0,8),rep(6,8)),5),rep(c(0,6), each = 3)), 
           col = half + as.numeric(factor(sampleID)), 
           row = c(rep(c("A","B","C","D","E","F","G","H"),10),
                   rep(c("A","D","H"), 2)))
pcr_plate_layout <- plan_a_pcr %>% filter(pcr_16S_plate == 1, dilution == -1) %>% 
    mutate(col = rep(c(6,12), 5), row = rep(c("F","B","C","E","G"),each = 2)) %>%
    bind_rows(pcr_plate_layout)

pcr_plate_layout %>% mutate(Row = as.numeric(factor(row))) %>% 
 ggplot(aes(x=col, y=Row)) +
    geom_point(data=expand.grid(seq(1,12), seq(1,8)), aes(x=Var1, y=Var2), 
               color='grey90', fill='white', shape=21, size=6) +
    geom_point(aes(color = sampleID), size=10) +
    geom_text(aes(label = dilution)) +
    coord_fixed(ratio=(13/12)/(9/8), xlim=c(0.5,12.5), ylim=c(0.5,8.5)) +
    scale_y_reverse(breaks=seq(1,8), labels=LETTERS[1:8]) +
    scale_x_continuous(breaks=seq(1,12)) +
    labs(title="PCR Layout") +
    theme_bw()
```

### Barcode layout
```{r}
sample_sheet %>% separate(pos, c("Row","Col")) %>%
 ggplot(aes(x = For_Index_ID, y = Rev_Index_ID)) +
    geom_point(aes(color = Row, shape = factor(pcr_16S_plate)), size = 4) +
    geom_text(aes(label = Col)) +
    labs(title="Barcode Layout") +
    facet_grid(pcr_16S_plate~barcode_lab) +
    theme_bw() + theme(axis.text.x = element_text(angle = 90)) + 
    labs(x = "Forward  Index", y = "Reverse Index", 
         fill = "Plate Row", shape = "Replicate PCR Plate")
```


#### Sample Sheet
```{r}
pre_df <- biosample_df %>% filter(treatment == "pre") %>% 
    rename(pre_labID = labID, pre_timepoint = timepoint) %>% select(-treatment)
post_df <- biosample_df %>% filter(treatment == "post") %>% 
    rename(post_labID = labID, post_timepoint = timepoint) %>% select(-treatment)
annotated_sample_sheet <- left_join(pre_df, post_df) %>% left_join(plan_a)
write_csv(annotated_sample_sheet, "mix_study_exp_design_sample_sheet_v3.csv")

pcr_sample_sheet <- left_join(pre_df, post_df) %>% 
    full_join(pcr_plate_layout) %>% select(-half)
write_csv(pcr_sample_sheet, "mix_study_exp_design_pcr_sample_sheet_v3.csv")
```


Sample Sheet
```{r, cache=FALSE}
datatable(annotated_sample_sheet)
```


#### Sample Sheet Provenance
Names of output data files and md5 sums for provenance checking

<!-- Name of input data files 
     Make sure to use full file path names for md5 sums
-->
```{r}
dat_files <- c(
    sample_sheet = "mix_study_exp_design_sample_sheet_v3.csv" 
)
md5table(dat_files) %>% kable()
```

#### To Do 
* RevComp barcodeseq for i7

### Session information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
kable(s_info$packages)
```

