---
title: "R Notebook"
output: html_notebook
---
Information from the SILVA TestPrime database
Can use different databases 
Primers GCCTACGGGNGGCWGCAG, TACHVGGGTATCTAATCC Need to get actual primers
https://www.arb-silva.de/browser/ssu-128/silva-ref-nr/testprime 

Primer Analysis SILVA 

ES Wright et al. (2014) "Automated Design of Probes for rRNA-Targeted Fluorescence In Situ Hybridization Reveals the Advantages of Using Dual Probes for Accurate Identification." Applied and Environmental Microbiology, doi:10.1128/AEM.01685-14. 
http://decipher.cee.wisc.edu/Download.html
http://decipher.cee.wisc.edu/cgi-bin/R.cgi/ProbeMelt2.R

```{r}
library(tidyverse)
```

```{r}
tax_list <- read_delim("~/Downloads/arb-silva.de_testprime_taxlist_417581.csv", delim = ";")
```

```{r}
tax_list %>% glimpse()
```

```{r}
tax_list
```

This information can potentially be aligned with Rep seq data.

```{r}
read_delim("~/Downloads/arb-silva.de_testprime_hitlist_417581.csv", delim = ";")
```
## Session information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
      knitr::kable()
``` 
