---
title: "Negative Binomial Linear Model Count Estimates"
author: "Nate Olson"
date: '`r Sys.Date()`'
always_allow_html: yes
output:
  pdf_document: default
  html_document: default
---

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(MASS)
library(tidyverse)
library(modelr)
count_df <- readRDS("data/raw_counts.RDS")

## Cleaning up data frame and calculating total abundance
count_df <- count_df %>% 
      filter(biosample_id != "NTC") %>% 
      select(pipe, biosample_id, id, pcr_rep, feature_id, t_fctr, count) %>% 
      group_by(biosample_id, pipe, id) %>%
      mutate(total_abu = sum(count))
```

## Summary
Using a negative binomial generalized linear model to calculate feature proportion estimates across the four PCR replicates accounting for differences in sampling depth.

```{r oldCode, warning=FALSE, message=FALSE, echo = FALSE}
## Old Code 
# library(ProjectTemplate)
# cwd <- getwd()
# setwd("../")
# load.project()
# setwd(cwd) 
# pipeline_dir <- "../../mgtst_pipelines"
# mrexp <- get_mrexp(pipeline_dir)
# get_count_df <- function(mrobj, agg_genus = FALSE, css = TRUE){
#       if(agg_genus){
#             mrobj <- aggregateByTaxonomy(mrobj, lvl = "Rank6", 
#                                          norm = FALSE, log = FALSE, sl = 1)
#       }
#       
#       if(css == TRUE){
#             mrobj <- cumNorm(mrobj, p = 0.75)
#             count_mat <- MRcounts(mrobj, norm = TRUE, log = FALSE, sl = 1000)
#       }else{
#             count_mat <- MRcounts(mrobj, norm = FALSE, log = FALSE, sl = 1)
#       }
#        count_mat %>%  
#             as.data.frame() %>% 
#             rownames_to_column(var = "feature_id") %>% 
#             gather("id","count", -feature_id)
# }
# 
# count_df <- mrexp %>% 
#       map_df(get_count_df, css = FALSE, .id = "pipe") %>% 
#       left_join(pData(mrexp$dada2)) %>% 
#       filter(biosample_id != "NTC") %>% 
#       dplyr::select(pipe, biosample_id, id, pcr_rep, feature_id, t_fctr, count) %>% 
#       group_by(biosample_id, pipe, id) %>% 
#       mutate(total_abu = sum(count))
# 
# ## Excluding features with no observed counts for an individual 
# nonzero_count_features <- count_df %>% 
#       mutate(t_fctr = factor(t_fctr, levels = c(0:5,10,15,20))) %>% 
#       group_by(pipe, biosample_id, feature_id) %>% 
#       summarise(total_count = sum(count)) %>% 
#       filter(total_count > 0) %>% 
#       dplyr::select(-total_count)
# 
# ## Excluding titrations with zero counts for all four PCR replicates
# titration_counts <- count_df %>% 
#       right_join(nonzero_count_features) %>% 
#        mutate(t_fctr = factor(t_fctr, levels = c(0:5,10,15,20))) %>% 
#       group_by(pipe, biosample_id, feature_id, t_fctr) %>% 
#       summarise(total_count = sum(count)) 
# 
# zero_titrations <- titration_counts %>%
#       filter(total_count == 0) %>% 
#       dplyr::select(-total_count) 
# 
# nonzero_titrations <- titration_counts %>% 
#       filter(total_count > 0) %>% 
#       dplyr::select(-total_count)
# 
# ### Filtering nonzero features and modifying data frame to fit model      
# nested_count_df <- count_df %>% 
#       mutate(t_fctr = factor(t_fctr, levels = c(0:5,10,15,20))) %>% 
#       right_join(nonzero_count_features) %>% 
#       right_join(nonzero_titrations) %>% 
#       group_by(pipe, biosample_id, feature_id) %>%
#       nest()
```



```{r nbModelFit}
calc_nb_prop <- function(count_df, fit_file, prop_file, fit_level = "feature"){
      if(file.exists(prop_file)){
            return(readRDS(prop_file))
      }
      
      ## Excluding features with no observed counts for an individual 
      nonzero_count_features <- count_df %>% 
            mutate(t_fctr = factor(t_fctr, levels = c(0:5,10,15,20))) %>% 
            group_by(pipe, biosample_id, feature_id) %>% 
            summarise(total_count = sum(count)) %>% 
            filter(total_count > 0) %>% 
            dplyr::select(-total_count)
      
      ## Excluding titrations with zero counts for all four PCR replicates
      titration_counts <- count_df %>% 
            right_join(nonzero_count_features) %>% 
            mutate(t_fctr = factor(t_fctr, levels = c(0:5,10,15,20))) %>% 
            group_by(pipe, biosample_id, feature_id, t_fctr) %>% 
            summarise(total_count = sum(count)) 
      
      zero_titrations <- titration_counts %>%
            filter(total_count == 0) %>% 
            dplyr::select(-total_count) 
      
      nonzero_titrations <- titration_counts %>% 
            filter(total_count > 0) %>% 
            dplyr::select(-total_count)
      
      ## Fit model at either feature level or titration level
      if(fit_level == "feature"){
            if(file.exists(fit_file)){
                  nb_fit <- readRDS(fit_file)
            }else{
                  
                  
                  ### Filtering nonzero features and modifying data frame to fit model      
                  nested_count_df <- count_df %>% 
                        mutate(t_fctr = factor(t_fctr, levels = c(0:5,10,15,20))) %>% 
                        right_join(nonzero_count_features) %>% 
                        right_join(nonzero_titrations) %>% 
                        group_by(pipe, biosample_id, feature_id) %>%
                        nest()
                  
                  ## Using safely to fit negative binomial to catch model fit errors
                  safe_glm.nb <- safely(glm.nb)
                  
                  nb_fit <- nested_count_df %>%
                        mutate(fit = map(data, 
                                         ~safe_glm.nb(count ~ t_fctr + offset(log(total_abu)), 
                                                      data = ., 
                                                      link = "log"))) 
                  
                  saveRDS(nb_fit, fit_file)
            }
            
            ### Filtering failed models
            fit_results <- nb_fit$fit %>% 
                  transpose() %>% 
                  as_data_frame()
            
            nb_fit_results <- nb_fit %>% 
                  dplyr::select(-fit) %>%
                  bind_cols(fit_results) %>% 
                  mutate(ok_model = map_lgl(error, is_null))  
            
            ## Extracting proportion estimates Using add_predictions to get
            ## proportions estimates. Removes need to account for differences in
            ## intercept when some or all of the titrations are excluded from the model
            ## fit due to no observed counts.
            nb_counts <- nb_fit_results %>% 
                  mutate(data = map(data, mutate, `offset(log(total_abu)))` = offset(log(total_abu)))) %>% 
                  mutate(data = map2(data, result, add_predictions)) %>%
                  dplyr::select(-result, -error, -ok_model) %>%
                  unnest() %>% 
                  mutate(nb_prop = exp(pred)/total_abu) %>% 
                  ## Proportion estimates for each PCR replicate - using mean as values are not unique
                  group_by(pipe, biosample_id, feature_id, t_fctr) %>% 
                  summarise(nb_prop = mean(nb_prop))
            
            
            ## set zero titrations to zero
            zero_titrations <- zero_titrations %>% mutate(nb_prop = 0)
            nb_counts <- nb_counts %>% full_join(zero_titrations)  
            
      }else if(fit_level == "titration"){
            if(file.exists(fit_file)){
                  nb_fit <- readRDS(fit_file)
            }else{
                  
                  ## Preparing data frame to fit negative binomial model to each
                  ## titration individually, essentially using the negative binomial
                  ## to calculate the average feature propportion across the PCR
                  ## replicates only considering one titration for a feature at a
                  ## time.
                  ### Filtering nonzero features and modifying data frame to fit model      
                  nested_count_df <- count_df %>% 
                        mutate(t_fctr = factor(t_fctr, levels = c(0:5,10,15,20))) %>% 
                        right_join(nonzero_count_features) %>% 
                        right_join(nonzero_titrations) %>% 
                        group_by(pipe, biosample_id, feature_id, t_fctr) %>% 
                        nest()
                  
                  ## Using safely to fit negative binomial to catch model fit errors
                  safe_glm.nb <- safely(glm.nb)
                  
                  nb_fit <- nested_count_df %>%
                        mutate(fit = map(data, 
                                         ~safe_glm.nb(count ~ offset(log(total_abu)), 
                                                      data = ., 
                                                      link = "log"))) 
                  
                  saveRDS(nb_fit, fit_file)
            }
            
            ### Filtering failed models
            fit_results <- nb_fit$fit %>% 
                  transpose() %>% 
                  as_data_frame()
            
            nb_fit_results <- nb_fit %>% 
                  dplyr::select(-fit) %>%
                  bind_cols(fit_results) %>% 
                  mutate(ok_model = map_lgl(error, is_null))  
            
            ## Extracting proportion estimates Using add_predictions to get
            ## proportions estimates. Removes need to account for differences in
            ## intercept when some or all of the titrations are excluded from the model
            ## fit due to no observed counts.
            nb_counts <- nb_fit_results %>% 
                  filter(ok_model == TRUE) %>% 
                  mutate(data = map(data, mutate, `offset(log(total_abu)))` = offset(log(total_abu)))) %>% 
                  mutate(data = map2(data, result, add_predictions)) %>%
                  dplyr::select(-result, -error, -ok_model) %>%
                  unnest() %>% 
                  mutate(nb_prop = exp(pred)/total_abu) %>% 
                  ## Proportion estimates for each PCR replicate - using mean as values are not unique
                  group_by(pipe, biosample_id, feature_id, t_fctr) %>% 
                  summarise(nb_prop = mean(nb_prop))
            
            
            ## set zero titrations to zero
            zero_titrations <- zero_titrations %>% mutate(nb_prop = 0)
            nb_counts <- nb_counts %>% full_join(zero_titrations)            
            
            
      }else{
            warning("fit_level not titration or feature, nb_prop not calculated")
            return()
      }
      
      
      
      saveRDS(nb_counts, prop_file)
      
      return(nb_counts)
}
```

Fitting titration level model
```{r calcNB, message = FALSE, warning=FALSE}
nb_counts_file <- "data/nb_counts_titrations.RDS"
if(file.exists(nb_counts_file)){
      nb_counts <- readRDS(nb_counts_file)
}else{
      dada_nb <- calc_nb_prop(count_df %>% filter(pipe == "dada2"),
                              fit_file = paste0("data/nb_fit_","dada2","_titration.RDS"), 
                              prop_file = paste0("data/nb_counts_","dada2","_titration.RDS"),
                              fit_level = "titration")
      
      qiime_nb <- calc_nb_prop(count_df %>% filter(pipe == "qiime"),
                               fit_file = paste0("data/nb_fit_","qiime","_titration.RDS"), 
                               prop_file = paste0("data/nb_counts_","qiime","_titration.RDS"),
                               fit_level = "titration")
      
      mothur_nb <- calc_nb_prop(count_df %>% filter(pipe == "mothur"),
                                fit_file = paste0("data/nb_fit_","mothur","_titration.RDS"), 
                                prop_file = paste0("data/nb_counts_","mothur","_titration.RDS"),
                                fit_level = "titration")
      
      unclustered_nb <- calc_nb_prop(count_df %>% filter(pipe == "unclustered"),
                                     fit_file = paste0("data/nb_fit_","unclustered","_titration.RDS"), 
                                     prop_file = paste0("data/nb_counts_","unclustered","_titration.RDS"),
                                     fit_level = "titration")
      
      nb_counts <- bind_rows(dada_nb, qiime_nb, mothur_nb, unclustered_nb)
      saveRDS(nb_counts, "data/nb_counts_titrations.RDS")
}
```

Sanity check for proportion estimates are between 0 and 1

```{r}
min(nb_counts$nb_prop) < 0
```

```{r}
max(nb_counts$nb_prop) > 1
```
```{r}
nb_counts %>% filter(nb_prop > 1)
```


# Notes from exploring model failuers during development
## Fitting model by features
## Issue with unmix specific features
No unmix specific features were observed for dada2 E01JH0017 and E01JH0038. 
```{r}
# zero_titrations %>% filter(feature_id == "SV1077", biosample_id == "E01JH0017")
```

Example where counts are only observed for one titration. 
```{r}
# test_df <- count_df %>% filter(feature_id == "SV1077", biosample_id == "E01JH0017") %>% filter(t_fctr == 20)
```

```{r}
# test_df
```

Unable to fit model when only one titration
```{r}
# glm.nb(count ~ t_fctr + offset(log(total_abu)), data = test_df, link = "log")
```
Able to fit model when only excluding t_fctr from the model. 
Potential alternative solution is to fit the negative binomial model to each titration individual instead of by feature. 
```{r}
# glm.nb(count ~ offset(log(total_abu)), data = test_df, link = "log")
```

Failed models have high proportion of zeros

Need approach to summarise failed model fits and zero counts
Compare with pa_summary pa totals
```{r}
# bad_models <- nb_fit_results %>% 
#       filter(ok_model == F) 
```


## Titration level model fit - Bad Models 
Model fit errors due to outliers - not sure how to address.
Fit errors due to model unable to come to convergence - providing starting value prevents error.
```{r}
# dada_fit <- readRDS("data/nb_fit_dada2_titration.RDS")
# dada_results <- dada_fit$fit %>% 
#       transpose() %>% 
#       as_data_frame()
# 
# dada_fit_results <- dada_fit %>% 
#       dplyr::select(-fit) %>%
#       bind_cols(dada_results) %>% 
#       mutate(ok_model = map_lgl(error, is_null)) 
# 
# dada_fail <- dada_fit_results %>% filter(ok_model == FALSE)
```

```{r}
# dada_fail
```
Unclear from the data frames why the models failed
```{r}
# dada_fail$data
```

```{r}
# dada_fail$data[[1]] %>% mutate(prop = count/total_abu)
```

```{r}
# glm.nb(count ~ offset(log(total_abu)), 
#        data = dada_fail$data[[1]], 
#        link = "log")
```


Removing low total abundance sample addresses error as the observer 0 count is an outlier
```{r}
# glm.nb(count ~ offset(log(total_abu)), 
#        data = dada_fail$data[[1]] %>% filter(total_abu > 10000), 
#        link = "log")
```


Providing starting value prevents error
```{r}
# dada_fail$data[[2]] %>% mutate(prop = count/total_abu) %>% .$prop %>% mean() %>% log()
```

```{r}
# glm.nb(count ~ offset(log(total_abu)), start = -5.654,  
#        data = dada_fail$data[[2]], link = "log")
```

Providing starting value prevents error
```{r}
# dada_fail$data[[3]]
```


```{r}
# st_value <- dada_fail$data[[3]] %>% mutate(prop = count/total_abu) %>% 
#       .$prop %>% mean() %>% log()
```

```{r}
# glm.nb(count ~ offset(log(total_abu)), start = st_value,  
#        data = dada_fail$data[[3]], link = "log")
```




```{r}
# mothur_fit <- readRDS("data/nb_fit_mothur_titration.RDS")
# mothur_results <- mothur_fit$fit %>% 
#       transpose() %>% 
#       as_data_frame()
# 
# mothur_fit_results <- mothur_fit %>% 
#       dplyr::select(-fit) %>%
#       bind_cols(mothur_results) %>% 
#       mutate(ok_model = map_lgl(error, is_null)) 
# 
# mothur_fail <- mothur_fit_results %>% filter(ok_model == FALSE)
```


```{r}
# mothur_fail
```

Again here one of the values is an outlier resulting in a failed model fit

```{r}
# mothur_fail$data[[1]]
```


```{r}
# st_value <- mothur_fail$data[[1]] %>% mutate(prop = count/total_abu) %>% 
#       .$prop %>% mean() %>% log()
```

This error was reported elsewhere, https://stackoverflow.com/questions/11749977/why-does-glm-nb-throw-a-missing-value-error-only-on-very-specific-inputs. Not sure if the solution is applicable
```{r}
# glm.nb(count ~ offset(log(total_abu)), start = st_value,  
#        data = mothur_fail$data[[1]], link = "log")
```

```{r}
# test_dat <- mothur_fail$data[[1]]
# test_dat[2,4] <-  test_dat[2,4] + 1
# test_dat
```

```{r}
# qiime_fit <- readRDS("data/nb_fit_qiime_titration.RDS")
# qiime_results <- qiime_fit$fit %>% 
#       transpose() %>% 
#       as_data_frame()
# 
# qiime_fit_results <- qiime_fit %>% 
#       dplyr::select(-fit) %>%
#       bind_cols(qiime_results) %>% 
#       mutate(ok_model = map_lgl(error, is_null)) 
# 
# qiime_fail <- qiime_fit_results %>% filter(ok_model == FALSE)
```

No failed models
```{r}
# qiime_fail
```

```{r}
# unclustered_fit <- readRDS("data/nb_fit_unclustered_titration.RDS")
# unclustered_results <- unclustered_fit$fit %>% 
#       transpose() %>% 
#       as_data_frame()
# 
# unclustered_fit_results <- unclustered_fit %>% 
#       dplyr::select(-fit) %>%
#       bind_cols(unclustered_results) %>% 
#       mutate(ok_model = map_lgl(error, is_null)) 
# 
# unclustered_fail <- unclustered_fit_results %>% filter(ok_model == FALSE)
```

Only two failed model fits
```{r}
# unclustered_fail
```

```{r}
# unclustered_fail$data
```

```{r}
# glm.nb(count ~ offset(log(total_abu)),  
#        data = unclustered_fail$data[[1]], link = "log")
```

```{r}
# st_value <- unclustered_fail$data[[1]] %>% mutate(prop = count/total_abu) %>% 
#       .$prop %>% mean() %>% log()
```

This error was reported elsewhere, https://stackoverflow.com/questions/11749977/why-does-glm-nb-throw-a-missing-value-error-only-on-very-specific-inputs. Not sure if the solution is applicable
```{r}
# glm.nb(count ~ offset(log(total_abu)), start = st_value,  
#        data = unclustered_fail$data[[1]], link = "log")
```