---
title: "dada2_pipeline"
author: "Nate Olson"
date: "November 7, 2016"
output: html_document
---

```{r}
library(dada2)
library(tidyverse)
library(readr)
library(stringr)
library(forcats)
```

## Pipeline description

1. Quality filter read pairs, remove reads with ambiguous bases (Ns) or more than 4 expected errors, trim ends of reads with quality score of 2 or to 290 bp and 220 bp for forward and reverse reads respectively, and the first 10 bases are trimmed.  
      * Number of expected errors calculated based on quality scores, `EE = sum(10^(-Q/10))`.   
      * Quality score of 2 used by Illumina to indicate end of good sequencing data.   

2. The forward and reverse reads are dereplicated. A consensus (average) quality score is assigned for each position.  
3. Sequence inference, denoising - error correction step.  
4. Merging forward and reverse read pairs. 
5. Chimera removal - filtering sequences with complementary regions matching more abundant sequences.


## Pipeline Budget
### Raw sqeunces
Starting number of sequences per sample.
```{r}
seq_meta <- readRDS("../data/seq_metadata_df.RDS")
```

```{r}
raw_count <- seq_meta %>% filter(Read == "R1") %>% select(ill_id,reads) %>% 
      dplyr::rename(id = ill_id, total = reads) %>% mutate(pipe_step = "raw")
```

## Quality Filter 

## Dereplicated Sequences 

## Merging 
Need to rerun pipeline to get sequence information
```{r}

```

