---
title: "Feature Classification"
author: "Nate Olson"
date: '`r Sys.Date()`'
always_allow_html: yes
output:
  pdf_document: default
  html_document: default
---
__REVISIONS__ 
- summary figure for publication
- justification of category definitions
      - pre-post specific features, different types of information
      - unclassified features have limited information

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
library(tidyverse)
feature_specificity_df <- readRDS("../data/feature_specificity_df.rds") 
feature_specificity_counts_df <- readRDS("../data/feature_specificity_counts_df.rds")
```

The information available to define the expected titration count values is dependent on the observed unmixed sample count values. 
The expected count values for features only found in unmixed pre- and post-treatment samples is dependent only treatment the features are observed in and the titration factor. 
Feature classification is defined for a biological replicate individually and classification values for one biological replicate is independent of the feature classification for another biological replicate. 
For features observed in both unmixed pre- and post-treatment samples the expected values is dependent on the titration factor and the unmixed pre- and post-treatment samples. 
Features present in 1-3 of the PCR replicates provide limited quantitative information for determining the expected count values but can be used for qualitative analysis and determining a method limit of detection.  



```{r fig.cap = "Number of features in each feature classification. Feature abundance is based on the number of pre- and post-treatment PCR replicates the feature was observed in for a biological replicate. Classification definitions: pre partial - not present in any post and present in 1-3 of the pre-treatment replicates; pre full - not present in any post and present in 4 of the pre-treatment replicate; post partial - not present in any pre and present in 1-3 of the post-treatment replicates; post partial - not present in any pre and present in 4 of the post-treatment replicates; mix specific - not present in any pre or post-treatment replicates but present in titration replicates; non-specific - present in less than 4 pre- and post-treatment replicates; non-specific full - present in all pre- and post-treatment replicates."}
feature_specificity_df %>% filter(specific_anno != "biorep_neg") %>% 
      ggplot() + geom_bar(aes(x = specific_anno)) + 
      facet_wrap(~pipe, ncol = 1, scale = "free_y") +
      theme_bw() + labs(x = "Feature Classification", y = "# of Features")
```

Expected high accuracy features based on titration expectations are non-specific, post_full, and pre_full features. 


```{r fig.cap = "Total reads for individual features by samples (titration and biological replicate)  grouped by feature classification."}
feature_specificity_counts_df %>%
      ggplot() + geom_jitter(aes(x = specific_anno, y = Reads), alpha = 0.25) + 
      facet_wrap(~pipe, ncol = 1, scale = "free_y") + theme_bw() + scale_y_log10() +
      labs(x = "Feature Classification")
```



# Session information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
      knitr::kable()
```