---
title: "qpcr_linear_model"
author: "Nate Olson"
date: '`r Sys.Date()`'
output:
  pdf_document: default
  html_document: default
---
```{r echo=FALSE, message=FALSE}
library(readxl)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(ggplot2)
library(broom)
library(modelr)
library(forcats)
library(knitr)
library(lme4)
library(purrr)
```


## Summary
* Objective - compare linear modeling methods for qPCR bacterial abundance data  
* Method - bacterial abundance quantified using zymo kit  
* Results  
      - $R^2$ standard curve 0.99  
      - Large fraction of samples (especially unmixed) with Ct values outside standard curve.

## Objective
The proportion of pre and post exposure samples in individual titrations is dependent on the ratios at which the two samples were mixed. 
This assumes that the pre and post samples have equivalent proportions of bacterial to non-bacterial DNA. 
To validate this assumption the concentration of bacterial DNA was assayed using qPCR. 
Additionally, the concentation of bacterial DNA in the titrations was assayed. 

## Methods 
* zymo qPCR assay - https://www.zymoresearch.com/dna/dna-analysis/femto-bacterial-dna-quantification-kit  
* 45 Samples - all mixed and unmixed  
* diluted samples - need to find out how they were diluted
* triplicates per sample - 135 reactions  
      * three qPCR plates, one replicate of each sample ran on each plate
* 7 concentration standard curve - for the assay  
      * issue with fourth standard  

### Munging qPCR Data
```{r}
bac_con <- read_excel(path = "../data/MixStudy_Nate_20160919.xls",
                      sheet = "QDNA_20160919",
                      skip = 11, na = "Undetermined", col_names = FALSE)
colnames(bac_con) <- c("well","sample_name",
                       "plate1_Ct","plate1_quant",
                       "plate2_Ct","plate2_quant",
                       "plate3_Ct","plate3_quant")
bac_con <- bac_con %>% gather("id","value", -well, -sample_name) %>% 
      separate(id, c("plate","var"), sep = "_") %>% 
      spread(var,value)
```

```{r}
bac_std <- read_excel(path = "../data/MixStudy_Nate_20160919.xls",
                      sheet = "QDNA_20160919",skip = 3, col_names = FALSE) %>% 
      select(-X12, -X13, -X5,-X7,-X9) %>% filter(X2 %in% paste0("Std",1:7))
colnames(bac_std) <- c("well","sample_name","conc","plate1","plate2","plate3")
bac_std <- bac_std %>% gather("plate","Ct",-well, -sample_name, -conc) %>% 
      mutate(conc = as.numeric(conc), Ct = as.numeric(Ct)) %>% filter(!is.na(Ct))
## NAs introduced when converting samples with undetermined and omit Ct values.
```

## Analysis of Standard Curve
Fitted model using the three plates as replicates and not fitting individual models for each plate. Using a random effect model is likely a better approach.

### Fitting the standard curve using linear models

```{r}
bac_std <- mutate(bac_std, log_conc = log10(conc))
std_fit1 <- lm(log_conc~Ct, data = bac_std)
# extracting values for plots
std_slope <- std_fit1$coefficients[2]
std_intercept <- std_fit1$coefficients[1]
fit1_df <- tibble(plate = paste0("plate",c(1:3)),
                  slope = std_fit1$coefficients[2], 
                  intercept = std_fit1$coefficients[1],
                  mod = "log_conc~Ct")
bac_fit <- bac_con %>% select(sample_name,plate, Ct) %>% add_predictions(std_fit1)
```

```{r}
fit1_df %>% knitr::kable()
```


```{r}
std_fit2 <- lm(log_conc~plate/Ct, data = bac_std)
bac_fit <- bac_con %>% select(sample_name,plate, Ct) %>% add_predictions(std_fit2)
std_coef <- std_fit2 %>% coefficients()
coef_df <- frame_data(
      ~plate, ~intercept, ~slope,
      "plate1",std_coef[1], std_coef[4],
      "plate2",std_coef[1] + std_coef[2], std_coef[5],
      "plate3",std_coef[1] + std_coef[3], std_coef[6]
)
fit2_df <- coef_df %>% mutate(mod = "log_conc~plate/Ct")
```
__Question__ Why do the "log_conc~plate/Ct" and "log_conc~plate*Ct" fit the data the same? Only appearant difference is that the t-test for the Ct:plate interactions for the "log_conc~plate/Ct" model are calculated independently and making them all significant. 

```{r}
fit2_df %>% knitr::kable()
```


```{r}
std_fit3 <- lm(log_conc~plate*Ct, data = bac_std)
std_coef <- std_fit3 %>% coefficients()
## not sure if I need to add Ct
fit3_df <- frame_data(
      ~plate, ~intercept, ~slope,
      "plate1", std_coef[1], std_coef[4],
      "plate2",std_coef[1] + std_coef[2], std_coef[4] + std_coef[5],
      "plate3", std_coef[1] + std_coef[3], std_coef[4] + std_coef[6]
      ) %>% 
      mutate(mod = "log_conc~plate*Ct")
```

```{r}
fit3_df %>% knitr::kable()
```

__Question__ Why do all of the plates have the same slope and intercept for random effects model?
```{r}
std_fit4 <- lmer(log_conc~ Ct + (1|plate), data = bac_std)
fit4_df <- tibble(plate = paste0("plate",c(1:3)),
                  slope = coef(std_fit4)$plate[1,2], 
                  intercept = coef(std_fit4)$plate[1,1],
                  mod = "log_conc~ Ct + (1|plate)")
```

```{r}
fit4_df %>% knitr::kable()
```


### Model Fit Summary
```{r}
fit_summary_df <-list(std_fit1, std_fit2, std_fit3, std_fit4) %>% 
      map_df(glance, .id = "fit")
fit_summary_df$mod <- c("log_conc~Ct","log_conc~plate/Ct",
                        "log_conc~plate*Ct","log_conc~ Ct + (1|plate)")
```

```{r}
fit_summary_df %>% knitr::kable()
```

#### Comparing Regression Models

```{r}
fit_coefs <- bind_rows(fit1_df, fit2_df, fit3_df,fit4_df)
ggplot(fit_coefs) + geom_point(data = bac_std, aes(y = log_conc, x = Ct)) +
      geom_abline(aes(slope = slope, intercept = intercept, color = mod, linetype = mod)) +
      facet_grid(.~plate) + theme_bw()
```

## Model Summaries
### log_conc~Ct
```{r}
summary(std_fit1)
```

### log_conc~plate/Ct
```{r}
summary(std_fit2)
```

### log_conc~plate*Ct
```{r}
summary(std_fit3)
```

### log_conc~ Ct + (1|plate)
```{r}
summary(std_fit4)
```

