---
title: "Measurement Assessment - Qualitative Analysis"
author: "Nate Olson"
date: '`r Sys.Date()`'
always_allow_html: yes
output:
  pdf_document: default
  html_document: default
---

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

feature_specificity_df <- readRDS("../data/feature_specificity_df.rds") 
```

Qualitative analysis: proportion of PCR replicates with 0 and non-zero count values. 

# Outline
* Mix specific - can the observed feature abundance be explained by random sampling alone {report: mix_specific_features}
* Features present in 1-3 PCR replicates, relate median count values to proportion detect to get a LOD 
* Fit a logistic regression for pre- and post- treatment specific features present in all four replicates,  proportion dect ~ expected count value 
* Comparison of LOD by pipeline

## Mix Specific Features  
* Not sure what the expectation is for the unmixed samples  
* Should be able to predict assuming multinomial sampling  
* The average read pre sample less than 10 agrees with prior expectation. The higher average number of reads per sample for DADA2 indicates to me that these mix-specific feature are artifacts of the feature inference procedure rather than a sampling artifact. 

```{r fig.cap = "Number of samples and average number of reads per sample for features only observed in the titrations and not the unmixed pre and post-treatment samples. Each point represent a titration specific feature in one of the biological replicates. Titration specific features can be observed in 28 samples, 4 PCR replicates for 7 titrations."}
feature_specificity_df %>% filter(specific_anno == "mix_specific") %>% 
      ggplot() + geom_point(aes(x = Samp, y = Reads/Samp, color = factor(NTC)), alpha = 0.25) +
      facet_grid(.~pipe) + scale_y_log10() + theme_bw() + 
      labs(x = "Number of Samples", y = "Average Reads per Sample")
```

## Estimating LOD
The median abundance and proportion of PCR replicates was used to estimate the method limit of detection (LOD). 
LOD is defined as the average number of reads per sample where the feature was present in 50\% of the PCR replicates. 
All features and samples (both unmixed and titrations) were used to estimate the LOD. 

```{r}
lod_fit <- glm(avg_read ~ detect_prop, data = lod_df, family = binomial(link='logit'))
```

```{r}
library(modelr)
lod_fit_df <- add_predictions(lod_df,lod_fit)
```

```{r}
ggplot(lod_fit_df) + geom_path(aes(x = avg_read, y = pred), color = "darkorange") + 
      geom_point(aes(x = avg_read, y = detect_prop), color = "darkorange")
```



# Session information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
      knitr::kable()
```