---
title: "Theta Estimate Results"
author: "Nate Olson"
date: '`r Sys.Date()`'
output:
  bookdown::pdf_document2: 
    toc: FALSE
  html_document: default
---


Estimating $\theta$, where $C_{obs_j}$ observed counts for mixture $j$, counts for unmixed $C_{pre_j}$ and $C_{post_j}$. 
The unmixed pre and post counts calculated using the count proportion for a feature calculated using weighted feature proportion estimates and the total abundance for sample $j$.  

$$C_{obs_j} = \theta_j (C_{post_j} - C_{pre_j}) + C_{pre_j}$$ 


```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
library(tidyverse)
library(modelr)
```

```{r loadData, warning=FALSE, message=FALSE, echo = FALSE}
## Extracting a tidy dataframe with count values from MRexpiment objects
get_count_df <- function(mrobj, agg_genus = FALSE){
      if(agg_genus){
            mrobj <- aggregateByTaxonomy(mrobj, lvl = "Rank6", 
                                         norm = FALSE, log = FALSE, sl = 1)
      }
      
      mrobj <- cumNorm(mrobj, p = 0.75)
      mrobj %>%
            # not sure whether or not to normalize counts prior to analysis
            MRcounts(norm = TRUE, log = TRUE, sl = 1000) %>%  
            as.data.frame() %>% 
            rownames_to_column(var = "feature_id") %>% 
            gather("id","count", -feature_id)
} 

count_df <- mrexp %>% map_df(get_count_df, .id = "pipe") %>% 
      left_join(pData(mrexp$dada2)) %>% 
      filter(biosample_id != "NTC") %>% 
      ungroup() %>% 
      mutate(t_fctr = fct_relevel(t_fctr, c(0:5, 10, 15, 20)),
             titration = as.numeric(as.character(t_fctr)), 
             theta = 2^(-titration),
             theta = if_else(theta == 2^-20, 0, theta))  %>% 
      filter(t_fctr %in% c(0:4,20))


pre_post_count <- count_df %>% filter(t_fctr %in% c(0,20)) %>% 
    dplyr::select(pipe, biosample_id, feature_id, t_fctr, count) %>% 
    mutate(t_fctr = if_else(t_fctr == 0, "count_post","count_pre")) #%>% 
    # spread(t_fctr, count)

count_df <- count_df %>% left_join(pre_post_count)

## Excluding features with median EO metric of 1 or -1 and logFC between pre and post < 1
## %%TODO%% replace with heuristic filter > 14 of the PCR replicates have observed counts, and only features with observed counts for unmixed samples.  
count_nb <- readRDS("../data/nb_expected_eo_metric_feature_df.rds") 

good_eo <- count_nb %>% group_by(pipe, biosample_id, feature_id) %>% 
      summarise(med_eo = median(eo_metric)) %>% 
      filter(abs(med_eo) != 1)

good_logfc <- readRDS("../data/pre_post_deseq_logFC.RDS") %>% 
    filter(abs(log2FoldChange) > 1, pvalue < 0.05) 


good_features <- count_nb %>% group_by(pipe, biosample_id, feature_id) %>% 
      summarise(med_eo = median(eo_metric)) %>% 
      filter(abs(med_eo) != 1)

count_good <- count_df %>% right_join(good_features)
```

```{r}
count_good
```


```{r fitModel, warning=FALSE, message=FALSE, echo = FALSE}
count_nest <- count_good %>% 
      group_by(pipe, biosample_id, t_fctr) %>% nest()

fit_df <- count_nest %>% 
      mutate(fit = map(data, ~lm(count ~ -1 + I(count_post - count_pre), offset = count_pre, data = .)),
             fit_tidy = map(fit, broom::tidy),
             fit_glance = map(fit, broom::glance))
```


```{r fitSummary, warning=FALSE, message=FALSE, echo = FALSE}
fit_est <- fit_df %>% dplyr::select(pipe, biosample_id, t_fctr, fit_tidy) %>% unnest() %>% 
    mutate(theta = 2^-as.numeric(t_fctr), t_fctr = factor(t_fctr, levels = c(1:5,10,15))) 

fit_glance <- fit_df %>% dplyr::select(pipe, biosample_id, t_fctr, fit_glance) %>% unnest()
```


Consistent deviation across pipeline and titration from the expected $\theta$ values indicates our assumptions regarding the composition of the mixtures are not valid (Fig. \@ref(fig:thetaHat)). 
Our assumptions are that the DNA from the pre- and post-treatment samples were titrated following a log2 dilution series and that the pre- and post-treatment samples contained similar proportions of bacterial DNA. 
The qPCR spike-in results indicate that the unmixed pre- and post-treatment samples were mixed volumetrically according to our expectations. 
Based on the qPCR bacterial DNA concentration assays, the pre-treatment E01JH0038 sample had the lowest concentration of bacterial DNA between the five biological replicates and the bacterial DNA concentration was higher in the pre-treatment sample compared to the post-treatment sample for only E01JH0017 (`In mixing_and_validating_titrations` artifact).

Another potential cause for the deviations in $\theta$ from our expectation is poor model fit. 
Biological replicate E01JH0038 and E01JH0017 had the largest deviation from the expected values of the five replicates. 
E01JH0038 and E01JH0017 had consistently high R2 values compared to the other biological replicates (Table \@ref(tab:thetaR2)).
These results indicate the poor model fit cannot account for the biological replicate specific differences in theta estimates from our expectations. 


```{r thetaHat, echo = FALSE, fig.cap = "Theta estimates by titration, biological replicate, and bioinformatic pipeline. Points indicates theta estimate and vertical lines plus-minus 2 X standard deviation. The black line indicates the expected theta values. Theta estimates below the expected theta indicate that the titrations contains less than expected bacterial DNA from the post-treatment sample. Theta estimates greater than the expected theta indicate the titration contains more bacterial DNA from the pre-treatment sample than expected."}
ggplot(fit_est) + 
      geom_point(aes(x = t_fctr, y = estimate, color = pipe, shape = pipe)) + 
      geom_errorbar(aes(x = t_fctr, ymax = estimate + 2*std.error, 
                        ymin = estimate - 2 * std.error, color = pipe), width = 0) + 
      geom_line(aes(x = as.numeric(t_fctr), y =  theta)) + 
      facet_grid(. ~ biosample_id, scales = "free_y") + theme_bw() +
    labs(x = "Titration Factor", y = expression(theta), color = "Pipeline", shape = "Pipeline") + 
    theme(legend.position = "bottom")
```


```{r thetaR2, echo = FALSE}
fit_glance %>% group_by(biosample_id) %>% 
    summarise(`Min.` = min(adj.r.squared),  
              `Median` = median(adj.r.squared), 
              `Max.` = max(adj.r.squared)) %>% 
    knitr::kable(caption = "Adjusted R2 values for model used to estimate theta.", digits = 3)
    
```

```{r}
saveRDS(fit_est, "../data/theta_estimates.RDS")
```


# Session information 

## Git repo commit information
```{r}
library(git2r)
repo <- repository(path = "../")
last_commit <- commits(repo)[[1]]
```

The current git commit of this file is `r last_commit@sha`, which is on the `r branches(repo)[[1]]@name` branch and was made by `r last_commit@committer@name` on `r when(last_commit)`. The current commit message is `r last_commit@summary`. The repository is online at https://github.com/nate-d-olson/mgtst-pub  


## Platform Information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
```


## Package Versions
```{r}
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
      knitr::kable()
```