---
title: "Wet Lab Quality Control"
author: "Nate Olson"
date: '`r Sys.Date()`'
always_allow_html: yes
output:
  pdf_document: default
  html_document: default
---
```{r echo=FALSE, message=FALSE}
library(ProjectTemplate)
cwd <- getwd()
setwd("../")
load.project()
setwd(cwd)
```


## Overview
Results of the quality control measurements made as part of the wet lab sample processing. 
The intial 16S rRNA PCR was verified using gel electrophoresis. 
DNA concentration measurements were made after the clean-up of the two PCR steps, 16S rRNA and barcoding, as well as after normalization. 
The pooled library was validated for DNA concentration and size using Qubit and Bioanalyzer. 

__Key Point__ What samples if any to exclude from analysis

### 16S rRNA PCR Gel 

__TODO__ Annotate wells
Wells 18 and 27 in second row are broken  

```{r}
knitr::include_graphics("../img/1st_16S_PCR_P1_P2_Gel_20160224.tif")
```

## DNA concentration Post 16S PCR
Post 16S PCR DNA concentration is tstWetLabQC - measurements were made at JHU
```{r echo = FALSE}
pcr_conc_16S <- tstWetLabQC %>% left_join(tstPcrSampleMetaData)

```
Three no template control samples have measureable but low DNA concentrations. 
This is not unexpected as contaminants in PCR reagents are common (Mark from Abbot reference).

```{r}
pcr_conc_16S %>% filter(biosample_id == "NTC", conc_ngul > 0) %>% 
      knitr::kable()
```

Of the 180 samples 5 had 16S rRNA PCR products with DNA concentration measurements less than 0.5 ng/ul. 
The DNA concenrtation was remeasured using Qubit for samples with DNA concentrations less than 0.5. 
Of the 5 samples quantified with Qubit, only one of the samples had a concentration below the limit of detection for the assay (1-F9). 
This sample is considered lost and excluded from downsteam analysis.  

__TODO__ See if this corresponds with the gel electrophoresis results.  

```{r}
pcr_conc_16S %>% filter(biosample_id != "NTC", conc_ngul < 0.5) %>% knitr::kable()
```

## Barcode and Normalization DNA concentration
__TODO__ Get values with JHU 

__TODO__ incorporate NIST values into ProjectTemplate 

## Pooled library QC results
__TODO__ Get from sequencing labs

# Session information
```{r}
s_info <- devtools::session_info()
print(s_info$platform)
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
      knitr::kable()
```

<!--
### 16S PCR Concentration Post-Cleanup
```{r echo = FALSE}
#tstWetLabQC %>% left_join(tstPcrSampleMetaData) %>% .$conc_ngml %>% summary()
## Joining, by = c("pcr_16S_plate", "pos")
##   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
##  -1.117  12.960  20.670  19.280  25.750  49.600 
```
The initial 16S PCR product was first verified for amplification and amplicon size using gel electrophoresis \@ref(fig:gel_16S).  
The concentration of the PCR product was also assessed using picogreen, samples with negative measured concentation values (excluding no template controls (NTC)) were also check using Qubit. 
The DNA concentration of the 16S PCR product after clean up  samples after ranged from below the limit of detection to 49.6 ng/ml \@ref(fig:16S_pcr_quant). 
The DNA concentration of samples with low DNA concentration measured using picrogreen ( < 1 $ng/\mu L$.) was also measured using Qubit (Table __QUBIT__). 
Only one sample had an undetectable DNA concentration using the Qubit. 
Gel electrophoresis of the 16S PCR products indicated that the sample failed to amplify during the initial PCR \@ref(fig:16S_pcr_gel). 
__TODO__ Need gel info - gel concentration, run info - V and time, ladder info - manufacturer and band sizes.
Despite failure to amplify the sample was processed through the sequence pipeline with the rest of the samples.

```{r gel_16S, echo=FALSE, fig.cap="Agarose gel electrophoresis of 16S PCR products."}
knitr::include_graphics("../img/1st_16S_PCR_P1_P2_Gel_20160224.tif")
```

```{r 16S_pcr_quant, fig.cap = "Picogreen DNA concentration measurements of 16S PCR product after clean-up."}
tstWetLabQC %>% left_join(tstPcrSampleMetaData) %>% 
      mutate(titration = as.factor(titration)) %>% 
      ggplot(aes(x = biosample_id, y = conc_ngml, color = titration)) + 
      geom_point() +
      theme_bw() + 
      labs(x = "Biological Replicate", 
           y = "DNA Concentration (ng/mL)",
           color = "Titration Factor")
```

```{r}
tstWetLabQC %>% left_join(tstPcrSampleMetaData) %>% 
      filter(!is.na(qubit), biosample_id != "NTC") %>% 
      select(biosample_id, titration, pcr_16S_plate, pos, conc_ngml, qubit) %>% 
      kable()
```


#### Index PCR  
* Post clean-up DNA concentrations
__TODO__ Script to extract concentration information

#### Normalization
* Picogreen concentration
__TODO__ Script to extract concentration information
excel file 16032_NIST_norm_15uL.xlsx

#### Library Pooling
* Assay methods   
* QC  
    * methods  
    * results 

-->